---
title: "A Conversation That Built a Model"
subtitle: "Five months, one city's school system, and a record of who did what"
author:
  - name: Adam Dennett FRGS FAcSS
    affiliations: "Bartlett Centre for Advanced Spatial Analysis, UCL"
date: "2026-10-01"
date-format: "dddd D MMMM YYYY"
format:
  revealjs:
    theme: [default, ucl-slides-theme.scss]
    logo: slides_assets/ucl-logo-purple.png
    slide-number: true
    progress: true
    hash: true
    navigation-mode: linear
    controls: true
    width: 1280
    height: 720
    margin: 0.08
    fig-cap-location: bottom
    fig-width: 10
    fig-height: 5.5
    fig-dpi: 150
    embed-resources: true
    auto-stretch: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
library(ggplot2)
library(dplyr)

INK    <- "#1a1a2e"
PURPLE <- "#361a54"
BRIGHT <- "#993bff"
WARM   <- "#f56300"
CYAN   <- "#30d6ff"
PALE   <- "#eedeff"
GREY   <- "#666688"

theme_talk <- function(base = 15) {
  theme_minimal(base_size = base) +
    theme(
      plot.title      = element_text(face = "bold", colour = PURPLE, size = rel(1.05)),
      plot.subtitle   = element_text(colour = GREY, size = rel(0.85)),
      axis.text       = element_text(colour = INK),
      axis.title      = element_text(colour = GREY, size = rel(0.85)),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_line(colour = "#e8e4f0", linewidth = 0.4),
      plot.background  = element_rect(fill = "#FAFAFA", colour = NA),
      panel.background = element_rect(fill = "#FAFAFA", colour = NA),
      legend.position = "top",
      legend.title    = element_blank()
    )
}
knitr::opts_chunk$set(dev.args = list(bg = "#FAFAFA"))

# ---------------------------------------------------------------------------
# The record. Counted from the Claude Code session transcripts held locally
# (~250 MB of JSONL across 10 sessions) and from git history across the
# project repositories. Snapshot taken 26 September 2026.
# ---------------------------------------------------------------------------

days <- as.Date(c(
  "2026-04-13","2026-04-14","2026-04-21","2026-04-26","2026-04-29","2026-04-30",
  "2026-05-01","2026-05-05","2026-05-06","2026-05-08","2026-05-11","2026-05-12",
  "2026-05-13","2026-05-14","2026-05-15","2026-05-16","2026-05-17","2026-05-19",
  "2026-05-20","2026-05-21","2026-05-22","2026-06-02","2026-06-09","2026-06-10",
  "2026-06-11","2026-06-30","2026-07-01","2026-07-02","2026-07-03","2026-07-04",
  "2026-07-08","2026-07-10","2026-07-15","2026-07-16","2026-07-21","2026-07-22",
  "2026-07-23","2026-07-24","2026-07-27","2026-07-29","2026-08-08","2026-08-09",
  "2026-08-14","2026-08-16","2026-08-17","2026-08-18","2026-08-19","2026-08-25",
  "2026-08-26","2026-08-27","2026-08-28","2026-09-01","2026-09-02","2026-09-03",
  "2026-09-04","2026-09-07","2026-09-08","2026-09-09","2026-09-10","2026-09-11",
  "2026-09-13","2026-09-15","2026-09-16","2026-09-17","2026-09-21","2026-09-22",
  "2026-09-24","2026-09-25","2026-09-26"))

tool_calls <- c(121,105,43,74,57,66,39,82,118,83,525,13,57,70,112,12,3,316,15,28,
                69,14,19,218,179,3,192,328,208,39,11,37,119,11,80,102,107,15,42,7,
                58,383,481,380,678,308,274,283,251,91,48,22,176,18,138,75,224,462,
                1140,558,624,597,686,96,408,321,441,170,321)

instructions <- c(13,30,4,30,14,14,5,22,16,17,45,1,4,11,21,1,1,26,1,8,11,4,2,16,15,
                  15,39,33,36,8,3,3,14,7,29,15,21,7,7,5,16,12,19,31,57,42,37,21,12,
                  5,6,11,17,8,9,12,12,31,44,26,41,52,73,5,68,35,16,11,17)

rec <- tibble(day = days, tool_calls = tool_calls, instructions = instructions)

N_INSTR   <- sum(instructions)   # 1320
N_TOOLS   <- sum(tool_calls)     # 13451
N_DAYS    <- length(days)        # 69
N_ELAPSED <- 166                 # days from first message to last
N_HOURS   <- 108                 # summing gaps under 10 minutes

fam <- tibble(
  what = factor(c("Running shell & R", "Writing code and prose", "Reading the codebase",
                  "Driving a browser", "Looking things up", "Other"),
                levels = rev(c("Running shell & R", "Writing code and prose",
                               "Reading the codebase", "Driving a browser",
                               "Looking things up", "Other"))),
  n = c(7953, 2919, 1088, 878, 196, 417))
```

## Who I am, and why this talk {.smaller}

::::: columns
::: {.column width="56%"}
I am a Professor of Urban Analytics at UCL's Centre for Advanced Spatial
Analysis. I build spatial models — of migration, of housing, of how people
move around cities.

I am also a parent in Brighton & Hove, a city that spent 2024 arguing
bitterly about which children should go to which secondary school.

Over five months I built a model of the whole system: every child, every
school, every journey. A public document, a technical guide, an interactive
simulator.

**I wrote almost none of the code.**
:::

::: {.column width="44%"}
::: key-finding
This talk is not "I used AI to write my code". It is an attempt to describe
honestly what a research collaboration with an AI system actually looked
like — with the receipts.
:::

Part of the **AI for Collective Intelligence** hub's interest in how AI
changes the practice of applied research.
:::
:::::

## What exists now {.smaller}

| The thing | Scale |
|:--|:--|
| A strategic analysis of the city's school system | 9,712 lines of Quarto |
| A technical guide auditing every model term | 4,736 lines |
| The model itself, plus data assembly | ~7,400 lines of R |
| An interactive policy simulator (Shiny) | ~3,000 lines |
| A private analysis repository | ~8,200 lines |
| Three slide decks | ~3,500 lines |
| Commits across six repositories | 805 |

::: {style="font-size:0.8em"}
A production-constrained spatial interaction model with a competing-destinations
term, calibrated per catchment, fitted against the council's published offer
data. It reproduces the September 2026 allocations with an R² of 0.95.
:::

## The record I am drawing on {.smaller}

::::: columns
::: {.column width="52%"}
Claude Code keeps a transcript of every session: every instruction I typed,
every action it took, timestamped.

For this talk I parsed **~250 MB of those logs** across ten sessions, plus
the git history of six repositories.

So the numbers that follow are counted, not remembered. They are also
**a snapshot** — the work is still going.
:::

::: {.column width="48%"}
::: key-finding
**1,320** instructions from me.

**13,451** tool calls from Claude.

**69** working days, spread over 166.
:::

About **ten actions for every sentence I typed**.
:::
:::::

## Five months, in one chart

```{r fig-timeline}
#| fig-width: 11.5
#| fig-height: 4.8
ggplot(rec, aes(day)) +
  geom_col(aes(y = tool_calls), fill = BRIGHT, width = 1.9, alpha = 0.9) +
  geom_line(aes(y = instructions * 10), colour = WARM, linewidth = 0.9, linetype = "31") +
  scale_y_continuous(
    name = "Claude tool calls per day",
    sec.axis = sec_axis(~ . / 10, name = "My instructions per day")) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(x = NULL,
       title = "The work was bursty, and it accelerated",
       subtitle = "Bars: actions taken by Claude. Dashed line: instructions from me.") +
  theme_talk(14) +
  theme(axis.title.y.right = element_text(colour = WARM),
        axis.title.y.left  = element_text(colour = BRIGHT))
```

::: {style="font-size:0.72em"}
The April–July work is a different project — a national school attainment
model. The Brighton system model starts on **8 September** and is finished
eighteen days later.
:::

## What the 13,451 actions actually were

```{r fig-families}
#| fig-width: 11
#| fig-height: 4.4
ggplot(fam, aes(n, what)) +
  geom_col(fill = PURPLE, width = 0.62) +
  geom_text(aes(label = format(n, big.mark = ",")), hjust = -0.15,
            size = 4.8, colour = INK) +
  scale_x_continuous(limits = c(0, 9200), expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = NULL,
       title = "Mostly running things and checking what happened",
       subtitle = "Every tool call across ten sessions, grouped by what it was for") +
  theme_talk(14) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(colour = "#e8e4f0", linewidth = 0.4))
```

::: {style="font-size:0.75em"}
The largest single category is not writing code. It is **running code and
reading the result** — the loop that turns a plausible script into a correct one.
:::

## The division of labour {.smaller}

::::: columns
::: {.column width="50%"}
**What I brought**

- The question, and why it mattered
- Knowledge of the school system: admissions law, catchments, the politics
- Judgement about what is defensible to publish
- Knowing when an answer was wrong
- The argument, and the voice it is written in
- Every decision about what to do next
:::

::: {.column width="50%"}
**What Claude brought**

- The model implementation, and its calibration
- Data wrangling across a dozen published sources
- R, Shiny, Quarto, `ggplot2`, `sf`, routing with `r5r`
- Debugging, and the patience for it
- Documentation written as it went
- Deployment, rendering, version control
:::
:::::

::: key-finding
The vision and the domain knowledge were mine. The technical expertise and
the implementation were its. Neither half would have produced this alone.
:::

## Worked example: a school in the wrong place {.smaller}

::::: columns
::: {.column width="55%"}
The model had Hove Park School at its Nevill Road site. Reasonable — that is
the address in the published data.

Five months in, I realised: the school has operated **two campuses**. Year 7
and 8 were at the Valley Campus, a mile away, until a closure proposal in 2026.

The model had been calibrated on the wrong location for the children who
actually generate the demand.
:::

::: {.column width="45%"}
::: key-finding
No amount of code quality would have caught this. It needed someone who
knew the city.
:::

Re-calibrating moved about **a quarter of Hove Park's modelled intake**
between wards, and changed the disadvantage figures materially.
:::
:::::

::: {style="font-size:0.75em"}
It also became a genuine test: the model could now predict what happens when
the school consolidates onto one site.
:::

## Worked example: a question only a human asks {.smaller}

Families choose schools on the published Attainment 8 score — which mostly
measures the intake, not the teaching.

I asked: **what if they chose on value added instead?**

::::: columns
::: {.column width="50%"}
That is not a coding task. It is a question about how a city might behave
differently, and it took about a day to answer properly: a new attractiveness
basis, a switch in the simulator, a measure of how evenly children would be
spread.
:::

::: {.column width="50%"}
::: key-finding
The answer: one struggling school's intake nearly doubles, the surplus stops
concentrating in three schools — and the total number of empty places does
not move at all.
:::
:::::

## Worked example: catching it being wrong {.smaller}

::::: columns
::: {.column width="54%"}
Claude reported a school's value-added score as being in the "top 100% in
England". It had applied an Attainment 8 percentile function to a
value-added figure. Silently wrong, confidently stated.

Later, the attainment charts appeared to show huge jumps when the
value-added option was selected — because they were reading a different
variable than I thought.

I caught both. It fixed both in minutes.
:::

::: {.column width="46%"}
::: key-finding
The failure mode is not gibberish. It is **plausible, well-formatted,
confidently-presented wrongness** — exactly what a domain expert catches
and a non-expert does not.
:::

This is the strongest argument I have for why the human in this loop has to
be someone who knows the subject.
:::
:::::

## The other failure modes, honestly {.smaller}

| What went wrong | What it cost |
|:--|:--|
| A shell quoting slip corrupted two files | 20 minutes, fixed by regenerating |
| An optimiser fell into the wrong basin | Caught by a sanity check on one school's intake |
| A variable name collision inside the app | Silent wrong behaviour until tested |
| Model outputs stale after re-calibration | Needed a discipline of re-running everything |
| Confident summaries of work not yet verified | Needed me to ask "did you actually check?" |

::: {style="font-size:0.78em"}
None of these are exotic. They are the ordinary failure modes of a fast,
capable, over-confident collaborator — and they are managed the same way you
manage those in a person: tests, checks, and someone reading the output.
:::

## The workflow {.smaller}

::::: columns
::: {.column width="52%"}
Everything is **Quarto**. The analysis document, the technical guide, the
slide decks — all `.qmd`, all executing live R against the same data.

The loop, many times a day:

1. I say what I want, usually in two or three sentences
2. Claude edits the source and runs it
3. It renders, screenshots, checks the numbers
4. It commits and pushes
5. I read the rendered output and react
:::

::: {.column width="48%"}
::: key-finding
The document **is** the analysis. There is no separate "writing up" —
changing the model changes the paragraph that describes it, because the
paragraph is generated from the model.
:::

Git is the memory. 805 commits, each with a message explaining why.
:::
:::::

## Why Quarto matters more than it sounds {.smaller}

When I asked "does the Hove Park change affect section 5.5?", that was
answerable — because section 5.5 is code, and the code re-runs.

When I asked for the deck to be rebuilt with the new numbers, no one had to
re-type any numbers.

::: key-finding
Reproducibility stops being a virtue you aspire to and becomes the only way
the thing can work at this speed. If the outputs were hand-copied into Word,
none of this would survive its own rate of change.
:::

## The recursion {.smaller}

The interesting part is not that it wrote code. It is the **loop**:

::: incremental
1. A model produces a result.
2. The result goes into a document, written to be read by councillors.
3. Writing it clearly exposes a hole in the argument.
4. The hole becomes a new question for the model.
5. Answering it changes the model — and the document rewrites itself.
:::

. . .

That cycle ran perhaps two hundred times. Each turn took minutes, not weeks.
**That compression is the whole story.**

## Eighteen days {.smaller}

```{r fig-sept}
#| fig-width: 11
#| fig-height: 4.0
sept <- rec |> filter(day >= as.Date("2026-09-07"))
ggplot(sept, aes(day, tool_calls)) +
  geom_col(fill = BRIGHT, width = 0.65) +
  geom_text(aes(label = tool_calls), vjust = -0.5, size = 4.2, colour = INK) +
  scale_x_date(date_breaks = "3 days", date_labels = "%d %b") +
  scale_y_continuous(limits = c(0, 1300)) +
  labs(x = NULL, y = "Tool calls",
       title = "From nothing to a working city model",
       subtitle = "8 September: first commit. 26 September: model, site, simulator, three decks.") +
  theme_talk(14)
```

## What I would say to a colleague {.smaller}

::::: columns
::: {.column width="50%"}
**It does not replace knowing things.** Every moment that mattered in this
project turned on domain knowledge I had and it did not.

**It does collapse the cost of trying.** The scenarios I tested would each
have been a week of work. Testing twenty of them was previously impossible,
so I would have tested two.
:::

::: {.column width="50%"}
**It writes better documentation than I do**, because it never gets bored
of writing it.

**It needs supervision that looks like peer review**, not like proofreading.

**The bottleneck moved.** It is no longer implementation. It is knowing
which question is worth asking.
:::
:::::

## Open questions {.smaller}

::: incremental
- **Authorship.** What is the honest way to describe this on a paper? I directed
  every step and verified every output — but I did not write the code.
- **Verification at speed.** I can generate analysis faster than I can
  properly check it. That is a new kind of risk.
- **Reproducibility of the process.** The code is reproducible. The
  *conversation* that produced it is not, in any meaningful sense.
- **Who else can do this?** This worked because I could tell right from wrong.
  What happens when the person in the loop cannot?
:::

## Links

::: {style="font-size:0.9em"}
- The analysis — <https://adamdennett.github.io/bh-school-system/>
- The simulator — <https://adamdennett-bh-school-system.share.connect.posit.cloud/>
- The national report — <https://adamdennett.github.io/school_attainment_tool/>
- Everything is open source on GitHub
:::

::: key-finding
Happy to be argued with — particularly about the open questions.
:::
