---
title: "Data Overview: School Attainment Analysis"
subtitle: "Sources, construction, and descriptive statistics for the KS4 panel dataset"
author: "Adam Dennett with some assistance from Claude Code"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    fig-width: 9
    fig-height: 6
    fig-dpi: 300
    df-print: kable
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 5
    geometry: margin=2.5cm
    keep-tex: false
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(here)
library(naniar)

source(here::here("R", "helpers.R"))

# Load final panel and intermediate data
panel_data <- readRDS(here::here("data", "panel_data.rds"))

# Ensure factor ordering
panel_data <- panel_data %>%
  mutate(
    year_label = factor(year_label,
                        levels = c("2021-22", "2022-23", "2023-24", "2024-25")),
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name = factor(gor_name)
  )

# Analysis sample: restrict to Academies and Maintained schools only
# (matches the filter in 05_fit_model.R prepare_model_data())
model_panel <- panel_data %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school"))

# Publication theme
theme_pub <- theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    strip.text    = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

# Colour palette
year_cols <- c("2021-22" = "#1b9e77", "2022-23" = "#d95f02",
               "2023-24" = "#7570b3", "2024-25" = "#e7298a")

# Output directory for saved figures
fig_dir <- here::here("output", "figures")
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)

# Output directory for saved tables
tbl_dir <- here::here("output", "tables")
dir.create(tbl_dir, showWarnings = FALSE, recursive = TRUE)

# Helper: save a ggplot as both PDF and JPEG
save_fig <- function(p, name, width = 7, height = 5) {
  ggsave(file.path(fig_dir, paste0(name, ".pdf")),  p,
         width = width, height = height, device = "pdf")
  ggsave(file.path(fig_dir, paste0(name, ".jpeg")), p,
         width = width, height = height, dpi = 300)
  invisible(p)
}

# Helper: save a kableExtra table as standalone HTML and PNG
save_tbl <- function(tbl, name) {
  save_kable(tbl, file = file.path(tbl_dir, paste0(name, ".html")),
             self_contained = TRUE)
  save_kable(tbl, file = file.path(tbl_dir, paste0(name, ".png")),
             zoom = 2, density = 300)
  tbl
}
```


# Introduction

This document describes the construction and characteristics of the school-level panel dataset used in the multilevel regression analysis of KS4 Attainment 8 scores. It covers:

1. **Data sources** — where each variable originates
2. **Panel construction** — how per-year data are harmonised and linked
3. **Cleaning and imputation** — handling of missing values and the 2024--25 carry-forward strategy
4. **Descriptive statistics** — distributions, correlations, and completeness of the analysis variables

All figures are saved as publication-ready PDF and JPEG files in `output/data_overview_files/`.


# Data Sources

@tbl-sources summarises the six primary data feeds that are combined into the final panel.

```{r tbl-sources}
#| label: tbl-sources
#| tbl-cap: "Primary data sources and their contribution to the panel"

tibble::tribble(
  ~Source, ~Provider, ~Coverage, ~`Key variables`,
  "KS4 Performance Tables", "DfE", "4 years (2021-22 to 2024-25)",
    "ATT8SCR, P8MEA, PTFSM6CLA1A, PTPRIORLO, ADMPOL_PT, TOTPUPS",
  "School Census", "DfE", "4 years (matched to KS4)",
    "PNUMEAL, PNUMENGFL, TPUP, TFSM6CLA1A",
  "Absence (perf. tables + API)", "DfE / EES", "4 years",
    "PERCTOT, PPERSABS10",
  "Ofsted Ratings", "Ofsted Official Statistics", "As at 31 Aug 2024",
    "OFSTEDRATING_1 (harmonised 1\u20134)",
  "Edubase", "DfE", "Oct 2024 snapshot",
    "gor_name (region), easting, northing",
  "School Workforce Census", "DfE / EES", "4 years (2024-25 delayed)",
    "Teacher retention, leadership pay %, sickness days"
) %>%
  kable(align = "llll") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  save_tbl("data_sources")
```


# Variable Definitions

@tbl-var-definitions lists all variables used in the multilevel model, together with their role and a brief description. This serves as a reference for the sections that follow.

```{r tbl-var-definitions}
#| label: tbl-var-definitions
#| tbl-cap: "Definitions of all variables used in the multilevel model"

model_vars_def <- c(
  "ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A",
  "PTFSM6CLA1A", "PERCTOT", "PNUMEAL", "PTPRIORLO",
  "ADMPOL_PT", "gorard_segregation",
  "OFSTEDRATING_1", "gor_name", "LANAME",
  "remained_in_the_same_school",
  "teachers_on_leadership_pay_range_percent",
  "average_number_of_days_taken",
  "has_imputed_predictors"
)

tibble(
  Variable = model_vars_def,
  Label = var_labels(model_vars_def),
  Role = c(rep("Outcome", 3),
           rep("Fixed effect", 6),
           rep("Random intercept", 3),
           rep("Fixed effect", 3),
           "Flag")
) %>%
  kable(align = "lll") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  pack_rows("Outcomes", 1, 3) %>%
  pack_rows("Fixed-effect predictors", 4, 9) %>%
  pack_rows("Random-effect grouping", 10, 12) %>%
  pack_rows("Workforce predictors", 13, 15) %>%
  pack_rows("Metadata", 16, 16) %>%
  save_tbl("variable_definitions")
```


# Panel Construction Pipeline

The panel is assembled through four sequential scripts, each adding data and producing an intermediate `.rds` file.

```{r tbl-pipeline}
#| label: tbl-pipeline
#| tbl-cap: "Data pipeline steps from raw ZIP files to the final analysis panel"

tibble::tribble(
  ~Step, ~Script, ~`Input`, ~`Output`, ~Description,
  "1", "01_extract_data.R",
    "Raw DfE ZIPs", "year_data_list.rds",
    "Unzip, read KS4 + census + absence + school info per year; join by URN; filter RECTYPE=1 (mainstream)",
  "2", "02_build_panel.R",
    "year_data_list.rds", "panel_raw.rds",
    "Download Edubase (region/coords), Ofsted ratings, absence API; join to each year; harmonise Ofsted text/numeric; stack into panel",
  "3", "03_add_workforce.R",
    "panel_raw.rds", "panel_with_workforce.rds",
    "Download 5 workforce datasets from EES (turnover, pay, sickness, size, ratios); join by URN + time_period",
  "4", "04_compute_derived.R",
    "panel_with_workforce.rds", "panel_data.rds",
    "Gorard segregation index; impute 2024-25 missing predictors; log transforms; final cleaning"
) %>%
  kable(align = "clllll") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 10) %>%
  column_spec(5, width = "25em") %>%
  save_tbl("pipeline_steps")
```


# Sample Overview

## Panel Dimensions

```{r tbl-sample-summary}
#| label: tbl-sample-summary
#| tbl-cap: "Panel dimensions by academic year"

sample_tbl <- panel_data %>%
  group_by(`Academic year` = year_label) %>%
  summarise(
    Schools = n_distinct(URN),
    `School-year obs.` = n(),
    LAs = n_distinct(LANAME),
    Regions = n_distinct(gor_name),
    .groups = "drop"
  )

totals <- tibble(
  `Academic year` = "Total / Overall",
  Schools = n_distinct(panel_data$URN),
  `School-year obs.` = nrow(panel_data),
  LAs = n_distinct(panel_data$LANAME),
  Regions = n_distinct(panel_data$gor_name)
)

bind_rows(sample_tbl, totals) %>%
  kable(align = "lrrrr", format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(nrow(sample_tbl) + 1, bold = TRUE) %>%
  save_tbl("panel_dimensions")
```


## Geographic Distribution

```{r fig-region-counts}
#| label: fig-region-counts
#| fig-cap: "Number of schools per region and academic year"
#| fig-height: 5

p_region <- panel_data %>%
  filter(!is.na(gor_name)) %>%
  count(gor_name, year_label) %>%
  ggplot(aes(x = fct_reorder(gor_name, n, .fun = sum), y = n, fill = year_label)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = year_cols, name = "Academic year") +
  labs(x = NULL, y = "Number of schools",
       title = "Schools per region by academic year") +
  theme_pub

save_fig(p_region, "region_counts", width = 8, height = 5)
p_region
```

The South East, North West, and London contribute the largest number of schools, consistent with their population sizes. School counts are highly stable across years, reflecting the fact that the panel tracks the same institutions over time with minimal entry or exit.

## School Type Composition

```{r fig-school-types}
#| label: fig-school-types
#| fig-cap: "School type composition across the panel"
#| fig-height: 4

p_types <- panel_data %>%
  filter(!is.na(MINORGROUP)) %>%
  count(MINORGROUP) %>%
  mutate(pct = n / sum(n) * 100,
         MINORGROUP = fct_reorder(MINORGROUP, n)) %>%
  ggplot(aes(x = MINORGROUP, y = pct)) +
  geom_col(fill = "#3575b5", width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", pct)), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(x = NULL, y = "% of school-year observations",
       title = "School type composition (all years pooled)") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_pub

save_fig(p_types, "school_types", width = 7, height = 4)
p_types
```

Academies form the largest group in the full panel, followed by Maintained schools and then Independent schools. The Independent school share --- while non-trivial in the raw data --- is excluded from the analysis sample for the reasons described in the callout below.

## Ofsted Rating Distribution

```{r fig-ofsted-dist}
#| label: fig-ofsted-dist
#| fig-cap: "Ofsted rating distribution by academic year"
#| fig-height: 4.5

p_ofsted <- model_panel %>%
  filter(!is.na(OFSTEDRATING_1)) %>%
  count(year_label, OFSTEDRATING_1) %>%
  group_by(year_label) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup() %>%
  ggplot(aes(x = year_label, y = pct, fill = OFSTEDRATING_1)) +
  geom_col(position = "stack", width = 0.65) +
  geom_text(aes(label = sprintf("%.0f%%", pct)),
            position = position_stack(vjust = 0.5), size = 3, colour = "white") +
  scale_fill_manual(
    values = c("Outstanding" = "#1a9641", "Good" = "#a6d96a",
               "Requires Improvement" = "#fdae61", "Inadequate" = "#d7191c"),
    name = "Ofsted rating"
  ) +
  labs(x = "Academic year", y = "% of schools",
       title = "Distribution of Ofsted ratings by year") +
  theme_pub

save_fig(p_ofsted, "ofsted_distribution", width = 7, height = 4.5)
p_ofsted
```

The vast majority of schools in the analysis sample are rated Good, followed by Outstanding. The proportion rated Requires Improvement or Inadequate is small and relatively stable across years. Because Ofsted ratings are carried forward until a school is re-inspected, the year-to-year changes are modest and reflect the pace of the inspection cycle rather than rapid shifts in school quality.

## Admissions Policy

```{r fig-admissions}
#| label: fig-admissions
#| fig-cap: "Admissions policy distribution"
#| fig-height: 3.5

if ("ADMPOL_PT" %in% names(model_panel)) {
  p_adm <- model_panel %>%
    filter(!is.na(ADMPOL_PT)) %>%
    mutate(ADMPOL_PT = factor(ADMPOL_PT)) %>%
    count(year_label, ADMPOL_PT) %>%
    group_by(year_label) %>%
    mutate(pct = n / sum(n) * 100) %>%
    ggplot(aes(x = year_label, y = pct, fill = ADMPOL_PT)) +
    geom_col(position = "stack", width = 0.65) +
    scale_fill_brewer(palette = "Set2", name = "Admissions policy") +
    labs(x = "Academic year", y = "% of schools",
         title = "Admissions policy composition by year") +
    theme_pub

  save_fig(p_adm, "admissions_policy", width = 7, height = 3.5)
  p_adm
}
```

The admissions policy variable captures whether a school selects pupils on academic ability. The overwhelming majority of mainstream state-funded schools operate non-selective admissions; the small selective share is concentrated in areas that retain grammar schools. This variable is included in the model to control for the compositional advantage that selective schools enjoy.

# Outcome Variables

::: {.callout-note}
## Analysis sample
From this section onward, all statistics, figures, and tables are based on the **analysis sample** of mainstream state-funded schools only --- Academies and Maintained schools. Independent schools (`r format(sum(panel_data$MINORGROUP == "Independent school", na.rm = TRUE), big.mark = ",")` observations), colleges (`r sum(panel_data$MINORGROUP == "College", na.rm = TRUE)`), and special schools (`r sum(panel_data$MINORGROUP == "Special school", na.rm = TRUE)`) are excluded because they operate under different funding, governance, and accountability frameworks that make direct comparison with the mainstream sector inappropriate. This reduces the panel from `r format(nrow(panel_data), big.mark = ",")` to **`r format(nrow(model_panel), big.mark = ",")`** school-year observations. Full details of the restriction are given in @tbl-school-type-filter.
:::

## Descriptive Statistics

```{r tbl-outcome-stats}
#| label: tbl-outcome-stats
#| tbl-cap: "Descriptive statistics for Attainment 8 outcome variables by academic year"

outcome_vars <- c("ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A")

outcome_stats <- model_panel %>%
  pivot_longer(all_of(outcome_vars), names_to = "Variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  group_by(Variable, `Year` = year_label) %>%
  summarise(
    N     = n(),
    Mean  = round(mean(value), 1),
    SD    = round(sd(value), 1),
    Min   = round(min(value), 1),
    Median = round(median(value), 1),
    Max   = round(max(value), 1),
    .groups = "drop"
  ) %>%
  mutate(Variable = var_label(Variable))

outcome_stats %>%
  kable(align = "llrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top") %>%
  save_tbl("outcome_statistics")
```

Mean Attainment 8 scores for all pupils sit in the mid-to-high 40s and have edged upward over the four-year window. The disadvantaged subgroup consistently scores around 10 points below the non-disadvantaged subgroup, and its standard deviation is slightly wider, reflecting greater heterogeneity in outcomes for this group. Scores for all three measures are approximately normally distributed, so the log transformation applied to the outcome in the multilevel model serves primarily to stabilise variance across the fitted-value range rather than to correct severe skew.

## Outcome Distributions

```{r fig-outcome-density}
#| label: fig-outcome-density
#| fig-cap: "Density plots of Attainment 8 scores by pupil group and year"
#| fig-height: 6

outcome_long <- model_panel %>%
  select(year_label, all_of(outcome_vars)) %>%
  pivot_longer(-year_label, names_to = "outcome", values_to = "score") %>%
  filter(!is.na(score)) %>%
  mutate(
    outcome = recode(outcome,
      ATT8SCR               = "All pupils",
      ATT8SCR_FSM6CLA1A     = "Disadvantaged",
      ATT8SCR_NFSM6CLA1A    = "Non-disadvantaged"
    ),
    outcome = factor(outcome,
                     levels = c("All pupils", "Disadvantaged", "Non-disadvantaged"))
  )

p_density <- ggplot(outcome_long, aes(x = score, fill = year_label)) +
  geom_density(alpha = 0.45, colour = NA) +
  facet_wrap(~ outcome, ncol = 1, scales = "free_y") +
  scale_fill_manual(values = year_cols, name = "Academic year") +
  labs(x = "Attainment 8 score", y = "Density",
       title = "Distribution of Attainment 8 scores") +
  theme_pub

save_fig(p_density, "outcome_density", width = 8, height = 7)
p_density
```

The density plots confirm that all three Attainment 8 measures are unimodal and broadly symmetric. The disadvantaged subgroup distribution is shifted noticeably to the left, with a slight positive tail. Year-on-year shapes are highly stable, suggesting that the national distribution of school-level attainment has not shifted dramatically over the period.

## Disadvantage Gap Over Time

```{r fig-gap-trend}
#| label: fig-gap-trend
#| fig-cap: "Attainment gap between non-disadvantaged and disadvantaged pupils by year"
#| fig-height: 4.5

gap_data <- model_panel %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
  mutate(gap = ATT8SCR_NFSM6CLA1A - ATT8SCR_FSM6CLA1A)

gap_summary <- gap_data %>%
  group_by(year_label) %>%
  summarise(
    mean_gap   = mean(gap, na.rm = TRUE),
    median_gap = median(gap, na.rm = TRUE),
    sd_gap     = sd(gap, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

p_gap <- ggplot(gap_data, aes(x = year_label, y = gap)) +
  geom_boxplot(aes(fill = year_label), alpha = 0.6, outlier.size = 0.5,
               show.legend = FALSE) +
  geom_point(data = gap_summary, aes(y = mean_gap),
             shape = 18, size = 3, colour = "red") +
  scale_fill_manual(values = year_cols) +
  labs(x = "Academic year",
       y = "ATT8 gap (non-disadvantaged \u2013 disadvantaged)",
       title = "School-level disadvantage gap in Attainment 8",
       subtitle = "Red diamond = annual mean") +
  theme_pub

save_fig(p_gap, "disadvantage_gap_trend", width = 7, height = 4.5)
p_gap
```

The school-level disadvantage gap is remarkably persistent, with median values of around 13--14 Attainment 8 points across all four years. The interquartile range is wide, indicating that some schools narrow the gap far more effectively than others --- a key motivation for the multilevel modelling approach. A small number of schools show negative gaps (i.e. disadvantaged pupils outperforming their non-disadvantaged peers), visible as outliers below zero.

## Outcomes by Region

```{r fig-outcome-region}
#| label: fig-outcome-region
#| fig-cap: "Attainment 8 distributions by region — the grouping variable for the multilevel model's regional random intercept"
#| fig-height: 6

p_outcome_region <- model_panel %>%
  filter(!is.na(gor_name)) %>%
  select(gor_name, all_of(outcome_vars)) %>%
  pivot_longer(-gor_name, names_to = "outcome", values_to = "score") %>%
  filter(!is.na(score)) %>%
  mutate(
    outcome = recode(outcome,
      ATT8SCR            = "All pupils",
      ATT8SCR_FSM6CLA1A  = "Disadvantaged",
      ATT8SCR_NFSM6CLA1A = "Non-disadvantaged"
    ),
    outcome = factor(outcome,
                     levels = c("All pupils", "Disadvantaged", "Non-disadvantaged")),
    gor_name = fct_reorder(gor_name, score, .fun = median)
  )

p_out_reg <- ggplot(p_outcome_region, aes(x = gor_name, y = score, fill = gor_name)) +
  geom_violin(alpha = 0.6, colour = NA, scale = "width") +
  geom_boxplot(width = 0.15, alpha = 0.8, outlier.size = 0.3, show.legend = FALSE) +
  facet_wrap(~ outcome, ncol = 3) +
  coord_flip() +
  labs(x = NULL, y = "Attainment 8 score",
       title = "Attainment 8 by region and pupil group",
       subtitle = "All years pooled") +
  theme_pub +
  theme(legend.position = "none",
        strip.text = element_text(size = 9))

save_fig(p_out_reg, "outcome_by_region", width = 11, height = 6)
p_out_reg
```

London stands out with the highest median scores across all three pupil groups --- a well-documented feature of the English school system. The North East and Yorkshire & The Humber cluster at the lower end, though there is substantial overlap between regions. Notably, regional rankings are broadly consistent across the three outcome groups, suggesting that the "London effect" benefits disadvantaged and non-disadvantaged pupils alike.

## Outcomes by Ofsted Rating

```{r fig-outcome-ofsted}
#| label: fig-outcome-ofsted
#| fig-cap: "Attainment 8 distributions by Ofsted rating — the grouping variable for the multilevel model's Ofsted random intercept"
#| fig-height: 5

p_outcome_ofsted <- model_panel %>%
  filter(!is.na(OFSTEDRATING_1)) %>%
  select(OFSTEDRATING_1, all_of(outcome_vars)) %>%
  pivot_longer(-OFSTEDRATING_1, names_to = "outcome", values_to = "score") %>%
  filter(!is.na(score)) %>%
  mutate(
    outcome = recode(outcome,
      ATT8SCR            = "All pupils",
      ATT8SCR_FSM6CLA1A  = "Disadvantaged",
      ATT8SCR_NFSM6CLA1A = "Non-disadvantaged"
    ),
    outcome = factor(outcome,
                     levels = c("All pupils", "Disadvantaged", "Non-disadvantaged"))
  )

ofsted_cols <- c("Outstanding" = "#1b7e2e", "Good" = "#6aaa47",
                 "Requires Improvement" = "#e8a838", "Inadequate" = "#c62828")

p_out_ofs <- ggplot(p_outcome_ofsted,
                    aes(x = OFSTEDRATING_1, y = score, fill = OFSTEDRATING_1)) +
  geom_violin(alpha = 0.6, colour = NA, scale = "width") +
  geom_boxplot(width = 0.15, alpha = 0.8, outlier.size = 0.3, show.legend = FALSE) +
  facet_wrap(~ outcome, ncol = 3) +
  scale_fill_manual(values = ofsted_cols) +
  labs(x = NULL, y = "Attainment 8 score",
       title = "Attainment 8 by Ofsted rating and pupil group",
       subtitle = "All years pooled") +
  theme_pub +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1, size = 8),
        strip.text = element_text(size = 9))

save_fig(p_out_ofs, "outcome_by_ofsted", width = 10, height = 5)
p_out_ofs
```

There is a clear gradient in attainment across Ofsted categories: Outstanding schools have the highest median scores and the tightest distributions, while Inadequate schools sit noticeably lower with wider spread. The step from Good to Requires Improvement is the sharpest for all three pupil groups.

::: {.callout-tip}
## Relation to the multilevel null model
These two plots illustrate the structure captured by the **random intercepts** in the multilevel model. Region (`gor_name`) and Ofsted rating (`OFSTEDRATING_1`) each enter as grouping variables, allowing the model to estimate a separate baseline (intercept) for each level. Before any school-level predictors are added, these group-level intercepts absorb the systematic variation visible in the violin plots above --- effectively representing a **null model** that accounts only for which region and Ofsted category a school belongs to. The fixed-effect predictors (deprivation, absence, EAL, etc.) then explain the remaining within-group variation around these baselines.
:::

# Predictor Variables

The multilevel model includes eight fixed-effect predictors capturing school composition, pupil characteristics, and workforce conditions. This section summarises their distributions and motivates the log transformations applied to the most heavily skewed variables.

## Descriptive Statistics

```{r tbl-predictor-stats}
#| label: tbl-predictor-stats
#| tbl-cap: "Descriptive statistics for predictor variables used in the multilevel model"

pred_vars <- c(
  "PTFSM6CLA1A", "PERCTOT", "PNUMEAL", "PTPRIORLO",
  "gorard_segregation",
  "remained_in_the_same_school",
  "teachers_on_leadership_pay_range_percent",
  "average_number_of_days_taken"
)
pred_vars <- intersect(pred_vars, names(model_panel))

pred_table <- map_dfr(pred_vars, function(v) {
  x <- model_panel[[v]]
  tibble(
    Variable = var_label(v),
    N        = sum(!is.na(x)),
    `% available` = round(sum(!is.na(x)) / nrow(model_panel) * 100, 1),
    Mean     = round(mean(x, na.rm = TRUE), 2),
    SD       = round(sd(x, na.rm = TRUE), 2),
    Min      = round(min(x, na.rm = TRUE), 2),
    Median   = round(median(x, na.rm = TRUE), 2),
    Max      = round(max(x, na.rm = TRUE), 2)
  )
})

pred_table %>%
  kable(align = "lrrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  pack_rows("School composition", 1, 4) %>%
  pack_rows("Segregation (LA-level)", 5, 5) %>%
  pack_rows("Workforce", 6, nrow(pred_table)) %>%
  save_tbl("predictor_statistics")
```

There is wide variation across schools in most composition variables. The proportion of disadvantaged pupils (PTFSM6CLA1A) averages around 30% but ranges from near-zero to over 90%, reflecting the substantial segregation that exists within the English school system. Overall absence rates (PERCTOT) are tightly clustered for most schools but exhibit a long right tail of high-absence outliers. The workforce variables --- teacher retention, leadership pay share, and average sickness days --- have somewhat lower coverage (particularly for 2024--25, which relies on carry-forward imputation; see @tbl-imputation).

## Predictor Distributions

```{r fig-predictor-hist}
#| label: fig-predictor-hist
#| fig-cap: "Distributions of key predictor variables (raw scale)"
#| fig-height: 8

pred_long <- model_panel %>%
  select(year_label, all_of(pred_vars)) %>%
  pivot_longer(-year_label, names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(variable = var_label(variable))

p_hist <- ggplot(pred_long, aes(x = value)) +
  geom_histogram(aes(fill = year_label), bins = 40, alpha = 0.65,
                 position = "identity", colour = NA) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = year_cols, name = "Academic year") +
  labs(x = NULL, y = "Frequency",
       title = "Predictor variable distributions") +
  theme_pub +
  theme(strip.text = element_text(size = 8))

save_fig(p_hist, "predictor_distributions", width = 10, height = 8)
p_hist
```

Several predictors exhibit clear right skew --- notably the proportion of pupils with English as an additional language (PNUMEAL) and average teacher sickness days. The FSM eligibility rate and overall absence rate are also right-skewed, though less severely. By contrast, the Gorard segregation index and teacher retention FTE are approximately symmetric. These distributional differences motivate the selective log-transformation strategy described below.

## Log-Transformed Predictors

Predictors with |skewness| > 1 are log-transformed for the multilevel model. `remained_in_the_same_school` (skewness = 0.47) is entered on the raw scale as it is near-symmetric. @fig-log-compare shows the raw and log-transformed distributions for the variables that undergo this transformation.

```{r fig-log-compare}
#| label: fig-log-compare
#| fig-cap: "Raw vs. log-transformed distributions for skewed predictor variables"
#| fig-height: 7

log_vars <- c("PTFSM6CLA1A", "PERCTOT", "PNUMEAL",
              "average_number_of_days_taken")
log_vars <- intersect(log_vars, names(model_panel))

raw_long <- model_panel %>%
  select(all_of(log_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  filter(!is.na(value), value > 0) %>%
  mutate(scale = "Raw")

log_long <- model_panel %>%
  select(all_of(log_vars)) %>%
  mutate(across(everything(), safe_log)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(scale = "Log-transformed")

compare_long <- bind_rows(raw_long, log_long) %>%
  mutate(
    variable = var_label(variable),
    scale = factor(scale, levels = c("Raw", "Log-transformed"))
  )

p_log <- ggplot(compare_long, aes(x = value)) +
  geom_histogram(bins = 40, fill = "#3575b5", alpha = 0.7, colour = "white",
                 linewidth = 0.2) +
  facet_grid(variable ~ scale, scales = "free") +
  labs(x = NULL, y = "Frequency",
       title = "Effect of log transformation on skewed predictors") +
  theme_pub +
  theme(strip.text = element_text(size = 9))

save_fig(p_log, "log_transform_comparison", width = 8, height = 7)
p_log
```

The side-by-side comparison confirms that the log transformation substantially reduces skewness for all four treated variables. After transformation, the distributions are much closer to the normal shape assumed by the linear mixed-effects model, improving the reliability of coefficient estimates and standard errors.

# Bivariate Relationships

The interactive scatter plot below allows exploration of the bivariate relationships between each Attainment 8 outcome and the predictor variables. Use the dropdowns to select the outcome (y-axis) and predictor (x-axis), and the checkboxes to independently toggle log-transformation on each axis. This lets you compare the raw-scale relationship with the model-scale relationship (where several variables enter as logs). The OLS regression line and R² update in real time.

```{r fig-bivariate-overview}
#| label: fig-bivariate-overview

library(plotly)
library(htmltools)
library(jsonlite)

# --- Build a clean data frame with both raw and log versions ---
bivar_df <- model_panel %>%
  filter(!is.na(ATT8SCR)) %>%
  mutate(
    log_ATT8SCR            = safe_log(ATT8SCR),
    log_ATT8SCR_FSM6CLA1A  = safe_log(ATT8SCR_FSM6CLA1A),
    log_ATT8SCR_NFSM6CLA1A = safe_log(ATT8SCR_NFSM6CLA1A),
    log_PTFSM6CLA1A        = safe_log(PTFSM6CLA1A),
    log_PERCTOT             = safe_log(PERCTOT),
    log_PNUMEAL             = safe_log(PNUMEAL),
    log_PTPRIORLO           = safe_log(PTPRIORLO),
    log_gorard_segregation  = safe_log(gorard_segregation),
    log_remained            = safe_log(remained_in_the_same_school),
    log_leadership_pay      = safe_log(teachers_on_leadership_pay_range_percent),
    log_sickness_days       = safe_log(average_number_of_days_taken)
  )

# Hover label
bivar_df$hover_label <- if ("SCHNAME" %in% names(bivar_df) && "LANAME" %in% names(bivar_df)) {
  paste0(bivar_df$SCHNAME, " (", bivar_df$LANAME, ")")
} else {
  paste0("URN ", bivar_df$URN)
}

# --- Define variable metadata ---
# Each entry: raw column, log column, display label
outcome_meta <- list(
  list(raw = "ATT8SCR",             log = "log_ATT8SCR",
       label = "Attainment 8 — All Pupils"),
  list(raw = "ATT8SCR_FSM6CLA1A",   log = "log_ATT8SCR_FSM6CLA1A",
       label = "Attainment 8 — Disadvantaged"),
  list(raw = "ATT8SCR_NFSM6CLA1A",  log = "log_ATT8SCR_NFSM6CLA1A",
       label = "Attainment 8 — Non-Disadvantaged")
)

pred_meta <- list(
  list(raw = "PTFSM6CLA1A", log = "log_PTFSM6CLA1A",
       label = "% Disadvantaged (FSM)"),
  list(raw = "PERCTOT", log = "log_PERCTOT",
       label = "% Overall Absence"),
  list(raw = "PNUMEAL", log = "log_PNUMEAL",
       label = "% EAL"),
  list(raw = "PTPRIORLO", log = "log_PTPRIORLO",
       label = "% Low Prior Attainment (KS2)"),
  list(raw = "gorard_segregation", log = "log_gorard_segregation",
       label = "Gorard Segregation Index"),
  list(raw = "remained_in_the_same_school", log = "log_remained",
       label = "Teachers Remaining (FTE)"),
  list(raw = "teachers_on_leadership_pay_range_percent",
       log = "log_leadership_pay",
       label = "% Teachers on Leadership Pay"),
  list(raw = "average_number_of_days_taken", log = "log_sickness_days",
       label = "Avg. Teacher Sickness Days")
)

# Filter out predictors with no data
pred_meta <- Filter(function(p) any(!is.na(bivar_df[[p$raw]])), pred_meta)

# Split by year
years <- as.character(sort(unique(bivar_df$year_label)))
year_dfs <- lapply(years, function(yr) bivar_df[bivar_df$year_label == yr, ])
names(year_dfs) <- years

yr_colours <- c("#1565c0", "#c62828", "#2e7d32", "#e65100")
n_tr <- length(years)

# --- Build base plotly (one trace per year + one OLS line) ---
fig <- plot_ly()
for (i in seq_along(years)) {
  df_i <- year_dfs[[i]]
  fig <- fig %>%
    add_markers(
      x = df_i[[pred_meta[[1]]$raw]],
      y = df_i[[outcome_meta[[1]]$raw]],
      text = df_i[["hover_label"]],
      name = years[i],
      marker = list(color = yr_colours[i], opacity = 0.35, size = 4),
      hovertemplate = paste0(
        "<b>%{text}</b><br>x: %{x:.3f}<br>y: %{y:.2f}",
        "<extra>", years[i], "</extra>"
      )
    )
}

# Regression line trace (hidden until JS computes OLS)
fig <- fig %>%
  add_lines(
    x = c(0, 1), y = c(0, 1),
    name = "OLS fit",
    line = list(color = "#c62828", width = 2),
    showlegend = FALSE, hoverinfo = "skip", visible = FALSE
  )

fig <- fig %>%
  layout(
    xaxis  = list(title = pred_meta[[1]]$label),
    yaxis  = list(title = outcome_meta[[1]]$label),
    legend = list(orientation = "h", x = 1, xanchor = "right",
                  y = 1.02, yanchor = "bottom"),
    margin = list(t = 50, l = 80, b = 60)
  )

fig$elementId <- "bivar-overview"

# --- Pack ALL column data (raw + log) into JSON ---
all_cols <- c(
  unlist(lapply(outcome_meta, function(o) c(o$raw, o$log))),
  unlist(lapply(pred_meta, function(p) c(p$raw, p$log)))
)

js_data <- list()
for (yr in years) {
  yr_list <- list()
  df <- year_dfs[[yr]]
  for (col in all_cols) {
    if (col %in% names(df)) {
      yr_list[[col]] <- round(df[[col]], 5)
    }
  }
  js_data[[yr]] <- yr_list
}

# --- HTML dropdown + checkbox controls ---
select_style <- paste0(
  "padding:6px 10px; border:1px solid #bbb; border-radius:4px;",
  "font-size:0.88em; background:#fff; cursor:pointer;"
)
label_style <- paste0(
  "font-weight:600; font-size:0.82em; display:block;",
  "margin-bottom:4px; color:#555;"
)
cb_style <- paste0(
  "display:flex; align-items:center; gap:6px; margin-top:6px;",
  "font-size:0.82em; color:#555;"
)

controls <- tags$div(
  style = "display:flex; gap:24px; margin-bottom:14px; flex-wrap:wrap; align-items:end;",
  tags$div(
    tags$label("Outcome (y-axis):", `for` = "bvo-outcome", style = label_style),
    tags$select(
      id = "bvo-outcome", style = paste0(select_style, "min-width:230px;"),
      lapply(seq_along(outcome_meta), function(k)
        tags$option(value = k, outcome_meta[[k]]$label))
    ),
    tags$div(style = cb_style,
      tags$input(type = "checkbox", id = "bvo-log-y"),
      tags$label("Log scale", `for` = "bvo-log-y")
    )
  ),
  tags$div(
    tags$label("Predictor (x-axis):", `for` = "bvo-predictor", style = label_style),
    tags$select(
      id = "bvo-predictor", style = paste0(select_style, "min-width:270px;"),
      lapply(seq_along(pred_meta), function(k)
        tags$option(value = k, pred_meta[[k]]$label))
    ),
    tags$div(style = cb_style,
      tags$input(type = "checkbox", id = "bvo-log-x"),
      tags$label("Log scale", `for` = "bvo-log-x")
    )
  ),
  tags$div(
    tags$label("Year:", `for` = "bvo-year", style = label_style),
    tags$select(
      id = "bvo-year", style = paste0(select_style, "min-width:140px;"),
      tags$option(value = "all", "All Years"),
      lapply(years, function(yr) tags$option(value = yr, yr))
    )
  )
)

# --- Metadata arrays for JS ---
outcome_js <- lapply(outcome_meta, function(o)
  list(raw = o$raw, log = o$log, label = o$label))
pred_js <- lapply(pred_meta, function(p)
  list(raw = p$raw, log = p$log, label = p$label))

# --- JavaScript: wire controls -> Plotly.restyle / relayout + live OLS ---
js_block <- tags$script(HTML(sprintf('
(function() {
  var DATA  = %s;
  var YEARS = %s;
  var N     = YEARS.length;
  var LINE  = N;

  var OUTCOMES   = %s;
  var PREDICTORS = %s;

  function fitOLS(xAll, yAll) {
    var x = [], y = [];
    for (var i = 0; i < xAll.length; i++) {
      if (xAll[i] != null && yAll[i] != null &&
          isFinite(xAll[i]) && isFinite(yAll[i])) {
        x.push(xAll[i]); y.push(yAll[i]);
      }
    }
    var n = x.length;
    if (n < 3) return null;
    var sx=0, sy=0, sxx=0, sxy=0, syy=0;
    for (var i = 0; i < n; i++) {
      sx += x[i]; sy += y[i];
      sxx += x[i]*x[i]; sxy += x[i]*y[i]; syy += y[i]*y[i];
    }
    var xb = sx/n, yb = sy/n;
    var Sxx = sxx - n*xb*xb, Sxy = sxy - n*xb*yb, Syy = syy - n*yb*yb;
    if (Sxx === 0) return null;
    var b1 = Sxy / Sxx, b0 = yb - b1 * xb;
    var SSres = 0;
    for (var i = 0; i < n; i++) {
      var r = y[i] - (b0 + b1*x[i]); SSres += r*r;
    }
    var r2 = (Syy > 0) ? 1 - SSres/Syy : 0;
    var xmin = x[0], xmax = x[0];
    for (var i = 1; i < n; i++) {
      if (x[i] < xmin) xmin = x[i];
      if (x[i] > xmax) xmax = x[i];
    }
    return { b0:b0, b1:b1, r2:r2, n:n, xmin:xmin, xmax:xmax };
  }

  function update() {
    var gd = document.getElementById("bivar-overview");
    if (!gd || !gd._fullLayout) return;

    var outIdx  = parseInt(document.getElementById("bvo-outcome").value) - 1;
    var predIdx = parseInt(document.getElementById("bvo-predictor").value) - 1;
    var yrVal   = document.getElementById("bvo-year").value;
    var logX    = document.getElementById("bvo-log-x").checked;
    var logY    = document.getElementById("bvo-log-y").checked;

    var outMeta  = OUTCOMES[outIdx];
    var predMeta = PREDICTORS[predIdx];
    var xCol = logX ? predMeta.log : predMeta.raw;
    var yCol = logY ? outMeta.log  : outMeta.raw;
    var xLabel = (logX ? "log(" : "") + predMeta.label + (logX ? ")" : "");
    var yLabel = (logY ? "log(" : "") + outMeta.label  + (logY ? ")" : "");

    var xArr = [], yArr = [], visArr = [];
    var allX = [], allY = [];
    for (var i = 0; i < N; i++) {
      var yr = YEARS[i];
      var xd = DATA[yr][xCol] || [];
      var yd = DATA[yr][yCol] || [];
      xArr.push(xd); yArr.push(yd);
      var vis = (yrVal === "all" || yr === yrVal);
      visArr.push(vis);
      if (vis) {
        for (var j = 0; j < xd.length; j++) {
          allX.push(xd[j]); allY.push(yd[j]);
        }
      }
    }

    var idx = []; for (var i = 0; i < N; i++) idx.push(i);
    Plotly.restyle(gd, { x: xArr, y: yArr, visible: visArr }, idx);

    var fit = fitOLS(allX, allY);
    if (fit) {
      Plotly.restyle(gd, {
        x: [[ fit.xmin, fit.xmax ]],
        y: [[ fit.b0 + fit.b1*fit.xmin, fit.b0 + fit.b1*fit.xmax ]],
        visible: [true]
      }, [LINE]);

      var sign = (fit.b1 >= 0) ? " + " : " \\u2212 ";
      var eqn  = "\\u0177 = " + fit.b0.toFixed(2) + sign +
                 Math.abs(fit.b1).toFixed(3) + "x" +
                 "\\u2003\\u2003R\\u00b2 = " + fit.r2.toFixed(3) +
                 "\\u2003\\u2003n = " + fit.n.toLocaleString();

      Plotly.relayout(gd, {
        "xaxis.title": xLabel,
        "yaxis.title": yLabel,
        annotations: [{
          text: eqn,
          xref: "paper", yref: "paper",
          x: 0.02, y: 0.98,
          xanchor: "left", yanchor: "top",
          showarrow: false,
          font: { size: 12.5, color: "#c62828", family: "monospace" },
          bgcolor: "rgba(255,255,255,0.88)",
          borderpad: 5
        }]
      });
    } else {
      Plotly.restyle(gd, { visible: [false] }, [LINE]);
      Plotly.relayout(gd, {
        "xaxis.title": xLabel,
        "yaxis.title": yLabel,
        annotations: []
      });
    }
  }

  ["bvo-outcome","bvo-predictor","bvo-year"].forEach(function(id) {
    document.getElementById(id).addEventListener("change", update);
  });
  ["bvo-log-x","bvo-log-y"].forEach(function(id) {
    document.getElementById(id).addEventListener("change", update);
  });

  var poll = setInterval(function() {
    var gd = document.getElementById("bivar-overview");
    if (gd && gd._fullLayout) { clearInterval(poll); update(); }
  }, 250);
})();
',
  toJSON(js_data, digits = 5, na = "null", auto_unbox = FALSE),
  toJSON(years),
  toJSON(outcome_js, auto_unbox = TRUE),
  toJSON(pred_js, auto_unbox = TRUE)
)))

htmltools::tagList(controls, fig, js_block)
```

Toggling the log checkboxes reveals how the transformation reshapes each relationship. For example, the raw `% Disadvantaged` vs raw `Attainment 8` plot shows a curved, heteroscedastic pattern; switching both axes to log scale linearises the relationship and stabilises the variance --- illustrating why the model uses `log(y) ~ log(x)` for these variables.


# Analysis Sample Restriction

The multilevel models are fitted to **mainstream state-funded schools only** (Academies and Maintained schools). Independent schools, colleges, and special schools are excluded because they operate under different funding, governance, and admissions structures that make direct comparison inappropriate.

@tbl-school-type-filter shows the effect of this restriction.

```{r tbl-school-type-filter}
#| label: tbl-school-type-filter
#| tbl-cap: "Sample restriction by school type (MINORGROUP)"

type_summary <- panel_data %>%
  mutate(
    in_model = MINORGROUP %in% c("Academy", "Maintained school"),
    MINORGROUP = ifelse(is.na(MINORGROUP), "(Missing)", MINORGROUP)
  ) %>%
  count(MINORGROUP, in_model) %>%
  arrange(desc(n)) %>%
  mutate(
    `% of panel` = round(n / sum(n) * 100, 1),
    Status = ifelse(in_model, "Included", "Excluded")
  ) %>%
  select(`School type` = MINORGROUP, `School-year obs.` = n,
         `% of panel`, Status)

type_summary %>%
  kable(align = "lrrl") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(which(type_summary$Status == "Excluded"),
           color = "#c62828", italic = TRUE) %>%
  save_tbl("school_type_filter")
```

```{r tbl-model-sample}
#| label: tbl-model-sample
#| tbl-cap: "Analysis sample dimensions after excluding non-mainstream schools"

model_summary <- model_panel %>%
  group_by(`Academic year` = year_label) %>%
  summarise(
    Schools = n_distinct(URN),
    `School-year obs.` = n(),
    LAs = n_distinct(LANAME),
    .groups = "drop"
  )

totals <- tibble(
  `Academic year` = "Total / Overall",
  Schools = n_distinct(model_panel$URN),
  `School-year obs.` = nrow(model_panel),
  LAs = n_distinct(model_panel$LANAME)
)

bind_rows(model_summary, totals) %>%
  kable(align = "lrrr", format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(nrow(model_summary) + 1, bold = TRUE) %>%
  save_tbl("analysis_sample_dimensions")
```

All descriptive statistics and figures from this point forward are based on the **analysis sample** (`r format(nrow(model_panel), big.mark = ",")` school-year observations from `r n_distinct(model_panel$URN)` mainstream state-funded schools).


# Missing Data

Understanding the pattern of missing data is important for assessing whether the analysis sample is representative and whether the imputation strategy (described in the next section) is appropriate.

## Completeness by Variable and Year

```{r tbl-completeness}
#| label: tbl-completeness
#| tbl-cap: "Data completeness (% non-missing) by variable and academic year"

all_model_vars <- c(
  "ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A",
  "PTFSM6CLA1A", "PERCTOT", "PPERSABS10", "PNUMEAL", "PTPRIORLO",
  "ADMPOL_PT", "gorard_segregation", "OFSTEDRATING_1",
  "remained_in_the_same_school",
  "teachers_on_leadership_pay_range_percent",
  "average_number_of_days_taken",
  "gor_name", "LANAME"
)
all_model_vars <- intersect(all_model_vars, names(model_panel))

completeness <- model_panel %>%
  group_by(year_label) %>%
  summarise(
    across(all_of(all_model_vars),
           ~ round(sum(!is.na(.)) / n() * 100, 1),
           .names = "{.col}"),
    .groups = "drop"
  ) %>%
  pivot_longer(-year_label, names_to = "Variable", values_to = "pct") %>%
  pivot_wider(names_from = year_label, values_from = pct) %>%
  mutate(Variable = var_label(Variable))

completeness %>%
  kable(align = "lrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  save_tbl("data_completeness")
```

Core attainment and composition variables (Attainment 8, FSM eligibility, absence, EAL) enjoy near-complete coverage across all four years. The notable exception is the workforce block --- teacher retention, leadership pay share, and sickness absence --- which drops to 0% availability in 2024--25 because the School Workforce Census for that year had not been published at the time of data extraction. KS2 prior attainment (PTPRIORLO) is also missing for 2024--25 due to the lingering effects of COVID-era assessment cancellations.

## Missing Data Heatmap

```{r fig-missing-heatmap}
#| label: fig-missing-heatmap
#| fig-cap: "Missing data pattern across model variables by academic year"
#| fig-height: 5

miss_long <- model_panel %>%
  select(year_label, all_of(all_model_vars)) %>%
  group_by(year_label) %>%
  summarise(
    across(everything(), ~ sum(is.na(.)) / n() * 100),
    .groups = "drop"
  ) %>%
  pivot_longer(-year_label, names_to = "variable", values_to = "pct_missing") %>%
  mutate(variable = var_label(variable))

p_miss <- ggplot(miss_long, aes(x = year_label,
                                 y = fct_reorder(variable, pct_missing, .fun = max),
                                 fill = pct_missing)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.0f%%", pct_missing)), size = 3) +
  scale_fill_gradient(low = "#f7fbff", high = "#c62828",
                      name = "% missing", limits = c(0, 100)) +
  labs(x = "Academic year", y = NULL,
       title = "Missing data by variable and year") +
  theme_pub +
  theme(panel.grid = element_blank())

save_fig(p_miss, "missing_heatmap", width = 8, height = 5)
p_miss
```

The heatmap makes the pattern stark: missingness is concentrated almost entirely in the workforce and prior attainment variables for the 2024--25 year. For the earlier three years, data completeness is consistently high across all variables. This structured, year-specific pattern supports the use of a carry-forward imputation strategy rather than more complex multiple imputation approaches.

# Imputation: 2024--25 Carry-Forward

Four predictor variables are unavailable for the 2024--25 academic year:

- **PTPRIORLO** — KS2 prior attainment was cancelled due to the lingering effects of COVID-19 disruption
- **remained_in_the_same_school** — workforce turnover data not yet published
- **teachers_on_leadership_pay_range_percent** — workforce pay data not yet published
- **average_number_of_days_taken** — workforce sickness data not yet published

These are imputed using a two-step approach:

1. **Carry-forward**: use the school's 2023--24 value
2. **Fallback**: if no 2023--24 value exists, use the school-level mean across all available years

A flag variable `has_imputed_predictors` is set to `TRUE` for any 2024--25 row that received at least one imputed value.

```{r tbl-imputation}
#| label: tbl-imputation
#| tbl-cap: "Imputation summary for 2024-25 rows"

impute_vars <- c("PTPRIORLO", "remained_in_the_same_school",
                 "teachers_on_leadership_pay_range_percent",
                 "average_number_of_days_taken")
impute_vars <- intersect(impute_vars, names(model_panel))

rows_2425 <- model_panel %>% filter(year_label == "2024-25")
n_2425 <- nrow(rows_2425)

imputation_summary <- map_dfr(impute_vars, function(v) {
  n_available <- sum(!is.na(rows_2425[[v]]))
  n_missing   <- n_2425 - n_available
  tibble(
    Variable          = var_label(v),
    `2024-25 rows`    = n_2425,
    `Non-missing`     = n_available,
    `Still missing`   = n_missing,
    `% imputed/available` = round(n_available / n_2425 * 100, 1)
  )
})

flagged <- sum(rows_2425$has_imputed_predictors, na.rm = TRUE)

imputation_summary %>%
  kable(align = "lrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  footnote(general = paste0(flagged, " of ", n_2425,
                            " 2024-25 rows flagged as having imputed predictors (",
                            round(flagged / n_2425 * 100, 1), "%).")) %>%
  save_tbl("imputation_summary")
```


```{r fig-imputed-vs-observed}
#| label: fig-imputed-vs-observed
#| fig-cap: "Comparison of observed (2023-24) vs. carried-forward (2024-25) values for imputed variables"
#| fig-height: 6

if (length(impute_vars) > 0) {
  compare_df <- model_panel %>%
    filter(year_label %in% c("2023-24", "2024-25")) %>%
    select(year_label, all_of(impute_vars)) %>%
    pivot_longer(-year_label, names_to = "variable", values_to = "value") %>%
    filter(!is.na(value)) %>%
    mutate(variable = var_label(variable))

  p_impute <- ggplot(compare_df, aes(x = value, fill = year_label)) +
    geom_density(alpha = 0.5, colour = NA) +
    facet_wrap(~ variable, scales = "free", ncol = 2) +
    scale_fill_manual(
      values = c("2023-24" = "#7570b3", "2024-25" = "#e7298a"),
      name = "Academic year"
    ) +
    labs(x = NULL, y = "Density",
         title = "Observed (2023-24) vs. imputed (2024-25) distributions",
         subtitle = "2024-25 values are carry-forwards from 2023-24 or school-level means") +
    theme_pub

  save_fig(p_impute, "imputed_vs_observed", width = 8, height = 6)
  p_impute
}
```

The density overlays show that the carried-forward 2024--25 distributions closely match the observed 2023--24 distributions, as expected. This provides reassurance that the imputation does not introduce systematic bias into the predictor values. The assumption underpinning this approach is that school-level workforce characteristics are slow-moving and unlikely to change dramatically within a single year.

## Year-on-Year Stability of KS2 Prior Attainment

A key question for the carry-forward strategy is whether imputed variables are sufficiently stable from one year to the next at the school level. @fig-ptpriorlo-scatter addresses this for PTPRIORLO (the percentage of pupils entering with low KS2 prior attainment) by plotting each school's value in one year against its value in the next. If the variable were perfectly stable, all points would fall on the 45-degree line.

```{r fig-ptpriorlo-scatter}
#| label: fig-ptpriorlo-scatter
#| fig-cap: "Year-on-year stability of KS2 prior attainment (% low prior) at the school level"
#| fig-height: 5
#| fig-width: 10

# Build wide data: one row per school with each year's PTPRIORLO
prior_wide <- model_panel %>%
  filter(!is.na(PTPRIORLO)) %>%
  select(URN, year_label, PTPRIORLO) %>%
  pivot_wider(names_from = year_label, values_from = PTPRIORLO,
              names_prefix = "y_")

# Pair 1: 2021-22 vs 2022-23
pair1 <- prior_wide %>%
  filter(!is.na(`y_2021-22`), !is.na(`y_2022-23`))
r1 <- cor(pair1$`y_2021-22`, pair1$`y_2022-23`)

# Pair 2: 2022-23 vs 2023-24
pair2 <- prior_wide %>%
  filter(!is.na(`y_2022-23`), !is.na(`y_2023-24`))
r2 <- cor(pair2$`y_2022-23`, pair2$`y_2023-24`)

p_scatter1 <- ggplot(pair1, aes(x = `y_2021-22`, y = `y_2022-23`)) +
  geom_point(alpha = 0.25, size = 1, colour = "#1b9e77") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey30") +
  annotate("text", x = Inf, y = -Inf,
           label = paste0("r = ", sprintf("%.3f", r1), "\nn = ", format(nrow(pair1), big.mark = ",")),
           hjust = 1.1, vjust = -0.5, size = 3.5, colour = "grey30") +
  labs(x = "% low prior attainment (2021-22)",
       y = "% low prior attainment (2022-23)",
       title = "2021-22 vs 2022-23") +
  coord_equal() +
  theme_pub

p_scatter2 <- ggplot(pair2, aes(x = `y_2022-23`, y = `y_2023-24`)) +
  geom_point(alpha = 0.25, size = 1, colour = "#d95f02") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey30") +
  annotate("text", x = Inf, y = -Inf,
           label = paste0("r = ", sprintf("%.3f", r2), "\nn = ", format(nrow(pair2), big.mark = ",")),
           hjust = 1.1, vjust = -0.5, size = 3.5, colour = "grey30") +
  labs(x = "% low prior attainment (2022-23)",
       y = "% low prior attainment (2023-24)",
       title = "2022-23 vs 2023-24") +
  coord_equal() +
  theme_pub

p_prior_scatter <- p_scatter1 + p_scatter2 +
  plot_annotation(
    title = "Year-on-year stability of PTPRIORLO at the school level",
    subtitle = "Dashed line = perfect carry-forward (y = x)",
    theme = theme(plot.title = element_text(face = "bold", size = 12),
                  plot.subtitle = element_text(colour = "grey40", size = 10))
  )

save_fig(p_prior_scatter, "ptpriorlo_year_scatter", width = 10, height = 5)
p_prior_scatter
```

Both panels show a tight clustering of schools along the 45-degree line, with high year-on-year correlations. This confirms that the proportion of pupils entering with low KS2 prior attainment is a stable, slow-moving school characteristic. The degree of scatter is modest and symmetric around the diagonal, indicating no systematic drift. This pattern provides strong support for the carry-forward imputation of PTPRIORLO into 2024--25: if a school's value barely changes from 2021--22 to 2022--23 or from 2022--23 to 2023--24, it is reasonable to assume a similar level of stability into the following year.

# Derived Variables

## Gorard Segregation Index

The Gorard segregation index measures how unevenly disadvantaged pupils are distributed across schools within each Local Authority ([calculation details](https://sru.soc.surrey.ac.uk/SRU30.html)). It is computed per LA per year as:

$$G_s = \frac{1}{2} \sum_{i=1}^{n} \left| \frac{F_i}{F_{\text{total}}} - \frac{T_i}{T_{\text{total}}} \right|$$

where $F_i$ is the number of disadvantaged pupils in school $i$, $F_{\text{total}}$ the LA total, and $T_i$, $T_{\text{total}}$ the corresponding all-pupil totals.

```{r fig-gorard}
#| label: fig-gorard
#| fig-cap: "Distribution of Gorard segregation index across LAs by year"
#| fig-height: 4

gorard_la <- model_panel %>%
  filter(!is.na(gorard_segregation)) %>%
  distinct(LANAME, year_label, gorard_segregation)

p_gorard <- ggplot(gorard_la, aes(x = gorard_segregation, fill = year_label)) +
  geom_histogram(bins = 30, alpha = 0.65, position = "identity", colour = "white",
                 linewidth = 0.2) +
  scale_fill_manual(values = year_cols, name = "Academic year") +
  labs(x = "Gorard segregation index", y = "Number of LAs",
       title = "Distribution of LA-level segregation index") +
  theme_pub

save_fig(p_gorard, "gorard_segregation", width = 7, height = 4)
p_gorard
```

Most Local Authorities have segregation index values between 0.10 and 0.35, indicating moderate unevenness in the distribution of disadvantaged pupils across schools. A small number of LAs show notably high segregation (above 0.40), often corresponding to areas with grammar schools or other selective provision. The distribution is fairly stable across years, consistent with the view that school-level segregation is a structural feature that changes slowly.

# Correlation Structure

```{r fig-correlation-matrix}
#| label: fig-correlation-matrix
#| fig-cap: "Pairwise Pearson correlations among the outcome and all fixed-effect predictors (log-transformed variables shown on the model scale)"
#| fig-height: 7
#| fig-width: 8

# Variables that enter the model as log(x)
log_transform_vars <- c("PTFSM6CLA1A", "PERCTOT", "PNUMEAL",
                         "average_number_of_days_taken")

cor_vars <- c("ATT8SCR", "PTFSM6CLA1A", "PERCTOT", "PNUMEAL",
              "PTPRIORLO", "gorard_segregation",
              "remained_in_the_same_school",
              "teachers_on_leadership_pay_range_percent",
              "average_number_of_days_taken")
cor_vars <- intersect(cor_vars, names(model_panel))

cor_data <- model_panel %>%
  select(all_of(cor_vars)) %>%
  drop_na() %>%
  mutate(across(all_of(intersect(log_transform_vars, names(.))), safe_log))

cor_mat <- cor(cor_data)

# Apply labels — prefix logged variables with "log()"
cor_labels <- sapply(cor_vars, function(v) {
  lbl <- var_label(v)
  if (v %in% log_transform_vars) paste0("log(", lbl, ")") else lbl
})
labs_vec <- str_wrap(cor_labels, width = 20)
rownames(cor_mat) <- labs_vec
colnames(cor_mat) <- labs_vec

cor_long <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "r") %>%
  mutate(
    var1 = factor(var1, levels = rev(labs_vec)),
    var2 = factor(var2, levels = labs_vec)
  )

p_cor <- ggplot(cor_long, aes(x = var2, y = var1, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.2) +
  scale_fill_gradient2(low = "#c62828", mid = "white", high = "#1565c0",
                       midpoint = 0, limits = c(-1, 1), name = "Pearson r") +
  labs(x = NULL, y = NULL,
       title = "Correlation matrix: outcome and predictors") +
  theme_pub +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    panel.grid = element_blank()
  )

save_fig(p_cor, "correlation_matrix", width = 9, height = 7)
p_cor
```

Correlations are computed on the model scale --- i.e. the four log-transformed predictors are shown after transformation, matching the scale on which they enter the multilevel model. Several expected patterns emerge. The proportion of disadvantaged pupils has a strong negative correlation with Attainment 8, confirming that school-level deprivation is the dominant compositional predictor. Absence is positively correlated with deprivation and negatively with attainment. The EAL proportion shows a modest positive association with attainment, likely reflecting the concentration of EAL pupils in London schools which tend to perform above the national average. Among the workforce variables, teacher retention shows a positive correlation with attainment, while sickness absence is negatively associated. Importantly, no pair of predictors is so highly correlated as to raise serious multicollinearity concerns for the multilevel model.

# Summary

The full panel dataset contains **`r format(nrow(panel_data), big.mark = ",")`** school-year observations from **`r n_distinct(panel_data$URN)`** unique schools. After restricting to mainstream state-funded schools (Academies and Maintained schools), the **analysis sample** comprises **`r format(nrow(model_panel), big.mark = ",")`** school-year observations from **`r n_distinct(model_panel$URN)`** schools across **`r n_distinct(model_panel$LANAME)`** Local Authorities and **4 academic years** (2021--22 to 2024--25). Key features:

- **Sample restriction**: Independent schools (`r format(sum(panel_data$MINORGROUP == "Independent school", na.rm = TRUE), big.mark = ",")` obs.), colleges, and special schools are excluded from the analysis.
- **Outcome completeness**: Attainment 8 scores (all pupils) are available for 100% of analysis sample observations by construction.
- **Predictor completeness**: Core predictors (disadvantage, absence, EAL) exceed 95% availability across all years. Workforce variables are unavailable for 2024--25 and are carry-forward imputed.
- **Imputation scope**: `r round(sum(model_panel$has_imputed_predictors, na.rm = TRUE) / nrow(model_panel) * 100, 1)`% of analysis sample observations (confined to 2024--25) contain at least one imputed predictor.
- **Transformations**: Four right-skewed predictors are log-transformed; the Gorard segregation index is computed at the LA-year level.
