---
title: "Financial Model Experiments"
subtitle: "Do school financial variables improve the attainment models?"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: ucl-theme.scss
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
    df-print: kable
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup

library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(knitr)
library(scales)
library(patchwork)
library(here)

source(here::here("R", "helpers.R"))
```

# Overview

This document takes the baseline multilevel models from `model_experiments.qmd`
and systematically tests whether adding school-level financial variables
(from CFR/AAR returns) improves model fit.

The baseline model predicts `log(ATT8SCR)` using 9 fixed effects with
nested random intercepts for year, Ofsted rating, and region/LA:

$$\log(\text{ATT8}) \sim \text{9 predictors} + (1|\text{year}) + (1|\text{Ofsted}) + (1|\text{region/LA})$$

We test financial variables one at a time, then in combination, comparing
marginal and conditional $R^2$.


# Data Preparation

```{r data-prep}
#| label: data-prep

# Load the main panel (has gorard_segregation, OFSTEDRATING_1, fin_* cols, etc.)
panel <- readRDS(here::here("data", "panel_data.rds"))

# Supplement with per-pupil columns not carried by 03a_add_finance.R
finance <- readRDS(here::here("data", "financial_returns.rds"))
finance_extra <- finance %>%
  transmute(URN, academic_year,
            fin_premises_costs_pp = premises_costs_pp)
panel <- panel %>%
  left_join(finance_extra, by = c("URN", "academic_year"))

cat("Panel:", nrow(panel), "rows\n")
cat("Financial match:", sum(!is.na(panel$fin_total_income_pp)), "/", nrow(panel), "\n")
```

## Model dataset

The full model dataset requires workforce variables (so drops 2024-25)
**and** financial data (which further reduces sample size).

```{r model-data}
#| label: model-data

model_data <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(
    !is.na(ATT8SCR), ATT8SCR > 0,
    PTFSM6CLA1A > 0, PERCTOT > 0, PNUMEAL > 0,
    !is.na(OFSTEDRATING_1), !is.na(gor_name), !is.na(LANAME),
    !is.na(remained_in_the_same_school),
    !is.na(teachers_on_leadership_pay_range_percent),
    average_number_of_days_taken > 0,
    !is.na(gorard_segregation)
  ) %>%
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

contrasts(model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(model_data$OFSTEDRATING_1))

cat("Full model dataset:", nrow(model_data), "rows\n")
cat("Years:", paste(levels(model_data$year_label), collapse = ", "), "\n")
cat("With financial data:", sum(!is.na(model_data$fin_total_income_pp)), "\n")
```

To ensure fair comparison, all models (baseline and extended) are fitted
on the **same rows** — only those with complete financial data.

```{r model-data-fin}
#| label: model-data-fin

# Dataset with complete financial data for fair comparison
d_fin <- model_data %>%
  filter(
    !is.na(fin_total_income_pp), fin_total_income_pp > 0,
    !is.na(fin_staff_income_ratio),
    !is.na(fin_balance_income_ratio)
  ) %>%
  droplevels()

contrasts(d_fin$OFSTEDRATING_1) <-
  contr.treatment(levels(d_fin$OFSTEDRATING_1))

cat("Model dataset with financial data:", nrow(d_fin), "rows\n")
cat("Years:", paste(sort(unique(as.character(d_fin$year_label))), collapse = ", "), "\n")
cat("LAs:", n_distinct(d_fin$LANAME), "\n")
```

## Candidate financial variables

```{r candidates}
#| label: tbl-candidates
#| tbl-cap: "Financial variables available for model experiments"

d_fin %>%
  summarise(
    across(
      starts_with("fin_") & where(is.numeric),
      list(
        n = ~sum(!is.na(.x)),
        Median = ~median(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE)
      ),
      .names = "{.col}|{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "\\|") %>%
  pivot_wider(names_from = Stat, values_from = value) %>%
  mutate(
    Label = var_label(Variable),
    Median = round(Median, 2),
    SD = round(SD, 2),
    n = comma(n)
  ) %>%
  select(Variable, Label, n, Median, SD) %>%
  kable()
```


# Baseline Model (no financial variables)

This exactly replicates Analysis A "All Pupils" from `model_experiments.qmd`,
but fitted on the financial-data-complete subset for fair comparison.

```{r baseline}
#| label: baseline

mod_baseline <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d_fin,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

r2_baseline <- r2(mod_baseline)
cat("Baseline R² (marginal):", round(r2_baseline$R2_marginal, 4), "\n")
cat("Baseline R² (conditional):", round(r2_baseline$R2_conditional, 4), "\n")
cat("Singular?", isSingular(mod_baseline), "\n")
```


# Single-Variable Additions

Each financial variable is added one at a time to the baseline model.
We compare the change in marginal and conditional $R^2$.

```{r single-vars}
#| label: single-vars

# Define candidate additions
# Per-pupil vars are log-transformed; ratios are left linear
candidates <- list(
  "log(fin_total_income_pp)"          = "Total income pp (log)",
  "log(fin_total_expenditure_pp)"     = "Total expenditure pp (log)",
  "log(fin_staff_costs_pp)"           = "Staff costs pp (log)",
  "log(fin_teaching_staff_pp)"        = "Teaching staff pp (log)",
  "log(fin_grant_funding_pp)"         = "Grant funding pp (log)",
  "log(fin_premises_costs_pp)"        = "Premises costs pp (log)",
  "fin_staff_income_ratio"            = "Staff/income ratio",
  "fin_teaching_expenditure_ratio"    = "Teaching/expenditure ratio",
  "fin_balance_income_ratio"          = "Balance/income ratio"
)

base_formula <- log(ATT8SCR) ~
  log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)

results <- tibble(
  term = character(),
  label = character(),
  r2_marginal = numeric(),
  r2_conditional = numeric(),
  delta_marginal = numeric(),
  delta_conditional = numeric(),
  coef = numeric(),
  se = numeric(),
  p_value = numeric(),
  singular = logical()
)

# Fit each candidate
for (term in names(candidates)) {

  # Build extended formula
  ext_formula <- update(base_formula, paste("~ . +", term))

  # Filter for valid values of the candidate
  # (log-transformed vars need > 0)
  d_sub <- d_fin
  if (grepl("^log\\(", term)) {
    inner_var <- gsub("^log\\((.+)\\)$", "\\1", term)
    d_sub <- d_sub %>% filter(!is.na(.data[[inner_var]]), .data[[inner_var]] > 0)
  } else {
    d_sub <- d_sub %>% filter(!is.na(.data[[term]]), is.finite(.data[[term]]))
  }
  d_sub <- droplevels(d_sub)
  contrasts(d_sub$OFSTEDRATING_1) <- contr.treatment(levels(d_sub$OFSTEDRATING_1))

  tryCatch({
    mod <- lmer(
      ext_formula, data = d_sub, REML = TRUE, na.action = na.exclude,
      control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
    )

    r2_ext <- r2(mod)
    coefs <- summary(mod)$coefficients

    # Find the row for our added term
    term_clean <- term
    row_match <- grep(gsub("[()]", "", term), rownames(coefs), fixed = FALSE)
    if (length(row_match) == 0) row_match <- nrow(coefs)

    results <- bind_rows(results, tibble(
      term = term,
      label = candidates[[term]],
      r2_marginal = r2_ext$R2_marginal,
      r2_conditional = r2_ext$R2_conditional,
      delta_marginal = r2_ext$R2_marginal - r2_baseline$R2_marginal,
      delta_conditional = r2_ext$R2_conditional - r2_baseline$R2_conditional,
      coef = coefs[row_match[1], "Estimate"],
      se = coefs[row_match[1], "Std. Error"],
      p_value = coefs[row_match[1], "Pr(>|t|)"],
      singular = isSingular(mod)
    ))

  }, error = function(e) {
    message("  Failed: ", term, " — ", e$message)
  })
}
```

## Results

```{r single-results}
#| label: tbl-single
#| tbl-cap: "Effect of adding each financial variable to the baseline model"

results %>%
  mutate(
    `R² (m)` = sprintf("%.4f", r2_marginal),
    `R² (c)` = sprintf("%.4f", r2_conditional),
    `ΔR² (m)` = sprintf("%+.4f", delta_marginal),
    `ΔR² (c)` = sprintf("%+.4f", delta_conditional),
    Coef = sprintf("%.4f", coef),
    SE = sprintf("%.4f", se),
    p = case_when(
      p_value < 0.001 ~ "< 0.001",
      p_value < 0.01  ~ sprintf("%.3f", p_value),
      p_value < 0.05  ~ sprintf("%.3f", p_value),
      TRUE            ~ sprintf("%.3f", p_value)
    ),
    Sig = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE            ~ ""
    )
  ) %>%
  select(Variable = label, `R² (m)`, `R² (c)`, `ΔR² (m)`, `ΔR² (c)`,
         Coef, SE, p, Sig) %>%
  kable()
```

```{r single-plot}
#| label: fig-single
#| fig-cap: "Change in marginal R² from adding each financial variable"
#| fig-height: 5

results %>%
  mutate(label = fct_reorder(label, delta_marginal)) %>%
  ggplot(aes(delta_marginal, label)) +
  geom_col(aes(fill = delta_marginal > 0), show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c(`TRUE` = "#2e6260", `FALSE` = "#7b132b")) +
  scale_x_continuous(labels = label_number(accuracy = 0.0001)) +
  labs(x = expression(Delta * R^2 ~ "(marginal)"),
       y = NULL,
       title = "Change in marginal R² from adding each financial variable")
```

```{r single-coef-plot}
#| label: fig-coefs
#| fig-cap: "Coefficient estimates and 95% CIs for financial variable additions"
#| fig-height: 5

results %>%
  filter(p_value < 0.1) %>%
  mutate(
    label = fct_reorder(label, coef),
    lower = coef - 1.96 * se,
    upper = coef + 1.96 * se
  ) %>%
  ggplot(aes(coef, label)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = lower, xmax = upper), colour = "#2e6260",
                  linewidth = 0.8, size = 0.8) +
  labs(x = "Coefficient estimate (95% CI)", y = NULL,
       title = "Significant or marginally significant financial predictors")
```


# Multi-Variable Models

Based on the single-variable results, we test combinations of the most
promising financial variables.

```{r multi-select}
#| label: multi-select

# Select variables that improved marginal R² or were significant
sig_vars <- results %>%
  filter(p_value < 0.05, delta_marginal > 0) %>%
  arrange(desc(delta_marginal))

cat("Significant improvers (p < 0.05, positive ΔR²):\n")
for (i in seq_len(nrow(sig_vars))) {
  cat(sprintf("  %s: ΔR²(m) = %+.4f, coef = %.4f, p = %.4f\n",
              sig_vars$label[i], sig_vars$delta_marginal[i],
              sig_vars$coef[i], sig_vars$p_value[i]))
}
```

## Model with all significant financial variables

```{r multi-all-sig}
#| label: multi-all-sig

if (nrow(sig_vars) > 0) {
  # Build formula with all significant vars
  fin_terms <- paste(sig_vars$term, collapse = " + ")
  multi_formula <- update(base_formula, paste("~ . +", fin_terms))

  # Filter data
  d_multi <- d_fin
  for (term in sig_vars$term) {
    if (grepl("^log\\(", term)) {
      inner_var <- gsub("^log\\((.+)\\)$", "\\1", term)
      d_multi <- d_multi %>% filter(!is.na(.data[[inner_var]]), .data[[inner_var]] > 0)
    } else {
      d_multi <- d_multi %>% filter(!is.na(.data[[term]]), is.finite(.data[[term]]))
    }
  }
  d_multi <- droplevels(d_multi)
  contrasts(d_multi$OFSTEDRATING_1) <- contr.treatment(levels(d_multi$OFSTEDRATING_1))

  mod_multi <- lmer(
    multi_formula, data = d_multi, REML = TRUE, na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  r2_multi <- r2(mod_multi)
  cat("Combined model R² (marginal):", round(r2_multi$R2_marginal, 4), "\n")
  cat("Combined model R² (conditional):", round(r2_multi$R2_conditional, 4), "\n")
  cat("Baseline R² (marginal):", round(r2_baseline$R2_marginal, 4), "\n")
  cat("Improvement:", sprintf("%+.4f", r2_multi$R2_marginal - r2_baseline$R2_marginal), "\n")
  cat("Singular?", isSingular(mod_multi), "\n")
  cat("Observations:", nrow(d_multi), "\n\n")

  summary(mod_multi)
} else {
  cat("No financial variables significantly improved the baseline model.\n")
}
```


# Disadvantaged Pupils

Do financial variables help predict outcomes for **disadvantaged** pupils
specifically?

```{r disadv-baseline}
#| label: disadv-baseline

d_disadv <- d_fin %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()
contrasts(d_disadv$OFSTEDRATING_1) <- contr.treatment(levels(d_disadv$OFSTEDRATING_1))

mod_disadv_base <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d_disadv, REML = TRUE, na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

r2_disadv_base <- r2(mod_disadv_base)
cat("Disadvantaged baseline R² (m):", round(r2_disadv_base$R2_marginal, 4), "\n")
cat("Disadvantaged baseline R² (c):", round(r2_disadv_base$R2_conditional, 4), "\n")
```

```{r disadv-tests}
#| label: disadv-tests

results_disadv <- tibble(
  term = character(), label = character(),
  r2_marginal = numeric(), r2_conditional = numeric(),
  delta_marginal = numeric(), delta_conditional = numeric(),
  coef = numeric(), se = numeric(), p_value = numeric()
)

disadv_base_formula <- log(ATT8SCR_FSM6CLA1A) ~
  log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)

for (term in names(candidates)) {
  ext_formula <- update(disadv_base_formula, paste("~ . +", term))

  d_sub <- d_disadv
  if (grepl("^log\\(", term)) {
    inner_var <- gsub("^log\\((.+)\\)$", "\\1", term)
    d_sub <- d_sub %>% filter(!is.na(.data[[inner_var]]), .data[[inner_var]] > 0)
  } else {
    d_sub <- d_sub %>% filter(!is.na(.data[[term]]), is.finite(.data[[term]]))
  }
  d_sub <- droplevels(d_sub)
  contrasts(d_sub$OFSTEDRATING_1) <- contr.treatment(levels(d_sub$OFSTEDRATING_1))

  tryCatch({
    mod <- lmer(
      ext_formula, data = d_sub, REML = TRUE, na.action = na.exclude,
      control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
    )
    r2_ext <- r2(mod)
    coefs <- summary(mod)$coefficients
    row_match <- grep(gsub("[()]", "", term), rownames(coefs), fixed = FALSE)
    if (length(row_match) == 0) row_match <- nrow(coefs)

    results_disadv <- bind_rows(results_disadv, tibble(
      term = term, label = candidates[[term]],
      r2_marginal = r2_ext$R2_marginal, r2_conditional = r2_ext$R2_conditional,
      delta_marginal = r2_ext$R2_marginal - r2_disadv_base$R2_marginal,
      delta_conditional = r2_ext$R2_conditional - r2_disadv_base$R2_conditional,
      coef = coefs[row_match[1], "Estimate"],
      se = coefs[row_match[1], "Std. Error"],
      p_value = coefs[row_match[1], "Pr(>|t|)"]
    ))
  }, error = function(e) message("  Failed: ", term))
}
```

```{r disadv-results}
#| label: tbl-disadv
#| tbl-cap: "Financial variable effects on disadvantaged pupil attainment"

results_disadv %>%
  mutate(
    `R² (m)` = sprintf("%.4f", r2_marginal),
    `ΔR² (m)` = sprintf("%+.4f", delta_marginal),
    Coef = sprintf("%.4f", coef),
    p = case_when(
      p_value < 0.001 ~ "< 0.001",
      TRUE ~ sprintf("%.3f", p_value)
    ),
    Sig = case_when(
      p_value < 0.001 ~ "***", p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*", p_value < 0.1 ~ ".", TRUE ~ ""
    )
  ) %>%
  select(Variable = label, `R² (m)`, `ΔR² (m)`, Coef, p, Sig) %>%
  kable()
```


# Non-Disadvantaged Pupils

```{r nondisadv-baseline}
#| label: nondisadv-baseline

d_nondisadv <- d_fin %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()
contrasts(d_nondisadv$OFSTEDRATING_1) <- contr.treatment(levels(d_nondisadv$OFSTEDRATING_1))

mod_nondisadv_base <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d_nondisadv, REML = TRUE, na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

r2_nondisadv_base <- r2(mod_nondisadv_base)
cat("Non-disadvantaged baseline R² (m):", round(r2_nondisadv_base$R2_marginal, 4), "\n")
cat("Non-disadvantaged baseline R² (c):", round(r2_nondisadv_base$R2_conditional, 4), "\n")
```

```{r nondisadv-tests}
#| label: nondisadv-tests

results_nondisadv <- tibble(
  term = character(), label = character(),
  r2_marginal = numeric(), r2_conditional = numeric(),
  delta_marginal = numeric(), delta_conditional = numeric(),
  coef = numeric(), se = numeric(), p_value = numeric()
)

nondisadv_base_formula <- log(ATT8SCR_NFSM6CLA1A) ~
  log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)

for (term in names(candidates)) {
  ext_formula <- update(nondisadv_base_formula, paste("~ . +", term))

  d_sub <- d_nondisadv
  if (grepl("^log\\(", term)) {
    inner_var <- gsub("^log\\((.+)\\)$", "\\1", term)
    d_sub <- d_sub %>% filter(!is.na(.data[[inner_var]]), .data[[inner_var]] > 0)
  } else {
    d_sub <- d_sub %>% filter(!is.na(.data[[term]]), is.finite(.data[[term]]))
  }
  d_sub <- droplevels(d_sub)
  contrasts(d_sub$OFSTEDRATING_1) <- contr.treatment(levels(d_sub$OFSTEDRATING_1))

  tryCatch({
    mod <- lmer(
      ext_formula, data = d_sub, REML = TRUE, na.action = na.exclude,
      control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
    )
    r2_ext <- r2(mod)
    coefs <- summary(mod)$coefficients
    row_match <- grep(gsub("[()]", "", term), rownames(coefs), fixed = FALSE)
    if (length(row_match) == 0) row_match <- nrow(coefs)

    results_nondisadv <- bind_rows(results_nondisadv, tibble(
      term = term, label = candidates[[term]],
      r2_marginal = r2_ext$R2_marginal, r2_conditional = r2_ext$R2_conditional,
      delta_marginal = r2_ext$R2_marginal - r2_nondisadv_base$R2_marginal,
      delta_conditional = r2_ext$R2_conditional - r2_nondisadv_base$R2_conditional,
      coef = coefs[row_match[1], "Estimate"],
      se = coefs[row_match[1], "Std. Error"],
      p_value = coefs[row_match[1], "Pr(>|t|)"]
    ))
  }, error = function(e) message("  Failed: ", term))
}
```

```{r nondisadv-results}
#| label: tbl-nondisadv
#| tbl-cap: "Financial variable effects on non-disadvantaged pupil attainment"

results_nondisadv %>%
  mutate(
    `R² (m)` = sprintf("%.4f", r2_marginal),
    `ΔR² (m)` = sprintf("%+.4f", delta_marginal),
    Coef = sprintf("%.4f", coef),
    p = case_when(
      p_value < 0.001 ~ "< 0.001",
      TRUE ~ sprintf("%.3f", p_value)
    ),
    Sig = case_when(
      p_value < 0.001 ~ "***", p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*", p_value < 0.1 ~ ".", TRUE ~ ""
    )
  ) %>%
  select(Variable = label, `R² (m)`, `ΔR² (m)`, Coef, p, Sig) %>%
  kable()
```


# Comparison Across Outcomes

```{r comparison}
#| label: fig-comparison
#| fig-cap: "Change in marginal R² by outcome group"
#| fig-height: 7

all_results <- bind_rows(
  results %>% mutate(outcome = "All Pupils"),
  results_disadv %>% mutate(outcome = "Disadvantaged"),
  results_nondisadv %>% mutate(outcome = "Non-Disadvantaged")
)

all_results %>%
  mutate(label = fct_reorder(label, delta_marginal, .fun = mean)) %>%
  ggplot(aes(delta_marginal, label, colour = outcome)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c("#2e6260", "#4e3c56", "#abc766", "#e16fca", "#f9dd73"), name = "Outcome") +
  scale_x_continuous(labels = label_number(accuracy = 0.0001)) +
  labs(x = expression(Delta * R^2 ~ "(marginal)"), y = NULL,
       title = "Financial variable impact varies by pupil group") +
  theme(legend.position = "bottom")
```

```{r sig-comparison}
#| label: tbl-sig-comparison
#| tbl-cap: "Significance of financial variables across all three outcomes"

all_results %>%
  mutate(
    sig = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE            ~ ""
    ),
    cell = paste0(sprintf("%+.4f", delta_marginal), " ", sig)
  ) %>%
  select(Variable = label, outcome, cell) %>%
  pivot_wider(names_from = outcome, values_from = cell) %>%
  kable()
```


# Best Combined Model (if improvements found)

```{r best-model}
#| label: best-model

# Collect all significant improvers across all outcomes
all_sig <- all_results %>%
  filter(p_value < 0.05, delta_marginal > 0)

if (nrow(all_sig) > 0) {
  best_terms <- all_sig %>%
    group_by(term, label) %>%
    summarise(
      mean_delta = mean(delta_marginal),
      n_outcomes = n(),
      .groups = "drop"
    ) %>%
    arrange(desc(n_outcomes), desc(mean_delta))

  cat("Financial variables that significantly improved at least one outcome:\n\n")
  for (i in seq_len(nrow(best_terms))) {
    cat(sprintf("  %s: improved %d/3 outcomes, mean ΔR²(m) = %+.4f\n",
                best_terms$label[i], best_terms$n_outcomes[i],
                best_terms$mean_delta[i]))
  }

  # Fit the best combined model for All Pupils
  cat("\n\n--- Best combined model for All Pupils ---\n\n")

  best_fin_terms <- paste(best_terms$term, collapse = " + ")
  best_formula <- update(base_formula, paste("~ . +", best_fin_terms))

  d_best <- d_fin
  for (term in best_terms$term) {
    if (grepl("^log\\(", term)) {
      inner_var <- gsub("^log\\((.+)\\)$", "\\1", term)
      d_best <- d_best %>% filter(!is.na(.data[[inner_var]]), .data[[inner_var]] > 0)
    } else {
      d_best <- d_best %>% filter(!is.na(.data[[term]]), is.finite(.data[[term]]))
    }
  }
  d_best <- droplevels(d_best)
  contrasts(d_best$OFSTEDRATING_1) <- contr.treatment(levels(d_best$OFSTEDRATING_1))

  mod_best <- lmer(
    best_formula, data = d_best, REML = TRUE, na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  r2_best <- r2(mod_best)
  cat("Best model R² (marginal):", round(r2_best$R2_marginal, 4), "\n")
  cat("Best model R² (conditional):", round(r2_best$R2_conditional, 4), "\n")
  cat("Baseline R² (marginal):", round(r2_baseline$R2_marginal, 4), "\n")
  cat("Total improvement:", sprintf("%+.4f", r2_best$R2_marginal - r2_baseline$R2_marginal), "\n")
  cat("Observations:", nrow(d_best), "\n")
  cat("Singular?", isSingular(mod_best), "\n\n")

  summary(mod_best)
} else {
  cat("No financial variables significantly improved any of the three baseline models.\n")
  cat("This suggests that school-level financial characteristics are largely\n")
  cat("orthogonal to attainment outcomes once the existing predictors\n")
  cat("(deprivation, absence, prior attainment, workforce, segregation)\n")
  cat("are accounted for.\n")
}
```


# Summary

```{r summary-table}
#| label: tbl-summary
#| tbl-cap: "R² comparison: baseline vs best financial model"

summary_rows <- tibble(
  Model = c("Baseline (no finance)", "Best financial model"),
  `R² marginal` = c(
    round(r2_baseline$R2_marginal, 4),
    if (exists("mod_best")) round(r2_best$R2_marginal, 4) else NA
  ),
  `R² conditional` = c(
    round(r2_baseline$R2_conditional, 4),
    if (exists("mod_best")) round(r2_best$R2_conditional, 4) else NA
  )
)

if (exists("mod_multi") && nrow(sig_vars) > 0) {
  summary_rows <- bind_rows(
    summary_rows[1, ],
    tibble(
      Model = "All significant financial vars",
      `R² marginal` = round(r2_multi$R2_marginal, 4),
      `R² conditional` = round(r2_multi$R2_conditional, 4)
    ),
    summary_rows[2, ]
  )
}

summary_rows %>%
  filter(!is.na(`R² marginal`)) %>%
  kable()
```

::: {.callout-note}
## Interpretation notes

- **Marginal R²** captures variance explained by fixed effects only.
  This is the main comparison metric since financial variables enter as
  fixed effects.
- **Conditional R²** includes random effects (Ofsted, region/LA, year).
  Changes here indicate whether financial data absorbs variance that was
  previously captured by the random structure.
- Even small R² improvements can be substantively interesting if the
  coefficient is interpretable and policy-relevant.
- Financial variables may be collinear with existing predictors
  (e.g. deprivation drives both FSM% and per-pupil funding via the
  National Funding Formula), so null results don't mean finance is
  unimportant — just that its signal is already captured.
:::


# SEN / EHCP Experiments {#sen-experiments}

Do special educational needs (SEN) and Education, Health and Care Plan
(EHCP) variables add predictive power beyond the 9-variable baseline?

The panel data includes four school-level SEN variables from the school
census:

| Variable | Description |
|----------|-------------|
| `PSENELSE` | % pupils with an EHC plan |
| `TSENELSE` | Number of pupils with an EHC plan |
| `PSENELK` | % pupils with SEN support (but no EHC plan) |
| `TSENELK` | Number of pupils with SEN support |

We test each variable as a single addition to the baseline model,
following the same methodology as the financial experiments above. The
baseline here uses `average_number_of_days_taken` (raw, not
log-transformed) to align with the main model specification.

## SEN data availability

```{r sen-availability}
#| label: tbl-sen-availability
#| tbl-cap: "SEN variable availability in the model dataset"

sen_vars <- c("PSENELSE", "TSENELSE", "PSENELK", "TSENELK")

model_data %>%
  summarise(
    across(
      all_of(sen_vars),
      list(
        n_valid = ~sum(!is.na(.x)),
        pct_valid = ~round(mean(!is.na(.x)) * 100, 1),
        Median = ~median(.x, na.rm = TRUE),
        SD = ~sd(.x, na.rm = TRUE)
      ),
      .names = "{.col}|{.fn}"
    )
  ) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "\\|") %>%
  pivot_wider(names_from = Stat, values_from = value) %>%
  mutate(
    Label = var_label(Variable),
    Median = round(Median, 2),
    SD = round(SD, 2),
    n_valid = comma(n_valid)
  ) %>%
  select(Variable, Label, n_valid, pct_valid, Median, SD) %>%
  kable()
```

## SEN dataset

We create a dataset with complete SEN data for fair comparison. The
baseline is re-fitted on this same subset using the raw (not
log-transformed) `average_number_of_days_taken`.

```{r sen-data}
#| label: sen-data

d_sen <- model_data %>%
  filter(
    !is.na(PSENELSE), !is.na(PSENELK),
    !is.na(TSENELSE), !is.na(TSENELK)
  ) %>%
  droplevels()

contrasts(d_sen$OFSTEDRATING_1) <-
  contr.treatment(levels(d_sen$OFSTEDRATING_1))

cat("SEN-complete dataset:", nrow(d_sen), "rows\n")
cat("Years:", paste(sort(unique(as.character(d_sen$year_label))), collapse = ", "), "\n")
cat("LAs:", n_distinct(d_sen$LANAME), "\n")
cat("Schools:", n_distinct(d_sen$URN), "\n")
```

## Baseline model (SEN subset)

```{r sen-baseline}
#| label: sen-baseline

mod_sen_baseline <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    average_number_of_days_taken +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d_sen,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

r2_sen_baseline <- r2(mod_sen_baseline)
cat("SEN baseline R² (marginal):", round(r2_sen_baseline$R2_marginal, 4), "\n")
cat("SEN baseline R² (conditional):", round(r2_sen_baseline$R2_conditional, 4), "\n")
cat("Singular?", isSingular(mod_sen_baseline), "\n")
```

## Single-variable SEN additions

```{r sen-single}
#| label: sen-single

sen_candidates <- c("PSENELSE", "TSENELSE", "PSENELK", "TSENELK")

base_formula_sen <- log(ATT8SCR) ~
  log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  average_number_of_days_taken +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)

sen_results <- map_dfr(sen_candidates, function(term) {
  fml <- update(base_formula_sen, paste("~ . +", term))

  mod <- tryCatch(
    lmer(fml, data = d_sen, REML = TRUE, na.action = na.exclude,
         control = lmerControl(optimizer = "bobyqa",
                               optCtrl = list(maxfun = 20000))),
    error = function(e) NULL
  )

  if (is.null(mod)) {
    return(tibble(term = term, label = var_label(term),
                  coef = NA, se = NA, t_value = NA, p_value = NA,
                  r2_marginal = NA, r2_conditional = NA,
                  delta_marginal = NA, delta_conditional = NA,
                  singular = NA))
  }

  r2_ext <- r2(mod)
  coefs <- summary(mod)$coefficients
  row <- which(rownames(coefs) == term)

  tibble(
    term           = term,
    label          = var_label(term),
    coef           = coefs[row, "Estimate"],
    se             = coefs[row, "Std. Error"],
    t_value        = coefs[row, "t value"],
    p_value        = coefs[row, "Pr(>|t|)"],
    r2_marginal    = r2_ext$R2_marginal,
    r2_conditional = r2_ext$R2_conditional,
    delta_marginal    = r2_ext$R2_marginal - r2_sen_baseline$R2_marginal,
    delta_conditional = r2_ext$R2_conditional - r2_sen_baseline$R2_conditional,
    singular       = isSingular(mod)
  )
})
```

### Results table

```{r sen-results-table}
#| label: tbl-sen-results
#| tbl-cap: "Effect of adding SEN/EHCP variables to the baseline model"

sen_results %>%
  mutate(
    coef  = round(coef, 5),
    se    = round(se, 5),
    t_value = round(t_value, 2),
    p_value = round(p_value, 4),
    r2_marginal    = round(r2_marginal, 4),
    r2_conditional = round(r2_conditional, 4),
    delta_marginal    = round(delta_marginal, 5),
    delta_conditional = round(delta_conditional, 5),
    sig = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE            ~ ""
    )
  ) %>%
  select(Variable = term, Label = label, Coef = coef, SE = se,
         `t` = t_value, `p` = p_value, Sig = sig,
         `R² marg.` = r2_marginal, `ΔR² marg.` = delta_marginal) %>%
  kable()
```

### R² improvement plot

```{r sen-delta-plot}
#| label: fig-sen-delta
#| fig-cap: "Change in marginal R² from adding each SEN/EHCP variable"
#| fig-height: 4

sen_results %>%
  filter(!is.na(delta_marginal)) %>%
  mutate(label = fct_reorder(label, delta_marginal)) %>%
  ggplot(aes(delta_marginal, label)) +
  geom_col(aes(fill = delta_marginal > 0), show.legend = FALSE) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c(`TRUE` = "#2e6260", `FALSE` = "#7b132b")) +
  scale_x_continuous(labels = label_number(accuracy = 0.0001)) +
  labs(x = expression(Delta * R^2 ~ "(marginal)"),
       y = NULL,
       title = "Change in marginal R² from adding each SEN/EHCP variable")
```

### Coefficient estimates

```{r sen-coef-plot}
#| label: fig-sen-coefs
#| fig-cap: "Coefficient estimates and 95% CIs for SEN/EHCP variable additions"
#| fig-height: 4

sen_results %>%
  filter(!is.na(coef)) %>%
  mutate(
    label = fct_reorder(label, coef),
    lower = coef - 1.96 * se,
    upper = coef + 1.96 * se
  ) %>%
  ggplot(aes(coef, label)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = lower, xmax = upper), colour = "#2e6260",
                  linewidth = 0.8, size = 0.8) +
  labs(x = "Coefficient estimate (95% CI)", y = NULL,
       title = "SEN/EHCP variable coefficients")
```

## Multi-variable SEN model

If any SEN variables are significant (p < 0.05) and improve marginal R²,
we fit them together.

```{r sen-multi}
#| label: sen-multi

sen_sig <- sen_results %>%
  filter(p_value < 0.05, delta_marginal > 0)

cat("Significant SEN improvers:", nrow(sen_sig), "\n")
if (nrow(sen_sig) > 0) {
  cat("Variables:", paste(sen_sig$term, collapse = ", "), "\n")
}

if (nrow(sen_sig) > 1) {
  multi_terms <- paste(sen_sig$term, collapse = " + ")
  fml_multi <- update(base_formula_sen, paste("~ . +", multi_terms))

  mod_sen_multi <- lmer(
    fml_multi, data = d_sen, REML = TRUE, na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  r2_sen_multi <- r2(mod_sen_multi)
  cat("Multi-SEN model R² (marginal):",
      round(r2_sen_multi$R2_marginal, 4), "\n")
  cat("Multi-SEN model R² (conditional):",
      round(r2_sen_multi$R2_conditional, 4), "\n")
  cat("ΔR² marginal:",
      round(r2_sen_multi$R2_marginal - r2_sen_baseline$R2_marginal, 5), "\n")

  cat("\nCoefficients:\n")
  print(summary(mod_sen_multi)$coefficients[sen_sig$term, , drop = FALSE])
} else if (nrow(sen_sig) == 1) {
  cat("\nOnly one significant SEN variable — no multi-variable model needed.\n")
  cat("The single-variable result above is the best SEN model.\n")
} else {
  cat("\nNo SEN variables significantly improve the baseline model.\n")
}
```

## SEN summary

```{r sen-summary-table}
#| label: tbl-sen-summary
#| tbl-cap: "R² comparison: baseline vs SEN-augmented models"

sen_summary <- tibble(
  Model = "Baseline (no SEN)",
  `R² marginal` = round(r2_sen_baseline$R2_marginal, 4),
  `R² conditional` = round(r2_sen_baseline$R2_conditional, 4)
)

# Add best single-variable result
best_single <- sen_results %>% filter(!is.na(delta_marginal)) %>%
  slice_max(delta_marginal, n = 1)

if (nrow(best_single) > 0 && best_single$delta_marginal > 0) {
  sen_summary <- bind_rows(
    sen_summary,
    tibble(
      Model = paste0("+ ", best_single$label, " (best single)"),
      `R² marginal` = round(best_single$r2_marginal, 4),
      `R² conditional` = round(best_single$r2_conditional, 4)
    )
  )
}

if (exists("mod_sen_multi") && nrow(sen_sig) > 1) {
  sen_summary <- bind_rows(
    sen_summary,
    tibble(
      Model = paste0("+ all significant SEN vars (",
                     paste(sen_sig$term, collapse = " + "), ")"),
      `R² marginal` = round(r2_sen_multi$R2_marginal, 4),
      `R² conditional` = round(r2_sen_multi$R2_conditional, 4)
    )
  )
}

sen_summary %>% kable()
```

::: {.callout-note}
## SEN interpretation notes

- The **percentage** variables (`PSENELSE`, `PSENELK`) are more
  interpretable than counts because they adjust for school size.
- SEN prevalence is strongly correlated with deprivation (`PTFSM6CLA1A`),
  so these variables may not add much beyond what FSM% already captures.
- A significant negative coefficient on `PSENELSE` (% with EHCP) would
  suggest that schools with more EHCP pupils have lower attainment *even
  after controlling for* deprivation, absence, prior attainment, and other
  predictors — potentially reflecting the severity of need rather than
  school effectiveness.
- Count variables (`TSENELSE`, `TSENELK`) may partly proxy for school
  size and should be interpreted with caution.
:::
