---
title: "KS2 Model Experiments"
subtitle: "Preliminary primary-phase models: exploration, non-linearities, and a like-for-like comparison with the KS4 models"
author:
  - name: Adam Dennett
    affiliations:
      - name: Bartlett Centre for Advanced Spatial Analysis, University College London
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
    df-print: kable
execute:
  echo: true
  warning: false
  message: false
---

# Purpose and approach

A first pass at modelling **school-level KS2 outcomes** from the new primary panel, designed to answer one question honestly: *how good are primary-phase models relative to the KS4 models we know are good (conditional R² ≈ 0.8)?* Every step — data preparation, exploration, each modelling phase — is in this document with folded code, so any result can be interrogated.

The design has five phases:

- **Phase A** — a *common specification* fitted identically to both the KS2 and KS4 panels: same predictors, standardised outcomes, same random-effects structure. This is the only fair way to compare phases, because the published KS4 models include a predictor KS2 cannot have (prior attainment).
- **Phase B** — nested KS2 models, adding predictor blocks stepwise.
- **Phase C** — non-linearity checks (the KS4 work found policy-relevant curvature).
- **Phase D** — disadvantaged-pupil outcomes (the KS4 work found absence ≫ disadvantage concentration; does primary invert this?).
- **Phase E** — the KS4 anchor: what prior attainment adds at KS4, i.e. the size of the hole KS2 models live in.

```{r setup}
library(tidyverse)
library(lme4)
library(performance)
library(knitr)
library(kableExtra)
library(splines)
library(scales)
library(here)

set.seed(42)

ks2_raw <- readRDS(here("data", "ks2", "ks2_panel.rds"))
ks4_raw <- readRDS(here("data", "panel_data.rds"))

OFSTED_LABELS <- c("1" = "Outstanding", "2" = "Good",
                   "3" = "Requires improvement", "4" = "Inadequate")
```

# Data preparation

State-funded mainstream schools with non-missing outcomes and core predictors. Workforce variables are kept but not required (they are absent for 2024-25 by publication lag, as in the KS4 panel).

```{r prep}
ks2 <- ks2_raw %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(!is.na(READ_AVERAGE), !is.na(MAT_AVERAGE), !is.na(GPS_AVERAGE),
         !is.na(PTRWM_EXP),
         !is.na(PERCTOT), PERCTOT > 0, !is.na(PPERSABS10),
         !is.na(PTFSM6CLA1A), !is.na(PNUMEAL), !is.na(PTMOBN),
         !is.na(TOTPUPS), TOTPUPS > 0, !is.na(LANAME)) %>%
  mutate(
    ofsted = factor(OFSTED_LABELS[as.character(OFSTEDRATING_external)],
                    levels = OFSTED_LABELS),
    fsm_quintile = ntile(PTFSM6CLA1A, 5),
    phase = "KS2"
  )

ks4 <- ks4_raw %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0,
         !is.na(PERCTOT), PERCTOT > 0, !is.na(PPERSABS10),
         !is.na(PTFSM6CLA1A), !is.na(PNUMEAL),
         !is.na(TOTPUPS), TOTPUPS > 0, !is.na(LANAME)) %>%
  mutate(phase = "KS4")

tibble(
  Panel = c("KS2 (model dataset)", "KS4 (model dataset)"),
  `School-years` = c(nrow(ks2), nrow(ks4)),
  Schools = c(n_distinct(ks2$URN), n_distinct(ks4$URN)),
  Years = c(paste(range(ks2$year_label), collapse = "–"),
            paste(range(ks4$year_label), collapse = "–")),
  LAs = c(n_distinct(ks2$LANAME), n_distinct(ks4$LANAME))
) %>% kable(caption = "Model datasets after filtering")
```

# Exploratory analysis

## Outcome distributions

```{r outcome-hists}
#| fig-height: 7
ks2 %>%
  select(year_label, READ_AVERAGE, MAT_AVERAGE, GPS_AVERAGE, PTRWM_EXP) %>%
  pivot_longer(-year_label) %>%
  mutate(name = recode(name,
    READ_AVERAGE = "Reading scaled score", MAT_AVERAGE = "Maths scaled score",
    GPS_AVERAGE = "GPS scaled score", PTRWM_EXP = "% reaching expected standard (RWM)")) %>%
  ggplot(aes(value, fill = year_label)) +
  geom_histogram(bins = 60, position = "identity", alpha = 0.55) +
  facet_wrap(~name, scales = "free") +
  scale_fill_brewer(palette = "Dark2", name = NULL) +
  labs(title = "KS2 outcome distributions by year",
       subtitle = "Scaled scores are roughly symmetric; the expected-standard percentage is the bounded, coarser measure",
       x = NULL, y = "Schools") +
  theme_minimal() + theme(legend.position = "top")
```

The scaled scores (theoretical range 80–120) are approximately normal with a mild ceiling; `PTRWM_EXP` is bounded at 0/100 with visible heaping — one reason to prefer scaled scores as the primary modelling outcome (continuous, Attainment 8-like behaviour).

## Predictor distributions

```{r predictor-hists}
#| fig-height: 8
ks2 %>%
  select(PERCTOT, PPERSABS10, PTFSM6CLA1A, PNUMEAL, PTMOBN, TOTPUPS,
         remained_in_the_same_school, average_number_of_days_taken) %>%
  pivot_longer(everything()) %>%
  filter(!is.na(value)) %>%
  mutate(name = recode(name,
    PERCTOT = "Overall absence %", PPERSABS10 = "Persistent absence %",
    PTFSM6CLA1A = "Disadvantaged (FSM6/CLA) %", PNUMEAL = "EAL %",
    PTMOBN = "Non-mobile pupils %", TOTPUPS = "School size (pupils)",
    remained_in_the_same_school = "Teachers remained %",
    average_number_of_days_taken = "Sickness days per teacher")) %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 60, fill = "steelblue") +
  facet_wrap(~name, scales = "free", ncol = 2) +
  labs(title = "KS2 predictor distributions",
       subtitle = "Note right skew in EAL and disadvantage; school size is multi-modal (forms of entry)",
       x = NULL, y = "Schools") +
  theme_minimal()
```

Skewness in EAL and disadvantage suggests trying logs; the multi-modal size distribution reflects one/two/three-form entry — size is structural, not continuous, at primary.

## Outcomes by group

```{r violins}
#| fig-height: 5
ks2 %>%
  filter(!is.na(ofsted)) %>%
  ggplot(aes(ofsted, READ_AVERAGE, fill = ofsted)) +
  geom_violin(alpha = 0.7, scale = "width") +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white") +
  scale_fill_manual(values = c("#1b7837", "#7fbf7b", "#fdae61", "#d73027"),
                    guide = "none") +
  labs(title = "Reading scaled score by (snapshot) Ofsted rating",
       subtitle = "Monotonic gradient, large overlap — ratings order schools but discriminate weakly",
       x = NULL, y = "Reading scaled score") +
  theme_minimal()
```

```{r box-fsm}
#| fig-height: 5
ks2 %>%
  ggplot(aes(factor(fsm_quintile), READ_AVERAGE, fill = factor(fsm_quintile))) +
  geom_boxplot(outlier.alpha = 0.15) +
  scale_fill_brewer(palette = "RdYlBu", direction = -1, guide = "none") +
  facet_wrap(~year_label) +
  labs(title = "Reading scaled score by disadvantage quintile and year",
       subtitle = "Quintile 5 = most disadvantaged intake. A steady ~3-point gradient, stable across years",
       x = "FSM6/CLA quintile (1 = least disadvantaged)", y = "Reading scaled score") +
  theme_minimal()
```

## Bivariate relationships and non-linearity

Each candidate predictor against the reading scaled score, with a linear fit (blue) and a loess curve (red) — divergence between them flags non-linearity. Points are a 30% sample for legibility; fits use all data.

```{r bivariate}
#| fig-height: 9
biv <- ks2 %>%
  select(READ_AVERAGE, PERCTOT, PPERSABS10, PTFSM6CLA1A, PNUMEAL, PTMOBN,
         TOTPUPS, remained_in_the_same_school) %>%
  pivot_longer(-READ_AVERAGE) %>%
  filter(!is.na(value)) %>%
  mutate(name = recode(name,
    PERCTOT = "Overall absence %", PPERSABS10 = "Persistent absence %",
    PTFSM6CLA1A = "Disadvantaged %", PNUMEAL = "EAL %",
    PTMOBN = "Non-mobile %", TOTPUPS = "School size",
    remained_in_the_same_school = "Teachers remained %"))

ggplot(biv, aes(value, READ_AVERAGE)) +
  geom_point(data = ~ slice_sample(.x, prop = 0.3), alpha = 0.05, size = 0.4) +
  geom_smooth(method = "lm", se = FALSE, colour = "steelblue", linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, colour = "red", linewidth = 0.8) +
  facet_wrap(~name, scales = "free_x", ncol = 2) +
  coord_cartesian(ylim = c(95, 115)) +
  labs(title = "Reading scaled score vs candidate predictors",
       subtitle = "Blue = linear, red = loess. Curvature is visible for absence (steepening) and EAL (non-monotonic)",
       x = NULL, y = "Reading scaled score") +
  theme_minimal()
```

```{r raw-vs-log}
#| fig-height: 4.5
ks2 %>%
  select(READ_AVERAGE, PERCTOT) %>%
  mutate(`log(absence)` = log(PERCTOT)) %>%
  rename(`absence % (raw)` = PERCTOT) %>%
  pivot_longer(-READ_AVERAGE) %>%
  ggplot(aes(value, READ_AVERAGE)) +
  geom_point(data = ~ slice_sample(.x, prop = 0.3), alpha = 0.05, size = 0.4) +
  geom_smooth(method = "lm", se = FALSE, colour = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, colour = "red") +
  facet_wrap(~name, scales = "free_x") +
  coord_cartesian(ylim = c(95, 115)) +
  labs(title = "Absence: raw vs log scale",
       subtitle = "The same diagnostic used at KS4 — logging straightens the relationship somewhat but curvature remains",
       x = NULL, y = "Reading scaled score") +
  theme_minimal()
```

## Predictor correlations

```{r cor-heatmap}
#| fig-height: 6
cm <- ks2 %>%
  select(`Reading` = READ_AVERAGE, `Maths` = MAT_AVERAGE,
         `Absence` = PERCTOT, `Persistent abs.` = PPERSABS10,
         `Disadvantaged` = PTFSM6CLA1A, `EAL` = PNUMEAL,
         `Non-mobile` = PTMOBN, `Size` = TOTPUPS,
         `Teachers remained` = remained_in_the_same_school) %>%
  cor(use = "pairwise.complete.obs")

as.data.frame(cm) %>%
  rownames_to_column("v1") %>%
  pivot_longer(-v1, names_to = "v2") %>%
  mutate(v1 = factor(v1, levels = rownames(cm)),
         v2 = factor(v2, levels = rev(rownames(cm)))) %>%
  ggplot(aes(v1, v2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", value)), size = 3) +
  scale_fill_gradient2(low = "#b2182b", mid = "white", high = "#2166ac",
                       limits = c(-1, 1), name = "r") +
  labs(title = "Correlation matrix (pairwise complete)",
       subtitle = "Absence and persistent absence are highly collinear (interpret their separate coefficients with care)",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

## How stable are the outcomes? The measurement-noise ceiling

Before fitting anything, one structural fact: a model cannot explain year-to-year variation that is mostly cohort noise. Primary cohorts are small (median `r median(ks2$TELIG, na.rm = TRUE)` pupils sat KS2 per school vs ~`r round(median(ks4$TOTPUPS, na.rm = TRUE)/5)` per KS4 year group), so outcomes bounce more between years.

```{r stability}
#| fig-height: 4.5
lag_pair <- function(df, outcome) {
  df %>%
    select(URN, year_numeric, value = all_of(outcome)) %>%
    filter(!is.na(value)) %>%
    arrange(URN, year_numeric) %>%
    group_by(URN) %>%
    mutate(value_next = lead(value, order_by = year_numeric),
           gap = lead(year_numeric) - year_numeric) %>%
    ungroup() %>%
    filter(gap == 1, !is.na(value_next))
}

p2 <- lag_pair(ks2, "READ_AVERAGE") %>% mutate(phase = "KS2 reading scaled score")
p4 <- lag_pair(ks4, "ATT8SCR")      %>% mutate(phase = "KS4 Attainment 8")

stab <- bind_rows(p2, p4) %>% group_by(phase) %>%
  summarise(r = cor(value, value_next), n = n(), .groups = "drop")

bind_rows(p2, p4) %>%
  group_by(phase) %>% slice_sample(prop = 0.3) %>% ungroup() %>%
  ggplot(aes(value, value_next)) +
  geom_point(alpha = 0.06, size = 0.4) +
  geom_smooth(method = "lm", colour = "red", se = FALSE) +
  geom_text(data = stab, aes(x = -Inf, y = Inf,
            label = sprintf("  r = %.2f", r)), hjust = 0, vjust = 1.5,
            colour = "red", fontface = "bold") +
  facet_wrap(~phase, scales = "free") +
  labs(title = "Same school, consecutive years: outcome stability",
       subtitle = "Lower persistence at KS2 = more cohort noise = a lower ceiling on any school-level model",
       x = "Outcome in year t", y = "Outcome in year t+1") +
  theme_minimal()
```

KS2 reading persists at r ≈ `r sprintf('%.2f', stab$r[stab$phase == 'KS2 reading scaled score'])` year-on-year against r ≈ `r sprintf('%.2f', stab$r[stab$phase == 'KS4 Attainment 8'])` for Attainment 8. Squaring these gives a rough upper bound on explainable between-school variance — **the KS2 ceiling is genuinely lower**, before any model is fitted. Keep this in mind when comparing R² across phases below.

# Phase A — the common specification, both phases

Identical model, identical predictor set, outcomes standardised within phase, schools nested in LAs, year fixed effects:

$$z(\text{outcome})_{it} = \beta_1 z(\text{abs}) + \beta_2 z(\text{persabs}) + \beta_3 z(\text{FSM}) + \beta_4 z(\text{EAL}) + \beta_5 z(\text{size}) + \gamma_t + u_{LA} + \varepsilon$$

No Ofsted (not comparably measured across phases), no workforce (missing 2024-25), and — crucially — **no prior attainment**, because KS2 has none. This is the like-for-like playing field.

```{r common-spec}
zs <- function(x) as.numeric(scale(x))

add_z <- function(df) df %>% mutate(
  z_abs = zs(PERCTOT), z_pers = zs(PPERSABS10), z_fsm = zs(PTFSM6CLA1A),
  z_eal = zs(PNUMEAL), z_size = zs(TOTPUPS),
  z_mob = if ("PTMOBN" %in% names(df)) zs(PTMOBN) else NA_real_)

ks2z <- add_z(ks2) %>% mutate(z_read = zs(READ_AVERAGE), z_mat = zs(MAT_AVERAGE),
                              z_gps = zs(GPS_AVERAGE), z_rwm = zs(PTRWM_EXP))
ks4z <- add_z(ks4) %>% mutate(z_att8 = zs(ATT8SCR))

rhs <- "z_abs + z_pers + z_fsm + z_eal + z_size + year_label + (1 | LANAME)"

fits <- list(
  "KS4: Attainment 8"   = lmer(as.formula(paste("z_att8 ~", rhs)), data = ks4z, REML = FALSE),
  "KS2: Reading scaled" = lmer(as.formula(paste("z_read ~", rhs)), data = ks2z, REML = FALSE),
  "KS2: Maths scaled"   = lmer(as.formula(paste("z_mat ~",  rhs)), data = ks2z, REML = FALSE),
  "KS2: GPS scaled"     = lmer(as.formula(paste("z_gps ~",  rhs)), data = ks2z, REML = FALSE),
  "KS2: % RWM expected" = lmer(as.formula(paste("z_rwm ~",  rhs)), data = ks2z, REML = FALSE)
)

r2_tab <- map_dfr(fits, function(m) {
  r <- suppressWarnings(r2(m))
  tibble(`Marginal R²` = round(r$R2_marginal, 3),
         `Conditional R²` = round(r$R2_conditional, 3),
         ICC = round(icc(m)$ICC_adjusted, 3),
         N = nobs(m))
}, .id = "Model")

kable(r2_tab, caption = "Common specification: variance explained, both phases")
```

```{r common-coefs}
#| fig-height: 5.5
coef_tab <- map_dfr(fits, function(m) {
  s <- summary(m)$coefficients
  as_tibble(s, rownames = "term") %>%
    filter(grepl("^z_", term)) %>%
    transmute(term, est = Estimate, se = `Std. Error`)
}, .id = "model") %>%
  mutate(
    term = recode(term, z_abs = "Absence", z_pers = "Persistent absence",
                  z_fsm = "Disadvantage", z_eal = "EAL", z_size = "School size"),
    term = factor(term, levels = rev(c("Absence", "Persistent absence",
                                       "Disadvantage", "EAL", "School size"))),
    phase = ifelse(grepl("^KS4", model), "KS4", "KS2"))

ggplot(coef_tab, aes(est, term, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = est - 1.96 * se, xmax = est + 1.96 * se),
                  position = position_dodge(width = 0.6)) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(title = "Standardised coefficients: the same model at two phases",
       subtitle = "At KS4, absence dominates disadvantage; at KS2 the ordering inverts — the levers change with phase",
       x = "Standardised coefficient (SD outcome per SD predictor)", y = NULL) +
  theme_minimal() + theme(legend.position = "top")
```

Two headline results. First, **predictability**: the identical specification explains far less at KS2 (conditional R² ≈ `r sprintf('%.2f', r2_tab[["Conditional R²"]][2])` for reading) than at KS4 (≈ `r sprintf('%.2f', r2_tab[["Conditional R²"]][1])`) — and note the KS4 figure here is *without* prior attainment; the published models with it reach ≈ 0.8 (Phase E quantifies this). Second, **the lever ordering inverts**: disadvantage concentration is the strongest predictor at KS2 where combined absence dominates at KS4.

# Phase B — nested KS2 models

Predictor blocks added stepwise to the reading model. Workforce enters in the final model on the 2022-24 subset (it is unpublished for 2024-25).

```{r nested}
m0 <- lmer(z_read ~ year_label + (1 | LANAME), data = ks2z, REML = FALSE)
m1 <- update(m0, . ~ . + z_fsm + z_eal + z_size + z_mob)
m2 <- update(m1, . ~ . + z_abs + z_pers)
m3 <- update(m2, . ~ . + ofsted)

ks2_wf <- ks2z %>%
  filter(!is.na(remained_in_the_same_school), !is.na(average_number_of_days_taken)) %>%
  mutate(z_remained = zs(remained_in_the_same_school),
         z_sick = zs(average_number_of_days_taken))
m4 <- lmer(z_read ~ year_label + z_fsm + z_eal + z_size + z_mob +
             z_abs + z_pers + ofsted + z_remained + z_sick + (1 | LANAME),
           data = ks2_wf, REML = FALSE)

nested <- list("M0 null" = m0, "M1 +composition" = m1, "M2 +absence" = m2,
               "M3 +Ofsted" = m3, "M4 +workforce (2022-24)" = m4)

map_dfr(nested, function(m) {
  r <- suppressWarnings(r2(m))
  tibble(`Marginal R²` = round(r$R2_marginal, 3),
         `Conditional R²` = round(r$R2_conditional, 3),
         AIC = round(AIC(m)), N = nobs(m))
}, .id = "Model") %>%
  kable(caption = "Nested models, z(reading scaled score). M4 is fitted on the two workforce years, so its AIC is not comparable with M0–M3")
```

```{r nested-coefs}
s3 <- summary(m3)$coefficients
as_tibble(s3, rownames = "Term") %>%
  filter(Term != "(Intercept)") %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  select(Term, Estimate, `Std. Error`, `t value`) %>%
  kable(caption = "M3 coefficients (full three-year sample)")
```

Composition does most of the lifting; absence adds a real but smaller block; Ofsted adds a visible gradient (interpreted with the usual caution — it is a snapshot rating and partly an outcome itself); workforce adds little on top, as at KS4.

# Phase C — non-linearity

Natural splines on absence and disadvantage versus the linear versions, then the fitted partial effects.

```{r nonlin}
mlin <- m2
mspl <- update(m1, . ~ . + ns(PERCTOT, 3) + ns(PPERSABS10, 3))
mfsm <- update(m2, . ~ . - z_fsm + ns(PTFSM6CLA1A, 3))

tibble(
  Model = c("Linear absence (M2)", "Spline absence", "Spline disadvantage"),
  AIC = round(c(AIC(mlin), AIC(mspl), AIC(mfsm))),
  `Marginal R²` = round(c(r2(mlin)$R2_marginal, suppressWarnings(r2(mspl))$R2_marginal,
                          suppressWarnings(r2(mfsm))$R2_marginal), 3)
) %>% kable(caption = "Linear vs spline specifications (lower AIC = better)")
```

```{r partial-effects}
#| fig-height: 4.5
grid_at <- function(var, df, model, n = 80) {
  rng <- quantile(df[[var]], c(0.02, 0.98), na.rm = TRUE)
  nd <- df %>%
    summarise(across(c(PERCTOT, PPERSABS10, PTFSM6CLA1A, z_abs, z_pers,
                       z_fsm, z_eal, z_size, z_mob), ~ mean(., na.rm = TRUE))) %>%
    select(-all_of(var)) %>%
    crossing(!!sym(var) := seq(rng[1], rng[2], length.out = n),
             year_label = "2023-24")
  nd$pred <- predict(model, newdata = nd, re.form = NA)
  nd
}

bind_rows(
  grid_at("PERCTOT", ks2z, mspl) %>%
    transmute(x = PERCTOT, pred, var = "Overall absence % (spline)"),
  grid_at("PTFSM6CLA1A", ks2z, mfsm) %>%
    transmute(x = PTFSM6CLA1A, pred, var = "Disadvantaged % (spline)")
) %>%
  ggplot(aes(x, pred)) +
  geom_line(colour = "red", linewidth = 1) +
  facet_wrap(~var, scales = "free_x") +
  labs(title = "Fitted partial effects (other predictors at their means)",
       subtitle = "Mild steepening at high absence; the disadvantage effect is close to linear at KS2",
       x = NULL, y = "Predicted z(reading)") +
  theme_minimal()
```

Splines beat the linear fit on AIC, but the curvature is **gentler than at KS4**: primary absence has a shorter tail (range compression), so the accelerating-returns logic — big gains from improving the worst — has less room to operate at KS2. A substantive cross-phase finding in itself.

# Phase D — disadvantaged pupils' outcomes

The KS4 result that mattered most for policy: for disadvantaged pupils, absence dominated and disadvantage *concentration* was slightly positive. The KS2 counterpart, where the disadvantaged-pupil outcome is published (≈ two-thirds of schools; suppressed in small cohorts):

```{r disadv}
ks2d <- ks2z %>%
  filter(!is.na(READ_AVERAGE_FSM6CLA1A)) %>%
  mutate(z_read_dis = zs(READ_AVERAGE_FSM6CLA1A))

md <- lmer(z_read_dis ~ z_abs + z_pers + z_fsm + z_eal + z_size + year_label +
             (1 | LANAME), data = ks2d, REML = FALSE)

rd <- suppressWarnings(r2(md))
sd_ <- summary(md)$coefficients
as_tibble(sd_, rownames = "Term") %>%
  filter(grepl("^z_", Term)) %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  select(Term, Estimate, `Std. Error`, `t value`) %>%
  kable(caption = sprintf(
    "Disadvantaged pupils' reading scaled score (N = %s schools-years; conditional R² = %.2f)",
    format(nobs(md), big.mark = ","), rd$R2_conditional))
```

```{r disadv-compare}
#| fig-height: 4
bind_rows(
  as_tibble(summary(fits[["KS2: Reading scaled"]])$coefficients, rownames = "term") %>%
    mutate(model = "All pupils"),
  as_tibble(sd_, rownames = "term") %>% mutate(model = "Disadvantaged pupils")
) %>%
  filter(grepl("^z_", term)) %>%
  mutate(term = recode(term, z_abs = "Absence", z_pers = "Persistent absence",
                       z_fsm = "Disadvantage concentration", z_eal = "EAL",
                       z_size = "School size")) %>%
  ggplot(aes(Estimate, term, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = Estimate - 1.96 * `Std. Error`,
                      xmax = Estimate + 1.96 * `Std. Error`),
                  position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c("grey30", "#d95f02"), name = NULL) +
  labs(title = "KS2 reading: all pupils vs disadvantaged pupils",
       x = "Standardised coefficient", y = NULL) +
  theme_minimal() + theme(legend.position = "top")
```

# Phase E — the KS4 anchor: what prior attainment is worth

The published KS4 models reach R² ≈ 0.8 largely because they control for the intake's KS2 prior attainment. Adding it to the common spec on the KS4 panel quantifies the hole that KS2 models sit in — *there is no earlier school-level baseline for primary*:

```{r anchor}
ks4p <- ks4z %>% filter(!is.na(PTPRIORLO), !is.na(PTPRIORHI)) %>%
  mutate(z_plo = zs(PTPRIORLO), z_phi = zs(PTPRIORHI))

a1 <- lmer(z_att8 ~ z_abs + z_pers + z_fsm + z_eal + z_size + year_label +
             (1 | LANAME), data = ks4p, REML = FALSE)
a2 <- update(a1, . ~ . + z_plo + z_phi)

map_dfr(list("KS4 common spec" = a1, "KS4 + prior attainment" = a2),
        function(m) {
          r <- suppressWarnings(r2(m))
          tibble(`Marginal R²` = round(r$R2_marginal, 3),
                 `Conditional R²` = round(r$R2_conditional, 3), N = nobs(m))
        }, .id = "Model") %>%
  kable(caption = "The prior-attainment dividend at KS4 (same sample both rows)")
```

# Phase F — Census 2021 neighbourhood context as a prior-attainment proxy

KS2 models lack the ingredient that powers KS4 models (Phase E). The nearest open-data substitute: the **socio-economic position of the neighbourhoods each school serves**, from Census 2021 at output-area level — parental human capital (highest qualification) and labour-market status — averaged over each school's k nearest OA population-weighted centroids (`R/13_census_context.R`; k = 10, inverse-distance-weighted).

```{r census-join}
ctx <- readRDS(here("data", "census", "school_census_context.rds"))

ks2c <- ks2z %>%
  inner_join(ctx %>% select(URN, starts_with("cens_"), mean_dist_k10), by = "URN") %>%
  mutate(z_l4plus = zs(cens_l4plus_idw10),
         z_ltsick = zs(cens_ltsick_idw10),
         z_cunemp = zs(cens_unemp_idw10),
         z_dist = zs(log(mean_dist_k10)))

tibble(
  Variable = c("% level 4+ qualified (k=10, IDW)", "% no qualifications",
               "% unemployed", "% economically inactive", "% long-term sick",
               "Mean distance to 10 nearest OAs (km)"),
  `r with reading score` = round(c(
    cor(ks2c$READ_AVERAGE, ks2c$cens_l4plus_idw10, use = "pair"),
    cor(ks2c$READ_AVERAGE, ks2c$cens_noquals_idw10, use = "pair"),
    cor(ks2c$READ_AVERAGE, ks2c$cens_unemp_idw10, use = "pair"),
    cor(ks2c$READ_AVERAGE, ks2c$cens_inactive_idw10, use = "pair"),
    cor(ks2c$READ_AVERAGE, ks2c$cens_ltsick_idw10, use = "pair"),
    cor(ks2c$READ_AVERAGE, ks2c$mean_dist_k10, use = "pair")), 2)
) %>% kable(caption = "Neighbourhood census indicators vs school reading score (bivariate)")
```

```{r census-bivariate}
#| fig-height: 4.5
ks2c %>%
  select(READ_AVERAGE, `% level 4+ qualified` = cens_l4plus_idw10,
         `% long-term sick` = cens_ltsick_idw10) %>%
  pivot_longer(-READ_AVERAGE) %>%
  ggplot(aes(value, READ_AVERAGE)) +
  geom_point(data = ~ slice_sample(.x, prop = 0.3), alpha = 0.05, size = 0.4) +
  geom_smooth(method = "lm", se = FALSE, colour = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, colour = "red") +
  facet_wrap(~name, scales = "free_x") +
  coord_cartesian(ylim = c(95, 115)) +
  labs(title = "Reading scaled score vs neighbourhood census context (10 nearest OAs)",
       x = NULL, y = "Reading scaled score") +
  theme_minimal()
```

Does it add anything *beyond* the school's own composition (which already encodes some of the same geography through intake)?

```{r census-models}
f0 <- lmer(z_read ~ z_abs + z_pers + z_fsm + z_eal + z_size + z_mob + ofsted +
             year_label + (1 | LANAME), data = ks2c, REML = FALSE)
f1 <- update(f0, . ~ . + z_l4plus + z_ltsick + z_cunemp + z_dist)

map_dfr(list("M3 (same sample)" = f0, "M3 + census context" = f1), function(m) {
  r <- suppressWarnings(r2(m))
  tibble(`Marginal R²` = round(r$R2_marginal, 3),
         `Conditional R²` = round(r$R2_conditional, 3),
         AIC = round(AIC(m)), N = nobs(m))
}, .id = "Model") %>%
  kable(caption = "The census-context dividend at KS2 (identical samples)")
```

```{r census-coefs}
sf1 <- summary(f1)$coefficients
as_tibble(sf1, rownames = "Term") %>%
  filter(grepl("^z_", Term)) %>%
  mutate(Term = recode(Term, z_abs = "Absence", z_pers = "Persistent absence",
    z_fsm = "Disadvantage (school intake)", z_eal = "EAL", z_size = "School size",
    z_mob = "Non-mobile %", z_l4plus = "Nbhd % level 4+ (census)",
    z_ltsick = "Nbhd % long-term sick (census)",
    z_cunemp = "Nbhd % unemployed (census)", z_dist = "log mean OA distance"),
    across(where(is.numeric), ~ round(., 3))) %>%
  select(Term, Estimate, `Std. Error`, `t value`) %>%
  kable(caption = "M3 + census context: standardised coefficients")
```

Read the comparison both ways: the R² gain tells you what neighbourhood context adds **net of intake composition**, and the change in the school-intake disadvantage coefficient between M3 and this model tells you how much of "disadvantage" was really *neighbourhood* all along — the residence-vs-school geography distinction from the spatial programme, here in regression form.

## Reading the census results correctly

Three apparent puzzles in the tables above deserve explicit resolution, because each is a finding rather than a problem.

```{r census-varcomp}
vc <- function(m) {
  v <- as.data.frame(VarCorr(m))
  c(LA = v$vcov[v$grp == "LANAME"], Residual = v$vcov[v$grp == "Residual"])
}
v0 <- vc(f0); v1 <- vc(f1)

tibble(
  Component = c("Between-LA variance (random effect)", "School-level residual variance",
                "School-intake disadvantage coefficient (z_fsm)",
                "Neighbourhood % level 4+ coefficient"),
  `M3 (f0)` = c(round(v0["LA"], 3), round(v0["Residual"], 3),
                round(fixef(f0)["z_fsm"], 3), NA),
  `M3 + census (f1)` = c(round(v1["LA"], 3), round(v1["Residual"], 3),
                         round(fixef(f1)["z_fsm"], 3), round(fixef(f1)["z_l4plus"], 3)),
  Change = c(sprintf("−%.0f%%", 100 * (1 - v1["LA"] / v0["LA"])),
             sprintf("−%.1f%%", 100 * (1 - v1["Residual"] / v0["Residual"])),
             sprintf("attenuates %.0f%%", 100 * (1 - fixef(f1)["z_fsm"] / fixef(f0)["z_fsm"])),
             "—")
) %>% kable(caption = "What actually moves when the census variables enter")
```

**Puzzle 1 — "conditional R² fell, so the census model is worse?" No.** The LA random effect is a 152-entry lookup table: it absorbs *every* between-LA difference flexibly, but explains none of it — it is labelled ignorance, not understanding. The census variables explain **`r sprintf('%.0f%%', 100 * (1 - v1['LA'] / v0['LA']))` of the between-LA variance** (the table above), but four linear terms cannot absorb quite as much as 152 free parameters, so total *accounted* variance (conditional R²) dips slightly even though the school-level residual variance falls and AIC improves by ~1,700. Marginal R² — explanation from *measured* quantities — rises 0.31 → 0.35. For prediction with LA labels in hand the change is a wash; for explanation, transportability and policy reasoning the census model is strictly better. This is also a live, in-miniature demonstration of the variance-partition argument in the spatial programme: a large share of what multilevel models call "the LA effect" is measurable geography wearing an administrative costume.

**Puzzle 2 — "neighbourhood context just duplicates school disadvantage?" Only partially.** The correlation between neighbourhood %-level-4+ and school FSM% is `r round(cor(ks2c$cens_l4plus_idw10, ks2c$PTFSM6CLA1A, use = "pair"), 2)` — related, far from identical. School-intake disadvantage attenuates by about a fifth when census enters (some of what looked like intake effect was neighbourhood all along) but survives at nearly equal magnitude to the census term. They are complementary constructs: FSM measures *income deprivation of the enrolled cohort*; %-level-4+ measures *adult human capital of the surrounding area* — parental education, plausibly operating through home learning environments rather than poverty per se. If they measured the same thing, one of them would have died.

**Puzzle 3 — "so the ~0.4 ceiling means we're missing pupil ability?" Half right.** Two distinct components fill the unexplained share, and they demand different responses:

- *Missing entry ability.* Phase E showed actual cohort prior attainment is worth ≈ +0.22 marginal R² at KS4; the census proxy recovers only ≈ +0.05 at KS2. Ecological neighbourhood averages are a weak substitute for knowing what the enrolled children could do at entry — and no school-level entry measure exists (EYFSP is not published at school level; KS1 assessment has ended).
- *Irreducible cohort noise.* The stability analysis (see *How stable are the outcomes?* above) found consecutive-year reading scores correlate at only r ≈ 0.61: roughly 40% of year-to-year variance in a primary school's score is transient — 30-pupil cohort composition, test-week luck. No school-level predictor, census included, can explain cohort-specific noise, so the realistic ceiling for *any* model of this kind at KS2 is ≈ 0.6, not the ≈ 0.87 achievable at KS4.

Informal decomposition: explained ≈ 0.35–0.41; ceiling ≈ 0.6; the gap to the ceiling is plausibly mostly unmeasured cohort ability; the ceiling itself is small-cohort measurement noise. The design implications differ: **Bayesian shrinkage and multi-year averaging** address the noise component; **feeder/pipeline modelling and (long-run) NPD-validated entry measures** address the ability component.

# Diagnostics for the headline KS2 model (M3)

```{r diagnostics}
#| fig-height: 4.5
diag_df <- tibble(fitted = fitted(m3), resid = resid(m3),
                  obs = fitted(m3) + resid(m3))

ggplot(diag_df, aes(fitted, obs)) +
  geom_bin2d(bins = 70) +
  geom_abline(colour = "red") +
  scale_fill_viridis_c(name = "Schools") +
  labs(title = "Observed vs predicted, z(reading), M3",
       x = "Predicted", y = "Observed") +
  theme_minimal()
```

```{r diagnostics2}
#| fig-height: 4
ggplot(diag_df, aes(sample = resid)) +
  stat_qq(alpha = 0.2, size = 0.5) + stat_qq_line(colour = "red") +
  labs(title = "Residual QQ plot — heavier tails than normal (small-cohort noise)",
       x = "Theoretical", y = "Sample") +
  theme_minimal()
```

```{r caterpillar}
#| fig-height: 6.5
re <- ranef(m3)$LANAME %>%
  rownames_to_column("LANAME") %>%
  rename(effect = `(Intercept)`) %>%
  arrange(effect)

bind_rows(head(re, 15) %>% mutate(end = "Lowest 15"),
          tail(re, 15) %>% mutate(end = "Highest 15")) %>%
  mutate(LANAME = fct_reorder(LANAME, effect)) %>%
  ggplot(aes(effect, LANAME, colour = end)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2) +
  scale_colour_manual(values = c("#1b7837", "#d73027"), guide = "none") +
  labs(title = "LA random effects, KS2 reading (M3): top and bottom 15",
       subtitle = "Conditional on composition, absence and Ofsted — the primary-phase counterpart of the KS4 LA effects",
       x = "LA effect (SD of z(reading))", y = NULL) +
  theme_minimal()
```

# Conclusions

1. **KS2 outcomes are much less predictable than KS4 outcomes from the same information set** — common-spec conditional R² ≈ `r sprintf('%.2f', r2_tab[["Conditional R²"]][2])` (reading) vs ≈ `r sprintf('%.2f', r2_tab[["Conditional R²"]][1])` (Attainment 8), and the year-on-year stability analysis shows a meaningful part of this gap is *structural*: small primary cohorts mean noisier outcomes and a lower explainable ceiling, not just missing predictors.
2. **The levers re-order across phases.** Disadvantage concentration is the strongest school-level correlate at KS2; absence (especially persistent absence) overtakes it by KS4. The cross-phase question at the heart of the funding bid is real and answerable.
3. **Prior attainment is the decisive KS4 ingredient** (Phase E: adding it lifts marginal R² substantially) — and it *cannot exist* at KS2 school level since KS1 assessment ended. Primary models will never look like secondary models; uncertainty-honest presentation (Bayesian shrinkage, intervals) matters more here.
4. **Non-linearity exists at KS2 but is gentler** — the accelerating-returns lever logic has less range to work with in primary absence.
5. **Practical implication for the intelligence system:** KS2 contextual benchmarking is feasible and useful (the LA caterpillar already shows credible variation), but should lean on multi-year averages and displayed uncertainty rather than single-year point estimates.
6. **Census neighbourhood context partially fills the prior-attainment hole** (Phase F). Adding the k-nearest-OA census indicators lifts marginal R² from 0.31 to 0.35 (AIC −1,722), and neighbourhood adult education (% level 4+) enters as the **single strongest predictor in the model** — larger than school-intake disadvantage, which it partially displaces. The conditional R² *falls* slightly because the census variables explain variance the LA random effect was previously absorbing blindly: geography measured rather than geography black-boxed. This validates the residence-vs-school geography agenda in the spatial programme, and suggests the same construction (k-nearest-OA context) belongs in the KS4 and KS5 models too.
