---
title: "KS5 Model Experiments"
subtitle: "A first post-16 provider model: route-specific outcomes, wide-catchment neighbourhood context, and LA pipeline effects"
author:
  - name: Adam Dennett
    affiliations:
      - name: Bartlett Centre for Advanced Spatial Analysis, University College London
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    embed-resources: true
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
    df-print: kable
execute:
  echo: true
  warning: false
  message: false
---

# Purpose and design

A quick test build of post-16 provider models, completing the cross-phase set (KS2 and KS4 experiments precede this). Three design points specific to KS5:

1. **Routes are not one outcome.** A-level, academic, applied general and T-level cohorts are different markets with different grading conventions — we model **A-level points per entry** as the headline and refit the same specification for the other routes rather than pooling.
2. **Catchments are wide and may be weak.** Students travel much further at 16+, and several provider types barely have local catchments at all. We test the census neighbourhood construction at **k = 10, 25, 50 and 100** nearest output areas and let AIC choose — extending the k-gradient logic that picked 10 at KS2 and 50 at KS4.
3. **No published intake baseline.** The KS5 files carry *no* cohort prior-attainment variable — the KS4 baseline exists only inside DfE's own value-added measures (`VA_INS_*`, 2023-24 onwards). So attainment models here have the same structural handicap as KS2 models (no baseline), and we use the VA outcomes as the "baseline-adjusted" robustness check. We also test **LA pipeline effects**: the LA effects estimated in the KS4 census model, and LA mean KS2 attainment, as group-level predictors — does the quality of a place's earlier phases follow its students into post-16?

```{r setup}
library(tidyverse)
library(lme4)
library(performance)
library(knitr)
library(splines)
library(here)

set.seed(42)

ks5_raw <- readRDS(here("data", "ks5", "ks5_panel.rds"))
ctx     <- readRDS(here("data", "census", "school_census_context.rds"))
gias    <- readRDS(here("data", "gias", "gias_establishments.rds"))
afr     <- readRDS(here("data", "cache", "analysis_f_results.rds"))
ks2     <- readRDS(here("data", "ks2", "ks2_panel.rds"))

zs <- function(x) as.numeric(scale(x))
pct <- function(x) round(100 * mean(!is.na(x)))
OFSTED_LABELS <- c("1" = "Outstanding", "2" = "Good",
                   "3" = "Requires improvement", "4" = "Inadequate")
```

# Data preparation

State-funded providers only (independents have no destinations data, different incentives and no LA accountability relationship); LA names from GIAS (the school-information join misses colleges). LA pipeline measures: the KS4 LA effect from the census model (Analysis F cache) and LA mean KS2 reading.

```{r prep}
la_lookup_gias <- gias %>% select(URN = urn, gias_la = la_name)

la_ks2 <- ks2 %>%
  group_by(gias_la = LANAME) %>%
  summarise(la_ks2_read = mean(READ_AVERAGE, na.rm = TRUE), .groups = "drop")

la_ks4 <- afr$la_cmp %>% select(gias_la = la_name, la_ks4_effect = effect_census)

ks5 <- ks5_raw %>%
  filter(!provider_group %in% c("Independent", "Other/unknown")) %>%
  left_join(la_lookup_gias, by = "URN") %>%
  left_join(ctx %>% select(URN, starts_with("cens_l4plus"), starts_with("mean_dist")),
            by = "URN") %>%
  left_join(la_ks2, by = "gias_la") %>%
  left_join(la_ks4, by = "gias_la") %>%
  mutate(
    dis_share = 100 * TALLPUP_1618_DIS / TALLPUP_1618,
    ofsted = factor(OFSTED_LABELS[as.character(OFSTEDRATING_external)],
                    levels = OFSTED_LABELS),
    provider_group = fct_relevel(factor(provider_group),
                                 "School sixth form (academy/free)")
  )

tibble(
  Check = c("Provider-years (state-funded)", "Providers", "With LA (GIAS)",
            "With census context", "With A-level PPE", "With disadvantage share",
            "With Ofsted", "With LA KS4 effect"),
  N = c(nrow(ks5), n_distinct(ks5$URN), sum(!is.na(ks5$gias_la)),
        sum(!is.na(ks5$cens_l4plus_k50)), sum(!is.na(ks5$TALLPPE_ALEV_1618)),
        sum(!is.na(ks5$dis_share)), sum(!is.na(ks5$ofsted)),
        sum(!is.na(ks5$la_ks4_effect)))
) %>% kable(caption = "Sample construction")
```

# Exploratory analysis

## Outcomes by route and provider type

```{r route-hists}
#| fig-height: 6.5
ks5 %>%
  select(year_label, `A level` = TALLPPE_ALEV_1618, Academic = TALLPPE_ACAD_1618,
         `Applied general` = TALLPPE_AGEN, `T level` = TALLPPE_TLEV) %>%
  pivot_longer(-year_label) %>%
  filter(!is.na(value)) %>%
  ggplot(aes(value, fill = year_label)) +
  geom_histogram(bins = 50, position = "identity", alpha = 0.5) +
  facet_wrap(~name, scales = "free") +
  scale_fill_brewer(palette = "Dark2", name = NULL) +
  labs(title = "Points per entry by qualification route and year",
       subtitle = "Different scales, different grading regimes — and visible 2021-22 TAG inflation in the A-level panel",
       x = "Average points per entry", y = "Providers") +
  theme_minimal() + theme(legend.position = "top")
```

```{r violin-providers}
#| fig-height: 5
ks5 %>%
  filter(!is.na(TALLPPE_ALEV_1618)) %>%
  ggplot(aes(provider_group, TALLPPE_ALEV_1618, fill = provider_group)) +
  geom_violin(alpha = 0.7, scale = "width") +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  coord_flip() +
  labs(title = "A-level points per entry by provider type",
       subtitle = "The provider landscape is the defining KS5 stratifier",
       x = NULL, y = "A-level points per entry") +
  theme_minimal()
```

```{r dis-share}
#| fig-height: 4.5
ks5 %>%
  filter(!is.na(dis_share), dis_share <= 60) %>%
  ggplot(aes(provider_group, dis_share, fill = provider_group)) +
  geom_boxplot(outlier.alpha = 0.15) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  coord_flip() +
  labs(title = "Disadvantaged share of the 16-18 cohort by provider type",
       subtitle = "FE colleges and 16-19 academies carry systematically more disadvantaged cohorts",
       x = NULL, y = "% of cohort disadvantaged (KS4-era FSM6/CLA)") +
  theme_minimal()
```

## Bivariate relationships

```{r bivariate}
#| fig-height: 7
ks5 %>%
  select(TALLPPE_ALEV_1618, `Disadvantaged %` = dis_share,
         `Cohort size` = TALLPUP_1618,
         `Nbhd % L4+ (k=50)` = cens_l4plus_k50,
         `Nbhd % L4+ (k=100)` = cens_l4plus_k100) %>%
  pivot_longer(-TALLPPE_ALEV_1618) %>%
  filter(!is.na(value), !is.na(TALLPPE_ALEV_1618)) %>%
  ggplot(aes(value, TALLPPE_ALEV_1618)) +
  geom_point(alpha = 0.07, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, colour = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, colour = "red") +
  facet_wrap(~name, scales = "free_x", ncol = 2) +
  labs(title = "A-level points per entry vs candidate predictors",
       x = NULL, y = "A-level points per entry") +
  theme_minimal()
```

## Catchment width: the correlation by k

```{r k-correlations}
tibble(
  k = c(10, 25, 50, 100),
  `r with A-level PPE` = c(
    cor(ks5$TALLPPE_ALEV_1618, ks5$cens_l4plus_k10, use = "pair"),
    cor(ks5$TALLPPE_ALEV_1618, ks5$cens_l4plus_k25, use = "pair"),
    cor(ks5$TALLPPE_ALEV_1618, ks5$cens_l4plus_k50, use = "pair"),
    cor(ks5$TALLPPE_ALEV_1618, ks5$cens_l4plus_k100, use = "pair"))
) %>%
  mutate(across(-k, ~ round(., 3))) %>%
  kable(caption = "Neighbourhood adult education vs A-level outcomes, by neighbourhood size")
```

## Outcome stability: the three-phase comparison

```{r stability}
#| fig-height: 4
lagged <- ks5 %>%
  select(URN, year_numeric, value = TALLPPE_ALEV_1618) %>%
  filter(!is.na(value)) %>%
  arrange(URN, year_numeric) %>%
  group_by(URN) %>%
  mutate(value_next = lead(value, order_by = year_numeric),
         gap = lead(year_numeric) - year_numeric) %>%
  ungroup() %>%
  filter(gap == 1, !is.na(value_next))

r_ks5 <- cor(lagged$value, lagged$value_next)

ggplot(lagged, aes(value, value_next)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "red", se = FALSE) +
  annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 1.5,
           label = sprintf("r = %.2f", r_ks5), colour = "red", fontface = "bold") +
  labs(title = "A-level points per entry: same provider, consecutive years",
       subtitle = sprintf("Lag-1 r = %.2f — between KS2 reading (0.61) and KS4 Attainment 8 (0.95)", r_ks5),
       x = "Year t", y = "Year t+1") +
  theme_minimal()
```

# Models

Headline outcome: standardised A-level points per entry. Schools/providers in LAs, year fixed effects (absorbing the TAG-era grade regime), provider type as the structural stratifier.

## Build-up and the k test

```{r models-core}
md <- ks5 %>%
  filter(!is.na(TALLPPE_ALEV_1618), !is.na(dis_share), !is.na(TALLPUP_1618),
         !is.na(gias_la), !is.na(cens_l4plus_k100), !is.na(ofsted)) %>%
  mutate(z_alev = zs(TALLPPE_ALEV_1618),
         z_dis = zs(dis_share),
         z_cohort = zs(log(TALLPUP_1618)),
         z_l4_k10 = zs(cens_l4plus_k10), z_l4_k25 = zs(cens_l4plus_k25),
         z_l4_k50 = zs(cens_l4plus_k50), z_l4_k100 = zs(cens_l4plus_k100),
         z_la_ks2 = zs(la_ks2_read), z_la_ks4 = zs(la_ks4_effect))

m0 <- lmer(z_alev ~ year_label + (1 | gias_la), data = md, REML = FALSE)
m1 <- update(m0, . ~ . + provider_group + z_dis + z_cohort)
m2 <- update(m1, . ~ . + ofsted)
mk <- map(c(10, 25, 50, 100),
          ~ update(m2, as.formula(paste(". ~ . + z_l4_k", .x, sep = ""))))
names(mk) <- paste0("M2 + census k=", c(10, 25, 50, 100))

map_dfr(c(list("M0 null" = m0, "M1 structure+composition" = m1,
              "M2 +Ofsted" = m2), mk),
        function(m) {
          r <- suppressWarnings(r2(m))
          tibble(`Marginal R²` = round(r$R2_marginal, 3),
                 `Conditional R²` = round(r$R2_conditional, 3),
                 AIC = round(AIC(m)), N = nobs(m))
        }, .id = "Model") %>%
  kable(caption = "Model build-up and the neighbourhood-size test (identical samples)")
```

## LA pipeline effects

```{r pipeline}
best_k_ks5 <- c("10", "25", "50", "100")[which.min(map_dbl(mk, AIC))]
m3 <- mk[[paste0("M2 + census k=", best_k_ks5)]]
m4 <- update(m3, . ~ . + z_la_ks2 + z_la_ks4)

map_dfr(list("M3 best census k" = m3,
             "M4 + LA pipeline (KS2 mean, KS4 LA effect)" = m4),
        function(m) {
          r <- suppressWarnings(r2(m))
          la_v <- as.data.frame(VarCorr(m)) %>% filter(grp == "gias_la") %>% pull(vcov)
          tibble(`Marginal R²` = round(r$R2_marginal, 3),
                 `Conditional R²` = round(r$R2_conditional, 3),
                 AIC = round(AIC(m)), `LA variance` = round(la_v, 4))
        }, .id = "Model") %>%
  kable(caption = "Do a place's earlier phases follow its students into post-16?")

s4 <- summary(m4)$coefficients
as_tibble(s4, rownames = "Term") %>%
  filter(grepl("^z_|provider_group|ofsted", Term)) %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  select(Term, Estimate, `Std. Error`, `t value`) %>%
  kable(caption = paste0("M4 coefficients (census at k = ", best_k_ks5, ")"))
```

## Route comparison: the same model for other qualification types

```{r routes}
fit_route <- function(outcome) {
  d <- ks5 %>%
    filter(!is.na(.data[[outcome]]), !is.na(dis_share), !is.na(TALLPUP_1618),
           !is.na(gias_la), !is.na(cens_l4plus_k100), !is.na(ofsted)) %>%
    mutate(z_out = zs(.data[[outcome]]), z_dis = zs(dis_share),
           z_cohort = zs(log(TALLPUP_1618)),
           z_l4 = zs(.data[[paste0("cens_l4plus_k", best_k_ks5)]]),
           z_la_ks2 = zs(la_ks2_read), z_la_ks4 = zs(la_ks4_effect))
  lmer(z_out ~ provider_group + z_dis + z_cohort + ofsted + z_l4 +
         z_la_ks2 + z_la_ks4 + year_label + (1 | gias_la),
       data = d, REML = FALSE)
}

routes <- list(`A level` = "TALLPPE_ALEV_1618",
               `Academic` = "TALLPPE_ACAD_1618",
               `Applied general` = "TALLPPE_AGEN")
route_fits <- map(routes, fit_route)

map_dfr(route_fits, function(m) {
  r <- suppressWarnings(r2(m))
  tibble(`Marginal R²` = round(r$R2_marginal, 3),
         `Conditional R²` = round(r$R2_conditional, 3), N = nobs(m))
}, .id = "Route") %>%
  kable(caption = "The same specification across qualification routes")
```

```{r route-coefs}
#| fig-height: 5
map_dfr(route_fits, function(m) {
  as_tibble(summary(m)$coefficients, rownames = "term") %>%
    filter(grepl("^z_", term))
}, .id = "route") %>%
  mutate(term = recode(term, z_dis = "Disadvantaged share", z_cohort = "log cohort size",
                       z_l4 = paste0("Nbhd % L4+ (k=", best_k_ks5, ")"),
                       z_la_ks2 = "LA mean KS2 reading",
                       z_la_ks4 = "LA KS4 effect (census model)")) %>%
  ggplot(aes(Estimate, term, colour = route)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = Estimate - 1.96 * `Std. Error`,
                      xmax = Estimate + 1.96 * `Std. Error`),
                  position = position_dodge(width = 0.55)) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(title = "Standardised coefficients by qualification route",
       x = "Standardised coefficient", y = NULL) +
  theme_minimal() + theme(legend.position = "top")
```

## The value-added check (2023-24 onwards)

DfE's `VA_INS_ALEV` already nets out each student's KS4 baseline — the missing-intake problem solved by construction. If neighbourhood and pipeline effects survive *here*, they operate beyond intake selection:

```{r va-check}
va <- ks5 %>%
  filter(!is.na(VA_INS_ALEV), !is.na(dis_share), !is.na(gias_la),
         !is.na(cens_l4plus_k100), !is.na(ofsted)) %>%
  mutate(z_va = zs(VA_INS_ALEV), z_dis = zs(dis_share),
         z_cohort = zs(log(TALLPUP_1618)),
         z_l4 = zs(.data[[paste0("cens_l4plus_k", best_k_ks5)]]),
         z_la_ks2 = zs(la_ks2_read), z_la_ks4 = zs(la_ks4_effect))

mva <- lmer(z_va ~ provider_group + z_dis + z_cohort + ofsted + z_l4 +
              z_la_ks2 + z_la_ks4 + year_label + (1 | gias_la),
            data = va, REML = FALSE)

rva <- suppressWarnings(r2(mva))
as_tibble(summary(mva)$coefficients, rownames = "Term") %>%
  filter(grepl("^z_", Term)) %>%
  mutate(across(where(is.numeric), ~ round(., 3))) %>%
  select(Term, Estimate, `Std. Error`, `t value`) %>%
  kable(caption = sprintf("A-level value added (KS4-baseline adjusted) as outcome — N = %s, marginal R² = %.2f",
                          format(nobs(mva), big.mark = ","), rva$R2_marginal))
```

## Provider effect decomposition

The cross-phase comparison the KS2/KS4 work set up — how much is the provider, at 16-18?

```{r decomposition}
m5 <- update(m4, . ~ . + (1 | URN))
vc <- as.data.frame(VarCorr(m5)) %>% filter(is.na(var2))
fixed_var <- var(predict(m5, re.form = NA), na.rm = TRUE)
total_var <- fixed_var + sum(vc$vcov)

decomp_ks5 <- vc %>%
  transmute(Component = recode(grp,
    "URN" = "Provider (persistent, unexplained)",
    "gias_la" = "LA", "Residual" = "Transient (cohort-year noise)"),
    Variance = vcov) %>%
  bind_rows(tibble(Component = "Fixed effects (provider type, composition, Ofsted, neighbourhood, LA pipeline)",
                   Variance = fixed_var)) %>%
  mutate(`Share of total %` = round(100 * Variance / total_var, 1)) %>%
  arrange(desc(Variance))

kable(decomp_ks5 %>% select(-Variance),
      caption = "Variance decomposition, A-level points per entry (M4 + provider random intercept)")
```

# Conclusions

```{r conclusions-numbers}
#| include: false
r2_m2 <- suppressWarnings(r2(m2)); r2_m4 <- suppressWarnings(r2(m4))
prov_share <- decomp_ks5 %>% filter(grepl("Provider", Component)) %>% pull(`Share of total %`)
trans_share <- decomp_ks5 %>% filter(grepl("Transient", Component)) %>% pull(`Share of total %`)
```

1. **Predictability sits where the stability analysis says it should.** A-level points per entry persists at r ≈ `r sprintf('%.2f', r_ks5)` year-on-year — between KS2 (0.61) and KS4 (0.95) — and the models explain marginal R² ≈ `r sprintf('%.2f', r2_m4$R2_marginal)` without any intake baseline, sitting accordingly between the two phases.
2. **The catchment hypothesis is tested — and the answer is *not* "wider is better".** Best k = `r best_k_ks5`, with k = 50 essentially tied and **k = 100 clearly worse**. Widening the neighbourhood past the town scale dilutes signal rather than capturing long-distance travel-to-learn. The k-sequence across phases (KS2: 10, KS4: 50, KS5: `r best_k_ks5`) is non-monotonic — a hint that at 16-18 no fixed-k neighbourhood approximates the catchment well, because travel behaviour differs radically by provider type. That is an argument for the calibrated route-specific travel-to-learn models in the spatial programme, not for bigger k.
3. **Routes differ.** The same specification fits the A-level, academic and applied-general markets differently — pooling them into one outcome would have been a mistake, vindicating the route-specific design.
4. **Place follows pupils across phases.** The LA pipeline terms (LA mean KS2 reading; LA KS4 effect) both add signal to attainment beyond provider type, composition and neighbourhood — the first direct evidence in this programme that the quality of a place's earlier phases is visible in its post-16 outcomes. The VA check refines the mechanism: against the KS4-baseline-adjusted outcome, the **LA KS4 effect disappears** (consistent with it operating through intake quality) while **neighbourhood education and LA KS2 attainment survive** — context effects that operate beyond intake selection.
5. **The provider effect at 16-18 looks large (≈ `r prov_share`% persistent) but is *not comparable* to the ~16% school effect at KS4** — that decomposition controlled cohort prior attainment in the fixed effects; this one cannot, so the provider intercept here bundles genuine institutional effects with stable intake selection (sixth-form entry requirements are explicit selection devices). The VA model is the better guide to institutional effects at this phase; the honest cross-phase decomposition awaits provider-level cohort baselines (NPD/SRS).
6. **Missing intake baseline remains the binding constraint**, exactly as at KS2: no published provider-level KS4 baseline exists outside the VA measures. The VA-outcome check is the partial workaround; provider-level cohort baselines are an NPD/SRS follow-on.
