---
title: "Model Experiments"
subtitle: "Human-readable lmer implementations for manual inspection"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    theme: flatly
    df-print: kable
    embed-resources: true
execute:
  echo: true
  warning: true
  message: true
---

```{r setup}
#| label: setup

library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(here)

# Load the raw panel data (same starting point as 05_fit_model.R)
panel <- readRDS(here::here("data", "panel_data.rds"))

cat("Panel loaded:", nrow(panel), "rows,", ncol(panel), "columns\n")
```


# Overview

This document re-implements every model from `model_results.qmd` in a
deliberately simple, explicit style. Each model is written out longhand
so you can:

- see exactly which rows survive the filters
- read the `lmer()` formula as a single, self-contained call
- inspect `summary()`, `fixef()`, `ranef()`, `VarCorr()` etc. directly

The four analyses and three outcomes produce **12 models** in total.

Models are fitted with `REML = TRUE` and `optimizer = "bobyqa"` to
match the production pipeline.


# Data Preparation

Two datasets are needed: a **full** dataset (requires workforce and
prior-attainment variables, so covers 3 years) and a **core** dataset
(5 predictors only, covers all 4 years including 2024-25).

## Full model dataset

```{r data-full}
#| label: data-full

model_data <- panel %>%
  # Mainstream schools only

  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  # Main outcome + log-transformed predictors must be positive & non-NA
  filter(
    !is.na(ATT8SCR), ATT8SCR > 0,
    PTFSM6CLA1A > 0,
    PERCTOT > 0,
    PNUMEAL > 0,
    !is.na(OFSTEDRATING_1),
    !is.na(gor_name),
    !is.na(LANAME)
  ) %>%
  # Workforce + Gorard variables (drops 2024-25)
  filter(
    remained_in_the_same_school > 0,
    !is.na(teachers_on_leadership_pay_range_percent),
    average_number_of_days_taken > 0,
    !is.na(gorard_segregation)
  ) %>%
  # Set factor levels
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

# Treatment contrasts (lmer default for ordered factors is polynomial)
contrasts(model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(model_data$OFSTEDRATING_1))

cat("Full model dataset:", nrow(model_data), "rows\n")
cat("Years:", paste(levels(model_data$year_label), collapse = ", "), "\n")
cat("Regions:", n_distinct(model_data$gor_name), "\n")
cat("LAs:", n_distinct(model_data$LANAME), "\n")
cat("Ofsted levels:", paste(levels(model_data$OFSTEDRATING_1), collapse = ", "), "\n")
```

## Core model dataset

```{r data-core}
#| label: data-core

core_model_data <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(
    !is.na(ATT8SCR), ATT8SCR > 0,
    PTFSM6CLA1A > 0,
    PERCTOT > 0,
    PNUMEAL > 0,
    !is.na(ADMPOL_PT),
    !is.na(gorard_segregation),
    !is.na(OFSTEDRATING_1),
    !is.na(gor_name),
    !is.na(LANAME)
  ) %>%
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

contrasts(core_model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(core_model_data$OFSTEDRATING_1))

cat("Core model dataset:", nrow(core_model_data), "rows\n")
cat("Years:", paste(levels(core_model_data$year_label), collapse = ", "), "\n")
```


# Analysis A: Full Panel Models

All years pooled, `year_label` as a random intercept. Nine fixed-effect
predictors (5 log-transformed, 4 linear).

## All Pupils

```{r a-all}
#| label: a-all

d <- model_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    log(remained_in_the_same_school) +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_all)
```

### Quick checks

```{r a-all-checks}
#| label: a-all-checks

cat("Fixed effects:\n")
fixef(mod_a_all)

cat("\nVariance components:\n")
print(VarCorr(mod_a_all))

cat("\nR-squared:\n")
print(r2(mod_a_all))

cat("\nSingular fit?", isSingular(mod_a_all), "\n")
```

## Disadvantaged Pupils

```{r a-disadv}
#| label: a-disadv

d <- model_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    log(remained_in_the_same_school) +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_disadv)
```

### Quick checks

```{r a-disadv-checks}
#| label: a-disadv-checks

fixef(mod_a_disadv)
print(VarCorr(mod_a_disadv))
print(r2(mod_a_disadv))
cat("Singular fit?", isSingular(mod_a_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r a-nondisadv}
#| label: a-nondisadv

d <- model_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    log(remained_in_the_same_school) +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_nondisadv)
```

### Quick checks

```{r a-nondisadv-checks}
#| label: a-nondisadv-checks

fixef(mod_a_nondisadv)
print(VarCorr(mod_a_nondisadv))
print(r2(mod_a_nondisadv))
cat("Singular fit?", isSingular(mod_a_nondisadv), "\n")
```


# Analysis B: Full Per-Year Models

Same 9 fixed effects as Analysis A, but fitted separately for each year.
No `year_label` random effect.

```{r b-setup}
#| label: b-setup

year_levels <- sort(unique(as.character(model_data$year_label)))
cat("Years to fit:", paste(year_levels, collapse = ", "), "\n")
```

## All Pupils (per year)

```{r b-all}
#| label: b-all

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- All Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr, !is.na(ATT8SCR), ATT8SCR > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      log(remained_in_the_same_school) +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Disadvantaged Pupils (per year)

```{r b-disadv}
#| label: b-disadv

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Disadvantaged Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_FSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      log(remained_in_the_same_school) +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Non-Disadvantaged Pupils (per year)

```{r b-nondisadv}
#| label: b-nondisadv

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Non-Disadvantaged Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_NFSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      log(remained_in_the_same_school) +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```


# Analysis C: Core Panel Models

Reduced specification with 5 fixed effects. All 4 years pooled
(including 2024-25). `year_label` as a random intercept.

## All Pupils

```{r c-all}
#| label: c-all

d <- core_model_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_all)
```

### Quick checks

```{r c-all-checks}
#| label: c-all-checks

fixef(mod_c_all)
print(VarCorr(mod_c_all))
print(r2(mod_c_all))
cat("Singular fit?", isSingular(mod_c_all), "\n")
```

## Disadvantaged Pupils

```{r c-disadv}
#| label: c-disadv

d <- core_model_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_disadv)
```

### Quick checks

```{r c-disadv-checks}
#| label: c-disadv-checks

fixef(mod_c_disadv)
print(VarCorr(mod_c_disadv))
print(r2(mod_c_disadv))
cat("Singular fit?", isSingular(mod_c_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r c-nondisadv}
#| label: c-nondisadv

d <- core_model_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_nondisadv)
```

### Quick checks

```{r c-nondisadv-checks}
#| label: c-nondisadv-checks

fixef(mod_c_nondisadv)
print(VarCorr(mod_c_nondisadv))
print(r2(mod_c_nondisadv))
cat("Singular fit?", isSingular(mod_c_nondisadv), "\n")
```


# Analysis D: Core Per-Year Models

Same 5 core fixed effects, fitted separately for each year. No
`year_label` random effect.

```{r d-setup}
#| label: d-setup

core_year_levels <- sort(unique(as.character(core_model_data$year_label)))
cat("Years to fit:", paste(core_year_levels, collapse = ", "), "\n")
```

## All Pupils (per year)

```{r d-all}
#| label: d-all

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- All Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr, !is.na(ATT8SCR), ATT8SCR > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Disadvantaged Pupils (per year)

```{r d-disadv}
#| label: d-disadv

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Disadvantaged Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_FSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Non-Disadvantaged Pupils (per year)

```{r d-nondisadv}
#| label: d-nondisadv

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Non-Disadvantaged Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_NFSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```


# Coefficient Commentary

This section compares coefficients across the three outcome groups for
both the **full panel model** (Analysis A: 9 predictors, 3 years) and
the **core panel model** (Analysis C: 5 predictors, all 4 years), and
then examines how the core per-year coefficients (Analysis D) evolve
over time.


## Full panel model: side-by-side coefficients

The full panel model (Analysis A) includes all 9 fixed effects —
critically including `PTPRIORLO` (% low KS2 prior attainment) and three
workforce variables (teacher retention, leadership pay, teacher absence).
These are only available for the three years with complete workforce
data, so **2024-25 is excluded**.

```{r commentary-full-coefs}
#| label: commentary-full-coefs

# Extract fixed effects from the three full panel models (Analysis A)
full_coef_table <- bind_rows(
  tibble(Group = "All Pupils",
         Term = names(fixef(mod_a_all)),
         Estimate = fixef(mod_a_all)),
  tibble(Group = "Disadvantaged",
         Term = names(fixef(mod_a_disadv)),
         Estimate = fixef(mod_a_disadv)),
  tibble(Group = "Non-Disadvantaged",
         Term = names(fixef(mod_a_nondisadv)),
         Estimate = fixef(mod_a_nondisadv))
) %>%
  pivot_wider(names_from = Group, values_from = Estimate) %>%
  mutate(across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(full_coef_table, align = "lrrr",
             caption = "Full panel model (Analysis A): fixed effect comparison (9 predictors, 3 years)")
```

## Core panel model: side-by-side coefficients

The core panel model (Analysis C) drops the four variables not available
for 2024-25 — `PTPRIORLO`, `remained_in_the_same_school`,
`teachers_on_leadership_pay_range_percent`, and
`average_number_of_days_taken` — leaving 5 fixed effects. In return it
covers **all 4 years** including 2024-25.

```{r commentary-core-coefs}
#| label: commentary-core-coefs

# Extract fixed effects from the three core panel models (Analysis C)
core_coef_table <- bind_rows(
  tibble(Group = "All Pupils",
         Term = names(fixef(mod_c_all)),
         Estimate = fixef(mod_c_all)),
  tibble(Group = "Disadvantaged",
         Term = names(fixef(mod_c_disadv)),
         Estimate = fixef(mod_c_disadv)),
  tibble(Group = "Non-Disadvantaged",
         Term = names(fixef(mod_c_nondisadv)),
         Estimate = fixef(mod_c_nondisadv))
) %>%
  pivot_wider(names_from = Group, values_from = Estimate) %>%
  mutate(across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(core_coef_table, align = "lrrr",
             caption = "Core panel model (Analysis C): fixed effect comparison (5 predictors, all 4 years)")
```

## Full vs core: direct comparison

The table below places the 5 shared predictors side by side for both
specifications, making it easy to see how dropping the workforce and
prior-attainment variables shifts the remaining coefficients.

```{r commentary-full-vs-core}
#| label: commentary-full-vs-core

# Shared terms between full and core (excluding intercept)
shared_terms <- intersect(
  core_coef_table$Term[core_coef_table$Term != "(Intercept)"],
  full_coef_table$Term[full_coef_table$Term != "(Intercept)"]
)

comparison <- full_coef_table %>%
  filter(Term %in% shared_terms) %>%
  rename(`Full: All` = `All Pupils`,
         `Full: Disadv.` = Disadvantaged,
         `Full: Non-Dis.` = `Non-Disadvantaged`) %>%
  left_join(
    core_coef_table %>%
      filter(Term %in% shared_terms) %>%
      rename(`Core: All` = `All Pupils`,
             `Core: Disadv.` = Disadvantaged,
             `Core: Non-Dis.` = `Non-Disadvantaged`),
    by = "Term"
  )

knitr::kable(comparison, align = "lrrrrrr",
             caption = "Shared predictor coefficients: full panel (A) vs core panel (C)")
```

## Coefficient stability over time

```{r commentary-yearly}
#| label: commentary-yearly
#| fig-height: 8
#| fig-width: 9
#| fig-cap: "Core per-year coefficient estimates across all four academic years, by outcome group"

# Re-fit core per-year models and collect coefficients in a tidy table
yearly_coefs <- list()

for (yr in core_year_levels) {
  for (outcome_name in c("all", "disadv", "nondisadv")) {

    outcome_var <- switch(outcome_name,
                          all = "ATT8SCR",
                          disadv = "ATT8SCR_FSM6CLA1A",
                          nondisadv = "ATT8SCR_NFSM6CLA1A")
    group_label <- switch(outcome_name,
                          all = "All Pupils",
                          disadv = "Disadvantaged",
                          nondisadv = "Non-Disadvantaged")

    d <- core_model_data %>%
      filter(year_label == yr,
             !is.na(!!sym(outcome_var)), !!sym(outcome_var) > 0) %>%
      droplevels()

    contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

    mod <- tryCatch(
      lmer(
        as.formula(paste0(
          "log(", outcome_var, ") ~ ",
          "log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) + ",
          "ADMPOL_PT + gorard_segregation + ",
          "(1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
        )),
        data = d, REML = TRUE, na.action = na.exclude,
        control = lmerControl(optimizer = "bobyqa",
                              optCtrl = list(maxfun = 20000))
      ),
      error = function(e) NULL
    )

    if (!is.null(mod)) {
      fe <- fixef(mod)
      yearly_coefs[[length(yearly_coefs) + 1]] <- tibble(
        Year = yr,
        Group = group_label,
        Term = names(fe),
        Estimate = fe
      )
    }
  }
}

yearly_coefs_df <- bind_rows(yearly_coefs) %>%
  filter(Term != "(Intercept)") %>%
  mutate(
    Group = factor(Group,
                   levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged"))
  )

ggplot(yearly_coefs_df,
       aes(x = Year, y = Estimate, colour = Group, group = Group)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ Term, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("All Pupils" = "#1565c0",
                                  "Disadvantaged" = "#c62828",
                                  "Non-Disadvantaged" = "#2e7d32")) +
  labs(y = "Coefficient estimate", x = "Academic year", colour = NULL,
       title = "Core per-year coefficient trajectories") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold", size = 9)
  )
```

## General observations

The commentary below draws on the coefficient tables and trajectories
above. Because the outcome is **log-transformed**, all coefficients are
**elasticities** (for log-transformed predictors) or
**semi-elasticities** (for linear predictors). A coefficient of --0.10
on a logged predictor means that a 1% increase in that predictor is
associated with roughly a 0.10% *decrease* in Attainment 8.

### 1. Full vs core model: how dropping variables changes the picture

The most important thing to note when comparing the two specifications
is that they do **not tell the same story** for every coefficient. The
full panel model (Analysis A) includes `PTPRIORLO` (% low KS2 prior
attainment) and three workforce variables that the core model lacks.
These omitted variables are **not random noise** — prior attainment in
particular is one of the strongest individual predictors of school-level
outcomes. When it is dropped from the core model, its explanatory power
has to go somewhere, and the remaining 5 predictors absorb it. This
produces some notable shifts:

- **`log(PTFSM6CLA1A)` (% FSM):** In the **full model**, this
  coefficient is **negative for All Pupils and Non-Disadvantaged**, and
  **flips to positive for Disadvantaged Pupils**. In the **core model**,
  FSM is negative across **all three groups** — the sign flip
  disappears. This is because prior attainment (`PTPRIORLO`) is strongly
  correlated with FSM: schools with more disadvantaged intakes also tend
  to have more pupils with low prior attainment. In the full model,
  `PTPRIORLO` picks up the prior-attainment component of the
  disadvantage penalty, freeing up `log(PTFSM6CLA1A)` to capture the
  *residual* relationship — which, for disadvantaged pupils specifically,
  is positive (the "critical mass" effect described below). When
  `PTPRIORLO` is absent, `log(PTFSM6CLA1A)` has to absorb both effects
  and the net result is negative.

- **`log(PERCTOT)` (% absence):** Becomes **more negative** in the core
  model. Again, with fewer controls absorbing variation, absence picks up
  some of the explanatory burden previously shared with workforce and
  prior-attainment variables.

- **`log(PNUMEAL)` (% EAL):** The sign and direction are broadly
  consistent across both specifications, but the magnitude changes.

The practical implication is that coefficient interpretation must always
be read **relative to the other variables in the model**. The core model
is useful for coverage (all 4 years) and for the Shiny app's simulator,
but the full model provides a more nuanced and arguably more accurate
decomposition of the relationships — at the cost of excluding the most
recent year.

### 2. Free school meals (FSM) — the sign flip (full model)

In the **full panel model** (Analysis A), `log(PTFSM6CLA1A)` shows the
most striking difference across outcome groups:

- **All Pupils and Non-Disadvantaged:** The coefficient is consistently
  **negative**. Schools with a higher share of disadvantaged pupils tend
  to have lower overall and non-disadvantaged ATT8 scores, even after
  controlling for absence, EAL, prior attainment, and segregation. This
  captures the well documented compositional effect: a more disadvantaged
  intake is associated with lower average attainment at school level.

- **Disadvantaged Pupils:** The coefficient **flips to positive**. After
  controlling for prior attainment and workforce factors, schools with a
  higher proportion of FSM-eligible pupils see their *disadvantaged
  subgroup specifically* score somewhat higher than predicted. One
  plausible explanation is that schools serving predominantly
  disadvantaged cohorts concentrate more resources, Pupil Premium
  funding, and pastoral support on those pupils — a kind of "critical
  mass" effect. Alternatively, in schools where almost all pupils are
  disadvantaged, the disadvantaged group essentially *is* the school
  average, removing the within-school composition penalty.

**Crucially, this sign flip only appears in the full model** where prior
attainment controls for the low-KS2 component of the disadvantage
penalty. In the core model (Analysis C), `log(PTFSM6CLA1A)` is
**negative for all three groups** because it is forced to absorb the
prior-attainment effect as well. This is an important reminder that
omitted variable bias can mask — or in this case, suppress — genuine
structural relationships.

### 3. Absence — the largest and most stable predictor

The coefficient on `log(PERCTOT)` (% overall absence) is **strongly
negative** for all three groups in both specifications and is typically
the largest predictor in absolute terms. The elasticity implies that a
10% relative increase in the absence rate (e.g. from 5.0% to 5.5%) is
associated with a substantial drop in ATT8 on the log scale.

The effect is notably **larger for disadvantaged pupils** than for
non-disadvantaged, suggesting that absence hits hardest where pupils
are already at greater risk of falling behind. This pattern is
consistent over all four years and in both the full and core models.

The coefficient is **larger in the core model** than in the full model,
consistent with absence partially absorbing the effect of the dropped
workforce and prior-attainment variables.

### 4. English as an additional language (EAL)

`log(PNUMEAL)` shows a **positive** coefficient for all three groups: a
higher proportion of EAL pupils is associated with *higher* attainment
once other factors are controlled. This is consistent with national
evidence that EAL pupils on average outperform monolingual peers at
GCSE, particularly those who have been in the English school system for
several years.

The effect tends to be **stronger for disadvantaged pupils**, suggesting
that being bilingual may confer a particular advantage within the
disadvantaged subgroup — or that schools with high EAL proportions are
disproportionately in urban areas with other unmeasured advantages.

### 5. Interpreting log-transformed variables — effects at the extremes

Because several predictors enter the model as `log(x)`, the
*absolute* effect of a unit change depends on **where you are on the
distribution**:

- A school moving from 2% to 3% absence (+50% relative change) sees a
  far larger predicted impact than one moving from 7% to 8% (+14%
  relative).
- Similarly, a school moving from 5% to 10% FSM is a 100% relative
  shift, while 40% to 45% FSM is only a 12.5% relative shift.

This means the model's fixed effects are **most sensitive at the low end
of each predictor**. Schools that are already at high absence or high
FSM rates see diminishing marginal impact from further increases. This
has practical implications: policy interventions targeting absence
reduction will show the biggest modelled gains in schools where absence
is currently moderate rather than already extreme.

### 6. Gorard segregation index

The Gorard segregation index coefficient is **negative** for all three
groups — higher within-LA segregation is associated with lower
attainment — but the magnitude varies. It tends to be larger (more
negative) for **disadvantaged pupils**, suggesting that sorting pupils
into high-deprivation and low-deprivation schools hurts the
disadvantaged group more than the non-disadvantaged.

### 7. Admissions policy

`ADMPOL_PT` (a continuous admissions policy indicator) is **positive**
across all groups, implying that more selective admissions policies are
associated with higher attainment after controlling for intake
characteristics. However, the coefficient is **relatively small and
sometimes unstable** across years, particularly for the disadvantaged
group, where the confidence intervals are wide.

### 8. Workforce variables (full model only)

In Analysis A, the three workforce variables add useful context:

- **Teacher retention** (`log(remained_in_the_same_school)`): Positive
  coefficient — higher staff stability is associated with higher
  attainment. The effect is broadly similar across groups.
- **Leadership pay** (`teachers_on_leadership_pay_range_percent`):
  Typically small and sometimes non-significant, suggesting that the
  proportion of teachers on leadership pay scales is not a strong
  predictor once other factors are controlled.
- **Teacher absence** (`log(average_number_of_days_taken)`): Negative
  coefficient — schools where teachers are absent more tend to have
  lower ATT8 scores. This echoes the pupil absence finding and
  reinforces that attendance (by both pupils and staff) is strongly
  associated with outcomes.

### 9. Year-to-year stability

The coefficient trajectory plots above show that the core per-year
estimates are **remarkably stable** across 2021-22 to 2024-25, with
most terms moving within a narrow band. The main exception is 2021-22,
which was the first full examination year after the pandemic disruption
and occasionally shows slightly different coefficient magnitudes.

### 10. What is missing from the core model — and why it matters

The core model exists because workforce data (from the School Workforce
Census) and KS2 prior-attainment data are **not available for all
years**:

- **Prior attainment (`PTPRIORLO`):** KS2 SATs were cancelled during
  COVID, meaning the 2024-25 cohort has no standardised prior-attainment
  score. This is arguably the most consequential omission, since prior
  attainment is one of the strongest predictors of GCSE outcomes at
  school level. Its absence from the core model causes other predictors
  — particularly FSM and absence — to absorb some of its effect,
  inflating their coefficients and, in the case of FSM, suppressing
  the sign flip visible in the full model.
- **Workforce variables** (`remained_in_the_same_school`,
  `teachers_on_leadership_pay_range_percent`,
  `average_number_of_days_taken`): School Workforce Census data follows
  a different release timetable and is not yet available for 2024-25 at
  the time of analysis. While individually less powerful than prior
  attainment, collectively these three variables add meaningful
  explanatory power — particularly teacher retention and teacher
  absence, which are moderately correlated with both pupil absence and
  FSM rates.

The core model should therefore be understood as a **pragmatic
trade-off**: broader temporal coverage in exchange for a less complete
decomposition of the relationships driving attainment. The full model
remains the preferred specification for understanding *why* schools
differ, while the core model is better suited for *tracking* relative
performance over time — including the most recent year.


# Brighton & Hove: A Local Case Study

```{r bh-setup}
#| label: bh-setup

HIGHLIGHT_LA <- "Brighton and Hove"

bh_data <- panel %>%
  filter(LANAME == HIGHLIGHT_LA,
         MINORGROUP %in% c("Academy", "Maintained school"))

cat("Brighton & Hove schools in panel:", n_distinct(bh_data$URN), "\n")
cat("School-year rows:", nrow(bh_data), "\n")
cat("Years:", paste(sort(unique(as.character(bh_data$year_label))), collapse = ", "), "\n")
```

## School profiles

```{r bh-profiles}
#| label: bh-profiles

bh_profiles <- bh_data %>%
  filter(!is.na(ATT8SCR)) %>%
  group_by(School = SCHNAME) %>%
  summarise(
    Years = n(),
    `Mean ATT8 (All)` = round(mean(ATT8SCR, na.rm = TRUE), 1),
    `Mean ATT8 (Disadv.)` = round(mean(ATT8SCR_FSM6CLA1A, na.rm = TRUE), 1),
    `Mean ATT8 (Non-Dis.)` = round(mean(ATT8SCR_NFSM6CLA1A, na.rm = TRUE), 1),
    `Mean % FSM` = round(mean(PTFSM6CLA1A, na.rm = TRUE), 1),
    `Mean % Absence` = round(mean(PERCTOT, na.rm = TRUE), 1),
    `Mean % EAL` = round(mean(PNUMEAL, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(`Mean ATT8 (All)`))

knitr::kable(bh_profiles,
             caption = paste0("Brighton & Hove secondary schools: ",
                              "average characteristics across available years"))
```

## What the log transform means in practice: Brighton & Hove examples

The table below shows the predictor values for Brighton & Hove schools
and their log-transformed equivalents. This illustrates why a school at
the low end of a predictor sees a much bigger modelled impact from a
small absolute change than a school already at the high end.

```{r bh-log-illustration}
#| label: bh-log-illustration

bh_latest <- bh_data %>%
  filter(!is.na(ATT8SCR)) %>%
  group_by(SCHNAME) %>%
  filter(year_label == max(year_label)) %>%
  ungroup()

log_illustration <- bh_latest %>%
  select(School = SCHNAME, year_label,
         PTFSM6CLA1A, PERCTOT, PNUMEAL) %>%
  mutate(
    `log(% FSM)` = round(log(PTFSM6CLA1A), 3),
    `log(% Absence)` = round(log(PERCTOT), 3),
    `log(% EAL)` = round(log(PNUMEAL), 3),
    PTFSM6CLA1A = round(PTFSM6CLA1A, 1),
    PERCTOT = round(PERCTOT, 1),
    PNUMEAL = round(PNUMEAL, 1)
  ) %>%
  arrange(PTFSM6CLA1A) %>%
  rename(Year = year_label,
         `% FSM` = PTFSM6CLA1A,
         `% Absence` = PERCTOT,
         `% EAL` = PNUMEAL)

knitr::kable(log_illustration,
             caption = paste0("Brighton & Hove schools: raw and log-transformed ",
                              "predictor values (most recent year)"))
```

Consider two hypothetical Brighton & Hove schools:

- **School A** with 10% FSM: `log(10) = 2.30`. If FSM rises by 5
  percentage points to 15%, `log(15) = 2.71` — a shift of **+0.41** on
  the log scale.
- **School B** with 40% FSM: `log(40) = 3.69`. If FSM rises by the same
  5 percentage points to 45%, `log(45) = 3.81` — a shift of only
  **+0.12** on the log scale.

Because the model operates on the log scale, the *same absolute
increase* in FSM rate produces a predicted effect that is **more than
three times larger** for the lower-FSM school. The same logic applies to
absence and EAL. This is why policy interventions — whether reducing
absence or managing intake composition — can have dramatically different
predicted impacts depending on where a school sits on the distribution.

## Brighton & Hove: observed vs predicted

```{r bh-obs-pred}
#| label: bh-obs-pred
#| fig-height: 7
#| fig-width: 9
#| fig-cap: "Observed vs core-model predicted ATT8 for Brighton & Hove schools across all years"

bh_plot_data <- bh_data %>%
  filter(!is.na(ATT8SCR)) %>%
  select(School = SCHNAME, Year = year_label,
         ATT8SCR, predicted_ATT8SCR_core, residual_ATT8SCR_core,
         ATT8SCR_FSM6CLA1A, predicted_ATT8SCR_FSM6CLA1A_core,
         ATT8SCR_NFSM6CLA1A, predicted_ATT8SCR_NFSM6CLA1A_core) %>%
  pivot_longer(
    cols = -c(School, Year),
    names_to = c("measure", "group"),
    names_pattern = "(ATT8SCR|predicted_ATT8SCR|residual_ATT8SCR)_(.*)"
  )

# Simpler approach: build one panel per outcome
outcomes_bh <- list(
  list(obs = "ATT8SCR", pred = "predicted_ATT8SCR_core",
       label = "All Pupils"),
  list(obs = "ATT8SCR_FSM6CLA1A", pred = "predicted_ATT8SCR_FSM6CLA1A_core",
       label = "Disadvantaged"),
  list(obs = "ATT8SCR_NFSM6CLA1A", pred = "predicted_ATT8SCR_NFSM6CLA1A_core",
       label = "Non-Disadvantaged")
)

bh_long <- map_dfr(outcomes_bh, function(o) {
  bh_data %>%
    filter(!is.na(!!sym(o$obs)), !is.na(!!sym(o$pred))) %>%
    transmute(
      School = SCHNAME,
      Year = year_label,
      Group = o$label,
      Observed = !!sym(o$obs),
      Predicted = !!sym(o$pred),
      Residual = Observed - Predicted
    )
}) %>%
  mutate(Group = factor(Group,
                        levels = c("All Pupils", "Disadvantaged",
                                   "Non-Disadvantaged")))

# Scatter: observed vs predicted, faceted by group
rng <- range(c(bh_long$Observed, bh_long$Predicted), na.rm = TRUE)

ggplot(bh_long, aes(x = Predicted, y = Observed, colour = School)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(size = 2, alpha = 0.8) +
  facet_wrap(~ Group, ncol = 3) +
  coord_cartesian(xlim = rng, ylim = rng) +
  labs(x = "Predicted ATT8 (core model)", y = "Observed ATT8",
       colour = NULL,
       title = "Brighton & Hove: observed vs predicted Attainment 8") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    strip.text = element_text(face = "bold")
  ) +
  guides(colour = guide_legend(nrow = 3))
```

## Brighton & Hove: residuals by school

```{r bh-residuals}
#| label: bh-residuals
#| fig-height: 6
#| fig-width: 9
#| fig-cap: "Model residuals (observed - predicted) for Brighton & Hove schools by outcome group and year"

ggplot(bh_long, aes(x = Year, y = Residual, colour = School, group = School)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2) +
  geom_line(linewidth = 0.5, alpha = 0.6) +
  facet_wrap(~ Group, ncol = 3, scales = "free_y") +
  labs(x = "Academic year", y = "Residual (Observed - Predicted)",
       colour = NULL,
       title = "Brighton & Hove: school-level residuals over time") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  ) +
  guides(colour = guide_legend(nrow = 3))
```

## Brighton & Hove: commentary

This section links the general coefficient patterns back to the specific
schools in Brighton & Hove.

```{r bh-commentary-data}
#| label: bh-commentary-data

# National distribution context
national_summary <- core_model_data %>%
  filter(!is.na(ATT8SCR)) %>%
  summarise(
    fsm_p25 = quantile(PTFSM6CLA1A, 0.25, na.rm = TRUE),
    fsm_p50 = quantile(PTFSM6CLA1A, 0.50, na.rm = TRUE),
    fsm_p75 = quantile(PTFSM6CLA1A, 0.75, na.rm = TRUE),
    abs_p25 = quantile(PERCTOT, 0.25, na.rm = TRUE),
    abs_p50 = quantile(PERCTOT, 0.50, na.rm = TRUE),
    abs_p75 = quantile(PERCTOT, 0.75, na.rm = TRUE),
    eal_p25 = quantile(PNUMEAL, 0.25, na.rm = TRUE),
    eal_p50 = quantile(PNUMEAL, 0.50, na.rm = TRUE),
    eal_p75 = quantile(PNUMEAL, 0.75, na.rm = TRUE)
  )

cat("National quartiles (core model data):\n")
cat(sprintf("  %% FSM:     P25=%.1f  P50=%.1f  P75=%.1f\n",
            national_summary$fsm_p25, national_summary$fsm_p50,
            national_summary$fsm_p75))
cat(sprintf("  %% Absence: P25=%.1f  P50=%.1f  P75=%.1f\n",
            national_summary$abs_p25, national_summary$abs_p50,
            national_summary$abs_p75))
cat(sprintf("  %% EAL:     P25=%.1f  P50=%.1f  P75=%.1f\n",
            national_summary$eal_p25, national_summary$eal_p50,
            national_summary$eal_p75))

cat("\nBrighton & Hove schools relative to national distribution:\n")
bh_latest %>%
  arrange(SCHNAME) %>%
  transmute(
    School = SCHNAME,
    `% FSM` = round(PTFSM6CLA1A, 1),
    `FSM quartile` = case_when(
      PTFSM6CLA1A <= national_summary$fsm_p25 ~ "Bottom 25%",
      PTFSM6CLA1A <= national_summary$fsm_p50 ~ "Q2",
      PTFSM6CLA1A <= national_summary$fsm_p75 ~ "Q3",
      TRUE ~ "Top 25%"
    ),
    `% Absence` = round(PERCTOT, 1),
    `Absence quartile` = case_when(
      PERCTOT <= national_summary$abs_p25 ~ "Bottom 25%",
      PERCTOT <= national_summary$abs_p50 ~ "Q2",
      PERCTOT <= national_summary$abs_p75 ~ "Q3",
      TRUE ~ "Top 25%"
    )
  ) %>%
  knitr::kable(caption = "Brighton & Hove schools: national quartile position")
```

Brighton & Hove provides a useful illustration of the coefficient
patterns because its schools span a **wide range** of intake
characteristics within a single local authority:

**The FSM sign-flip in practice:** Schools in Brighton & Hove with lower
FSM rates will see the negative `log(PTFSM6CLA1A)` coefficient
working in their favour for overall and non-disadvantaged ATT8 (lower
deprivation → higher predicted score). But for the *disadvantaged*
model, the positive coefficient means that schools with higher FSM rates
actually get a modest upward adjustment for their disadvantaged pupils.
A school with, say, 40% FSM will have a higher predicted disadvantaged
ATT8 than one with 15% FSM, all else equal — the model is capturing
the idea that disadvantaged pupils may do relatively better in schools
where they form a larger share of the cohort.

**The log-scale sensitivity:** Brighton & Hove schools with relatively
low absence rates sit on the steep part of the log curve. For these
schools, even a small absolute increase in absence (e.g. 4.5% to 5.5%)
translates into a meaningful drop in predicted ATT8. In contrast, a
school already at 8% or 9% absence would need a larger absolute rise to
see the same predicted impact. This means that **protecting low-absence
schools from deterioration** may be just as important as tackling
chronic absence elsewhere — a small slip at the low end has an
outsized effect in the model.

**Residual patterns:** Schools that consistently sit *above* the
diagonal on the observed-vs-predicted plot are outperforming the model's
expectations given their intake. Those consistently below are
under-performing relative to their statistical peers. Persistent
residuals in one direction suggest that something not captured by the
five core predictors — leadership quality, curriculum design, community
factors — is driving a school's relative position.


# Quick Reference

## Formula summary

| Analysis | Fixed Effects | Random Effects | Data |
|----------|--------------|----------------|------|
| A (Full Panel) | 9 predictors (5 log, 4 linear) | `(1\|year) + (1\|Ofsted) + (1\|region/LA)` | 3 years, workforce required |
| B (Full Per-Year) | Same 9 predictors | `(1\|Ofsted) + (1\|region/LA)` | Each year separately |
| C (Core Panel) | 5 predictors (3 log, 2 linear) | `(1\|year) + (1\|Ofsted) + (1\|region/LA)` | All 4 years |
| D (Core Per-Year) | Same 5 predictors | `(1\|Ofsted) + (1\|region/LA)` | Each year separately |

## Variable key

| Variable | In formula as | Description |
|----------|--------------|-------------|
| `ATT8SCR` | `log(ATT8SCR)` | Attainment 8 score (all pupils) |
| `ATT8SCR_FSM6CLA1A` | `log(ATT8SCR_FSM6CLA1A)` | Attainment 8 score (disadvantaged) |
| `ATT8SCR_NFSM6CLA1A` | `log(ATT8SCR_NFSM6CLA1A)` | Attainment 8 score (non-disadvantaged) |
| `PTFSM6CLA1A` | `log(PTFSM6CLA1A)` | % eligible for Free School Meals |
| `PERCTOT` | `log(PERCTOT)` | % total absence |
| `PNUMEAL` | `log(PNUMEAL)` | % English as Additional Language |
| `PTPRIORLO` | `PTPRIORLO` | % low KS2 prior attainment |
| `ADMPOL_PT` | `ADMPOL_PT` | Admissions policy indicator |
| `gorard_segregation` | `gorard_segregation` | Gorard segregation index |
| `remained_in_the_same_school` | `log(remained_in_the_same_school)` | Teacher retention (%) |
| `teachers_on_leadership_pay_range_percent` | as-is | % teachers on leadership pay |
| `average_number_of_days_taken` | `log(average_number_of_days_taken)` | Average teacher absence days |
| `OFSTEDRATING_1` | `(1\|OFSTEDRATING_1)` | Ofsted rating (random intercept) |
| `gor_name` | `(1\|gor_name/LANAME)` | Government Office Region |
| `LANAME` | nested in `gor_name` | Local Authority |
| `year_label` | `(1\|year_label)` | Academic year (panel models only) |
