---
title: "School Attainment Multilevel Model Results"
subtitle: "Multilevel analysis of KS4 Attainment 8 scores across English secondary schools"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: ucl-theme.scss
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
    df-print: kable
    embed-resources: true
    include-in-header: header-logos.html
    include-after-body: footer-logos.html
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 5
    geometry: margin=2.5cm
    keep-tex: false
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(here)

source(here::here("R", "helpers.R"))

# ---- Load all saved outputs ----
panel_data       <- readRDS(here::here("data", "panel_data.rds"))
model_data       <- readRDS(here::here("data", "model_data.rds"))
panel_models     <- readRDS(here::here("data", "models.rds"))
panel_diag       <- readRDS(here::here("data", "model_diagnostics.rds"))
yearly_models    <- readRDS(here::here("data", "models_yearly.rds"))
yearly_diag      <- readRDS(here::here("data", "model_diagnostics_yearly.rds"))

# Core models (reduced specification, all 4 years including 2024-25)
core_model_data  <- readRDS(here::here("data", "model_data_core.rds"))
core_models      <- readRDS(here::here("data", "models_core.rds"))
core_diag        <- readRDS(here::here("data", "model_diagnostics_core.rds"))
core_yearly_models <- readRDS(here::here("data", "models_yearly_core.rds"))
core_yearly_diag <- readRDS(here::here("data", "model_diagnostics_yearly_core.rds"))

# Imputed full models (Analysis E: 9 predictors, all 4 years with carry-forward imputed 2024-25)
imputed_model_data <- readRDS(here::here("data", "model_data_imputed.rds"))
imputed_models     <- readRDS(here::here("data", "models_imputed.rds"))
imputed_diag       <- readRDS(here::here("data", "model_diagnostics_imputed.rds"))

# Outcome configuration
OUTCOMES <- list(
  all               = "ATT8SCR",
  disadvantaged     = "ATT8SCR_FSM6CLA1A",
  non_disadvantaged = "ATT8SCR_NFSM6CLA1A"
)

outcome_labels <- c(
  all               = "All Pupils",
  disadvantaged     = "Disadvantaged",
  non_disadvantaged = "Non-Disadvantaged"
)

# Ensure year_label is a factor for consistent ordering
model_data <- model_data %>%
  mutate(year_label = factor(year_label))

# Theme for publication-quality plots
theme_pub <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```


# Data Overview

For full details on data sources, panel construction, variable definitions, descriptive statistics, distributions, and bivariate relationships, see the [Data Overview report](https://adamdennett.github.io/school_attainment_tool/data_overview.html).

::: {.callout-tip}
## Model code & case study
The full `lmer()` code for all model specifications (Analyses A–E) is available in the [Model Experiments report](https://adamdennett.github.io/school_attainment_tool/model_experiments.html), which provides human-readable, self-contained implementations for manual inspection and reproducibility. A detailed application of the model to a single local authority — including maps, non-linear effect interpretation, and policy recommendations — is presented in the [Brighton & Hove Case Study](https://adamdennett.github.io/school_attainment_tool/brighton_case_study.html).
:::


::: {.callout-note}
## Sample size note

The full 9-predictor models (Analyses A, B, and E) are fitted on
approximately 12,200 school-year observations from ~13,000 mainstream
(Academy + Maintained) school-year records. The main source of attrition
is the `average_number_of_days_taken` (teacher sickness absence) variable,
which drops ~570 observations — predominantly in 2021-22 and 2022-23
where the workforce census sickness data was patchier (likely due to
COVID-era reporting disruption). Around 420 of these are NA values and
~150 are zeros excluded by the `> 0` requirement for log transformation.
Academies account for ~80% of the losses. See the
[dropped-rows diagnostic](model_experiments.html#dropped-rows-diagnostic)
in the Model Experiments report for a full breakdown.
:::

# Panel Models (Analysis A)

The panel models pool all four academic years, with `year_label` entering as a random intercept alongside Ofsted rating and local authority (nested within region). This specification treats year-to-year variation as a variance component rather than a fixed trend.

$$
\log(y_{ij}) = \mathbf{X}_{ij}\boldsymbol{\beta} + u_{\text{year}} + u_{\text{Ofsted}} + u_{\text{region}} + u_{\text{LA}} + \varepsilon_{ij}
$$


## Fixed Effects

```{r tbl-panel-fixed}
#| label: tbl-panel-fixed
#| tbl-cap: "Fixed effect estimates from panel models (all years pooled, year as random effect)"

# Build combined table
panel_fe <- map_dfr(names(panel_models), function(mn) {
  m <- panel_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    transmute(
      Model = outcome_labels[mn],
      Term = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      Estimate = round(Estimate, 4),
      `Std. Error` = round(`Std. Error`, 4),
      `t value` = round(`t value`, 2),
      `p value` = ifelse(`Pr(>|t|)` < 0.001, "< 0.001",
                         ifelse(`Pr(>|t|)` < 0.01, "< 0.01",
                                ifelse(`Pr(>|t|)` < 0.05, "< 0.05",
                                       round(`Pr(>|t|)`, 3))))
    )
})

panel_fe %>%
  kable(align = "llrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Coefficient Comparison Plot

```{r fig-panel-coefs}
#| label: fig-panel-coefs
#| fig-cap: "Fixed effect estimates with 95% confidence intervals (panel models)"
#| fig-height: 7

panel_coef_data <- map_dfr(names(panel_models), function(mn) {
  m <- panel_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      model = outcome_labels[mn],
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

ggplot(panel_coef_data,
       aes(x = estimate, y = reorder(term_label, estimate),
           colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi),
                  position = position_dodge(width = 0.6), size = 0.4) +
  scale_colour_manual(values = c("All Pupils" = "#2e6260",
                                 "Disadvantaged" = "#4e3c56",
                                 "Non-Disadvantaged" = "#abc766"),
                      name = NULL) +
  labs(x = "Estimate (log scale)", y = NULL,
       subtitle = "Points = estimate; bars = 95% CI") +
  theme_pub
```


## Random Effects & Model Fit

The multilevel model includes four levels of **random intercepts**, each allowing the baseline log(ATT8) to vary across groups:

1. **Academic year** (`year_label`) — captures system-wide year-on-year shifts in attainment (e.g. grade inflation, COVID recovery effects) that affect all schools equally within a given year.
2. **Ofsted rating** (`OFSTEDRATING_1`, harmonised to four levels) — allows schools with different inspection outcomes to have systematically different attainment baselines, beyond what the fixed-effect predictors explain.
3. **Government Office Region** (`gor_name`) — accounts for broad regional differences in attainment (e.g. London vs. the North East), which may reflect regional labour markets, funding patterns, or cultural factors.
4. **Local authority nested within region** (`LANAME:gor_name`) — the finest grouping level, capturing LA-specific effects such as local policy environments, demographic composition, or authority-level support services, over and above the regional baseline.

The table below reports the estimated variance for each random effect, along with the residual variance and overall model fit (marginal R², which reflects fixed effects only, and conditional R², which includes both fixed and random effects).

```{r tbl-panel-random}
#| label: tbl-panel-random
#| tbl-cap: "Random effect variance components and model fit statistics (panel models)"

panel_fit <- map_dfr(names(panel_models), function(mn) {
  m <- panel_models[[mn]]
  d <- panel_diag[[mn]]
  vc <- as.data.frame(VarCorr(m))

  # Extract variance for each grouping
  var_year   <- vc$vcov[vc$grp == "year_label"]
  var_ofsted <- vc$vcov[vc$grp == "OFSTEDRATING_1"]
  var_la     <- vc$vcov[vc$grp == "LANAME:gor_name"]
  var_region <- vc$vcov[vc$grp == "gor_name"]
  var_resid  <- d$sigma^2

  tibble(
    Model = outcome_labels[mn],
    var_year   = round(var_year, 6),
    var_ofsted = round(var_ofsted, 6),
    var_la     = round(var_la, 6),
    var_region = round(var_region, 6),
    var_resid  = round(var_resid, 6),
    r2m = round(d$r2_marginal, 4),
    r2c = round(d$r2_conditional, 4),
    N = d$n_obs
  )
})

names(panel_fit)[2:8] <- c(
  paste0("\u03c3\u00b2", "<sub>year</sub>"),
  paste0("\u03c3\u00b2", "<sub>Ofsted</sub>"),
  paste0("\u03c3\u00b2", "<sub>LA</sub>"),
  paste0("\u03c3\u00b2", "<sub>region</sub>"),
  paste0("\u03c3\u00b2", "<sub>resid</sub>"),
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>"
)

panel_fit %>%
  kable(format = "html", align = "lrrrrrrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


## Random Intercepts

```{r fig-random-intercepts}
#| label: fig-random-intercepts
#| fig-cap: "Random intercept estimates by grouping level (All Pupils panel model)"
#| fig-height: 8

m_all <- panel_models$all
re <- ranef(m_all)

# Year random effects
re_year <- tibble(
  year = rownames(re$year_label),
  intercept = re$year_label[, 1]
)

p_year <- ggplot(re_year, aes(x = reorder(year, intercept), y = intercept)) +
  geom_col(fill = "#2e6260", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title = "Year Random Intercepts", x = NULL, y = "Intercept (log scale)") +
  theme_pub

# Ofsted random effects (harmonised)
re_ofsted <- tibble(
  rating = rownames(re$OFSTEDRATING_1),
  intercept = re$OFSTEDRATING_1[, 1]
)

p_ofsted <- ggplot(re_ofsted, aes(x = reorder(rating, -intercept), y = intercept)) +
  geom_col(fill = "#abc766", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title = "Ofsted Rating Random Intercepts (harmonised)", x = NULL, y = "Intercept (log scale)") +
  theme_pub

# Region random effects
re_region <- tibble(
  region = rownames(re$gor_name),
  intercept = re$gor_name[, 1]
)

p_region <- ggplot(re_region, aes(x = reorder(region, intercept), y = intercept)) +
  geom_col(fill = "#4e3c56", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title = "Region Random Intercepts", x = NULL, y = "Intercept (log scale)") +
  theme_pub

p_year / p_ofsted / p_region
```


## Null Model: Variance Structure

The violin and box plots below show the distribution of Attainment 8 scores across the random-effect grouping variables. These visualise the between-group variance that the null (intercept-only) multilevel model captures — the same structure shown in more detail in the variance components table above.

```{r fig-null-violin-ofsted}
#| label: fig-null-violin-ofsted
#| fig-cap: "Distribution of Attainment 8 scores by Ofsted rating — visualising the null model variance structure"
#| fig-height: 6

# Fit null model: ATT8 with only Ofsted as random intercept
null_ofsted <- lmer(log(ATT8SCR) ~ 1 + (1 | OFSTEDRATING_1), data = model_data)
grand_mean_ofsted <- exp(fixef(null_ofsted)[1])

# Extract shrunken group means (BLUPs)
re_ofsted_null <- ranef(null_ofsted)$OFSTEDRATING_1 %>%
  as.data.frame() %>%
  rownames_to_column("OFSTEDRATING_1") %>%
  rename(blup = `(Intercept)`) %>%
  mutate(shrunken_mean = exp(fixef(null_ofsted)[1] + blup))

# Raw group means
raw_means_ofsted <- model_data %>%
  filter(!is.na(OFSTEDRATING_1), !is.na(ATT8SCR)) %>%
  group_by(OFSTEDRATING_1) %>%
  summarise(raw_mean = mean(ATT8SCR, na.rm = TRUE), n = n(), .groups = "drop") %>%
  left_join(re_ofsted_null, by = "OFSTEDRATING_1")

model_data %>%
  filter(!is.na(OFSTEDRATING_1), !is.na(ATT8SCR)) %>%
  ggplot(aes(x = OFSTEDRATING_1, y = ATT8SCR, fill = OFSTEDRATING_1)) +
  geom_violin(alpha = 0.3, colour = NA) +
  geom_boxplot(width = 0.15, alpha = 0.7, outlier.size = 0.5, outlier.alpha = 0.3) +
  geom_hline(yintercept = grand_mean_ofsted, linetype = "dashed",
             colour = "black", linewidth = 0.8) +
  geom_point(data = raw_means_ofsted, aes(y = raw_mean),
             shape = 4, size = 3, colour = "#7b132b", stroke = 1.2) +
  geom_point(data = raw_means_ofsted, aes(y = shrunken_mean),
             shape = 18, size = 4, colour = "#2e6260") +
  annotate("text", x = 0.55, y = grand_mean_ofsted + 1.5,
           label = paste0("Grand mean = ", round(grand_mean_ofsted, 1)),
           hjust = 0, size = 3.2, colour = "grey30") +
  labs(title = "ATT8 Distribution by Ofsted Rating (Null Model Structure)",
       subtitle = "Dashed = grand mean | Teal diamond = shrunken mean (BLUP) | Dark red cross = raw mean",
       x = "Ofsted Rating (harmonised)", y = "Attainment 8 Score") +
  theme_pub +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r fig-null-violin-region}
#| label: fig-null-violin-region
#| fig-cap: "Distribution of Attainment 8 scores by Government Office Region — null model variance structure"
#| fig-height: 6

null_region <- lmer(log(ATT8SCR) ~ 1 + (1 | gor_name), data = model_data)
grand_mean_region <- exp(fixef(null_region)[1])

re_region_null <- ranef(null_region)$gor_name %>%
  as.data.frame() %>%
  rownames_to_column("gor_name") %>%
  rename(blup = `(Intercept)`) %>%
  mutate(shrunken_mean = exp(fixef(null_region)[1] + blup))

raw_means_region <- model_data %>%
  filter(!is.na(gor_name), !is.na(ATT8SCR)) %>%
  group_by(gor_name) %>%
  summarise(raw_mean = mean(ATT8SCR, na.rm = TRUE), n = n(), .groups = "drop") %>%
  left_join(re_region_null, by = "gor_name")

model_data %>%
  filter(!is.na(gor_name), !is.na(ATT8SCR)) %>%
  ggplot(aes(x = reorder(gor_name, ATT8SCR, FUN = median),
             y = ATT8SCR, fill = gor_name)) +
  geom_violin(alpha = 0.3, colour = NA) +
  geom_boxplot(width = 0.15, alpha = 0.7, outlier.size = 0.5, outlier.alpha = 0.3) +
  geom_hline(yintercept = grand_mean_region, linetype = "dashed",
             colour = "black", linewidth = 0.8) +
  geom_point(data = raw_means_region,
             aes(x = reorder(gor_name, raw_mean), y = raw_mean),
             shape = 4, size = 3, colour = "#7b132b", stroke = 1.2) +
  geom_point(data = raw_means_region,
             aes(x = reorder(gor_name, raw_mean), y = shrunken_mean),
             shape = 18, size = 4, colour = "#2e6260") +
  coord_flip() +
  labs(title = "ATT8 Distribution by Region (Null Model Structure)",
       subtitle = "Dashed = grand mean | Teal diamond = shrunken mean (BLUP) | Dark red cross = raw mean",
       x = NULL, y = "Attainment 8 Score") +
  theme_pub +
  theme(legend.position = "none")
```

```{r fig-null-violin-ofsted-year}
#| label: fig-null-violin-ofsted-year
#| fig-cap: "Distribution of Attainment 8 by Ofsted rating, faceted by academic year"
#| fig-height: 8

model_data %>%
  filter(!is.na(OFSTEDRATING_1), !is.na(ATT8SCR)) %>%
  ggplot(aes(x = OFSTEDRATING_1, y = ATT8SCR, fill = OFSTEDRATING_1)) +
  geom_violin(alpha = 0.3, colour = NA) +
  geom_boxplot(width = 0.15, alpha = 0.7, outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_wrap(~ year_label, ncol = 2) +
  labs(title = "ATT8 by Ofsted Rating — by Academic Year",
       subtitle = "Violin + boxplot shows the distributional shape within each Ofsted category",
       x = "Ofsted Rating (harmonised)", y = "Attainment 8 Score") +
  theme_pub +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```


## Caterpillar Plot: LA Random Intercepts

The caterpillar plot below displays the estimated random intercepts for each local authority (nested within region), ordered by magnitude. Error bars show 95% confidence intervals (±1.96 posterior SE). Local authorities whose interval does not cross zero are meaningfully different from the overall mean, after accounting for fixed effects and other random effects. Brighton and Hove is highlighted in pink.

```{r fig-caterpillar-la}
#| label: fig-caterpillar-la
#| fig-cap: "Caterpillar plot of LA random intercepts (imputed full panel model, All Pupils). Top and bottom 10 LAs labelled; Brighton and Hove in pink."
#| fig-height: 8
#| fig-width: 10

library(ggrepel)

# Extract LA random effects with conditional variances
re_list_imp <- ranef(imputed_models$all, condVar = TRUE)
re_la <- as.data.frame(re_list_imp$`LANAME:gor_name`)
re_la$group <- rownames(re_list_imp$`LANAME:gor_name`)

# Posterior standard errors
se_la <- sqrt(attr(re_list_imp$`LANAME:gor_name`, "postVar")[1, 1, ])
re_la$se <- se_la

re_la <- re_la %>%
  mutate(
    la_name = sub(":.*$", "", group),
    ci_lo = `(Intercept)` - 1.96 * se,
    ci_hi = `(Intercept)` + 1.96 * se,
    sig = ci_lo > 0 | ci_hi < 0,
    highlight = ifelse(grepl("Brighton", la_name), "Brighton and Hove", "Other")
  )

# Top and bottom 10 for labelling (+ always include Brighton)
extreme_la <- c(
  head(re_la$la_name[order(re_la$`(Intercept)`)], 10),
  tail(re_la$la_name[order(re_la$`(Intercept)`)], 10)
)
if (any(grepl("Brighton", re_la$la_name))) {
  extreme_la <- unique(c(extreme_la,
                          re_la$la_name[grepl("Brighton", re_la$la_name)]))
}

ggplot(re_la, aes(x = reorder(group, `(Intercept)`), y = `(Intercept)`)) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0, colour = "grey60", linewidth = 0.3) +
  geom_point(aes(colour = highlight), size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_text_repel(
    data = re_la %>% filter(la_name %in% extreme_la),
    aes(label = la_name),
    size = 2.5, max.overlaps = 25, force = 4, nudge_y = 0.005,
    segment.colour = "grey60", segment.size = 0.3
  ) +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                  "Brighton and Hove" = "#e16fca"),
                      guide = "none") +
  labs(title = "LA Random Intercepts — Imputed Full Panel Model (All Pupils)",
       subtitle = "Ordered by estimate; error bars = 95% CI. LAs whose CI excludes 0 differ meaningfully from average.",
       x = "Local Authority (ordered by random intercept)",
       y = "Random Intercept (log scale)") +
  theme_pub +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r fig-caterpillar-compare}
#| label: fig-caterpillar-compare
#| fig-cap: "Caterpillar plot of LA random intercepts compared across model specifications (All Pupils). Brighton and Hove highlighted in pink."
#| fig-height: 12
#| fig-width: 10

# Helper: extract LA random effects from any model
extract_la_re <- function(model, model_name) {
  re_list <- ranef(model, condVar = TRUE)
  re_df <- as.data.frame(re_list$`LANAME:gor_name`)
  re_df$group <- rownames(re_list$`LANAME:gor_name`)
  se <- sqrt(attr(re_list$`LANAME:gor_name`, "postVar")[1, 1, ])
  re_df %>%
    mutate(
      model = model_name,
      la_name = sub(":.*$", "", group),
      se = se,
      ci_lo = `(Intercept)` - 1.96 * se,
      ci_hi = `(Intercept)` + 1.96 * se,
      highlight = ifelse(grepl("Brighton", la_name), "Brighton and Hove", "Other")
    )
}

# Extract for all three panel specifications
re_compare <- bind_rows(
  extract_la_re(panel_models$all, "Analysis A: Full Panel (3 yrs)"),
  extract_la_re(core_models$all,  "Analysis C: Core Panel (4 yrs)"),
  extract_la_re(imputed_models$all, "Analysis E: Imputed Full Panel (4 yrs)")
)

# Order LAs by the imputed model's estimate
la_order <- re_compare %>%
  filter(grepl("Analysis E", model)) %>%
  arrange(`(Intercept)`) %>%
  pull(group)

re_compare <- re_compare %>%
  mutate(
    group = factor(group, levels = la_order),
    model = factor(model, levels = c(
      "Analysis A: Full Panel (3 yrs)",
      "Analysis C: Core Panel (4 yrs)",
      "Analysis E: Imputed Full Panel (4 yrs)"
    ))
  )

ggplot(re_compare, aes(x = group, y = `(Intercept)`)) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0, colour = "grey70", linewidth = 0.25) +
  geom_point(aes(colour = highlight), size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  facet_wrap(~ model, ncol = 1, scales = "free_y") +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                  "Brighton and Hove" = "#e16fca"),
                      guide = "none") +
  labs(title = "LA Random Intercepts Compared Across Model Specifications (All Pupils)",
       subtitle = "All panels ordered by Analysis E estimate; error bars = 95% CI",
       x = "Local Authority (ordered)",
       y = "Random Intercept (log scale)") +
  theme_pub +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        strip.text = element_text(size = 10))
```


## Residual Diagnostics

Having examined the random effects structure, we now turn to the model's residuals to check whether the key assumptions of the multilevel model are met: linearity, homoscedasticity (constant variance), and approximate normality of residuals.

```{r fig-panel-residuals}
#| label: fig-panel-residuals
#| fig-cap: "Residual diagnostics for the All Pupils panel model"
#| fig-height: 8

m_all <- panel_models$all

resid_data <- tibble(
  fitted = fitted(m_all),
  resid  = residuals(m_all, type = "pearson"),
  resid_std = resid / sd(resid)
)

p1 <- ggplot(resid_data, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.1, size = 0.5, colour = "#2e6260") +
  geom_hline(yintercept = 0, colour = "#7b132b", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, colour = "#7b132b", linewidth = 0.7) +
  labs(title = "Residuals vs Fitted", x = "Fitted values (log scale)", y = "Pearson residuals") +
  theme_pub

p2 <- ggplot(resid_data, aes(sample = resid_std)) +
  stat_qq(alpha = 0.1, size = 0.5, colour = "#2e6260") +
  stat_qq_line(colour = "#7b132b", linewidth = 0.7) +
  labs(title = "Normal Q-Q", x = "Theoretical quantiles", y = "Standardised residuals") +
  theme_pub

p3 <- ggplot(resid_data, aes(x = resid)) +
  geom_histogram(bins = 80, fill = "#2e6260", alpha = 0.7, colour = "white") +
  geom_vline(xintercept = 0, colour = "#7b132b", linetype = "dashed") +
  labs(title = "Residual Distribution", x = "Pearson residuals", y = "Frequency") +
  theme_pub

p4 <- ggplot(resid_data, aes(x = fitted, y = sqrt(abs(resid_std)))) +
  geom_point(alpha = 0.1, size = 0.5, colour = "#2e6260") +
  geom_smooth(method = "loess", se = FALSE, colour = "#7b132b", linewidth = 0.7) +
  labs(title = "Scale-Location", x = "Fitted values (log scale)",
       y = expression(sqrt("|Standardised residuals|"))) +
  theme_pub

(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Panel Model Diagnostics: All Pupils",
    subtitle = "log(ATT8SCR) ~ fixed effects + (1|year) + (1|Ofsted harmonised) + (1|region/LA)"
  )
```


# Per-Year Models (Analysis B)

Separate models are fitted for each academic year with the same fixed-effect specification but without a year term. This allows all coefficients to vary freely across years, revealing whether relationships between predictors and attainment are stable or shifting over time.


## Model Fit Comparison Across Years

```{r tbl-yearly-fit}
#| label: tbl-yearly-fit
#| tbl-cap: "Model fit statistics by year and outcome"

yearly_fit <- map_dfr(names(yearly_diag), function(yr) {
  map_dfr(names(yearly_diag[[yr]]), function(mn) {
    d <- yearly_diag[[yr]][[mn]]
    tibble(
      Year = yr,
      Outcome = outcome_labels[mn],
      N = ifelse(is.na(d$r2_marginal), NA_integer_, d$n_obs),
      r2m = round(d$r2_marginal, 4),
      r2c = round(d$r2_conditional, 4),
      sigma = round(d$sigma, 4)
    )
  })
})

names(yearly_fit)[4:6] <- c(
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>",
  "\u03c3"
)

yearly_fit %>%
  kable(format = "html", align = "llrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Coefficient Stability Across Years

```{r fig-yearly-coefs}
#| label: fig-yearly-coefs
#| fig-cap: "How fixed-effect estimates change across academic years (All Pupils models)"
#| fig-height: 9

yearly_coef_data <- map_dfr(names(yearly_models), function(yr) {
  m <- yearly_models[[yr]]$all
  if (is.null(m)) return(NULL)

  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      year = yr,
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

if (nrow(yearly_coef_data) > 0) {
  ggplot(yearly_coef_data,
         aes(x = year, y = estimate, group = 1)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), alpha = 0.15, fill = "#2e6260") +
    geom_line(colour = "#2e6260", linewidth = 0.7) +
    geom_point(colour = "#2e6260", size = 2) +
    facet_wrap(~ term_label, scales = "free_y", ncol = 3) +
    labs(x = "Academic Year", y = "Estimate (log scale)",
         subtitle = "Shaded band = 95% CI") +
    theme_pub +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
}
```


## Side-by-Side Coefficient Table (Per-Year, All Pupils)

```{r tbl-yearly-coefs}
#| label: tbl-yearly-coefs
#| tbl-cap: "Fixed effect estimates by year (All Pupils outcome). Significance: *** p<0.001, ** p<0.01, * p<0.05"

yearly_coef_wide <- yearly_coef_data %>%
  mutate(
    # Look up p-values from the original models
    sig = map2_chr(year, term_raw, function(yr, term) {
      m <- yearly_models[[yr]]$all
      if (is.null(m)) return("")
      fe <- as.data.frame(summary(m)$coefficients)
      p <- fe[term, "Pr(>|t|)"]
      if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else ""
    }),
    cell = paste0(round(estimate, 4), sig)
  ) %>%
  select(term_label, year, cell) %>%
  pivot_wider(names_from = year, values_from = cell)

yearly_coef_wide %>%
  rename(Variable = term_label) %>%
  kable(align = c("l", rep("r", ncol(yearly_coef_wide) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


# Core Models (All 4 Years)

School Workforce Census variables (`remained_in_the_same_school`, `teachers_on_leadership_pay_range_percent`, `average_number_of_days_taken`) and KS2 prior attainment (`PTPRIORLO`) are not yet available for 2024-25. To enable cross-year comparison including the latest year, a **core specification** is fitted using only the 5 predictors available across all four years:

$$
\log(y_{ij}) = \beta_0 + \beta_1\log(\text{FSM\%}) + \beta_2\log(\text{Absence\%}) + \beta_3\log(\text{EAL\%}) + \beta_4\text{Admissions} + \beta_5\text{Gorard} + u_{\text{year}} + u_{\text{Ofsted}} + u_{\text{region}} + u_{\text{LA}} + \varepsilon_{ij}
$$


## Core Panel: Fixed Effects

```{r tbl-core-panel-fixed}
#| label: tbl-core-panel-fixed
#| tbl-cap: "Fixed effect estimates from core panel models (all 4 years, reduced specification)"

core_panel_fe <- map_dfr(names(core_models), function(mn) {
  m <- core_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    transmute(
      Model = outcome_labels[mn],
      Term = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      Estimate = round(Estimate, 4),
      `Std. Error` = round(`Std. Error`, 4),
      `t value` = round(`t value`, 2),
      `p value` = ifelse(`Pr(>|t|)` < 0.001, "< 0.001",
                         ifelse(`Pr(>|t|)` < 0.01, "< 0.01",
                                ifelse(`Pr(>|t|)` < 0.05, "< 0.05",
                                       round(`Pr(>|t|)`, 3))))
    )
})

core_panel_fe %>%
  kable(align = "llrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Core Panel: Random Effects & Model Fit

```{r tbl-core-panel-random}
#| label: tbl-core-panel-random
#| tbl-cap: "Random effect variance components and model fit statistics (core panel models)"

core_panel_fit <- map_dfr(names(core_models), function(mn) {
  m <- core_models[[mn]]
  d <- core_diag[[mn]]
  vc <- as.data.frame(VarCorr(m))

  var_year   <- vc$vcov[vc$grp == "year_label"]
  var_ofsted <- vc$vcov[vc$grp == "OFSTEDRATING_1"]
  var_la     <- vc$vcov[vc$grp == "LANAME:gor_name"]
  var_region <- vc$vcov[vc$grp == "gor_name"]
  var_resid  <- d$sigma^2

  tibble(
    Model = outcome_labels[mn],
    var_year   = round(var_year, 6),
    var_ofsted = round(var_ofsted, 6),
    var_la     = round(var_la, 6),
    var_region = round(var_region, 6),
    var_resid  = round(var_resid, 6),
    r2m = round(d$r2_marginal, 4),
    r2c = round(d$r2_conditional, 4),
    N = d$n_obs
  )
})

names(core_panel_fit)[2:8] <- c(
  paste0("\u03c3\u00b2", "<sub>year</sub>"),
  paste0("\u03c3\u00b2", "<sub>Ofsted</sub>"),
  paste0("\u03c3\u00b2", "<sub>LA</sub>"),
  paste0("\u03c3\u00b2", "<sub>region</sub>"),
  paste0("\u03c3\u00b2", "<sub>resid</sub>"),
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>"
)

core_panel_fit %>%
  kable(format = "html", align = "lrrrrrrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


## Core Per-Year: Model Fit Comparison

```{r tbl-core-yearly-fit}
#| label: tbl-core-yearly-fit
#| tbl-cap: "Core model fit statistics by year and outcome (all 4 years)"

core_yearly_fit <- map_dfr(names(core_yearly_diag), function(yr) {
  map_dfr(names(core_yearly_diag[[yr]]), function(mn) {
    d <- core_yearly_diag[[yr]][[mn]]
    tibble(
      Year = yr,
      Outcome = outcome_labels[mn],
      N = ifelse(is.na(d$r2_marginal), NA_integer_, d$n_obs),
      r2m = round(d$r2_marginal, 4),
      r2c = round(d$r2_conditional, 4),
      sigma = round(d$sigma, 4)
    )
  })
})

names(core_yearly_fit)[4:6] <- c(
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>",
  "\u03c3"
)

core_yearly_fit %>%
  kable(format = "html", align = "llrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Core Per-Year: Coefficient Stability

```{r fig-core-yearly-coefs}
#| label: fig-core-yearly-coefs
#| fig-cap: "How core fixed-effect estimates change across all 4 academic years (All Pupils)"
#| fig-height: 7

core_yearly_coef_data <- map_dfr(names(core_yearly_models), function(yr) {
  m <- core_yearly_models[[yr]]$all
  if (is.null(m)) return(NULL)

  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      year = yr,
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

if (nrow(core_yearly_coef_data) > 0) {
  ggplot(core_yearly_coef_data,
         aes(x = year, y = estimate, group = 1)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), alpha = 0.15, fill = "#e16fca") +
    geom_line(colour = "#e16fca", linewidth = 0.7) +
    geom_point(colour = "#e16fca", size = 2) +
    facet_wrap(~ term_label, scales = "free_y", ncol = 3) +
    labs(x = "Academic Year", y = "Estimate (log scale)",
         subtitle = "Shaded band = 95% CI — core specification (5 predictors)") +
    theme_pub +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
}
```


## Core Per-Year: Side-by-Side Coefficients

```{r tbl-core-yearly-coefs}
#| label: tbl-core-yearly-coefs
#| tbl-cap: "Core fixed effect estimates by year (All Pupils). Significance: *** p<0.001, ** p<0.01, * p<0.05"

core_yearly_coef_wide <- core_yearly_coef_data %>%
  mutate(
    sig = map2_chr(year, term_raw, function(yr, term) {
      m <- core_yearly_models[[yr]]$all
      if (is.null(m)) return("")
      fe <- as.data.frame(summary(m)$coefficients)
      p <- fe[term, "Pr(>|t|)"]
      if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else ""
    }),
    cell = paste0(round(estimate, 4), sig)
  ) %>%
  select(term_label, year, cell) %>%
  pivot_wider(names_from = year, values_from = cell)

core_yearly_coef_wide %>%
  rename(Variable = term_label) %>%
  kable(align = c("l", rep("r", ncol(core_yearly_coef_wide) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


# Imputed Full Panel Models (Analysis E)

The imputed full model uses the same 9-predictor specification as Analysis A, but extends coverage to all four academic years by carry-forward imputing the four variables unavailable for 2024-25 (`PTPRIORLO`, `remained_in_the_same_school`, `teachers_on_leadership_pay_range_percent`, `average_number_of_days_taken`). Each school's 2023-24 value is used (or the 3-year mean as a fallback). This is the model used in the Shiny app.

$$
\log(y_{ij}) = \mathbf{X}_{ij}\boldsymbol{\beta} + u_{\text{year}} + u_{\text{Ofsted}} + u_{\text{region}} + u_{\text{LA}} + \varepsilon_{ij}
$$


## Imputed Panel: Fixed Effects

```{r tbl-imputed-panel-fixed}
#| label: tbl-imputed-panel-fixed
#| tbl-cap: "Fixed effect estimates from imputed full panel models (Analysis E: 9 predictors, all 4 years)"

imputed_fe <- map_dfr(names(imputed_models), function(mn) {
  m <- imputed_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    transmute(
      Model = outcome_labels[mn],
      Term = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      Estimate = round(Estimate, 4),
      `Std. Error` = round(`Std. Error`, 4),
      `t value` = round(`t value`, 2),
      `p value` = ifelse(`Pr(>|t|)` < 0.001, "< 0.001",
                         ifelse(`Pr(>|t|)` < 0.01, "< 0.01",
                                ifelse(`Pr(>|t|)` < 0.05, "< 0.05",
                                       round(`Pr(>|t|)`, 3))))
    )
})

imputed_fe %>%
  kable(align = "llrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Imputed Panel: Coefficient Comparison Plot

```{r fig-imputed-coefs}
#| label: fig-imputed-coefs
#| fig-cap: "Fixed effect estimates with 95% confidence intervals (imputed full panel models, Analysis E)"
#| fig-height: 7

imputed_coef_data <- map_dfr(names(imputed_models), function(mn) {
  m <- imputed_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      model = outcome_labels[mn],
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

ggplot(imputed_coef_data,
       aes(x = estimate, y = reorder(term_label, estimate),
           colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi),
                  position = position_dodge(width = 0.6), size = 0.4) +
  scale_colour_manual(values = c("All Pupils" = "#2e6260",
                                 "Disadvantaged" = "#4e3c56",
                                 "Non-Disadvantaged" = "#abc766"),
                      name = NULL) +
  labs(x = "Estimate (log scale)", y = NULL,
       subtitle = "Points = estimate; bars = 95% CI. Analysis E (imputed full, all 4 years)") +
  theme_pub
```


## Imputed Panel: Random Effects & Model Fit

```{r tbl-imputed-panel-random}
#| label: tbl-imputed-panel-random
#| tbl-cap: "Random effect variance components and model fit statistics (imputed full panel models, Analysis E)"

imputed_fit <- map_dfr(names(imputed_models), function(mn) {
  m <- imputed_models[[mn]]
  d <- imputed_diag[[mn]]
  vc <- as.data.frame(VarCorr(m))

  var_year   <- vc$vcov[vc$grp == "year_label"]
  var_ofsted <- vc$vcov[vc$grp == "OFSTEDRATING_1"]
  var_la     <- vc$vcov[vc$grp == "LANAME:gor_name"]
  var_region <- vc$vcov[vc$grp == "gor_name"]
  var_resid  <- d$sigma^2

  tibble(
    Model = outcome_labels[mn],
    var_year   = round(var_year, 6),
    var_ofsted = round(var_ofsted, 6),
    var_la     = round(var_la, 6),
    var_region = round(var_region, 6),
    var_resid  = round(var_resid, 6),
    r2m = round(d$r2_marginal, 4),
    r2c = round(d$r2_conditional, 4),
    N = d$n_obs
  )
})

names(imputed_fit)[2:8] <- c(
  paste0("\u03c3\u00b2", "<sub>year</sub>"),
  paste0("\u03c3\u00b2", "<sub>Ofsted</sub>"),
  paste0("\u03c3\u00b2", "<sub>LA</sub>"),
  paste0("\u03c3\u00b2", "<sub>region</sub>"),
  paste0("\u03c3\u00b2", "<sub>resid</sub>"),
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>"
)

imputed_fit %>%
  kable(format = "html", align = "lrrrrrrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


## Analysis A vs Analysis E: Coefficient Comparison

Adding one year of carry-forward imputed data (Analysis E) produces coefficients that are very close to the original 3-year full model (Analysis A). This is reassuring: three-quarters of the data is identical, and the imputed fourth year uses conservative carry-forward values.

```{r tbl-a-vs-e}
#| label: tbl-a-vs-e
#| tbl-cap: "Coefficient comparison: Analysis A (3-year full) vs Analysis E (4-year imputed full)"

a_vs_e <- bind_rows(
  tibble(Analysis = "A (Full, 3 yrs)",
         Term = names(fixef(panel_models$all)),
         `All Pupils` = fixef(panel_models$all),
         Disadvantaged = fixef(panel_models$disadvantaged),
         `Non-Disadvantaged` = fixef(panel_models$non_disadvantaged)),
  tibble(Analysis = "E (Imputed, 4 yrs)",
         Term = names(fixef(imputed_models$all)),
         `All Pupils` = fixef(imputed_models$all),
         Disadvantaged = fixef(imputed_models$disadvantaged),
         `Non-Disadvantaged` = fixef(imputed_models$non_disadvantaged))
) %>%
  mutate(across(where(is.numeric), \(x) round(x, 5)))

a_vs_e %>%
  kable(align = "llrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


# Observed vs Predicted

The scatter plots below show observed Attainment 8 scores against imputed-full-model fitted values (Analysis E) for each of the four academic years. Schools in **Brighton and Hove** are highlighted in pink. A linear regression line with R² is overlaid on each panel; the dashed grey line marks perfect prediction.

## All Pupils

```{r fig-obs-vs-pred-all}
#| label: fig-obs-vs-pred-all
#| fig-cap: "Observed vs predicted Attainment 8 — All Pupils (imputed full model, by year). Brighton and Hove schools in pink."
#| fig-height: 8
#| fig-width: 10

HIGHLIGHT_LA <- "Brighton and Hove"

# Prepare data with highlight flag
pred_data <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_imputed), !is.na(ATT8SCR)) %>%
  mutate(
    residual = ATT8SCR - predicted_ATT8SCR_imputed,
    abs_resid = abs(residual),
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA))
  ) %>%
  arrange(highlight)   # draw highlighted points on top

# Per-year R² labels
r2_labels <- pred_data %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR, predicted_ATT8SCR_imputed, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R\u00b2 = ", round(r2, 3), "  (n = ", scales::comma(n), ")"))

ggplot(pred_data, aes(x = predicted_ATT8SCR_imputed, y = ATT8SCR)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  geom_text(data = r2_labels,
            aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.05, vjust = 1.4,
            size = 3.3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                  "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.12,
                                 "Brighton and Hove" = 0.85), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 (imputed full model)", y = "Observed ATT8",
       colour = NULL) +
  coord_equal() +
  theme_pub +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11))
```

```{r}
library(plotly)
library(dplyr)
library(scales)

# 1. Prepare data (same as your original logic)
pred_data <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_imputed), !is.na(ATT8SCR)) %>%
  mutate(
    residual = ATT8SCR - predicted_ATT8SCR_imputed,
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA)),
    # Custom tooltip text
    tooltip_text = paste0(
      "LA: ", LANAME,
      "<br>Sch:", SCHNAME,
      "<br>Observed: ", round(ATT8SCR, 2),
      "<br>Predicted: ", round(predicted_ATT8SCR_imputed, 2),
      "<br>Residual: ", round(residual, 2)
    )
  ) %>%
  arrange(highlight)

# 2. Calculate R² labels
r2_labels <- pred_data %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR, predicted_ATT8SCR_imputed, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 3), " (n = ", comma(n), ")"))

# 3. Build the ggplot object
p <- ggplot(pred_data, aes(x = predicted_ATT8SCR_imputed, y = ATT8SCR)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  # We add 'text' aesthetic here for the Plotly tooltip
  geom_point(aes(colour = highlight, alpha = highlight, text = tooltip_text), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  # Note: geom_text can be finicky in ggplotly; using explicit coordinates helps
  geom_text(data = r2_labels,
            aes(x = 20, y = 85, label = label),
            inherit.aes = FALSE, size = 3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                 "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.2,
                                "Brighton and Hove" = 0.9), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 (imputed full model)", y = "Observed ATT8", colour = NULL) +
  theme_minimal() + # Use minimal for cleaner plotly conversion
  theme(legend.position = "bottom")

# 4. Convert to Plotly
ggplotly(p, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", x = 0.4, y = -0.1),
    margin = list(t = 50) # Space for facet titles
  )

```



## Disadvantaged Pupils

```{r fig-obs-vs-pred-disadv}
#| label: fig-obs-vs-pred-disadv
#| fig-cap: "Observed vs predicted Attainment 8 — Disadvantaged Pupils (imputed full model, by year). Brighton and Hove schools in pink."
#| fig-height: 8
#| fig-width: 10

pred_disadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_FSM6CLA1A_imputed), !is.na(ATT8SCR_FSM6CLA1A)) %>%
  mutate(
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA))
  ) %>%
  arrange(highlight)

r2_disadv <- pred_disadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_FSM6CLA1A, predicted_ATT8SCR_FSM6CLA1A_imputed,
             use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R\u00b2 = ", round(r2, 3), "  (n = ", scales::comma(n), ")"))

ggplot(pred_disadv, aes(x = predicted_ATT8SCR_FSM6CLA1A_imputed,
                         y = ATT8SCR_FSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  geom_text(data = r2_disadv,
            aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.05, vjust = 1.4,
            size = 3.3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                  "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.12,
                                 "Brighton and Hove" = 0.85), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Disadvantaged (imputed full model)",
       y = "Observed ATT8 — Disadvantaged",
       colour = NULL) +
  coord_equal() +
  theme_pub +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11))
```

```{r}
library(plotly)
library(dplyr)
library(scales)

# 1. Prepare data for Disadvantaged subset
pred_disadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_FSM6CLA1A_imputed), !is.na(ATT8SCR_FSM6CLA1A)) %>%
  mutate(
    residual = ATT8SCR_FSM6CLA1A - predicted_ATT8SCR_FSM6CLA1A_imputed,
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA)),
    # Enhanced tooltip for disadvantaged context
    tooltip_text = paste0(
      "<b>", SCHNAME, "</b><br>",  # Assumes SCHNAME exists in your data
      "LA: ", LANAME,
      "<br>Sch:", SCHNAME,
      "<br>Observed (Disadv): ", round(ATT8SCR_FSM6CLA1A, 2),
      "<br>Predicted (Disadv): ", round(predicted_ATT8SCR_FSM6CLA1A_imputed, 2),
      "<br>Residual: ", round(residual, 2)
    )
  ) %>%
  arrange(highlight)

# 2. Calculate R² for the Disadvantaged model
r2_disadv_stats <- pred_disadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_FSM6CLA1A, predicted_ATT8SCR_FSM6CLA1A_imputed, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 3), " (n = ", comma(n), ")"))

# 3. Build the ggplot object
p_disadv <- ggplot(pred_disadv, aes(x = predicted_ATT8SCR_FSM6CLA1A_imputed,
                                   y = ATT8SCR_FSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight, text = tooltip_text), size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  # Positioning labels at safe numeric coordinates (e.g., x=10, y=85)
  geom_text(data = r2_disadv_stats,
            aes(x = 10, y = 85, label = label),
            inherit.aes = FALSE, size = 3, colour = "grey30", hjust = 0) +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                 "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.15,
                                "Brighton and Hove" = 0.9), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Disadvantaged",
       y = "Observed ATT8 — Disadvantaged",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))

# 4. Final Plotly conversion
ggplotly(p_disadv, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", x = 0.3, y = -0.15),
    margin = list(t = 60)
  )
```


## Non-Disadvantaged Pupils

```{r fig-obs-vs-pred-nondisadv}
#| label: fig-obs-vs-pred-nondisadv
#| fig-cap: "Observed vs predicted Attainment 8 — Non-Disadvantaged Pupils (imputed full model, by year). Brighton and Hove schools in pink."
#| fig-height: 8
#| fig-width: 10

pred_nondisadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_NFSM6CLA1A_imputed), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
  mutate(
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA))
  ) %>%
  arrange(highlight)

r2_nondisadv <- pred_nondisadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_NFSM6CLA1A, predicted_ATT8SCR_NFSM6CLA1A_imputed,
             use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R\u00b2 = ", round(r2, 3), "  (n = ", scales::comma(n), ")"))

ggplot(pred_nondisadv, aes(x = predicted_ATT8SCR_NFSM6CLA1A_imputed,
                            y = ATT8SCR_NFSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  geom_text(data = r2_nondisadv,
            aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.05, vjust = 1.4,
            size = 3.3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                  "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.12,
                                 "Brighton and Hove" = 0.85), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Non-Disadvantaged (imputed full model)",
       y = "Observed ATT8 — Non-Disadvantaged",
       colour = NULL) +
  coord_equal() +
  theme_pub +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11))
```

```{r}
library(plotly)
library(dplyr)
library(scales)

# 1. Prepare data for Non-Disadvantaged subset
pred_nondisadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_NFSM6CLA1A_imputed), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
  mutate(
    residual = ATT8SCR_NFSM6CLA1A - predicted_ATT8SCR_NFSM6CLA1A_imputed,
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA)),
    # Custom tooltip
    tooltip_text = paste0(
      "<b>", SCHNAME, "</b><br>",
      "LA: ", LANAME,
      "<br>Sch:", SCHNAME,
      "<br>Observed (Non-Disadv): ", round(ATT8SCR_NFSM6CLA1A, 2),
      "<br>Predicted (Non-Disadv): ", round(predicted_ATT8SCR_NFSM6CLA1A_imputed, 2),
      "<br>Residual: ", round(residual, 2)
    )
  ) %>%
  arrange(highlight)

# 2. Calculate R² for the Non-Disadvantaged model
r2_nondisadv_stats <- pred_nondisadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_NFSM6CLA1A, predicted_ATT8SCR_NFSM6CLA1A_imputed, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 3), " (n = ", comma(n), ")"))

# 3. Build the ggplot object
p_nondisadv <- ggplot(pred_nondisadv, aes(x = predicted_ATT8SCR_NFSM6CLA1A_imputed,
                                         y = ATT8SCR_NFSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight, text = tooltip_text), size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  # Placed at higher coordinates as Non-Disadv scores are typically higher
  geom_text(data = r2_nondisadv_stats,
            aes(x = 20, y = 88, label = label),
            inherit.aes = FALSE, size = 3, colour = "grey30", hjust = 0) +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                 "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.15,
                                "Brighton and Hove" = 0.9), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Non-Disadvantaged",
       y = "Observed ATT8 — Non-Disadvantaged",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))

# 4. Final Plotly conversion
ggplotly(p_nondisadv, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", x = 0.3, y = -0.15),
    margin = list(t = 60)
  )
```


# Residual Geography

```{r fig-residual-by-region}
#| label: fig-residual-by-region
#| fig-cap: "Distribution of model residuals by Government Office Region"
#| fig-height: 6

if ("gor_name" %in% names(pred_data) && sum(!is.na(pred_data$gor_name)) > 0) {

  region_resid <- pred_data %>%
    filter(!is.na(gor_name)) %>%
    group_by(gor_name) %>%
    summarise(
      mean_resid = mean(residual, na.rm = TRUE),
      median_resid = median(residual, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    arrange(mean_resid)

  ggplot(pred_data %>% filter(!is.na(gor_name)),
         aes(x = reorder(gor_name, residual, FUN = median), y = residual)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_boxplot(fill = "#2e6260", alpha = 0.3, outlier.size = 0.3, outlier.alpha = 0.2) +
    coord_flip() +
    labs(x = NULL, y = "Residual (Observed - Predicted ATT8)",
         subtitle = "Positive = outperforming model; Negative = underperforming") +
    theme_pub
}
```


# Disadvantage Gap

```{r fig-disadvantage-gap}
#| label: fig-disadvantage-gap
#| fig-cap: "Disadvantage attainment gap by academic year"
#| fig-height: 5

if (all(c("ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A") %in% names(model_data))) {
  gap_data <- model_data %>%
    filter(!is.na(ATT8SCR_FSM6CLA1A), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
    mutate(gap = ATT8SCR_NFSM6CLA1A - ATT8SCR_FSM6CLA1A)

  gap_summary <- gap_data %>%
    group_by(year_label) %>%
    summarise(
      mean_gap = mean(gap, na.rm = TRUE),
      median_gap = median(gap, na.rm = TRUE),
      q25 = quantile(gap, 0.25, na.rm = TRUE),
      q75 = quantile(gap, 0.75, na.rm = TRUE),
      .groups = "drop"
    )

  ggplot(gap_data, aes(x = year_label, y = gap)) +
    geom_boxplot(fill = "#4e3c56", alpha = 0.3, outlier.size = 0.3, outlier.alpha = 0.2) +
    geom_point(data = gap_summary, aes(y = mean_gap),
               colour = "#4e3c56", size = 3, shape = 18) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(x = "Academic Year",
         y = "ATT8 Gap (Non-Disadvantaged - Disadvantaged)",
         subtitle = "Diamond = mean gap; Box = IQR") +
    theme_pub
}
```


# Appendix: Model Formulae {.appendix}

## Panel Model (Analysis A)

All four years pooled with `year_label` as a random intercept:

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Per-Year Model (Analysis B)

Separate model for each academic year (no year term):

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Core Panel Model (Analysis C)

Reduced specification using only the 5 predictors available for all 4 years (including 2024-25). Drops workforce and prior-attainment variables:

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  ADMPOL_PT + gorard_segregation +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Core Per-Year Model (Analysis D)

Same 5 core fixed effects, separate model per year (no year term):

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  ADMPOL_PT + gorard_segregation +
  (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Imputed Full Panel Model (Analysis E)

Same 9-predictor specification as Analysis A, but covering all 4 years (2024-25 uses carry-forward imputed workforce and prior-attainment values). This is the model used in the Shiny app and for Observed vs Predicted plots.

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Variable Definitions

```{r tbl-variable-defs}
#| label: tbl-variable-defs
#| tbl-cap: "Variable definitions from DfE metadata"

model_vars <- c(
  "ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A",
  "PTFSM6CLA1A", "PERCTOT", "PNUMEAL", "PTPRIORLO",
  "ADMPOL_PT", "gorard_segregation",
  "remained_in_the_same_school",
  "teachers_on_leadership_pay_range_percent",
  "average_number_of_days_taken",
  "OFSTEDRATING_1", "gor_name", "LANAME", "year_label"
)

tibble(
  `Variable Code` = model_vars,
  Description = var_labels(model_vars),
  `In Formula As` = case_when(
    model_vars %in% c("PTFSM6CLA1A", "PERCTOT", "PNUMEAL",
                       "remained_in_the_same_school",
                       "average_number_of_days_taken",
                       "ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A") ~
      paste0("log(", model_vars, ")"),
    model_vars %in% c("OFSTEDRATING_1", "gor_name", "LANAME", "year_label") ~
      paste0("(1 | ", model_vars, ")"),
    TRUE ~ model_vars
  ),
  Source = case_when(
    model_vars %in% c("ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A",
                       "PTFSM6CLA1A", "PTPRIORLO", "ADMPOL_PT") ~ "KS4 Performance Tables",
    model_vars %in% c("PERCTOT") ~ "Absence Statistics",
    model_vars %in% c("PNUMEAL") ~ "School Census",
    model_vars %in% c("OFSTEDRATING_1") ~ "Ofsted Inspections (harmonised)",
    model_vars %in% c("gorard_segregation") ~ "Derived (Gorard index)",
    model_vars %in% c("remained_in_the_same_school",
                       "teachers_on_leadership_pay_range_percent",
                       "average_number_of_days_taken") ~ "School Workforce Census",
    model_vars %in% c("gor_name", "LANAME") ~ "Edubase",
    model_vars == "year_label" ~ "Constructed",
    TRUE ~ ""
  )
) %>%
  kable(align = "llll") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 10)
```
