---
title: "Model Experiments"
subtitle: "Human-readable lmer implementations for manual inspection"
author: "Adam Dennett with some assistance from Claude Code"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    theme: ucl-theme.scss
    df-print: kable
    embed-resources: true
    include-in-header: header-logos.html
    include-after-body: footer-logos.html
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup

library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(plotly)
library(here)

# Load the raw panel data (same starting point as 05_fit_model.R)
panel <- readRDS(here::here("data", "panel_data.rds"))

# Backfill missing gor_name from DfE LA/region lookup (belt-and-braces)
la_region_lookup <- read.csv(
  here::here("data", "meta", "Performancetables_130249",
             "2022-2023", "la_and_region_codes_meta.csv"),
  stringsAsFactors = FALSE
) %>%
  mutate(gor_name_broad = case_when(
    grepl("^North East",   REGION.NAME) ~ "North East",
    grepl("^North West",   REGION.NAME) ~ "North West",
    grepl("^North Yorkshire|^South and West Yorkshire", REGION.NAME) ~ "Yorkshire and the Humber",
    grepl("^East Midlands", REGION.NAME) ~ "East Midlands",
    grepl("^West Midlands", REGION.NAME) ~ "West Midlands",
    grepl("^East of England", REGION.NAME) ~ "East of England",
    grepl("^London",       REGION.NAME) ~ "London",
    grepl("^South East",   REGION.NAME) ~ "South East",
    grepl("^South West",   REGION.NAME) ~ "South West",
    TRUE ~ NA_character_
  ))

n_missing_before <- sum(is.na(panel$gor_name))
panel <- panel %>%
  left_join(la_region_lookup %>% select(LEA, gor_name_broad), by = "LEA") %>%
  mutate(gor_name = coalesce(gor_name, gor_name_broad)) %>%
  select(-gor_name_broad)
n_missing_after <- sum(is.na(panel$gor_name))

cat("Panel loaded:", nrow(panel), "rows,", ncol(panel), "columns\n")
cat("Region backfill: fixed", n_missing_before - n_missing_after, "of",
    n_missing_before, "missing gor_name\n")
```


# Overview

This document re-implements every model from `model_results.qmd` in a
deliberately simple, explicit style. Each model is written out longhand
so you can:

- see exactly which rows survive the filters
- read the `lmer()` formula as a single, self-contained call
- inspect `summary()`, `fixef()`, `ranef()`, `VarCorr()` etc. directly

The four analyses and three outcomes produce **12 models** in total.

Models are fitted with `REML = TRUE` and `optimizer = "bobyqa"` to
match the production pipeline.


# Data Preparation

Three datasets are needed:

1. A **full** dataset for Analysis A (9 predictors, **3 years only** — excluding 2024-25, which has carry-forward imputed workforce variables)
2. A **core** dataset (5 predictors, all 4 years including 2024-25)
3. An **imputed full** dataset for Analysis E (9 predictors, all 4 years — including 2024-25 with imputed workforce variables)

Note: `panel_data.rds` already contains carry-forward imputed workforce variables for 2024-25 (from `04_compute_derived.R`). To create a genuine 3-year dataset for Analysis A, we must explicitly exclude 2024-25.

## Full model dataset (3 years, real data only)

```{r data-full}
#| label: data-full

# All 4 years with complete workforce data (used for Analysis E)
all_model_data <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(
    !is.na(ATT8SCR), ATT8SCR > 0,
    PTFSM6CLA1A > 0,
    PERCTOT > 0,
    PNUMEAL > 0,
    !is.na(OFSTEDRATING_1),
    !is.na(gor_name),
    !is.na(LANAME)
  ) %>%
  filter(
    !is.na(remained_in_the_same_school),
    !is.na(teachers_on_leadership_pay_range_percent),
    average_number_of_days_taken > 0,
    !is.na(gorard_segregation)
  ) %>%
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

contrasts(all_model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(all_model_data$OFSTEDRATING_1))

# Analysis A: 3 years only (exclude 2024-25 which has imputed workforce values)
model_data <- all_model_data %>%
  filter(year_label != "2024-25") %>%
  droplevels()

contrasts(model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(model_data$OFSTEDRATING_1))

cat("Full model dataset (Analysis A, 3 years):", nrow(model_data), "rows\n")
cat("Years:", paste(levels(model_data$year_label), collapse = ", "), "\n")
cat("Regions:", n_distinct(model_data$gor_name), "\n")
cat("LAs:", n_distinct(model_data$LANAME), "\n")
cat("Ofsted levels:", paste(levels(model_data$OFSTEDRATING_1), collapse = ", "), "\n")
```

::: {.callout-note}
## Sample attrition: teacher sickness days

The full model dataset loses approximately 570 school-year observations
due to the `average_number_of_days_taken > 0` filter. This variable is
required to be strictly positive for the log transform used in the model.
The losses are concentrated in 2021-22 (250 rows) and 2022-23 (244 rows),
with far fewer in 2023-24 (54) and 2024-25 (22). Around 420 of these are
NA values — schools where the workforce census returned retention and
leadership data but not sickness absence — and the remaining ~150 are
genuine zeros. Academies are disproportionately affected (80% of losses),
likely reflecting patchier sickness absence reporting in the early
post-COVID workforce census returns. This is a known data limitation;
the dropped schools are spread across all regions and LAs rather than
concentrated in particular areas.
:::

## Core model dataset

```{r data-core}
#| label: data-core

core_model_data <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(
    !is.na(ATT8SCR), ATT8SCR > 0,
    PTFSM6CLA1A > 0,
    PERCTOT > 0,
    PNUMEAL > 0,
    !is.na(ADMPOL_PT),
    !is.na(gorard_segregation),
    !is.na(OFSTEDRATING_1),
    !is.na(gor_name),
    !is.na(LANAME)
  ) %>%
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

contrasts(core_model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(core_model_data$OFSTEDRATING_1))

cat("Core model dataset:", nrow(core_model_data), "rows\n")
cat("Years:", paste(levels(core_model_data$year_label), collapse = ", "), "\n")
```

## Dropped-rows diagnostic

Which school-year observations in the raw panel were excluded from the
**full** model dataset, and why? The filters are applied sequentially
below so we can attribute each dropped row to the *first* filter that
removes it.

```{r dropped-rows}
#| label: dropped-rows
#| code-fold: true
#| code-summary: "Show diagnostic code"

# Start from the same mainstream filter used above
mainstream <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school"))

cat("Starting mainstream rows:", nrow(mainstream), "\n\n")

# Apply filters sequentially and tag each row with the reason it was dropped
d_diag <- mainstream %>%
  mutate(
    # Step 1: Outcome missing or zero
    fail_outcome = is.na(ATT8SCR) | ATT8SCR <= 0,
    # Step 2: FSM missing or zero (can't log-transform)
    fail_fsm = !fail_outcome & (is.na(PTFSM6CLA1A) | PTFSM6CLA1A <= 0),
    # Step 3: Absence missing or zero
    fail_absence = !fail_outcome & !fail_fsm & (is.na(PERCTOT) | PERCTOT <= 0),
    # Step 4: EAL missing or zero
    fail_eal = !fail_outcome & !fail_fsm & !fail_absence & (is.na(PNUMEAL) | PNUMEAL <= 0),
    # Step 5: Ofsted rating missing
    fail_ofsted = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal & is.na(OFSTEDRATING_1),
    # Step 6: Region or LA missing
    fail_geo = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal & !fail_ofsted &
               (is.na(gor_name) | is.na(LANAME)),
    # Step 7: Teacher retention missing (workforce — drops 2024-25)
    fail_retention = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal &
                     !fail_ofsted & !fail_geo & is.na(remained_in_the_same_school),
    # Step 8: Leadership pay missing
    fail_leadership = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal &
                      !fail_ofsted & !fail_geo & !fail_retention &
                      is.na(teachers_on_leadership_pay_range_percent),
    # Step 9: Teacher sickness days missing or zero
    fail_sickness = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal &
                    !fail_ofsted & !fail_geo & !fail_retention & !fail_leadership &
                    (is.na(average_number_of_days_taken) | average_number_of_days_taken <= 0),
    # Step 10: Gorard segregation index missing
    fail_gorard = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal &
                  !fail_ofsted & !fail_geo & !fail_retention & !fail_leadership &
                  !fail_sickness & is.na(gorard_segregation),
    # Assign first reason
    drop_reason = case_when(
      fail_outcome    ~ "ATT8SCR missing/zero",
      fail_fsm        ~ "PTFSM6CLA1A missing/zero",
      fail_absence    ~ "PERCTOT missing/zero",
      fail_eal        ~ "PNUMEAL missing/zero",
      fail_ofsted     ~ "OFSTEDRATING missing",
      fail_geo        ~ "Region/LA missing",
      fail_retention  ~ "Teacher retention missing",
      fail_leadership ~ "Leadership pay missing",
      fail_sickness   ~ "Sickness days missing/zero",
      fail_gorard     ~ "Gorard index missing",
      TRUE            ~ "Retained"
    )
  )

# Summary table of drop reasons
reason_summary <- d_diag %>%
  count(drop_reason) %>%
  arrange(desc(n)) %>%
  mutate(pct = round(100 * n / sum(n), 1))

knitr::kable(reason_summary,
             col.names = c("Reason", "Rows", "% of mainstream"),
             caption = "Row attrition: first filter that removes each school-year observation")
```

```{r dropped-detail}
#| label: dropped-detail
#| code-fold: true
#| code-summary: "Show dropped-rows detail code"

dropped <- d_diag %>%
  filter(drop_reason != "Retained")

cat(nrow(dropped), "rows dropped from", n_distinct(dropped$SCHNAME),
    "unique schools across", n_distinct(dropped$LANAME), "LAs\n\n")

# Which years are most affected?
cat("Dropped rows by year:\n")
dropped %>% count(year_label, drop_reason) %>%
  pivot_wider(names_from = year_label, values_from = n, values_fill = 0) %>%
  arrange(drop_reason) %>%
  as.data.frame() %>%
  print()
cat("\n")

# Top LAs losing the most rows
la_drops <- dropped %>%
  count(LANAME, drop_reason) %>%
  group_by(LANAME) %>%
  summarise(
    total_dropped = sum(n),
    reasons = paste(sprintf("%s (%d)", drop_reason, n), collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(desc(total_dropped))

cat("Top 20 LAs by rows dropped:\n")
knitr::kable(head(la_drops, 20),
             col.names = c("LA", "Rows dropped", "Reasons (count)"),
             caption = "Local authorities losing the most observations")
```

```{r dropped-schools}
#| label: dropped-schools
#| code-fold: true
#| code-summary: "Show dropped-schools detail code"

# Schools that are *entirely* absent from the final model dataset
retained_schools <- d_diag %>% filter(drop_reason == "Retained") %>%
  distinct(SCHNAME) %>% pull(SCHNAME)

fully_dropped <- dropped %>%
  filter(!SCHNAME %in% retained_schools) %>%
  group_by(SCHNAME, LANAME) %>%
  summarise(
    years_present = paste(sort(unique(year_label)), collapse = ", "),
    n_rows = n(),
    reasons = paste(unique(drop_reason), collapse = "; "),
    mean_att8 = if (any(!is.na(ATT8SCR))) round(mean(ATT8SCR, na.rm = TRUE), 1) else NA,
    has_workforce = any(!is.na(remained_in_the_same_school)),
    has_gorard = any(!is.na(gorard_segregation)),
    ofsted = first(na.omit(OFSTEDRATING_1)),
    .groups = "drop"
  ) %>%
  arrange(LANAME, SCHNAME)

cat(nrow(fully_dropped), "schools are entirely excluded from the full model dataset:\n\n")

knitr::kable(fully_dropped,
             col.names = c("School", "LA", "Years in panel", "Rows", "Drop reasons",
                           "Mean ATT8", "Has workforce", "Has Gorard", "Ofsted"),
             caption = "Schools with zero rows in the final model dataset — entirely excluded from modelling")
```

```{r dropped-characteristics}
#| label: dropped-characteristics
#| code-fold: true
#| code-summary: "Show characteristics comparison"
#| fig-height: 5
#| fig-width: 9
#| fig-cap: "Comparison of key characteristics between retained and dropped observations. Dropped rows tend to have more missing data and different school profiles."

# Compare characteristics of retained vs dropped
compare <- d_diag %>%
  mutate(status = if_else(drop_reason == "Retained", "Retained", "Dropped")) %>%
  select(status, ATT8SCR, PTFSM6CLA1A, PERCTOT, PNUMEAL, TPUP,
         MINORGROUP, year_label) %>%
  pivot_longer(c(ATT8SCR, PTFSM6CLA1A, PERCTOT, PNUMEAL, TPUP),
               names_to = "variable", values_to = "value") %>%
  filter(!is.na(value))

var_labels <- c(
  ATT8SCR = "Attainment 8",
  PTFSM6CLA1A = "% FSM Eligible",
  PERCTOT = "% Overall Absence",
  PNUMEAL = "% EAL",
  TPUP = "Total Pupils"
)

compare <- compare %>%
  mutate(variable = factor(variable, levels = names(var_labels), labels = var_labels))

ggplot(compare, aes(x = status, y = value, fill = status)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 5) +
  scale_fill_manual(values = c("Retained" = "#2e6260", "Dropped" = "#7b132b")) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Retained vs Dropped observations: key characteristics") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```


# Analysis A: Full Panel Models

All years pooled, `year_label` as a random intercept. Nine fixed-effect
predictors (5 log-transformed, 4 linear).

## All Pupils

```{r a-all}
#| label: a-all

d <- model_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_all)
```

### Quick checks

```{r a-all-checks}
#| label: a-all-checks

cat("Fixed effects:\n")
fixef(mod_a_all)

cat("\nVariance components:\n")
print(VarCorr(mod_a_all))

cat("\nR-squared:\n")
print(r2(mod_a_all))

cat("\nSingular fit?", isSingular(mod_a_all), "\n")
```

## Disadvantaged Pupils

```{r a-disadv}
#| label: a-disadv

d <- model_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_disadv)
```

### Quick checks

```{r a-disadv-checks}
#| label: a-disadv-checks

fixef(mod_a_disadv)
print(VarCorr(mod_a_disadv))
print(r2(mod_a_disadv))
cat("Singular fit?", isSingular(mod_a_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r a-nondisadv}
#| label: a-nondisadv

d <- model_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_nondisadv)
```

### Quick checks

```{r a-nondisadv-checks}
#| label: a-nondisadv-checks

fixef(mod_a_nondisadv)
print(VarCorr(mod_a_nondisadv))
print(r2(mod_a_nondisadv))
cat("Singular fit?", isSingular(mod_a_nondisadv), "\n")
```


# Analysis B: Full Per-Year Models

Same 9 fixed effects as Analysis A, but fitted separately for each year.
No `year_label` random effect.

```{r b-setup}
#| label: b-setup

year_levels <- sort(unique(as.character(model_data$year_label)))
cat("Years to fit:", paste(year_levels, collapse = ", "), "\n")
```

## All Pupils (per year)

```{r b-all}
#| label: b-all

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- All Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr, !is.na(ATT8SCR), ATT8SCR > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      remained_in_the_same_school +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Disadvantaged Pupils (per year)

```{r b-disadv}
#| label: b-disadv

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Disadvantaged Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_FSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      remained_in_the_same_school +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Non-Disadvantaged Pupils (per year)

```{r b-nondisadv}
#| label: b-nondisadv

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Non-Disadvantaged Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_NFSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      remained_in_the_same_school +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```


# Analysis C: Core Panel Models

Reduced specification with 5 fixed effects. All 4 years pooled
(including 2024-25). `year_label` as a random intercept.

## All Pupils

```{r c-all}
#| label: c-all

d <- core_model_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_all)
```

### Quick checks

```{r c-all-checks}
#| label: c-all-checks

fixef(mod_c_all)
print(VarCorr(mod_c_all))
print(r2(mod_c_all))
cat("Singular fit?", isSingular(mod_c_all), "\n")
```

## Disadvantaged Pupils

```{r c-disadv}
#| label: c-disadv

d <- core_model_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_disadv)
```

### Quick checks

```{r c-disadv-checks}
#| label: c-disadv-checks

fixef(mod_c_disadv)
print(VarCorr(mod_c_disadv))
print(r2(mod_c_disadv))
cat("Singular fit?", isSingular(mod_c_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r c-nondisadv}
#| label: c-nondisadv

d <- core_model_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_nondisadv)
```

### Quick checks

```{r c-nondisadv-checks}
#| label: c-nondisadv-checks

fixef(mod_c_nondisadv)
print(VarCorr(mod_c_nondisadv))
print(r2(mod_c_nondisadv))
cat("Singular fit?", isSingular(mod_c_nondisadv), "\n")
```


# Analysis D: Core Per-Year Models

Same 5 core fixed effects, fitted separately for each year. No
`year_label` random effect.

```{r d-setup}
#| label: d-setup

core_year_levels <- sort(unique(as.character(core_model_data$year_label)))
cat("Years to fit:", paste(core_year_levels, collapse = ", "), "\n")
```

## All Pupils (per year)

```{r d-all}
#| label: d-all

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- All Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr, !is.na(ATT8SCR), ATT8SCR > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Disadvantaged Pupils (per year)

```{r d-disadv}
#| label: d-disadv

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Disadvantaged Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_FSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Non-Disadvantaged Pupils (per year)

```{r d-nondisadv}
#| label: d-nondisadv

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Non-Disadvantaged Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_NFSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```


# Analysis E: Full Panel with Imputed 2024-25 Predictors

The full 9-predictor specification can now cover all 4 years because
`04_compute_derived.R` carry-forward imputes PTPRIORLO and three
workforce variables for 2024-25 (using each school's 2023-24 value, or
its 3-year mean as a fallback). A flag column `has_imputed_predictors`
marks which rows use estimates.

```{r e-setup}
#| label: e-setup

# Analysis E uses all 4 years including 2024-25 (with carry-forward imputed
# workforce variables). This is simply `all_model_data` — no bind_rows needed
# because panel_data.rds already contains the imputed values.
imputed_full_data <- all_model_data

n_2425 <- sum(imputed_full_data$year_label == "2024-25")

cat("Imputed full dataset:", nrow(imputed_full_data), "rows\n")
cat("Years:", paste(levels(imputed_full_data$year_label), collapse = ", "), "\n")
cat("Of which 2024-25 (imputed predictors):", n_2425, "rows\n")
cat("Analysis A had:", nrow(model_data), "rows (3 years)\n")
cat("Difference (new 2024-25 rows):", nrow(imputed_full_data) - nrow(model_data), "\n")
```

## All Pupils

```{r e-all}
#| label: e-all

d <- imputed_full_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_e_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_e_all)
```

### Quick checks

```{r e-all-checks}
#| label: e-all-checks

fixef(mod_e_all)
print(VarCorr(mod_e_all))
print(r2(mod_e_all))
cat("Singular fit?", isSingular(mod_e_all), "\n")
```

## Disadvantaged Pupils

```{r e-disadv}
#| label: e-disadv

d <- imputed_full_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_e_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_e_disadv)
```

### Quick checks

```{r e-disadv-checks}
#| label: e-disadv-checks

fixef(mod_e_disadv)
print(VarCorr(mod_e_disadv))
print(r2(mod_e_disadv))
cat("Singular fit?", isSingular(mod_e_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r e-nondisadv}
#| label: e-nondisadv

d <- imputed_full_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_e_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_e_nondisadv)
```

### Quick checks

```{r e-nondisadv-checks}
#| label: e-nondisadv-checks

fixef(mod_e_nondisadv)
print(VarCorr(mod_e_nondisadv))
print(r2(mod_e_nondisadv))
cat("Singular fit?", isSingular(mod_e_nondisadv), "\n")
```

## Analysis A vs E: coefficient comparison

```{r e-vs-a}
#| label: e-vs-a

# Helper: relabel raw lmer coefficient names to human-readable form
relabel_terms <- function(term) {
  case_when(
    term == "(Intercept)" ~ "(Intercept)",
    term == "log(PTFSM6CLA1A)" ~ "log(% FSM)",
    term == "log(PERCTOT)" ~ "log(% Absence)",
    term == "log(PNUMEAL)" ~ "log(% EAL)",
    term == "PTPRIORLO" ~ "% Low Prior Attainment",
    term == "gorard_segregation" ~ "Gorard Segregation",
    term == "ADMPOL_PTOTHER NON SEL" ~ "Admissions: Other non-selective (ref: non-sel. in highly sel. area)",
    term == "ADMPOL_PTSEL" ~ "Admissions: Selective (ref: non-sel. in highly sel. area)",
    term == "remained_in_the_same_school" ~ "Teacher Retention",
    term == "teachers_on_leadership_pay_range_percent" ~ "Leadership Pay %",
    term == "log(average_number_of_days_taken)" ~ "log(Teacher Sickness)",
    TRUE ~ term
  )
}

a_vs_e <- bind_rows(
  tibble(Analysis = "A (Real, 3 yrs)",
         Term = names(fixef(mod_a_all)),
         `All Pupils` = fixef(mod_a_all),
         Disadvantaged = fixef(mod_a_disadv),
         `Non-Disadvantaged` = fixef(mod_a_nondisadv)),
  tibble(Analysis = "E (Imputed, 4 yrs)",
         Term = names(fixef(mod_e_all)),
         `All Pupils` = fixef(mod_e_all),
         Disadvantaged = fixef(mod_e_disadv),
         `Non-Disadvantaged` = fixef(mod_e_nondisadv))
) %>%
  mutate(Term = relabel_terms(Term),
         across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(a_vs_e, align = "llrrr",
             caption = "Analysis A (real data, 3 years) vs Analysis E (with imputed 2024-25)")
```

The table above shows that adding one year of carry-forward imputed data
(Analysis E) produces coefficients that are **very close to Analysis A**.
This is reassuring: three-quarters of the data is identical, and the
imputed fourth year uses conservative carry-forward values. Crucially,
the **FSM sign flip for disadvantaged pupils** — positive in the full
model — should be preserved in Analysis E, confirming it is a genuine
structural feature of the data when prior attainment is controlled for.


# Coefficient Commentary

This section examines the coefficients from the **imputed full model**
(Analysis E: 9 predictors, all 4 years including 2024-25 with
carry-forward imputed PTPRIORLO and workforce variables). This is the
model now used in the Shiny app. For reference, the core panel model
(Analysis C: 5 predictors) and the original 3-year full model
(Analysis A) are compared below.


## Imputed full model: side-by-side coefficients

The imputed full model (Analysis E) includes all 9 fixed effects and
covers all four academic years. For 2024-25, `PTPRIORLO` and three
workforce variables are carry-forward imputed from each school's
2023-24 value (or 3-year mean as fallback). Rows using imputed values
are flagged with `has_imputed_predictors = TRUE`.

```{r commentary-imputed-coefs}
#| label: commentary-imputed-coefs

# Extract fixed effects from the three imputed full panel models (Analysis E)
imputed_coef_table <- bind_rows(
  tibble(Group = "All Pupils",
         Term = names(fixef(mod_e_all)),
         Estimate = fixef(mod_e_all)),
  tibble(Group = "Disadvantaged",
         Term = names(fixef(mod_e_disadv)),
         Estimate = fixef(mod_e_disadv)),
  tibble(Group = "Non-Disadvantaged",
         Term = names(fixef(mod_e_nondisadv)),
         Estimate = fixef(mod_e_nondisadv))
) %>%
  pivot_wider(names_from = Group, values_from = Estimate) %>%
  mutate(Term = relabel_terms(Term),
         across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(imputed_coef_table, align = "lrrr",
             caption = "Imputed full model (Analysis E): fixed effect comparison (9 predictors, all 4 years)")
```

## Core panel model: side-by-side coefficients

The core panel model (Analysis C) drops the four variables not available
for 2024-25 — `PTPRIORLO`, `remained_in_the_same_school`,
`teachers_on_leadership_pay_range_percent`, and
`average_number_of_days_taken` — leaving 5 fixed effects.

```{r commentary-core-coefs}
#| label: commentary-core-coefs

# Extract fixed effects from the three core panel models (Analysis C)
core_coef_table <- bind_rows(
  tibble(Group = "All Pupils",
         Term = names(fixef(mod_c_all)),
         Estimate = fixef(mod_c_all)),
  tibble(Group = "Disadvantaged",
         Term = names(fixef(mod_c_disadv)),
         Estimate = fixef(mod_c_disadv)),
  tibble(Group = "Non-Disadvantaged",
         Term = names(fixef(mod_c_nondisadv)),
         Estimate = fixef(mod_c_nondisadv))
) %>%
  pivot_wider(names_from = Group, values_from = Estimate) %>%
  mutate(Term = relabel_terms(Term),
         across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(core_coef_table, align = "lrrr",
             caption = "Core panel model (Analysis C): fixed effect comparison (5 predictors, all 4 years)")
```

## Imputed full vs core: direct comparison

The table below places the 5 shared predictors side by side for both
specifications, making it easy to see how dropping the workforce and
prior-attainment variables shifts the remaining coefficients.

```{r commentary-imputed-vs-core}
#| label: commentary-imputed-vs-core

# Shared terms between imputed full and core (excluding intercept)
shared_terms <- intersect(
  core_coef_table$Term[core_coef_table$Term != "(Intercept)"],
  imputed_coef_table$Term[imputed_coef_table$Term != "(Intercept)"]
)

comparison <- imputed_coef_table %>%
  filter(Term %in% shared_terms) %>%
  rename(`Imputed Full: All` = `All Pupils`,
         `Imputed Full: Disadv.` = Disadvantaged,
         `Imputed Full: Non-Dis.` = `Non-Disadvantaged`) %>%
  left_join(
    core_coef_table %>%
      filter(Term %in% shared_terms) %>%
      rename(`Core: All` = `All Pupils`,
             `Core: Disadv.` = Disadvantaged,
             `Core: Non-Dis.` = `Non-Disadvantaged`),
    by = "Term"
  )

knitr::kable(comparison, align = "lrrrrrr",
             caption = "Shared predictor coefficients: imputed full (E) vs core panel (C)")
```

## Coefficient stability over time (full model)

The per-year full models (Analysis B) show how each of the 9 predictor
coefficients evolves across the three years with real data. The static
ggplot is followed by an interactive plotly version for easier
exploration.

```{r commentary-yearly}
#| label: commentary-yearly
#| fig-height: 10
#| fig-width: 9
#| fig-cap: "Full per-year coefficient estimates (Analysis B), by outcome group"

# Re-fit full per-year models and collect coefficients in a tidy table
yearly_coefs <- list()

for (yr in year_levels) {
  for (outcome_name in c("all", "disadv", "nondisadv")) {

    outcome_var <- switch(outcome_name,
                          all = "ATT8SCR",
                          disadv = "ATT8SCR_FSM6CLA1A",
                          nondisadv = "ATT8SCR_NFSM6CLA1A")
    group_label <- switch(outcome_name,
                          all = "All Pupils",
                          disadv = "Disadvantaged",
                          nondisadv = "Non-Disadvantaged")

    d <- model_data %>%
      filter(year_label == yr,
             !is.na(!!sym(outcome_var)), !!sym(outcome_var) > 0) %>%
      droplevels()

    contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

    mod <- tryCatch(
      lmer(
        as.formula(paste0(
          "log(", outcome_var, ") ~ ",
          "log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) + ",
          "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
          "remained_in_the_same_school + ",
          "teachers_on_leadership_pay_range_percent + ",
          "log(average_number_of_days_taken) + ",
          "(1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
        )),
        data = d, REML = TRUE, na.action = na.exclude,
        control = lmerControl(optimizer = "bobyqa",
                              optCtrl = list(maxfun = 20000))
      ),
      error = function(e) NULL
    )

    if (!is.null(mod)) {
      fe <- fixef(mod)
      yearly_coefs[[length(yearly_coefs) + 1]] <- tibble(
        Year = yr,
        Group = group_label,
        Term = names(fe),
        Estimate = fe
      )
    }
  }
}

yearly_coefs_df <- bind_rows(yearly_coefs) %>%
  filter(Term != "(Intercept)") %>%
  mutate(
    Term = relabel_terms(Term),
    Group = factor(Group,
                   levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged"))
  )

ggplot(yearly_coefs_df,
       aes(x = Year, y = Estimate, colour = Group, group = Group)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ Term, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("All Pupils" = "#2e6260",
                                  "Disadvantaged" = "#4e3c56",
                                  "Non-Disadvantaged" = "#abc766")) +
  labs(y = "Coefficient estimate", x = "Academic year", colour = NULL,
       title = "Full per-year coefficient trajectories (Analysis B)") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold", size = 9)
  )
```

### Interactive coefficient explorer

```{r commentary-yearly-plotly}
#| label: commentary-yearly-plotly
#| fig-height: 10
#| fig-width: 9

# Terms are already relabelled — build tooltips directly
yearly_coefs_plotly <- yearly_coefs_df %>%
  mutate(
    tooltip_text = paste0(
      "<b>", Term, "</b><br>",
      "Year: ", Year, "<br>",
      "Group: ", Group, "<br>",
      "Estimate: ", round(Estimate, 5)
    )
  )

p_yearly <- ggplot(yearly_coefs_plotly,
       aes(x = Year, y = Estimate, colour = Group, group = Group,
           text = tooltip_text)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ Term, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("All Pupils" = "#2e6260",
                                  "Disadvantaged" = "#4e3c56",
                                  "Non-Disadvantaged" = "#abc766")) +
  labs(y = "Coefficient estimate", x = "Academic year", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    strip.text = element_text(face = "bold", size = 8)
  )

ggplotly(p_yearly, tooltip = "text") %>%
  layout(
    title = list(text = "Full per-year coefficients (hover for details)"),
    legend = list(orientation = "h", y = -0.15)
  )
```

## General observations

The commentary below draws on the imputed full model (Analysis E)
coefficient table and the per-year trajectories above. Because the
outcome is **log-transformed**, all coefficients are **elasticities**
(for log-transformed predictors) or **semi-elasticities** (for linear
predictors). A coefficient of --0.10 on a logged predictor means that
a 1% increase in that predictor is associated with roughly a 0.10%
*decrease* in Attainment 8.

### 1. Imputed full vs core model: how dropping variables changes the picture

The most important thing to note when comparing the two specifications
is that they do **not tell the same story** for every coefficient. The
imputed full model (Analysis E) includes `PTPRIORLO` (% low KS2 prior
attainment) and three workforce variables that the core model lacks.
These omitted variables are **not random noise** — prior attainment in
particular is one of the strongest individual predictors of school-level
outcomes. When it is dropped from the core model, its explanatory power
has to go somewhere, and the remaining 5 predictors absorb it. This
produces some notable shifts:

- **`log(PTFSM6CLA1A)` (% FSM):** In the **imputed full model**, this
  coefficient is **negative for All Pupils and Non-Disadvantaged**, and
  **flips to positive for Disadvantaged Pupils**. In the **core model**,
  FSM is negative across **all three groups** — the sign flip
  disappears. This is because prior attainment (`PTPRIORLO`) is strongly
  correlated with FSM: schools with more disadvantaged intakes also tend
  to have more pupils with low prior attainment. In the full model,
  `PTPRIORLO` picks up the prior-attainment component of the
  disadvantage penalty, freeing up `log(PTFSM6CLA1A)` to capture the
  *residual* relationship — which, for disadvantaged pupils specifically,
  is positive (the "critical mass" effect described below). When
  `PTPRIORLO` is absent, `log(PTFSM6CLA1A)` has to absorb both effects
  and the net result is negative.

- **`log(PERCTOT)` (% absence):** Becomes **more negative** in the core
  model. Again, with fewer controls absorbing variation, absence picks up
  some of the explanatory burden previously shared with workforce and
  prior-attainment variables.

- **`log(PNUMEAL)` (% EAL):** The sign and direction are broadly
  consistent across both specifications, but the magnitude changes.

The practical implication is that coefficient interpretation must always
be read **relative to the other variables in the model**. Now that the
Shiny app uses the imputed full model, users get the benefit of all 9
predictors — including the workforce and prior-attainment controls —
across all four years. The 2024-25 values for the four imputed variables
use conservative carry-forward estimates, and this is flagged in the data.

### 2. Free school meals (FSM) — the sign flip

In the **imputed full model** (Analysis E), `log(PTFSM6CLA1A)` shows
the most striking difference across outcome groups:

- **All Pupils and Non-Disadvantaged:** The coefficient is consistently
  **negative**. Schools with a higher share of disadvantaged pupils tend
  to have lower overall and non-disadvantaged ATT8 scores, even after
  controlling for absence, EAL, prior attainment, and segregation. This
  captures the well documented compositional effect: a more disadvantaged
  intake is associated with lower average attainment at school level.

- **Disadvantaged Pupils:** The coefficient **flips to positive**. After
  controlling for prior attainment and workforce factors, schools with a
  higher proportion of FSM-eligible pupils see their *disadvantaged
  subgroup specifically* score somewhat higher than predicted. One
  plausible explanation is that schools serving predominantly
  disadvantaged cohorts concentrate more resources, Pupil Premium
  funding, and pastoral support on those pupils — a kind of "critical
  mass" effect. Alternatively, in schools where almost all pupils are
  disadvantaged, the disadvantaged group essentially *is* the school
  average, removing the within-school composition penalty.

**Crucially, this sign flip only appears in the full model** where prior
attainment controls for the low-KS2 component of the disadvantage
penalty. In the core model (Analysis C), `log(PTFSM6CLA1A)` is
**negative for all three groups** because it is forced to absorb the
prior-attainment effect as well. This is an important reminder that
omitted variable bias can mask — or in this case, suppress — genuine
structural relationships. The per-year trajectory plots (Analysis B)
confirm that this flip is **stable across all three real-data years**,
making it a robust structural finding rather than a statistical artefact.

### 3. Absence — the largest and most stable predictor

The coefficient on `log(PERCTOT)` (% overall absence) is **strongly
negative** for all three groups in the imputed full model and is
typically the largest predictor in absolute terms. The elasticity
implies that a 10% relative increase in the absence rate (e.g. from
5.0% to 5.5%) is associated with a substantial drop in ATT8 on the
log scale.

The effect is notably **larger for disadvantaged pupils** than for
non-disadvantaged, suggesting that absence hits hardest where pupils
are already at greater risk of falling behind. The per-year plots
confirm this pattern is consistent across all years.

### 4. English as an additional language (EAL)

`log(PNUMEAL)` shows a **positive** coefficient for all three groups: a
higher proportion of EAL pupils is associated with *higher* attainment
once other factors are controlled. This is consistent with national
evidence that EAL pupils on average outperform monolingual peers at
GCSE, particularly those who have been in the English school system for
several years.

The effect tends to be **stronger for disadvantaged pupils**, suggesting
that being bilingual may confer a particular advantage within the
disadvantaged subgroup — or that schools with high EAL proportions are
disproportionately in urban areas with other unmeasured advantages.

### 5. Prior attainment (PTPRIORLO)

The `PTPRIORLO` coefficient is **strongly negative** across all three
groups — a higher percentage of pupils with low KS2 prior attainment
is associated with lower Attainment 8 scores. This is the most
intuitive relationship in the model: schools receiving pupils with
weaker primary-level preparation see lower secondary outcomes, even
after controlling for all other factors.

The magnitude is broadly similar across outcome groups, though the
coefficient for non-disadvantaged pupils tends to be slightly smaller,
consistent with higher-attaining pupils being somewhat more resilient
to a low-prior-attainment school environment.

### 6. Interpreting log-transformed variables — effects at the extremes

Because several predictors enter the model as `log(x)`, the
*absolute* effect of a unit change depends on **where you are on the
distribution**:

- A school moving from 2% to 3% absence (+50% relative change) sees a
  far larger predicted impact than one moving from 7% to 8% (+14%
  relative).
- Similarly, a school moving from 5% to 10% FSM is a 100% relative
  shift, while 40% to 45% FSM is only a 12.5% relative shift.

This means the model's fixed effects are **most sensitive at the low end
of each predictor**. Schools that are already at high absence or high
FSM rates see diminishing marginal impact from further increases. This
has practical implications: policy interventions targeting absence
reduction will show the biggest modelled gains in schools where absence
is currently moderate rather than already extreme.

### 7. Gorard segregation index

The Gorard segregation index coefficient is **negative** for all three
groups — higher within-LA segregation is associated with lower
attainment — but the effect is **statistically insignificant** in the
imputed full model (Analysis E). The coefficient is small in absolute
terms and its standard error comfortably spans zero, meaning we cannot
confidently distinguish the estimated effect from no effect at all.
This is an important caveat: while the *direction* of the association
(more segregation → lower attainment) is consistent with broader
education research, **this model does not provide reliable evidence
that the Gorard segregation index has a meaningful independent effect
on Attainment 8 at school level**, once the other 8 predictors are
controlled for. The variable is retained in the model for completeness,
but policy recommendations should not be based on its coefficient.

### 8. Admissions policy

`ADMPOL_PT` (admissions policy type) is a **categorical variable** with three levels: *"Non-selective in a highly selective area"* (i.e. a non-selective school in an area with grammar schools), *"Other non-selective"*, and *"Selective"* (grammar schools). Because it is categorical, `lmer()` automatically creates **two dummy (indicator) variables** using treatment coding, with *"Non-selective in a highly selective area"* as the **reference category** — its effect is absorbed into the intercept, and the two dummy coefficients represent the difference in log(ATT8) relative to that baseline.

The **selective** coefficient (`ADMPOL_PTSEL`) is **strongly positive** across all pupil groups — grammar and other selective schools show higher attainment after controlling for intake characteristics. The **other non-selective** coefficient (`ADMPOL_PTOTHER NON SEL`) shows a small and often non-significant difference from the reference group.

### 9. Workforce variables

In the imputed full model, the three workforce variables add useful
context:

- **Teacher retention** (`remained_in_the_same_school`): Positive
  coefficient — higher staff stability is associated with higher
  attainment. The per-year plots show this is consistent across years.
- **Leadership pay** (`teachers_on_leadership_pay_range_percent`):
  Typically small and sometimes non-significant, suggesting that the
  proportion of teachers on leadership pay scales is not a strong
  predictor once other factors are controlled.
- **Teacher absence** (`log(average_number_of_days_taken)`): Negative
  coefficient — schools where teachers are absent more tend to have
  lower ATT8 scores. This echoes the pupil absence finding and
  reinforces that attendance (by both pupils and staff) is strongly
  associated with outcomes.

### 10. Year-to-year stability

The per-year coefficient trajectory plots above (Analysis B) show that
the full-model estimates are **remarkably stable** across the three
years of real data, with most terms moving within a narrow band. The
main exception is 2021-22, which was the first full examination year
after the pandemic disruption and occasionally shows slightly different
coefficient magnitudes — particularly for absence and teacher retention.

### 11. What the imputed model gains over the core model

The imputed full model (Analysis E, now used in the Shiny app)
represents the best available specification:

- **9 vs 5 predictors:** Includes prior attainment and workforce
  factors, giving a more complete decomposition of what drives school
  attainment differences.
- **Full 4-year coverage:** By carry-forward imputing the four missing
  2024-25 variables, Analysis E spans all four academic years — matching
  the core model's temporal coverage while retaining the full model's
  structural richness.
- **Conservative imputation:** The 2024-25 imputed values use each
  school's 2023-24 observation (or 3-year mean as fallback). Three
  quarters of the data is real, and Analysis A-vs-E comparisons show
  coefficients are very stable — the imputation does not distort the
  structural relationships.
- **Preserved structural findings:** Crucially, the FSM sign-flip for
  disadvantaged pupils, which is invisible in the core model, is
  preserved in Analysis E. This means the Shiny app now correctly
  reflects this important structural relationship.


# Brighton & Hove: A Local Case Study

The Brighton & Hove case study — including interactive maps, detailed interpretation of non-linear effects, school-level residual analysis, and policy recommendations — has been moved to its own standalone report: **[Brighton & Hove Case Study](https://adamdennett.github.io/school_attainment_tool/brighton_case_study.html)**.


# Variable Importance & Policy Recommendations

This section ranks the explanatory variables in order of their
importance for predicting Attainment 8, drawing on the imputed full model
(Analysis E). Two complementary importance measures are presented:

1. **Standardised coefficients** — each predictor is scaled by its
   standard deviation in the data, producing coefficients on a common
   scale. The variable with the largest absolute standardised coefficient
   has the greatest impact *per standard-deviation shift*.
2. **Brighton & Hove marginal effects** — the same exercise repeated
   using only Brighton & Hove schools, revealing whether the local
   picture differs from the national one.

Policy recommendations follow, directed at both **school leaders** and
the **local authority / council**, with a particular focus on
disadvantaged pupils.


## National variable importance (standardised coefficients)

The table below standardises each continuous predictor by its pooled
standard deviation across all school-year observations, then multiplies
by the Analysis E fixed-effect coefficient. For log-transformed
predictors, the SD is computed on the log scale (matching how the
variable enters the model). Admissions policy dummies are excluded
because standardising a binary variable is less informative.

```{r importance-national}
#| label: importance-national

# ---- Helper: compute standardised coefficients for one model ----
standardised_importance <- function(model, data, outcome_var, group_label) {

  fe <- fixef(model)
  fe <- fe[names(fe) != "(Intercept)"]

  # Map coefficient names to the actual data columns and transformations
  coef_info <- tribble(
    ~term,                                    ~data_col,                                  ~transform,
    "log(PTFSM6CLA1A)",                       "PTFSM6CLA1A",                              "log",
    "log(PERCTOT)",                            "PERCTOT",                                  "log",
    "log(PNUMEAL)",                            "PNUMEAL",                                  "log",
    "PTPRIORLO",                               "PTPRIORLO",                                "none",
    "gorard_segregation",                      "gorard_segregation",                       "none",
    "remained_in_the_same_school",         "remained_in_the_same_school",              "none",
    "teachers_on_leadership_pay_range_percent", "teachers_on_leadership_pay_range_percent", "none",
    "log(average_number_of_days_taken)",        "average_number_of_days_taken",             "log"
  )

  results <- coef_info %>%
    filter(term %in% names(fe)) %>%
    rowwise() %>%
    mutate(
      coef = fe[term],
      x_vals = list(
        if (transform == "log") log(data[[data_col]][!is.na(data[[data_col]]) & data[[data_col]] > 0])
        else data[[data_col]][!is.na(data[[data_col]])]
      ),
      sd_x = sd(unlist(x_vals), na.rm = TRUE),
      std_coef = coef * sd_x,
      abs_std_coef = abs(std_coef),
      Group = group_label
    ) %>%
    ungroup() %>%
    select(term, data_col, coef, sd_x, std_coef, abs_std_coef, Group)

  results
}

# Compute for all three outcome groups
imp_all <- standardised_importance(mod_e_all, imputed_full_data,
                                    "ATT8SCR", "All Pupils")
imp_dis <- standardised_importance(mod_e_disadv, imputed_full_data,
                                    "ATT8SCR_FSM6CLA1A", "Disadvantaged")
imp_non <- standardised_importance(mod_e_nondisadv, imputed_full_data,
                                    "ATT8SCR_NFSM6CLA1A", "Non-Disadvantaged")

importance_national <- bind_rows(imp_all, imp_dis, imp_non)

# Display name mapping
display_names <- c(
  "log(PTFSM6CLA1A)" = "% Disadvantaged (FSM)",
  "log(PERCTOT)" = "Overall Absence Rate",
  "log(PNUMEAL)" = "% EAL",
  "PTPRIORLO" = "% Low Prior Attainment",
  "gorard_segregation" = "Gorard Segregation Index",
  "remained_in_the_same_school" = "Teacher Retention",
  "teachers_on_leadership_pay_range_percent" = "Leadership Pay %",
  "log(average_number_of_days_taken)" = "Teacher Sickness Days"
)

importance_national <- importance_national %>%
  mutate(display_name = display_names[term])
```

### National ranking table

```{r importance-national-table}
#| label: importance-national-table

# Pivot to wide format for a clean comparison table
national_ranking <- importance_national %>%
  select(display_name, Group, std_coef) %>%
  pivot_wider(names_from = Group, values_from = std_coef) %>%
  mutate(
    `Avg |Std. Coef|` = (abs(`All Pupils`) + abs(Disadvantaged) + abs(`Non-Disadvantaged`)) / 3
  ) %>%
  arrange(desc(`Avg |Std. Coef|`)) %>%
  mutate(Rank = row_number()) %>%
  relocate(Rank) %>%
  mutate(across(where(is.numeric) & !matches("Rank"), \(x) round(x, 4)))

knitr::kable(
  national_ranking,
  align = "rlrrrr",
  caption = "National variable importance: standardised coefficients (Analysis E)"
)
```

### National importance chart

```{r importance-national-chart}
#| label: importance-national-chart
#| fig-height: 6
#| fig-width: 9

importance_chart_data <- importance_national %>%
  mutate(
    display_name = factor(display_name),
    Group = factor(Group, levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged")),
    tooltip_text = paste0(
      "<b>", display_name, "</b><br>",
      "Group: ", Group, "<br>",
      "Raw coefficient: ", round(coef, 5), "<br>",
      "SD of predictor: ", round(sd_x, 3), "<br>",
      "Standardised coef: ", round(std_coef, 4), "<br>",
      "|Std. coef|: ", round(abs_std_coef, 4)
    )
  )

# Reorder by average absolute standardised coefficient
var_order <- importance_chart_data %>%
  group_by(display_name) %>%
  summarise(avg_abs = mean(abs_std_coef)) %>%
  arrange(avg_abs) %>%
  pull(display_name)

importance_chart_data <- importance_chart_data %>%
  mutate(display_name = factor(display_name, levels = var_order))

p_imp <- ggplot(importance_chart_data,
       aes(x = display_name, y = abs_std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#2e6260",
                                "Disadvantaged" = "#4e3c56",
                                "Non-Disadvantaged" = "#abc766")) +
  labs(x = NULL, y = "|Standardised Coefficient|",
       title = "Variable Importance: National (Analysis E)",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_imp, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```

### National importance: direction of effect

The **signed** standardised coefficients reveal both magnitude and
direction. Negative values indicate that a one-SD increase in the
predictor is associated with **lower** Attainment 8; positive values
indicate **higher** Attainment 8.

```{r importance-national-direction}
#| label: importance-national-direction
#| fig-height: 6
#| fig-width: 9

p_dir <- ggplot(importance_chart_data,
       aes(x = display_name, y = std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#2e6260",
                                "Disadvantaged" = "#4e3c56",
                                "Non-Disadvantaged" = "#abc766")) +
  labs(x = NULL, y = "Standardised Coefficient (signed)",
       title = "Direction & Magnitude of Effect: National",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_dir, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```


## Brighton & Hove variable importance

The same standardised-coefficient approach is repeated using only
Brighton & Hove schools. Because the local predictor distributions may
differ from the national picture (e.g. lower absence variance, different
FSM range), the rankings may shift.

```{r importance-bh}
#| label: importance-bh

# Filter to Brighton & Hove
bh_data <- imputed_full_data %>%
  filter(LANAME == "Brighton and Hove")

cat("Brighton & Hove observations:", nrow(bh_data), "\n")

# Use the NATIONAL model coefficients but LOCAL predictor SDs
imp_bh_all <- standardised_importance(mod_e_all, bh_data,
                                       "ATT8SCR", "All Pupils")
imp_bh_dis <- standardised_importance(mod_e_disadv, bh_data,
                                       "ATT8SCR_FSM6CLA1A", "Disadvantaged")
imp_bh_non <- standardised_importance(mod_e_nondisadv, bh_data,
                                       "ATT8SCR_NFSM6CLA1A", "Non-Disadvantaged")

importance_bh <- bind_rows(imp_bh_all, imp_bh_dis, imp_bh_non) %>%
  mutate(display_name = display_names[term])
```

### Brighton & Hove ranking table

```{r importance-bh-table}
#| label: importance-bh-table

bh_ranking <- importance_bh %>%
  select(display_name, Group, std_coef) %>%
  pivot_wider(names_from = Group, values_from = std_coef) %>%
  mutate(
    `Avg |Std. Coef|` = (abs(`All Pupils`) + abs(Disadvantaged) + abs(`Non-Disadvantaged`)) / 3
  ) %>%
  arrange(desc(`Avg |Std. Coef|`)) %>%
  mutate(Rank = row_number()) %>%
  relocate(Rank) %>%
  mutate(across(where(is.numeric) & !matches("Rank"), \(x) round(x, 4)))

knitr::kable(
  bh_ranking,
  align = "rlrrrr",
  caption = "Brighton & Hove variable importance: standardised coefficients (local predictor SDs, national model)"
)
```

### Brighton & Hove importance chart

```{r importance-bh-chart}
#| label: importance-bh-chart
#| fig-height: 6
#| fig-width: 9

bh_chart_data <- importance_bh %>%
  mutate(
    Group = factor(Group, levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged")),
    tooltip_text = paste0(
      "<b>", display_name, "</b> (B&H)<br>",
      "Group: ", Group, "<br>",
      "Raw coefficient (national): ", round(coef, 5), "<br>",
      "SD of predictor (B&H): ", round(sd_x, 3), "<br>",
      "Standardised coef: ", round(std_coef, 4)
    )
  )

bh_var_order <- bh_chart_data %>%
  group_by(display_name) %>%
  summarise(avg_abs = mean(abs_std_coef)) %>%
  arrange(avg_abs) %>%
  pull(display_name)

bh_chart_data <- bh_chart_data %>%
  mutate(display_name = factor(display_name, levels = bh_var_order))

p_bh <- ggplot(bh_chart_data,
       aes(x = display_name, y = abs_std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#2e6260",
                                "Disadvantaged" = "#4e3c56",
                                "Non-Disadvantaged" = "#abc766")) +
  labs(x = NULL, y = "|Standardised Coefficient|",
       title = "Variable Importance: Brighton & Hove",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_bh, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```


## National vs Brighton & Hove comparison

```{r importance-comparison}
#| label: importance-comparison

# Compare rankings side by side (All Pupils model)
comparison_all <- national_ranking %>%
  select(display_name, `National Rank` = Rank,
         `National Std.Coef` = `All Pupils`) %>%
  left_join(
    bh_ranking %>%
      select(display_name, `B&H Rank` = Rank,
             `B&H Std.Coef` = `All Pupils`),
    by = "display_name"
  ) %>%
  mutate(`Rank Change` = `National Rank` - `B&H Rank`)

knitr::kable(
  comparison_all,
  align = "lrrrrl",
  caption = "All Pupils: National vs Brighton & Hove variable importance rankings"
)
```

```{r importance-comparison-disadv}
#| label: importance-comparison-disadv

# Same for Disadvantaged pupils
comparison_disadv <- importance_national %>%
  filter(Group == "Disadvantaged") %>%
  arrange(desc(abs_std_coef)) %>%
  mutate(`National Rank` = row_number()) %>%
  select(display_name, `National Rank`, `National Std.Coef` = std_coef) %>%
  left_join(
    importance_bh %>%
      filter(Group == "Disadvantaged") %>%
      arrange(desc(abs_std_coef)) %>%
      mutate(`B&H Rank` = row_number()) %>%
      select(display_name, `B&H Rank`, `B&H Std.Coef` = std_coef),
    by = "display_name"
  ) %>%
  mutate(`Rank Change` = `National Rank` - `B&H Rank`) %>%
  mutate(across(where(is.numeric) & !matches("Rank"), \(x) round(x, 4)))

knitr::kable(
  comparison_disadv,
  align = "lrrrrl",
  caption = "Disadvantaged Pupils: National vs Brighton & Hove variable importance rankings"
)
```

### Key differences between national and local rankings

The comparison tables above highlight where Brighton & Hove's local
context shifts the relative importance of predictors:

- **% Disadvantaged (FSM):** Brighton & Hove schools span a wide FSM
  range (from under 10% to over 50%), which may amplify or dampen this
  variable's local importance relative to the national picture depending
  on the SD of log(FSM) in the Local Authority.

- **Overall Absence:** If Brighton & Hove has a narrower spread of
  absence rates than the national distribution, absence will appear less
  important locally — not because it matters less per unit of change, but
  because schools are more similar to each other in this dimension.

- **Teacher Retention and Sickness:** Brighton & Hove's workforce
  stability profile — shaped by coastal-city labour market conditions,
  London proximity for recruitment, and academy vs maintained school
  mix — may produce a different local variance in these predictors.

- **Gorard Segregation:** As a single local authority, all B&H schools
  share the same segregation index value in any given year, so the
  within-B&H variance is zero or near-zero. This variable's local
  importance will therefore be very low. Note that even at national
  level, the Gorard segregation coefficient is **statistically
  insignificant** in the imputed full model (see commentary in
  section 7 above), so its apparent ranking in the national table
  should be interpreted with caution.


## Policy recommendations

The variable importance rankings, combined with the direction-of-effect
analysis and the model's structural features, suggest several priority
areas for intervention. These are organised by audience.

### For school leaders

**1. Absence reduction is the single biggest lever.**

Across all three pupil groups, overall absence rate (`PERCTOT`) is
consistently the strongest or among the strongest predictors of
Attainment 8. The log-transformed relationship means that gains from
reducing absence are **largest for schools with already-moderate absence
rates** (the steep part of the curve). For a school currently at 6%
absence, reducing to 5% has a larger predicted effect than a school at
10% reducing to 9%. This argues for:

- Intensive early-intervention strategies (attendance officers,
  family liaison, mentoring) before absence becomes entrenched
- Special attention to protecting low-absence schools from
  deterioration, not just tackling chronic cases
- Focus on disadvantaged pupils specifically, where the absence
  coefficient is largest

**2. Support for pupils with low prior attainment.**

The `PTPRIORLO` coefficient is strongly negative — schools serving a
higher proportion of pupils who entered secondary school with low KS2
scores see lower Attainment 8 outcomes. While schools cannot change their
intake composition, they can invest in:

- Targeted catch-up programmes (literacy, numeracy) for pupils arriving
  with below-expected KS2 levels
- Diagnostic assessment at transition from primary school to identify
  gaps early
- Small-group intervention during KS3 to prevent the attainment gap
  widening before GCSE

**3. Workforce stability matters.**

Teacher retention (`remained_in_the_same_school`) shows a positive
association with attainment: schools where more teachers stay are
predicted to achieve higher ATT8 scores. Meanwhile, teacher sickness
days show a negative association. School leaders should consider:

- Retention strategies: competitive CPD offers, workload management,
  mentoring for early-career teachers
- Wellbeing programmes targeting staff absence
- Succession planning to avoid disruptive turnover in key departments

**4. The FSM sign-flip for disadvantaged pupils.**

Schools with a higher share of disadvantaged pupils see a modest
*positive* effect on their disadvantaged subgroup's attainment (after
controlling for prior attainment). This suggests that concentrating
Pupil Premium resources — rather than spreading them thinly across a
school where disadvantaged pupils are a small minority — may be more
effective. Schools with lower FSM proportions should consider whether
their disadvantaged pupils receive sufficiently targeted support, or
whether resources are diluted across the whole cohort.


### For Brighton & Hove Council and local authority

**1. Tackle absence strategically across the Local Authority.**

The model confirms that absence is the most actionable predictor of
attainment at school level. The council can support schools through:

- Local Authority-wide attendance campaigns and shared best practice
- Coordinated data sharing to identify pupils with poor attendance
  across multiple schools (e.g. following managed moves)
- Targeted outreach to communities with high persistent-absence rates
- Ensuring that alternative provision maintains strong attendance
  tracking

**2. Address the disadvantaged attainment gap directly.**

The model shows that the coefficient magnitudes for disadvantaged pupils
are often **larger** than for non-disadvantaged — absence and prior
attainment in particular hit harder for the disadvantaged group.
Policy priorities include:

- Ring-fencing and monitoring the impact of Pupil Premium spending
  across all maintained and academy schools
- Commissioning cross-school programmes (e.g. summer schools, mentoring
  networks) that pool resources for disadvantaged pupils across the LA
- Tracking the FSM attainment gap as a headline KPI for the local
  authority, using the model's predicted vs actual framework to identify
  schools outperforming or underperforming expectations

**3. A note on segregation.**

The Gorard segregation index enters the model with a negative
coefficient (higher segregation associated with lower attainment), but
this effect is **statistically insignificant** — the standard error is
large relative to the estimate, and we cannot confidently distinguish
it from zero. While reducing segregation may be a worthwhile policy
goal for broader equity reasons, **this model does not provide reliable
evidence that reducing the Gorard index would improve Attainment 8
scores** at school level. The council should therefore not prioritise
segregation reduction *on the basis of this model alone*, though it
may of course remain a valid objective on other grounds (social
cohesion, equal access, community trust).

**4. Support workforce stability across schools.**

Teacher retention and absence are now confirmed as predictors of pupil
outcomes. The council and local partnerships can help by:

- Facilitating Local Authority-wide teacher recruitment and retention
  initiatives (e.g. housing support, shared CPD networks)
- Monitoring teacher turnover and sickness across schools, identifying
  outliers that may need additional support
- Ensuring that schools in challenging contexts (high FSM, high
  absence) receive additional staffing flexibility or funding to
  counteract higher turnover pressures

**5. Use the model to target support.**

The Shiny app's policy simulator allows the council to explore
"what-if" scenarios for individual schools. The variable importance
rankings show which levers move the needle most. By combining the
model's predictions with local knowledge, the council can:

- Prioritise school improvement resources toward the interventions with
  the largest predicted impact
- Use the residual analysis to identify schools that are outperforming
  expectations (positive residuals) and learn from their practices
- Track whether policy changes (e.g. attendance initiatives) produce
  the predicted improvements over subsequent years


### Summary: the biggest levers

The table below summarises the top policy priorities based on the
variable importance analysis:

```{r policy-summary-table}
#| label: policy-summary-table
#| echo: false

policy_summary <- tribble(
  ~Priority, ~`Variable`, ~`Who Can Act`, ~`Key Action`,
  1, "Overall Absence Rate", "School leaders + Council", "Intensive early intervention; protect low-absence schools",
  2, "% Low Prior Attainment", "School leaders", "Targeted KS2-KS3 transition catch-up programmes",
  3, "% Disadvantaged (FSM)", "School leaders + Council", "Concentrated Pupil Premium spend; monitor FSM gap as KPI",
  4, "Teacher Retention", "School leaders + Council", "Retention strategies; CPD investment; workload management",
  5, "Teacher Sickness Days", "School leaders", "Staff wellbeing programmes; absence monitoring",
  6, "% EAL", "School leaders", "Build on EAL advantage; ensure support for new arrivals",
  7, "Leadership Pay %", "School leaders", "Review leadership structure (low model impact)",
  8, "Gorard Segregation Index*", "Council", "Effect statistically insignificant — not a reliable lever in this model"
)

knitr::kable(
  policy_summary,
  align = "rlllc",
  caption = "Policy priorities ranked by variable importance"
)
```

\* The Gorard segregation index coefficient is statistically
insignificant in the imputed full model (Analysis E). It is included in
the table for completeness but should not be treated as an actionable
policy lever on the basis of this model.

> **Important caveat:** These recommendations are based on *associations*
> identified by the multilevel model. The model controls for observable
> confounders (intake, prior attainment, workforce, area characteristics)
> but cannot establish causation. A school reducing absence may not see
> exactly the predicted ATT8 gain if other unmeasured factors change
> simultaneously. The rankings and effect sizes should be treated as
> evidence-informed starting points for policy discussion, not as causal
> guarantees. Variables whose coefficients are statistically insignificant
> (notably the Gorard segregation index) should be interpreted with
> particular caution.


# Quick Reference

## Formula summary

| Analysis | Fixed Effects | Random Effects | Data | Used in App? |
|----------|--------------|----------------|------|:---:|
| A (Full Panel) | 9 predictors (5 log, 4 linear) | `(1\|year) + (1\|Ofsted) + (1\|region/LA)` | 3 years, workforce required | No |
| B (Full Per-Year) | Same 9 predictors | `(1\|Ofsted) + (1\|region/LA)` | Each year separately | No |
| C (Core Panel) | 5 predictors (3 log, 2 linear) | `(1\|year) + (1\|Ofsted) + (1\|region/LA)` | All 4 years | No |
| D (Core Per-Year) | Same 5 predictors | `(1\|Ofsted) + (1\|region/LA)` | Each year separately | No |
| **E (Imputed Full)** | **9 predictors (5 log, 4 linear)** | **`(1\|year) + (1\|Ofsted) + (1\|region/LA)`** | **All 4 years (2024-25 imputed)** | **Yes** |

## Variable key

| Variable | In formula as | Description |
|----------|--------------|-------------|
| `ATT8SCR` | `log(ATT8SCR)` | Attainment 8 score (all pupils) |
| `ATT8SCR_FSM6CLA1A` | `log(ATT8SCR_FSM6CLA1A)` | Attainment 8 score (disadvantaged) |
| `ATT8SCR_NFSM6CLA1A` | `log(ATT8SCR_NFSM6CLA1A)` | Attainment 8 score (non-disadvantaged) |
| `PTFSM6CLA1A` | `log(PTFSM6CLA1A)` | % eligible for Free School Meals |
| `PERCTOT` | `log(PERCTOT)` | % total absence |
| `PNUMEAL` | `log(PNUMEAL)` | % English as Additional Language |
| `PTPRIORLO` | `PTPRIORLO` | % low KS2 prior attainment |
| `ADMPOL_PT` | `ADMPOL_PT` | Admissions policy indicator |
| `gorard_segregation` | `gorard_segregation` | Gorard segregation index |
| `remained_in_the_same_school` | `remained_in_the_same_school` | Teacher retention (%) |
| `teachers_on_leadership_pay_range_percent` | as-is | % teachers on leadership pay |
| `average_number_of_days_taken` | `log(average_number_of_days_taken)` | Average teacher absence days |
| `OFSTEDRATING_1` | `(1\|OFSTEDRATING_1)` | Ofsted rating (random intercept) |
| `gor_name` | `(1\|gor_name/LANAME)` | Government Office Region |
| `LANAME` | nested in `gor_name` | Local Authority |
| `year_label` | `(1\|year_label)` | Academic year (panel models only) |


# Disadvantaged Attainment and Concentrations of Disadvantage: Directionality Test {#sec-directionality}

In Analysis E, the coefficient on `log(PTFSM6CLA1A)` flips from negative (All Pupils, Non-Disadvantaged) to **positive** for Disadvantaged pupils once prior attainment and other controls are included. This section tests whether the sign-flip is robust to alternative operationalisations of both the outcome (disadvantaged attainment) and the predictor (school-level disadvantage concentration), or whether it is an artefact of the specific FSM6CLA1A classification and Attainment 8 metric.

**Strategy:** Hold the model specification constant (same random effects, same controls) and systematically vary:

1. The **Y variable** — alternative measures of disadvantaged pupil attainment
2. The **X variable** — alternative measures of school-level disadvantage concentration

If the sign-flip appears consistently across different measures, it is more likely to reflect a genuine structural relationship than a measurement artefact.

## Setup: baseline model and data

```{r directionality-setup}
#| label: directionality-setup

# Use the same imputed full dataset as Analysis E
d_dir <- imputed_full_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0,
         PTFSM6CLA1A > 0, PERCTOT > 0, PNUMEAL > 0,
         !is.na(PTPRIORLO),
         !is.na(OFSTEDRATING_1), !is.na(gor_name), !is.na(LANAME),
         !is.na(remained_in_the_same_school),
         !is.na(teachers_on_leadership_pay_range_percent),
         average_number_of_days_taken > 0,
         !is.na(gorard_segregation)) %>%
  droplevels()

contrasts(d_dir$OFSTEDRATING_1) <- contr.treatment(levels(d_dir$OFSTEDRATING_1))

cat("Directionality test dataset:", nrow(d_dir), "rows\n")
cat("Years:", paste(unique(d_dir$year_label), collapse = ", "), "\n")

# Reference: the Analysis E disadvantaged coefficient
ref_coef <- fixef(mod_e_disadv)["log(PTFSM6CLA1A)"]
cat(sprintf("\nReference (Analysis E): log(PTFSM6CLA1A) coefficient = %+.5f\n", ref_coef))
```

## Test 1: Alternative Y variables (disadvantaged attainment)

Holding `log(PTFSM6CLA1A)` as the deprivation measure, we swap the outcome variable across a comprehensive set of alternatives. These fall into four groups:

1. **ATT8 subject components for disadvantaged pupils** — English, Maths, EBacc, Open (total), Open GCSE, and Open non-GCSE. These test whether the sign-flip is specific to the overall ATT8 aggregate or appears consistently across curriculum domains.
2. **% disadvantaged achieving L2 Basics (grade 4+)** — a binary threshold measure rather than a continuous score. If the sign-flip appears here, it means higher FSM concentration is associated with a *higher pass rate* for disadvantaged pupils, not just a higher average score.
3. **ATT8 by prior attainment band (all pupils)** — Low, Middle, and High prior attainment groups. These are not limited to FSM-eligible pupils, but if the sign-flip appears for low-prior-attainment pupils (who overlap substantially with disadvantaged pupils), it provides triangulating evidence from a different subgroup definition.
4. **The disadvantaged–non-disadvantaged ATT8 gap** (`DIFFN_ATT8`) — the difference itself as the outcome. A positive coefficient here would mean higher FSM concentration *narrows* the attainment gap.

Note: `PTFSM6CLA1ABASICS_94` is a percentage (0–100) and `DIFFN_ATT8` is typically negative (disadvantaged score minus non-disadvantaged); both are modelled without log transformation.

```{r directionality-y-vars}
#| label: directionality-y-vars

# Define alternative Y variables to test
y_tests <- tribble(
  ~y_var,                      ~label,                                         ~transform,  ~filter_positive,
  # --- ATT8 components for disadvantaged pupils ---
  "ATT8SCR_FSM6CLA1A",        "ATT8 Disadvantaged (reference)",               "log",       TRUE,
  "ATT8SCRENG_FSM6CLA1A",     "ATT8 English (Disadvantaged)",                 "log",       TRUE,
  "ATT8SCRMAT_FSM6CLA1A",     "ATT8 Maths (Disadvantaged)",                   "log",       TRUE,
  "ATT8SCREBAC_FSM6CLA1A",    "ATT8 EBacc (Disadvantaged)",                   "log",       TRUE,
  "ATT8SCROPEN_FSM6CLA1A",    "ATT8 Open (Disadvantaged)",                    "log",       TRUE,
  "ATT8SCROPENG_FSM6CLA1A",   "ATT8 Open GCSE (Disadvantaged)",               "log",       TRUE,
  "ATT8SCROPENNG_FSM6CLA1A",  "ATT8 Open non-GCSE (Disadvantaged)",           "log",       TRUE,
  # --- L2 Basics threshold for disadvantaged ---
  "PTFSM6CLA1ABASICS_94",     "% Disadv. achieving L2 Basics (grade 4+)",     "none",      FALSE,
  # --- ATT8 by prior attainment band (all pupils) ---
  "ATT8SCR_LO",               "ATT8 Low Prior Attainment (all pupils)",        "log",       TRUE,
  "ATT8SCR_MID",              "ATT8 Middle Prior Attainment (all pupils)",     "log",       TRUE,
  "ATT8SCR_HI",               "ATT8 High Prior Attainment (all pupils)",       "log",       TRUE,
  # --- The gap itself ---
  "DIFFN_ATT8",               "ATT8 Gap (Disadv. minus Non-Disadv.)",         "none",      FALSE
)

# Fit each model and extract the FSM coefficient
y_results <- map_dfr(seq_len(nrow(y_tests)), function(i) {
  y_var <- y_tests$y_var[i]
  label <- y_tests$label[i]
  transform <- y_tests$transform[i]
  filter_pos <- y_tests$filter_positive[i]

  d_test <- d_dir
  y_vals <- d_test[[y_var]]

  # Filter to valid observations
  if (filter_pos) {
    d_test <- d_test[!is.na(y_vals) & y_vals > 0, ]
  } else {
    d_test <- d_test[!is.na(y_vals), ]
  }
  d_test <- droplevels(d_test)
  contrasts(d_test$OFSTEDRATING_1) <- contr.treatment(levels(d_test$OFSTEDRATING_1))

  # Build formula
  if (transform == "log") {
    lhs <- paste0("log(", y_var, ")")
  } else {
    lhs <- y_var
  }

  formula_str <- paste0(
    lhs, " ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) + ",
    "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
    "remained_in_the_same_school + teachers_on_leadership_pay_range_percent + ",
    "log(average_number_of_days_taken) + ",
    "(1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
  )

  mod <- tryCatch(
    lmer(as.formula(formula_str), data = d_test, REML = TRUE,
         na.action = na.exclude,
         control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))),
    error = function(e) NULL
  )

  if (is.null(mod)) {
    return(tibble(Y_Variable = label, N = nrow(d_test),
                  Coefficient = NA_real_, SE = NA_real_,
                  t_value = NA_real_, Sign = "FAILED"))
  }

  # lmerTest::summary gives Satterthwaite p-values
  coefs <- summary(mod)$coefficients
  row_name <- "log(PTFSM6CLA1A)"
  if (row_name %in% rownames(coefs)) {
    p_val <- if ("Pr(>|t|)" %in% colnames(coefs)) coefs[row_name, "Pr(>|t|)"] else NA_real_
    tibble(
      Y_Variable = label,
      N = nrow(d_test),
      Coefficient = coefs[row_name, "Estimate"],
      SE = coefs[row_name, "Std. Error"],
      t_value = coefs[row_name, "t value"],
      p_value = p_val,
      Sign = ifelse(coefs[row_name, "Estimate"] > 0, "POSITIVE", "negative")
    )
  } else {
    tibble(Y_Variable = label, N = nrow(d_test),
           Coefficient = NA_real_, SE = NA_real_,
           t_value = NA_real_, p_value = NA_real_, Sign = "NOT FOUND")
  }
})

# Add significance stars
y_results <- y_results %>%
  mutate(
    Sig = case_when(
      is.na(p_value) ~ "",
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE ~ ""
    )
  )

knitr::kable(
  y_results %>%
    mutate(Coefficient = sprintf("%+.5f", Coefficient),
           SE = sprintf("%.5f", SE),
           t_value = sprintf("%+.2f", t_value),
           p_value = ifelse(p_value < 0.001, "< 0.001", sprintf("%.3f", p_value))) %>%
    select(Y_Variable, N, Coefficient, SE, t_value, p_value, Sig, Sign),
  caption = "Test 1: Coefficient on log(% FSM) across alternative disadvantaged attainment measures (Satterthwaite p-values)",
  align = "lrrrrrll"
)
```

```{r directionality-y-narrative}
#| label: directionality-y-narrative
#| output: asis

n_positive <- sum(y_results$Sign == "POSITIVE", na.rm = TRUE)
n_total <- sum(!is.na(y_results$Coefficient))
n_sig <- sum(y_results$Sign == "POSITIVE" & y_results$p_value < 0.05, na.rm = TRUE)
n_marginal <- sum(y_results$Sign == "POSITIVE" & y_results$p_value >= 0.05 & y_results$p_value < 0.1, na.rm = TRUE)
n_neg_sig <- sum(y_results$Sign == "negative" & y_results$p_value < 0.05, na.rm = TRUE)

cat(sprintf("Of the %d outcome measures tested, **%d show a positive coefficient** on log(%% FSM)", n_total, n_positive))
if (n_sig > 0 || n_marginal > 0) {
  parts <- c()
  if (n_sig > 0) parts <- c(parts, sprintf("%d statistically significant at p < 0.05", n_sig))
  if (n_marginal > 0) parts <- c(parts, sprintf("%d marginally significant (p < 0.1)", n_marginal))
  cat(sprintf(", of which %s", paste(parts, collapse = " and ")))
}
if (n_neg_sig > 0) {
  cat(sprintf(". **%d show a significantly negative coefficient**", n_neg_sig))
}
cat(".\n\n")

# --- Group-by-group commentary ---
# ATT8 disadvantaged components
disadv_rows <- y_results %>% filter(grepl("Disadvantaged|Open GCSE|Open non-GCSE", Y_Variable))
n_pos_disadv <- sum(disadv_rows$Sign == "POSITIVE", na.rm = TRUE)
n_sig_disadv <- sum(disadv_rows$Sign == "POSITIVE" & disadv_rows$p_value < 0.05, na.rm = TRUE)
cat(sprintf("**ATT8 components for disadvantaged pupils:** %d of %d show a positive coefficient (%d significant). ",
    n_pos_disadv, nrow(disadv_rows), n_sig_disadv))
if (n_pos_disadv == nrow(disadv_rows)) {
  cat("The sign-flip is consistent across **every subject domain** — English, Maths, EBacc, Open (total and split into GCSE/non-GCSE) — making it unlikely to be an artefact of the Attainment 8 weighting formula.\n\n")
} else {
  cat("The sign-flip varies across subject domains.\n\n")
}

# Open GCSE vs non-GCSE deep dive
openg_row <- y_results %>% filter(grepl("Open GCSE \\(", Y_Variable))
openng_row <- y_results %>% filter(grepl("Open non-GCSE", Y_Variable))
open_total_row <- y_results %>% filter(Y_Variable == "ATT8 Open (Disadvantaged)")

if (nrow(openng_row) > 0 && !is.na(openng_row$Coefficient) &&
    nrow(openg_row) > 0 && !is.na(openg_row$Coefficient)) {

  cat("::: {.callout-note}\n")
  cat("## The Open non-GCSE result: qualification pathways and the sign-flip\n\n")

  cat(sprintf("The coefficient on `log(%% FSM)` is strikingly different between the two Open sub-components: **%+.5f** (t = %+.2f) for Open GCSE qualifications versus **%+.5f** (t = %+.2f) for Open non-GCSE qualifications.\n\n",
      openg_row$Coefficient, openg_row$t_value,
      openng_row$Coefficient, openng_row$t_value))

  cat("**What Open non-GCSE measures:** The ATT8 Open bucket is split into GCSEs (e.g. Art, Music, Computer Science, Design & Technology) and non-GCSE qualifications (e.g. BTECs, Cambridge Nationals, vocational and technical awards). The non-GCSE component captures attainment in alternative qualification pathways that many schools use to provide accessible, applied routes to KS4 achievement.\n\n")

  if (openng_row$Coefficient > openg_row$Coefficient) {
    cat("The much larger positive coefficient for non-GCSE qualifications suggests that **schools with higher FSM concentrations achieve disproportionately better outcomes for their disadvantaged pupils specifically in vocational/technical pathways**. This finding admits two quite different interpretations:\n\n")

    cat("**Interpretation 1: Strategic curriculum design (strengthens the sign-flip argument).** High-FSM schools have learned, through experience and institutional focus, to build effective vocational pathways that genuinely serve their disadvantaged cohorts well. They invest Pupil Premium in well-resourced BTEC/vocational programmes, employ specialist staff, and develop employer links. The positive coefficient reflects real, meaningful attainment gains delivered through appropriate curriculum design — a form of the \"critical mass\" effect where concentrated disadvantage drives institutional specialisation.\n\n")

    cat("**Interpretation 2: Qualification gaming (weakens the sign-flip argument).** High-FSM schools may be systematically entering disadvantaged pupils for non-GCSE qualifications where grade boundaries or point-score equivalences are more generous, inflating the ATT8 score without necessarily delivering equivalent learning. If non-GCSE qualifications carry ATT8 point values that are more achievable per unit of effort, the positive coefficient may reflect strategic entry patterns rather than genuine attainment differences.\n\n")

    cat("**How to distinguish:** If the sign-flip were *only* visible in the non-GCSE Open component, Interpretation 2 would be concerning. But the sign-flip also appears in the GCSE-based components (English, Maths, EBacc")
    if (openg_row$Coefficient > 0) {
      cat(", and Open GCSE")
    }
    cat("), where qualification gaming is not possible — GCSEs are GCSEs. The non-GCSE result therefore most likely reflects a **combination** of both mechanisms: high-FSM schools genuinely serve their disadvantaged pupils better across all domains (Interpretation 1), *and* they additionally benefit from strategic use of vocational pathways (Interpretation 2) — which may itself be a legitimate form of institutional responsiveness to cohort needs.\n\n")

    cat(sprintf("The practical implication is that the **overall ATT8 sign-flip (coefficient = %+.5f) is likely an amalgam**: a modest genuine effect across all subjects, amplified by effective vocational provision in high-FSM schools. The GCSE-only components provide the cleaner test of the core structural relationship, while the non-GCSE component captures an additional pathway-specific effect.\n",
        y_results$Coefficient[y_results$Y_Variable == "ATT8 Disadvantaged (reference)"]))
  } else {
    cat("Interestingly, the Open GCSE component shows a stronger positive coefficient than the non-GCSE component, suggesting the sign-flip is not driven by vocational qualification pathways.\n\n")
  }
  cat(":::\n\n")
}

# L2 Basics
basics_row <- y_results %>% filter(grepl("L2 Basics", Y_Variable))
if (nrow(basics_row) > 0 && !is.na(basics_row$Coefficient)) {
  cat(sprintf("**%% Disadvantaged achieving L2 Basics (grade 4+):** Coefficient = %+.5f (t = %+.2f, p = %s). ",
      basics_row$Coefficient, basics_row$t_value,
      ifelse(basics_row$p_value < 0.001, "< 0.001", sprintf("%.3f", basics_row$p_value))))
  if (basics_row$Coefficient > 0) {
    cat("The positive sign means higher FSM concentration is associated with a higher *pass rate* for disadvantaged pupils — the effect is not limited to average scores but extends to the proportion clearing the key grade 4 threshold.\n\n")
  } else {
    cat("The negative sign here contrasts with the ATT8 results. This could reflect floor effects: schools with very high FSM may have more pupils clustered just below the grade boundary, where average ATT8 can rise slightly without pushing more pupils over the threshold.\n\n")
  }
}

# Prior attainment bands
prior_rows <- y_results %>% filter(grepl("Prior Attainment", Y_Variable))
if (nrow(prior_rows) > 0) {
  lo_row <- prior_rows %>% filter(grepl("Low", Y_Variable))
  mid_row <- prior_rows %>% filter(grepl("Middle", Y_Variable))
  hi_row <- prior_rows %>% filter(grepl("High", Y_Variable))

  cat("**ATT8 by prior attainment band (all pupils):** ")
  if (nrow(lo_row) > 0 && !is.na(lo_row$Coefficient)) {
    cat(sprintf("Low prior: %+.5f (t = %+.2f, p = %s); ",
        lo_row$Coefficient, lo_row$t_value,
        ifelse(lo_row$p_value < 0.001, "< 0.001", sprintf("%.3f", lo_row$p_value))))
  }
  if (nrow(mid_row) > 0 && !is.na(mid_row$Coefficient)) {
    cat(sprintf("Middle: %+.5f (t = %+.2f); ", mid_row$Coefficient, mid_row$t_value))
  }
  if (nrow(hi_row) > 0 && !is.na(hi_row$Coefficient)) {
    cat(sprintf("High: %+.5f (t = %+.2f). ", hi_row$Coefficient, hi_row$t_value))
  }

  if (nrow(lo_row) > 0 && !is.na(lo_row$Coefficient) && lo_row$Coefficient > 0) {
    cat("The positive coefficient for **low prior attainment pupils** is particularly telling — these pupils overlap substantially with the disadvantaged subgroup, providing triangulating evidence from a different (ability-based rather than income-based) definition of vulnerability. ")
  }
  if (nrow(hi_row) > 0 && !is.na(hi_row$Coefficient) && hi_row$Coefficient < 0) {
    cat("The negative coefficient for **high prior attainment** pupils makes intuitive sense: high-ability pupils in high-FSM schools may face less academic stretch and fewer high-achieving peers, consistent with a composition effect that works in the opposite direction for the most able.")
  }
  cat("\n\n")
}

# The gap
gap_row <- y_results %>% filter(grepl("Gap", Y_Variable))
if (nrow(gap_row) > 0 && !is.na(gap_row$Coefficient)) {
  cat(sprintf("**The ATT8 gap (Disadvantaged minus Non-Disadvantaged):** Coefficient = %+.5f (t = %+.2f, p = %s). ",
      gap_row$Coefficient, gap_row$t_value,
      ifelse(gap_row$p_value < 0.001, "< 0.001", sprintf("%.3f", gap_row$p_value))))
  if (gap_row$Coefficient > 0) {
    cat("A positive coefficient here means the gap **narrows** (becomes less negative) in schools with higher FSM concentrations. This is the most direct test of the policy-relevant question: does concentrating disadvantaged pupils reduce the attainment gap? The model suggests it does, modestly.\n\n")
  } else {
    cat("A negative coefficient means the gap **widens** in schools with higher FSM concentrations, which would undermine the sign-flip narrative. However, note that the gap variable captures the *difference* between two scores that are both affected by FSM in different ways.\n\n")
  }
}

# Significance patterns
sig_rows <- y_results %>% filter(Sign == "POSITIVE", p_value < 0.05)
nonsig_pos_rows <- y_results %>% filter(Sign == "POSITIVE", p_value >= 0.05)
if (nrow(sig_rows) > 0 && nrow(nonsig_pos_rows) > 0) {
  cat("**Significance patterns:** The coefficient is statistically significant for: ",
      paste0("*", sig_rows$Y_Variable, "*", collapse = ", "), ". ", sep = "")
  cat("It is positive but non-significant for: ",
      paste0("*", nonsig_pos_rows$Y_Variable, "*", collapse = ", "), ". ", sep = "")
  cat("This suggests the effect is real but modest — the direction is consistent but precision varies with sample size and variance in each outcome measure.\n\n")
} else if (nrow(sig_rows) == n_positive && n_positive > 0) {
  cat("**All positive coefficients are statistically significant**, providing strong evidence that the sign-flip is not a chance finding.\n\n")
} else if (n_positive > 0 && nrow(sig_rows) == 0) {
  cat("**None of the positive coefficients reach statistical significance** at p < 0.05. The consistent direction is suggestive but the effect may be too small to detect reliably.\n\n")
}
```

## Test 2: Alternative X variables (school-level disadvantage concentration)

Now we hold the outcome fixed at `log(ATT8SCR_FSM6CLA1A)` and swap the deprivation predictor. Each alternative X replaces `log(PTFSM6CLA1A)` in the model while keeping all other predictors unchanged.

The interpretation of these tests depends on understanding how each alternative X variable relates to the two key controls already in the model — **% low prior attainment** (`PTPRIORLO`) and **% overall absence** (`log(PERCTOT)`):

- **% FSM6CLA1A** (reference): the DfE's KS4-cohort disadvantage flag (FSM-eligible in the past 6 years or looked-after). Strongly correlated with both prior attainment and absence — the sign-flip only appears once these pathways are controlled for.

- **% ever FSM** (`PNUMFSMEVER`): a census-based measure covering all pupils on roll, not just the KS4 cohort. This is a broader economic deprivation measure and will be highly correlated with `PTFSM6CLA1A` but not identical — it includes younger year groups and uses a different snapshot. If the sign-flip persists, it confirms the finding is not an artefact of the specific KS4 cohort definition. Because `PNUMFSMEVER` is correlated with both `PTPRIORLO` and `PERCTOT` in similar ways to `PTFSM6CLA1A`, we would expect a similar decomposition.

- **% any SEN** (`PSEN_ALL4`): captures a different dimension of disadvantage — educational needs rather than economic deprivation. SEN rates are moderately correlated with FSM (schools in deprived areas tend to have higher SEN identification rates) but the overlap is far from complete. Crucially, the relationship between SEN concentration and prior attainment is more direct: schools with high SEN rates will mechanically have more pupils with low KS2 scores. If the sign-flip appears here too, it suggests that the "critical mass" or resource concentration effect operates across different types of disadvantage, not just economic deprivation.

- **% persistent absentees** (`PPERSABS10`): the proportion of pupils missing 10%+ of sessions. This is a behavioural measure that overlaps with but is conceptually distinct from FSM. The key complication is that persistent absence is *also* captured (partially) by the overall absence rate `PERCTOT` already in the model. A school with 15% persistent absentees and 7% overall absence is in a different situation from one with 15% persistent absentees and 5% overall absence. The coefficient on `PPERSABS10` therefore captures the *additional* effect of having a concentrated tail of chronic non-attenders beyond what the school's mean absence rate already explains.

- **% EAL not yet fluent** (`PTEALGRP2`): measures linguistic disadvantage independently of economic deprivation. EAL rates have a complex relationship with both prior attainment (newly arrived pupils may have low KS2 scores or no KS2 data) and absence (EAL pupils' absence patterns vary widely by community). This is the most conceptually distinct alternative — if the sign-flip appears with EAL concentration, it would provide strong evidence for a general "concentration benefit" rather than anything specific to FSM.

```{r directionality-x-correlations}
#| label: directionality-x-correlations
#| fig-height: 4
#| fig-width: 8
#| fig-cap: "Pairwise correlations between alternative X variables and the key model controls (PTPRIORLO and PERCTOT). Higher correlations mean the X variable shares more variance with the controls, and its coefficient captures a more residual effect."

# Compute correlations between X variables and the two key controls
x_var_names <- c("PTFSM6CLA1A", "PNUMFSMEVER", "PSEN_ALL4", "PPERSABS10", "PTEALGRP2")
x_var_labels <- c("% FSM", "% ever FSM", "% any SEN", "% persist. absent", "% EAL not fluent")
control_names <- c("PTPRIORLO", "PERCTOT")
control_labels <- c("% Low Prior Attainment", "% Overall Absence")

cor_data <- expand_grid(
  x_idx = seq_along(x_var_names),
  c_idx = seq_along(control_names)
) %>%
  mutate(
    x_var = x_var_names[x_idx],
    x_label = x_var_labels[x_idx],
    control = control_names[c_idx],
    control_label = control_labels[c_idx],
    r = map2_dbl(x_var, control, function(xv, cv) {
      vals <- d_dir[, c(xv, cv)] %>% drop_na()
      if (nrow(vals) < 30) return(NA_real_)
      cor(log(pmax(vals[[xv]], 0.01)), vals[[cv]], use = "complete.obs")
    })
  )

ggplot(cor_data, aes(x = x_label, y = control_label, fill = r)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = sprintf("r = %.2f", r)),
            size = 3.5, fontface = "bold",
            colour = ifelse(abs(cor_data$r) > 0.5, "white", "grey30")) +
  scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1), name = "Correlation") +
  labs(title = "How alternative X variables correlate with model controls",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

```{r directionality-x-cor-narrative}
#| label: directionality-x-cor-narrative
#| output: asis

# Find the highest and lowest correlations with PTPRIORLO
cor_prior <- cor_data %>% filter(control == "PTPRIORLO")
cor_abs <- cor_data %>% filter(control == "PERCTOT")

cat("The correlation matrix reveals important differences in how each X variable relates to the model's controls:\n\n")

for (i in seq_along(x_var_names)) {
  r_prior <- cor_prior$r[cor_prior$x_var == x_var_names[i]]
  r_abs <- cor_abs$r[cor_abs$x_var == x_var_names[i]]
  if (!is.na(r_prior) && !is.na(r_abs)) {
    cat(sprintf("- **%s**: r = %.2f with prior attainment, r = %.2f with absence\n",
        x_var_labels[i], r_prior, r_abs))
  }
}

cat("\nVariables with high correlations to both controls (e.g. % FSM, % ever FSM) will show the sign-flip most cleanly because the model's controls effectively remove the indirect pathways through prior attainment and absence. Variables with lower correlations to the controls (e.g. % EAL) are a tougher test — the sign-flip appearing with these would be stronger evidence of a genuine structural effect.\n\n")
```

```{r directionality-x-vars}
#| label: directionality-x-vars

# Define alternative X variables to test
x_tests <- tribble(
  ~x_var,            ~label,                                 ~transform,
  "PTFSM6CLA1A",    "% FSM6CLA1A (reference)",              "log",
  "PNUMFSMEVER",    "% ever FSM (census, broader defn.)",   "log",
  "PSEN_ALL4",      "% any SEN",                            "log",
  "PPERSABS10",     "% persistent absentees (10%+)",        "log",
  "PTEALGRP2",      "% EAL not yet fluent",                 "log"
)

x_results <- map_dfr(seq_len(nrow(x_tests)), function(i) {
  x_var <- x_tests$x_var[i]
  label <- x_tests$label[i]
  transform <- x_tests$transform[i]

  d_test <- d_dir
  x_vals <- d_test[[x_var]]

  # Filter to valid, positive values for log transform
  d_test <- d_test[!is.na(x_vals) & x_vals > 0, ]
  d_test <- droplevels(d_test)
  contrasts(d_test$OFSTEDRATING_1) <- contr.treatment(levels(d_test$OFSTEDRATING_1))

  # Build formula: swap x_var for PTFSM6CLA1A
  x_term <- if (transform == "log") paste0("log(", x_var, ")") else x_var
  coef_name <- if (transform == "log") paste0("log(", x_var, ")") else x_var

  formula_str <- paste0(
    "log(ATT8SCR_FSM6CLA1A) ~ ", x_term, " + log(PERCTOT) + log(PNUMEAL) + ",
    "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
    "remained_in_the_same_school + teachers_on_leadership_pay_range_percent + ",
    "log(average_number_of_days_taken) + ",
    "(1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
  )

  mod <- tryCatch(
    lmer(as.formula(formula_str), data = d_test, REML = TRUE,
         na.action = na.exclude,
         control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))),
    error = function(e) NULL
  )

  if (is.null(mod)) {
    return(tibble(X_Variable = label, N = nrow(d_test),
                  Coefficient = NA_real_, SE = NA_real_,
                  t_value = NA_real_, Sign = "FAILED"))
  }

  coefs <- summary(mod)$coefficients
  if (coef_name %in% rownames(coefs)) {
    p_val <- if ("Pr(>|t|)" %in% colnames(coefs)) coefs[coef_name, "Pr(>|t|)"] else NA_real_
    tibble(
      X_Variable = label,
      N = nrow(d_test),
      Coefficient = coefs[coef_name, "Estimate"],
      SE = coefs[coef_name, "Std. Error"],
      t_value = coefs[coef_name, "t value"],
      p_value = p_val,
      Sign = ifelse(coefs[coef_name, "Estimate"] > 0, "POSITIVE", "negative")
    )
  } else {
    tibble(X_Variable = label, N = nrow(d_test),
           Coefficient = NA_real_, SE = NA_real_,
           t_value = NA_real_, p_value = NA_real_, Sign = "NOT FOUND")
  }
})

x_results <- x_results %>%
  mutate(
    Sig = case_when(
      is.na(p_value) ~ "",
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      p_value < 0.1   ~ ".",
      TRUE ~ ""
    )
  )

knitr::kable(
  x_results %>%
    mutate(Coefficient = sprintf("%+.5f", Coefficient),
           SE = sprintf("%.5f", SE),
           t_value = sprintf("%+.2f", t_value),
           p_value = ifelse(p_value < 0.001, "< 0.001", sprintf("%.3f", p_value))) %>%
    select(X_Variable, N, Coefficient, SE, t_value, p_value, Sig, Sign),
  caption = "Test 2: Coefficient on alternative deprivation measures predicting log(ATT8 Disadvantaged) (Satterthwaite p-values)",
  align = "lrrrrrll"
)
```

```{r directionality-x-narrative}
#| label: directionality-x-narrative
#| output: asis

n_positive_x <- sum(x_results$Sign == "POSITIVE", na.rm = TRUE)
n_total_x <- sum(!is.na(x_results$Coefficient))

n_sig_x <- sum(x_results$Sign == "POSITIVE" & x_results$p_value < 0.05, na.rm = TRUE)
n_neg_sig_x <- sum(x_results$Sign == "negative" & x_results$p_value < 0.05, na.rm = TRUE)

cat(sprintf("Of the %d deprivation measures tested, **%d show a positive coefficient** when predicting disadvantaged attainment", n_total_x, n_positive_x))
if (n_sig_x > 0) {
  cat(sprintf(" (%d significant at p < 0.05)", n_sig_x))
}
if (n_neg_sig_x > 0) {
  cat(sprintf(". **%d show a significantly negative coefficient**, indicating the sign-flip does not generalise to all conceptualisations of disadvantage", n_neg_sig_x))
}
cat(".\n\n")

cat("As discussed above, each alternative X captures a different dimension of disadvantage concentration, with different degrees of overlap with the model's prior attainment and absence controls. The correlation analysis showed which variables share the most variance with those controls — and therefore which provide the most independent test of the sign-flip.\n\n")

# Classify results by correlation strength
cat("Interpreting the results in light of the correlations:\n\n")

for (i in seq_len(nrow(x_results))) {
  xv <- x_tests$x_var[i]
  r_prior_val <- cor_data %>% filter(x_var == xv, control == "PTPRIORLO") %>% pull(r)
  r_abs_val <- cor_data %>% filter(x_var == xv, control == "PERCTOT") %>% pull(r)
  sign_text <- x_results$Sign[i]

  if (length(r_prior_val) > 0 && !is.na(r_prior_val) && !is.na(x_results$Coefficient[i])) {
    overlap <- if (abs(r_prior_val) > 0.5 || abs(r_abs_val) > 0.5) {
      "high overlap with controls"
    } else if (abs(r_prior_val) > 0.3 || abs(r_abs_val) > 0.3) {
      "moderate overlap with controls"
    } else {
      "low overlap with controls — a more independent test"
    }
    cat(sprintf("- **%s** (%s): coefficient is **%s** (%+.5f, t = %+.2f)\n",
        x_results$X_Variable[i], overlap,
        tolower(sign_text), x_results$Coefficient[i], x_results$t_value[i]))
  }
}

cat("\n")

if (n_positive_x >= 3) {
  cat("The consistency of the positive sign across multiple, conceptually distinct measures of disadvantage concentration — including those with lower correlations to the model's controls — strengthens the case that this is a **genuine structural relationship** rather than an artefact of the FSM6CLA1A classification. The sign-flip survives even when the deprivation measure shares less variance with prior attainment and absence, suggesting the \"critical mass\" or resource concentration effect is not simply an artefact of how those confounders are partialled out.\n\n")
} else if (n_positive_x >= 1) {
  cat("The sign-flip appears with some but not all alternative measures. Notably, it tends to be strongest with X variables that are **highly correlated with the controls** (prior attainment and absence), suggesting the sign-flip may depend on the model successfully removing those indirect pathways. Where the X variable is more independent of the controls, the raw negative association between disadvantage and attainment may dominate.\n\n")
} else {
  cat("The positive sign appears **only with the original FSM6CLA1A measure**, raising the possibility that the sign-flip may be partly a measurement artefact. The alternative deprivation measures, which have different correlation structures with prior attainment and absence, do not replicate the finding. Further investigation with individual pupil-level data would be needed.\n\n")
}
```

## Test 3: Cross-tabulation (Y x X combinations)

As a final check, we fit the most informative subset of Y × X combinations to see whether the pattern is consistent across the full grid.

```{r directionality-grid}
#| label: directionality-grid
#| fig-height: 7
#| fig-width: 9

# Select key Y and X combinations
grid_y <- tribble(
  ~y_var,                      ~y_label,              ~y_transform,  ~y_filter_pos,
  "ATT8SCR_FSM6CLA1A",        "ATT8 Disadv.",        "log",         TRUE,
  "ATT8SCRENG_FSM6CLA1A",     "ATT8 Eng Dis.",       "log",         TRUE,
  "ATT8SCRMAT_FSM6CLA1A",     "ATT8 Mat Dis.",       "log",         TRUE,
  "ATT8SCREBAC_FSM6CLA1A",    "ATT8 EBacc Dis.",     "log",         TRUE,
  "ATT8SCROPEN_FSM6CLA1A",    "ATT8 Open Dis.",      "log",         TRUE,
  "PTFSM6CLA1ABASICS_94",     "% Dis. L2 Basics",    "none",        FALSE,
  "ATT8SCR_LO",               "ATT8 Low Prior",      "log",         TRUE,
  "DIFFN_ATT8",               "ATT8 Gap",            "none",        FALSE
)

grid_x <- tribble(
  ~x_var,            ~x_label,              ~x_transform,
  "PTFSM6CLA1A",    "% FSM",               "log",
  "PNUMFSMEVER",    "% ever FSM",          "log",
  "PSEN_ALL4",      "% any SEN",           "log",
  "PPERSABS10",     "% persist. absent",   "log"
)

grid_results <- expand_grid(
  y_idx = seq_len(nrow(grid_y)),
  x_idx = seq_len(nrow(grid_x))
) %>%
  mutate(result = map2(y_idx, x_idx, function(yi, xi) {
    y_var <- grid_y$y_var[yi]
    x_var <- grid_x$x_var[xi]
    y_transform <- grid_y$y_transform[yi]
    x_transform <- grid_x$x_transform[xi]
    filter_pos <- grid_y$y_filter_pos[yi]

    d_test <- d_dir
    y_vals <- d_test[[y_var]]
    x_vals <- d_test[[x_var]]

    # Filter
    valid <- !is.na(y_vals) & !is.na(x_vals) & x_vals > 0
    if (filter_pos) valid <- valid & y_vals > 0
    d_test <- d_test[valid, ] %>% droplevels()
    contrasts(d_test$OFSTEDRATING_1) <- contr.treatment(levels(d_test$OFSTEDRATING_1))

    lhs <- if (y_transform == "log") paste0("log(", y_var, ")") else y_var
    x_term <- if (x_transform == "log") paste0("log(", x_var, ")") else x_var
    coef_name <- if (x_transform == "log") paste0("log(", x_var, ")") else x_var

    formula_str <- paste0(
      lhs, " ~ ", x_term, " + log(PERCTOT) + log(PNUMEAL) + ",
      "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
      "remained_in_the_same_school + teachers_on_leadership_pay_range_percent + ",
      "log(average_number_of_days_taken) + ",
      "(1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
    )

    mod <- tryCatch(
      lmer(as.formula(formula_str), data = d_test, REML = TRUE,
           na.action = na.exclude,
           control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))),
      error = function(e) NULL
    )

    if (is.null(mod)) return(tibble(coef = NA_real_, t = NA_real_, p = NA_real_, n = nrow(d_test)))

    coefs <- summary(mod)$coefficients
    if (coef_name %in% rownames(coefs)) {
      p_val <- if ("Pr(>|t|)" %in% colnames(coefs)) coefs[coef_name, "Pr(>|t|)"] else NA_real_
      tibble(coef = coefs[coef_name, "Estimate"],
             t = coefs[coef_name, "t value"],
             p = p_val,
             n = nrow(d_test))
    } else {
      tibble(coef = NA_real_, t = NA_real_, p = NA_real_, n = nrow(d_test))
    }
  })) %>%
  unnest(result) %>%
  mutate(
    y_label = grid_y$y_label[y_idx],
    x_label = grid_x$x_label[x_idx],
    sign = case_when(
      is.na(coef) ~ "NA",
      coef > 0 ~ "+",
      TRUE ~ "–"
    ),
    sig_marker = case_when(
      is.na(p) ~ "",
      p < 0.001 ~ "***",
      p < 0.01  ~ "**",
      p < 0.05  ~ "*",
      p < 0.1   ~ "\u2020",
      TRUE ~ ""
    )
  )

# Heatmap with significance markers
ggplot(grid_results, aes(x = x_label, y = y_label, fill = coef)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = sprintf("%+.4f %s\n(t=%+.1f)", coef, sig_marker, t)),
            size = 3.0, colour = "white", fontface = "bold") +
  scale_fill_gradient2(low = "#7b132b", mid = "#f5f5f5", high = "#2e6260",
                       midpoint = 0, name = "Coefficient") +
  labs(title = "Disadvantage concentration coefficient across Y x X combinations",
       subtitle = "Green = positive (sign-flip); Red = negative. Significance: *** p<0.001, ** p<0.01, * p<0.05, \u2020 p<0.1",
       x = "Deprivation measure (X variable)",
       y = "Disadvantaged attainment measure (Y variable)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

```{r directionality-grid-narrative}
#| label: directionality-grid-narrative
#| output: asis

n_pos_grid <- sum(grid_results$coef > 0, na.rm = TRUE)
n_total_grid <- sum(!is.na(grid_results$coef))
n_sig_pos <- sum(grid_results$coef > 0 & grid_results$p < 0.05, na.rm = TRUE)
n_sig_neg <- sum(grid_results$coef < 0 & grid_results$p < 0.05, na.rm = TRUE)
n_marginal_pos <- sum(grid_results$coef > 0 & grid_results$p >= 0.05 & grid_results$p < 0.1, na.rm = TRUE)
n_nonsig <- sum(grid_results$p >= 0.05, na.rm = TRUE)

cat(sprintf("\nAcross the **%d-cell grid** of Y × X combinations, **%d (%d%%)** show a positive coefficient on the deprivation concentration measure.\n\n",
    n_total_grid, n_pos_grid, round(100 * n_pos_grid / n_total_grid)))

cat("### Statistical significance breakdown\n\n")
cat(sprintf("| Category | Count | %% of grid |\n|----------|------:|----------:|\n"))
cat(sprintf("| Positive & significant (p < 0.05) | %d | %d%% |\n", n_sig_pos, round(100 * n_sig_pos / n_total_grid)))
cat(sprintf("| Positive & marginal (p < 0.1) | %d | %d%% |\n", n_marginal_pos, round(100 * n_marginal_pos / n_total_grid)))
cat(sprintf("| Positive & non-significant | %d | %d%% |\n", n_pos_grid - n_sig_pos - n_marginal_pos, round(100 * (n_pos_grid - n_sig_pos - n_marginal_pos) / n_total_grid)))
cat(sprintf("| Negative & significant (p < 0.05) | %d | %d%% |\n", n_sig_neg, round(100 * n_sig_neg / n_total_grid)))
cat(sprintf("| Negative & non-significant | %d | %d%% |\n\n", n_total_grid - n_pos_grid - n_sig_neg, round(100 * (n_total_grid - n_pos_grid - n_sig_neg) / n_total_grid)))

# Analyse patterns by X variable
cat("### Patterns by deprivation measure\n\n")
for (xlab in unique(grid_results$x_label)) {
  x_sub <- grid_results %>% filter(x_label == xlab)
  n_pos_x <- sum(x_sub$coef > 0, na.rm = TRUE)
  n_sig_x <- sum(x_sub$coef > 0 & x_sub$p < 0.05, na.rm = TRUE)
  median_t <- median(x_sub$t, na.rm = TRUE)
  cat(sprintf("- **%s**: %d/%d positive (of which %d significant); median t = %+.1f\n",
      xlab, n_pos_x, nrow(x_sub), n_sig_x, median_t))
}

cat("\n### Patterns by attainment measure\n\n")
for (ylab in unique(grid_results$y_label)) {
  y_sub <- grid_results %>% filter(y_label == ylab)
  n_pos_y <- sum(y_sub$coef > 0, na.rm = TRUE)
  n_sig_y <- sum(y_sub$coef > 0 & y_sub$p < 0.05, na.rm = TRUE)
  median_t <- median(y_sub$t, na.rm = TRUE)
  cat(sprintf("- **%s**: %d/%d positive (of which %d significant); median t = %+.1f\n",
      ylab, n_pos_y, nrow(y_sub), n_sig_y, median_t))
}

cat("\n## Overall assessment\n\n")

# More nuanced assessment based on both direction and significance
pct_pos <- n_pos_grid / n_total_grid
pct_sig_pos <- n_sig_pos / n_total_grid

if (pct_pos >= 0.75 && pct_sig_pos >= 0.4) {
  cat("::: {.callout-important}\n")
  cat("## The sign-flip is robust and statistically significant\n\n")
  cat(sprintf("The positive coefficient appears in %d%% of combinations, and is statistically significant (p < 0.05) in %d%% of them. ", round(100 * pct_pos), round(100 * pct_sig_pos)))
  cat("This provides strong evidence that the relationship between disadvantage concentration and disadvantaged attainment is genuine and not an artefact of a specific measure. The effect is consistent across different ATT8 subject components and across conceptually distinct deprivation measures, including those with low correlations to the model's prior attainment and absence controls.\n\n")
  cat("The combination of directional consistency and statistical significance across diverse operationalisations makes this one of the more robust findings in the model experiments. It is consistent with mechanisms such as Pupil Premium resource concentration, institutional focus on disadvantaged cohorts, and reduced within-school stigmatisation of disadvantaged status.\n")
  cat(":::\n")
} else if (pct_pos >= 0.75 && pct_sig_pos < 0.4) {
  cat("::: {.callout-note}\n")
  cat("## The sign-flip is directionally robust but often not statistically significant\n\n")
  cat(sprintf("The positive coefficient appears in %d%% of combinations — a clear directional pattern — but is statistically significant (p < 0.05) in only %d%% of them. ", round(100 * pct_pos), round(100 * pct_sig_pos)))
  cat("This suggests a real but **small** effect: the sign-flip is not a chance finding (its consistency across measures is hard to attribute to noise), but its magnitude is modest enough that it does not always clear conventional significance thresholds.\n\n")
  cat("For policy purposes, the directional consistency matters more than the p-values in individual cells. The finding that higher concentrations of disadvantage are associated with *slightly better* (or at worst, no worse) outcomes for disadvantaged pupils — once prior attainment and absence are controlled — holds across multiple operationalisations. What varies is the precision of the estimate, not its direction.\n\n")
  cat("However, the modest significance should temper the strength of causal claims. The effect is likely real, but it is small relative to the dominant predictors in the model (absence, prior attainment), and could be confounded by unmeasured school-level factors correlated with intake composition.\n")
  cat(":::\n")
} else if (pct_pos >= 0.4) {
  cat("::: {.callout-note}\n")
  cat("## The sign-flip is partially robust\n\n")
  cat(sprintf("The positive coefficient appears in %d%% of combinations, with %d%% reaching statistical significance. ", round(100 * pct_pos), round(100 * pct_sig_pos)))
  cat("The sign-flip is not universal — it tends to be strongest with FSM-based deprivation measures (which share the most variance with the model's controls) and weaker or reversed with more conceptually distinct measures.\n\n")
  cat("This pattern is consistent with two interpretations: (a) the effect is genuine but specific to economic deprivation (FSM-type measures), not disadvantage in general; or (b) the sign-flip partly depends on the model's ability to remove indirect pathways through prior attainment and absence, and this works better for some X variables than others.\n\n")
  cat("The practical implication remains: there is no evidence that concentrating disadvantaged pupils in certain schools *harms* their attainment, and suggestive evidence that it may modestly *help* — but this finding is sensitive to how both disadvantage and attainment are measured.\n")
  cat(":::\n")
} else {
  cat("::: {.callout-warning}\n")
  cat("## The sign-flip may be fragile\n\n")
  cat(sprintf("The positive coefficient appears in only %d%% of combinations (%d%% significant). ", round(100 * pct_pos), round(100 * pct_sig_pos)))
  cat("The sign-flip does not generalise well beyond the original FSM6CLA1A × ATT8 combination used in Analysis E. Alternative measures of either disadvantage concentration or disadvantaged attainment tend to produce negative or non-significant coefficients.\n\n")
  cat("This does not necessarily mean the original finding is wrong — it may reflect something specific to how FSM6CLA1A interacts with the model's controls — but it does caution against strong claims that disadvantage concentration benefits disadvantaged pupils as a general principle. Individual pupil-level data would be needed to make further progress.\n")
  cat(":::\n")
}
```


# Two-Stage Absence Decomposition: Separating Exogenous from School-Controllable Absence {#sec-two-stage-absence}

## Motivation

The headline Analysis E model treats `log(PERCTOT)` (% absence) as a regressor on the same footing as intake variables like FSM, EAL and prior attainment. Conceptually this is awkward, because absence sits in two roles at once:

- **Partly exogenous to the school** --- driven by disadvantage, family circumstances, local health, SEN status, EAL, and the broader area context the school inherits.
- **Partly an outcome the school helps produce** --- through ethos, pastoral systems, follow-up on persistent absentees, attendance officers, parental engagement, and so on.

If we **control for raw absence**, we strip both components from the value-added residual. Schools that lift attendance through deliberate effort get no credit for it --- their work is absorbed into the absence coefficient and removed from what remains in the residual.

If we **don't control for absence at all**, we treat all attendance variation as the school's doing, and over-penalise schools whose intake brings high absence for reasons they cannot influence.

What we want for an honest value-added measure is something closer to the *total school effect* --- direct effect on attainment plus indirect effect via attendance --- conditioned on the parts of absence that are genuinely outside the school's control.

This section implements the **two-stage decomposition** that delivers that.

::: {.callout-note}
## The decomposition in brief

1. **Stage 1.** Regress school-year-level absence on **exogenous intake variables only** (FSM, EAL, prior attainment, area-level structural factors). Save the fitted values --- *expected absence* --- and the residuals --- *residual absence*, which captures the school-controllable component plus noise.

2. **Stage 2.** Predict attainment from intake variables and **expected absence** (not raw absence), and **drop** workforce variables (which are also school-controllable). Residual absence is *not* a regressor: its variance flows into the attainment residual, where it can show up as part of the value-added signal.

The school-level random effect from stage 2 is then a more honest value-added measure: it reflects what the school adds over and above what its intake (including the disadvantage-driven portion of absence) would predict, while still giving credit for the school's contribution to attendance and workforce stability.
:::

A practical caveat: **if disadvantage is essentially the only exogenous predictor of absence, expected absence becomes collinear with disadvantage and the two-stage approach collapses to "just don't control for absence"**. The decomposition only buys something when stage 1 can put the absence variable on a different footing from the variables already entering stage 2 directly. We rely on the **LA-level and region-level random effects** in stage 1 to soak up area health, deprivation and demographic structure that the attainment model otherwise treats only via fixed effects on FSM/EAL/prior attainment.


## Stage 1: Modelling expected absence

We regress `log(PERCTOT)` on the intake variables we treat as exogenous to the school's own behaviour, plus year and place random effects:

```{r stage1-fit}
#| label: stage1-fit

# Stage 1 dataset: any school-year with absence + the predictors
stage1_data <- imputed_full_data %>%
  filter(!is.na(PERCTOT), PERCTOT > 0,
         !is.na(PTFSM6CLA1A), PTFSM6CLA1A > 0,
         !is.na(PNUMEAL), PNUMEAL > 0,
         !is.na(PTPRIORLO),
         !is.na(gorard_segregation)) %>%
  droplevels()

cat("Stage 1 observations:", nrow(stage1_data), "\n\n")

mod_stage1 <- lmer(
  log(PERCTOT) ~
    log(PTFSM6CLA1A) + log(PNUMEAL) +
    PTPRIORLO + gorard_segregation +
    (1 | year_label) + (1 | gor_name/LANAME),
  data = stage1_data,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_stage1)
```

```{r stage1-checks}
#| label: stage1-checks

cat("Stage 1 R²:\n")
print(r2(mod_stage1))
cat("\nVariance components:\n")
print(VarCorr(mod_stage1))
```

The marginal R² tells us how much absence is explained by intake variables alone (fixed effects); the conditional R² adds the place + year random effects --- i.e. the share of absence variance that can be attributed to factors **outside the school's direct day-to-day control**. The gap between the two is the share of "expected absence" coming through area context rather than measured intake.

```{r stage1-decompose}
#| label: stage1-decompose

# Compute expected absence (fitted) and residual absence (residuals on the log scale)
stage1_data <- stage1_data %>%
  mutate(
    log_expected_absence  = predict(mod_stage1, newdata = ., re.form = NULL,
                                    allow.new.levels = TRUE),
    expected_absence      = exp(log_expected_absence),
    residual_absence_log  = log(PERCTOT) - log_expected_absence
  )

# How does observed absence split between expected and residual?
exp_summary <- stage1_data %>%
  summarise(
    var_log_obs       = var(log(PERCTOT)),
    var_log_expected  = var(log_expected_absence),
    var_log_residual  = var(residual_absence_log),
    cor_obs_expected  = cor(log(PERCTOT), log_expected_absence)
  )
print(exp_summary)

cat(sprintf(
  "\nApprox. share of log-absence variance attributed to exogenous (expected) component: %.1f%%\n",
  100 * exp_summary$var_log_expected /
        (exp_summary$var_log_expected + exp_summary$var_log_residual)
))
```


## Stage 2: Three attainment specifications

We now refit Analysis E's *all-pupils* attainment model under three specifications on the **same dataset**, so that residuals and rankings are directly comparable:

- **M0 --- No absence control, no workforce.** Drops `log(PERCTOT)` and the three workforce variables entirely. Treats both attendance and workforce stability as part of the school's value-added.
- **M1 --- Raw absence + workforce (= Analysis E).** The current production specification.
- **M2 --- Expected absence, no workforce.** Replaces `log(PERCTOT)` with `log(expected_absence)` from stage 1 and drops the workforce predictors. Residual absence and workforce variation flow into the value-added residual.

```{r stage2-data}
#| label: stage2-data

# Join expected absence back onto the full attainment dataset
stage2_data <- imputed_full_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  inner_join(
    stage1_data %>%
      select(URN, year_label,
             expected_absence, log_expected_absence, residual_absence_log),
    by = c("URN", "year_label")
  ) %>%
  droplevels()

contrasts(stage2_data$OFSTEDRATING_1) <- contr.treatment(levels(stage2_data$OFSTEDRATING_1))

cat("Stage 2 observations:", nrow(stage2_data), "\n")
```

```{r stage2-fits}
#| label: stage2-fits

# M0: no absence, no workforce
mod_M0 <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = stage2_data,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# M1: raw absence + workforce (Analysis E spec, refit on stage2_data)
mod_M1 <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = stage2_data,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

# M2: expected absence, no workforce
mod_M2 <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log_expected_absence + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = stage2_data,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)
```

```{r stage2-fit-summary}
#| label: stage2-fit-summary

mod_list <- list(M0 = mod_M0, M1 = mod_M1, M2 = mod_M2)

fit_summary <- map_dfr(names(mod_list), function(nm) {
  m <- mod_list[[nm]]
  r <- performance::r2(m)
  tibble(
    model         = nm,
    description   = c(M0 = "No absence, no workforce",
                       M1 = "Raw absence + workforce (Analysis E)",
                       M2 = "Expected absence, no workforce")[nm],
    n_obs         = nobs(m),
    n_fixed       = length(fixef(m)),
    R2_marginal   = as.numeric(r$R2_marginal),
    R2_conditional = as.numeric(r$R2_conditional),
    AIC           = AIC(m),
    BIC           = BIC(m)
  )
})

knitr::kable(fit_summary, digits = 3,
             caption = "Fit statistics across the three Stage 2 specifications")
```


## Coefficient comparison

```{r stage2-coefs}
#| label: stage2-coefs

extract_fixef_table <- function(m, label) {
  s <- summary(m)$coefficients
  tibble(
    term = rownames(s),
    estimate = s[, "Estimate"],
    std_error = s[, "Std. Error"],
    p_value = s[, "Pr(>|t|)"],
    model = label
  )
}

coef_long <- bind_rows(
  extract_fixef_table(mod_M0, "M0"),
  extract_fixef_table(mod_M1, "M1"),
  extract_fixef_table(mod_M2, "M2")
)

coef_wide <- coef_long %>%
  mutate(value = sprintf("%.3f%s",
                         estimate,
                         case_when(p_value < 0.001 ~ "***",
                                   p_value < 0.01  ~ "**",
                                   p_value < 0.05  ~ "*",
                                   TRUE            ~ ""))) %>%
  select(term, model, value) %>%
  pivot_wider(names_from = model, values_from = value, values_fill = "—")

knitr::kable(coef_wide,
             caption = "Fixed-effects coefficients across specifications (* p<.05, ** p<.01, *** p<.001)")
```

::: {.callout-note}
## Reading the coefficient table

There are two diagnostic comparisons worth making.

**1. M0 vs (M1, M2) on FSM.** M0 has no absence control, so the FSM coefficient there absorbs both the *direct* disadvantage → attainment channel and the *indirect* disadvantage → absence → attainment channel. Adding any absence variable --- raw (M1) or expected (M2) --- should shrink FSM substantially. The size of that shrinkage is a rough indicator of how much of FSM's apparent influence on attainment runs *through* attendance rather than directly. In our results, the FSM elasticity is roughly **-0.115 in M0** and **-0.067 / -0.069 in M1 / M2** --- the absence channel accounts for almost half of FSM's "raw" association with attainment.

**2. M1 vs M2 on FSM.** This is where the two-stage decomposition either earns its keep or doesn't. If raw absence and expected absence behave very differently in the attainment model, then the *school-controllable* component of absence (the part removed in M1 but left in the residual in M2) carries unique predictive power, and the decomposition is doing real work. If M1 and M2 produce near-identical FSM coefficients (as they do here, -0.067 vs -0.069), then within this fixed-effects view of attainment, **raw and expected absence are near-substitutes** --- residual absence carries little marginal predictive power beyond what expected absence already captures.

This second finding is the more interesting one. It does *not* mean the decomposition is pointless --- it means its main effect lands on the **value-added residual at the school level** (where residual absence flows into M2's residual but not M1's), not on the population-level coefficients. The downstream school-by-school comparisons below are where the decomposition pays off, not the fixed-effects table above.

A third comparison --- the coefficient on `log(PERCTOT)` in M1 vs the coefficient on `log_expected_absence` in M2 --- tells us whether the elasticity-on-absence reading we have been using elsewhere in the analysis is robust to the decomposition. If they are close, the "1pp absence reduction → ATT8 gain" story holds whether you take "absence" to mean raw absence or the intake-predicted component.
:::


## School-level value-added under each specification

For each model we extract the school-level random effects (the `LANAME:gor_name`-nested intercepts pull out *LA × region* effects, but the school's own value-added shows up via the fitted-vs-observed residual at the school-year level). We aggregate to school level by taking each school's mean residual on the log scale across the years it appears.

```{r school-vadd}
#| label: school-vadd

# Per-school average of the school-year residual on the log scale,
# converted to an approximate "GCSE-points value-added" via the school's mean ATT8.
school_value_added <- function(m, label) {
  d_used <- model.frame(m)
  d_used$log_resid <- residuals(m)
  # We need URN/SCHNAME/LANAME — pull them from stage2_data via row index
  d_used <- bind_cols(
    stage2_data %>%
      slice(as.integer(rownames(d_used))) %>%
      select(URN, SCHNAME, LANAME, gor_name, ATT8SCR),
    d_used %>% select(log_resid)
  )
  d_used %>%
    group_by(URN, SCHNAME, LANAME, gor_name) %>%
    summarise(
      mean_log_resid = mean(log_resid, na.rm = TRUE),
      mean_att8      = mean(ATT8SCR, na.rm = TRUE),
      n_years        = n(),
      .groups = "drop"
    ) %>%
    mutate(
      vadd_points = mean_att8 * (exp(mean_log_resid) - 1),
      model = label
    )
}

va_M0 <- school_value_added(mod_M0, "M0")
va_M1 <- school_value_added(mod_M1, "M1")
va_M2 <- school_value_added(mod_M2, "M2")

va_all <- bind_rows(va_M0, va_M1, va_M2)

va_wide <- va_all %>%
  select(URN, SCHNAME, LANAME, model, vadd_points) %>%
  pivot_wider(names_from = model, values_from = vadd_points,
              names_prefix = "vadd_") %>%
  mutate(
    rank_M0 = rank(-vadd_M0, na.last = "keep"),
    rank_M1 = rank(-vadd_M1, na.last = "keep"),
    rank_M2 = rank(-vadd_M2, na.last = "keep"),
    n_schools = sum(!is.na(vadd_M1))
  )

cat("Schools with value-added under all three models:", sum(complete.cases(va_wide[, c("vadd_M0","vadd_M1","vadd_M2")])), "\n")
```

```{r vadd-correlations}
#| label: vadd-correlations

va_complete <- va_wide %>%
  filter(!is.na(vadd_M0), !is.na(vadd_M1), !is.na(vadd_M2))

cor_table <- tibble(
  pair = c("M0 vs M1", "M0 vs M2", "M1 vs M2"),
  pearson_vadd = c(
    cor(va_complete$vadd_M0, va_complete$vadd_M1),
    cor(va_complete$vadd_M0, va_complete$vadd_M2),
    cor(va_complete$vadd_M1, va_complete$vadd_M2)
  ),
  spearman_rank = c(
    cor(va_complete$rank_M0, va_complete$rank_M1, method = "spearman"),
    cor(va_complete$rank_M0, va_complete$rank_M2, method = "spearman"),
    cor(va_complete$rank_M1, va_complete$rank_M2, method = "spearman")
  )
)

knitr::kable(cor_table, digits = 3,
             caption = "Cross-specification correlations of school value-added and rank")
```

If the rank correlation between **M1 and M2** is very high (say > 0.95), the absence-vs-expected-absence choice barely moves the rankings overall --- the two specifications largely agree about who is doing well. The interesting cases sit where they disagree.


## Where the specifications disagree most

```{r vadd-shifts}
#| label: vadd-shifts

shifts <- va_complete %>%
  mutate(
    rank_shift_M2_vs_M1 = rank_M1 - rank_M2,    # positive = M2 ranks them better
    rank_shift_M0_vs_M1 = rank_M1 - rank_M0
  ) %>%
  arrange(desc(abs(rank_shift_M2_vs_M1)))

cat("Top 15 schools where M2 (expected absence) ranks the school higher than M1:\n")
shifts %>%
  filter(rank_shift_M2_vs_M1 > 0) %>%
  slice_head(n = 15) %>%
  select(SCHNAME, LANAME, rank_M1, rank_M2,
         rank_shift_M2_vs_M1, vadd_M1, vadd_M2) %>%
  knitr::kable(digits = 2)

cat("\nTop 15 schools where M2 (expected absence) ranks the school *lower* than M1:\n")
shifts %>%
  filter(rank_shift_M2_vs_M1 < 0) %>%
  arrange(rank_shift_M2_vs_M1) %>%
  slice_head(n = 15) %>%
  select(SCHNAME, LANAME, rank_M1, rank_M2,
         rank_shift_M2_vs_M1, vadd_M1, vadd_M2) %>%
  knitr::kable(digits = 2)
```

The pattern that should emerge: **schools that work hard to lift attendance** (low residual absence given their intake) rank higher under M2 than under M1, because M2 returns the credit for that work to the school's residual. **Schools coasting on a low-absence intake** see no such uplift --- their attendance is already explained by who walks through the door.


## Brighton & Hove: school-by-school under all three specifications

```{r bh-vadd}
#| label: bh-vadd

bh_va <- va_complete %>%
  filter(grepl("Brighton", LANAME)) %>%
  arrange(desc(vadd_M1))

# Pull each school's residual absence (school mean, log-scale, then converted to %-pts)
bh_resid_abs <- stage2_data %>%
  filter(grepl("Brighton", LANAME)) %>%
  group_by(URN, SCHNAME) %>%
  summarise(
    mean_obs_abs       = mean(PERCTOT, na.rm = TRUE),
    mean_expected_abs  = mean(expected_absence, na.rm = TRUE),
    mean_resid_abs_log = mean(residual_absence_log, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(resid_abs_pp = mean_obs_abs - mean_expected_abs)

bh_combined <- bh_va %>%
  left_join(bh_resid_abs, by = c("URN", "SCHNAME")) %>%
  select(SCHNAME,
         vadd_M0, vadd_M1, vadd_M2,
         mean_obs_abs, mean_expected_abs, resid_abs_pp)

knitr::kable(bh_combined, digits = 2,
             caption = "Brighton & Hove schools: value-added (ATT8 points) under each specification, plus the residual-absence indicator")
```

The `resid_abs_pp` column is the **school-controllable absence indicator** described in the motivation: positive values mean the school has more absence than the stage-1 model would expect from its intake; negative values mean the school is doing better on attendance than its intake would predict.

The combination of **`vadd_M2` plus `resid_abs_pp`** is the transparent two-component view: how much the school adds to attainment (with attendance management already credited), and *separately* how much of that comes through unusually good (or bad) attendance management.


## Sensitivity: bootstrap stability of B&H rankings

We resample LAs-with-replacement (clustered bootstrap, since school-years within an LA are not independent) and refit stages 1 and 2. For each replicate we record each B&H school's value-added rank under M1 and M2.

```{r boot-setup}
#| label: boot-setup

set.seed(42)
n_boot <- 50  # keep small to bound render time; bump for production use

la_pool <- unique(stage2_data$LANAME)

boot_rank <- function() {
  # Sample LAs with replacement
  la_sample <- sample(la_pool, length(la_pool), replace = TRUE)
  d_boot <- map_dfr(seq_along(la_sample), function(i) {
    stage2_data %>% filter(LANAME == la_sample[i]) %>% mutate(.boot_la = i)
  })
  if (nrow(d_boot) < 1000) return(NULL)

  # Refit stage 1 on the bootstrap sample
  s1_b <- try(lmer(
    log(PERCTOT) ~ log(PTFSM6CLA1A) + log(PNUMEAL) + PTPRIORLO +
      gorard_segregation +
      (1 | year_label) + (1 | gor_name),
    data = d_boot, REML = TRUE,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  ), silent = TRUE)
  if (inherits(s1_b, "try-error")) return(NULL)

  d_boot$log_expected_absence <- predict(s1_b, newdata = d_boot,
                                          re.form = NULL,
                                          allow.new.levels = TRUE)

  # Refit M1 and M2
  m1_b <- try(lmer(
    log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      remained_in_the_same_school +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | year_label) + (1 | gor_name),
    data = d_boot, REML = TRUE,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  ), silent = TRUE)

  m2_b <- try(lmer(
    log(ATT8SCR) ~ log(PTFSM6CLA1A) + log_expected_absence + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      (1 | year_label) + (1 | gor_name),
    data = d_boot, REML = TRUE,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  ), silent = TRUE)

  if (inherits(m1_b, "try-error") || inherits(m2_b, "try-error")) return(NULL)

  d_boot$resid_M1 <- residuals(m1_b)
  d_boot$resid_M2 <- residuals(m2_b)

  d_boot %>%
    group_by(URN, SCHNAME, LANAME) %>%
    summarise(va_M1 = mean(resid_M1), va_M2 = mean(resid_M2), .groups = "drop") %>%
    mutate(rank_M1 = rank(-va_M1), rank_M2 = rank(-va_M2))
}

boot_results <- map_dfr(1:n_boot, function(i) {
  res <- boot_rank()
  if (!is.null(res)) res$rep <- i
  res
})

cat("Successful bootstrap replicates:", length(unique(boot_results$rep)), "/", n_boot, "\n")
```

```{r boot-summary}
#| label: boot-summary

bh_boot <- boot_results %>%
  filter(grepl("Brighton", LANAME)) %>%
  group_by(SCHNAME) %>%
  summarise(
    rank_M1_median = median(rank_M1),
    rank_M1_q05    = quantile(rank_M1, 0.05),
    rank_M1_q95    = quantile(rank_M1, 0.95),
    rank_M2_median = median(rank_M2),
    rank_M2_q05    = quantile(rank_M2, 0.05),
    rank_M2_q95    = quantile(rank_M2, 0.95),
    .groups = "drop"
  ) %>%
  arrange(rank_M2_median)

knitr::kable(bh_boot, digits = 0,
             caption = "Brighton & Hove school rank uncertainty under M1 and M2 (clustered bootstrap, n_boot replicates above)")
```

The width of the 5--95% interval gives an honest sense of how much each school's position can move purely through sampling. Schools whose interval brackets stay tight under both M1 and M2 are *robustly* ranked; wide intervals signal cases where the methodological choice or a small change in sample matters.


## Take-aways

::: {.callout-tip}
## What this experiment tells us

1. **Most rankings are likely robust to the absence specification.** If M1 and M2 produce a Spearman rank correlation above ~0.9, then for parental-choice purposes the simpler Analysis E specification is fine.
2. **Where they disagree, the disagreement is informative.** A school whose M2 rank is materially better than its M1 rank is plausibly *managing attendance unusually well* given its intake; the reverse pattern flags a school whose M1 advantage rests on an attendance profile its intake would have predicted anyway.
3. **The combined indicator (value-added + residual absence) is the most transparent reporting unit.** It separates "this school adds points to attainment" from "this school's pupils attend more than their intake predicts" --- two distinct signals that the single-residual M1 reporting fuses together.
4. **Stage 1 is the binding constraint.** With only FSM, EAL, low prior attainment and segregation as fixed exogenous predictors, stage 1's discriminating power leans heavily on the LA random effect. Adding richer area-level controls (IDACI, area health indicators, SEN/EHCP rates, neighbourhood deprivation deciles) would sharpen the decomposition substantially.
5. **Bootstrap confidence intervals are wide for many schools.** A single point estimate of value-added is almost always more precise than the underlying data warrants. Reporting the interval alongside the point estimate is honest practice.
:::

::: {.callout-warning}
## What this experiment does not do

- **It does not establish causality.** All three specifications are observational. The two-stage approach reduces one specific bias (over-attributing exogenous absence to school behaviour) but does not eliminate omitted-variable bias from unmeasured intake characteristics.
- **It does not address measurement error in stage 1.** Expected absence enters stage 2 as if it were known with certainty. Properly accounting for the generated-regressor problem requires either a full bootstrap of standard errors (the small simulation above is a start) or an instrumental-variable / structural-equation framing.
- **It does not test pupil-level decompositions.** Pupil-level NPD data with linked attendance, intake and outcomes would allow a much sharper version of this analysis, including who within each school benefits from attendance management.
:::


## Direct test: expected vs residual absence in the same model

The M1↔M2 FSM equality is partly mechanical (stage 1 residuals are orthogonal to stage 1 regressors by construction), so the cleanest test of whether the school-controllable absence component matters at population level is to put **both components into the same attainment model** and read off their coefficients side by side.

```{r m-combined}
#| label: m-combined

mod_M_combined <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log_expected_absence + residual_absence_log +
    log(PNUMEAL) + PTPRIORLO + ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = stage2_data,
  REML = TRUE,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_M_combined)
```

```{r m-combined-compare}
#| label: m-combined-compare

# Pull the absence-related coefficients into a single table
absence_coef_table <- bind_rows(
  extract_fixef_table(mod_M1, "M1: raw absence")          %>%
    filter(term == "log(PERCTOT)") %>% mutate(component = "raw absence"),
  extract_fixef_table(mod_M2, "M2: expected absence")     %>%
    filter(term == "log_expected_absence") %>% mutate(component = "expected absence"),
  extract_fixef_table(mod_M_combined, "M_combined: both") %>%
    filter(term %in% c("log_expected_absence", "residual_absence_log")) %>%
    mutate(component = ifelse(term == "log_expected_absence",
                               "expected absence", "residual absence"))
) %>%
  select(model, component, estimate, std_error, p_value)

knitr::kable(absence_coef_table, digits = 4,
             caption = "Absence-coefficient comparison across M1, M2 and the combined model. The M_combined row for residual absence is the direct test of whether school-controllable absence carries marginal predictive power.")
```

```{r m-combined-r2}
#| label: m-combined-r2

cat("Model fit comparison:\n")
fit_combined <- tibble(
  model         = c("M1", "M2", "M_combined"),
  n_obs         = c(nobs(mod_M1), nobs(mod_M2), nobs(mod_M_combined)),
  R2_marginal   = c(as.numeric(performance::r2(mod_M1)$R2_marginal),
                     as.numeric(performance::r2(mod_M2)$R2_marginal),
                     as.numeric(performance::r2(mod_M_combined)$R2_marginal)),
  R2_conditional = c(as.numeric(performance::r2(mod_M1)$R2_conditional),
                      as.numeric(performance::r2(mod_M2)$R2_conditional),
                      as.numeric(performance::r2(mod_M_combined)$R2_conditional)),
  AIC           = c(AIC(mod_M1), AIC(mod_M2), AIC(mod_M_combined))
)
knitr::kable(fit_combined, digits = 4,
             caption = "Fit statistics: M_combined vs M1 vs M2")
```

::: {.callout-note}
## How to read the combined-model output

- **If the residual-absence coefficient is small relative to the expected-absence coefficient** (and/or non-significant), the school-controllable component of absence carries little marginal predictive power for attainment at the population level after intake is accounted for. Read this as: at national average, the bulk of the absence → attainment association runs through structural/intake factors that schools don't directly control.
- **If the two coefficients are similar in magnitude**, both components matter independently --- intake-driven absence and school-management absence have comparable per-percentage-point effects on attainment.
- **If residual absence has a larger coefficient than expected absence**, then conditional on intake, what schools *do* about attendance matters more than the intake-driven baseline. This would be a strong finding for actionable policy.

Note that the residual-absence coefficient here is the **partial** effect: the marginal contribution of school-controllable absence after intake-driven absence is already in the model. It is not the same as the unconditional correlation between residual absence and attainment.
:::


## Brighton & Hove combined view: value-added + residual absence

The local question is sharper than the national one. Even if at the population level residual absence carries modest marginal effect, individual B&H schools can sit at very different points on the residual-absence distribution, and those differences are the actionable signal for local policy.

```{r bh-combined-view}
#| label: bh-combined-view

# Re-pull B&H school-level summaries with both signals on one row
bh_local_view <- va_complete %>%
  filter(grepl("Brighton", LANAME)) %>%
  left_join(
    stage2_data %>%
      filter(grepl("Brighton", LANAME)) %>%
      group_by(URN, SCHNAME) %>%
      summarise(
        mean_obs_abs       = mean(PERCTOT, na.rm = TRUE),
        mean_expected_abs  = mean(expected_absence, na.rm = TRUE),
        n_years_obs        = n(),
        .groups = "drop"
      ) %>%
      mutate(
        resid_abs_pp = mean_obs_abs - mean_expected_abs,
        # Approximate ATT8 implication of that residual absence under the
        # M2 expected-absence elasticity (illustrative only)
        absence_implied_att8_swing = NA_real_  # placeholder; populated below
      ),
    by = c("URN", "SCHNAME")
  )

# Use the M2 expected-absence elasticity to translate residual absence into
# an indicative attainment-points equivalent, IF residual absence behaved like
# expected absence (this is illustrative — see caveat above).
beta_exp_abs <- fixef(mod_M2)["log_expected_absence"]

bh_local_view <- bh_local_view %>%
  mutate(
    # mean_att8 from earlier work; recompute robustly here
    mean_att8 = NA_real_
  )

bh_att8_means <- stage2_data %>%
  filter(grepl("Brighton", LANAME)) %>%
  group_by(URN) %>%
  summarise(mean_att8 = mean(ATT8SCR, na.rm = TRUE), .groups = "drop")

bh_local_view <- bh_local_view %>%
  select(-mean_att8) %>%
  left_join(bh_att8_means, by = "URN") %>%
  mutate(
    # log-scale change in absence implied by residual_abs_pp at school's mean
    delta_log_abs = log(mean_obs_abs) - log(mean_expected_abs),
    # Indicative ATT8 points equivalent if M2 elasticity applied
    absence_implied_att8_swing = mean_att8 *
      (exp(beta_exp_abs * delta_log_abs) - 1)
  ) %>%
  arrange(desc(vadd_M2))

bh_summary_table <- bh_local_view %>%
  select(SCHNAME,
         vadd_M1, vadd_M2,
         mean_obs_abs, mean_expected_abs, resid_abs_pp,
         absence_implied_att8_swing)

knitr::kable(bh_summary_table, digits = 2,
             caption = "Brighton & Hove combined view: model value-added (M1 vs M2) alongside residual absence (pp). The 'absence_implied_att8_swing' column gives an indicative translation of residual absence into ATT8 points under the M2 expected-absence elasticity --- illustrative only, since residual absence's true elasticity is what the combined model above estimates.")
```

::: {.callout-tip}
## Reading this table for B&H

For each school, three signals sit side by side:

1. **`vadd_M2`** --- attainment value-added under the expected-absence specification (the "school does more than its intake would predict" indicator, with attendance management credited).
2. **`resid_abs_pp`** --- the school-controllable absence indicator. Positive = more absence than intake predicts; negative = less.
3. **`absence_implied_att8_swing`** --- the indicative ATT8 implication of that residual absence, *if* residual absence were translated through the same elasticity as expected absence. This is illustrative; the actual residual-absence elasticity is the one in the combined-model table above.

Schools where `vadd_M2` is high *and* `resid_abs_pp` is negative are doing well on attainment partly *through* unusually good attendance --- a coherent management story. Schools where `vadd_M2` is high but `resid_abs_pp` is positive are adding value *despite* a worse-than-expected absence profile --- pedagogically impressive, with attendance still the obvious lever to pull. Schools where `vadd_M2` is negative *and* `resid_abs_pp` is positive have attendance as the most plausible single point of intervention.
:::


## Worked examples: reading the B&H combined view school-by-school

To make the combined view concrete, here are worked readings for the city's secondaries, grouped to highlight the different combinations of value-added and attendance management. Numbers below are pulled programmatically from `bh_local_view` so they stay in sync with the model fit.

```{r bh-worked-examples}
#| label: bh-worked-examples
#| output: asis

# Helper: pull a single school's row by partial name match, return a list
get_school <- function(pattern) {
  row <- bh_local_view %>% filter(grepl(pattern, SCHNAME, ignore.case = TRUE))
  if (nrow(row) == 0) return(NULL)
  if (nrow(row) > 1) row <- row %>% slice(1)
  list(
    name      = row$SCHNAME,
    vadd_M1   = row$vadd_M1,
    vadd_M2   = row$vadd_M2,
    obs_abs   = row$mean_obs_abs,
    exp_abs   = row$mean_expected_abs,
    resid_pp  = row$resid_abs_pp,
    swing     = row$absence_implied_att8_swing
  )
}

# Helper: format a value-added comment
va_label <- function(v) {
  if (is.na(v)) return("data missing")
  if (v >  1.5) return(sprintf("strongly positive (+%.1f pts)", v))
  if (v >  0.3) return(sprintf("positive (+%.1f pts)",          v))
  if (v > -0.3) return(sprintf("around zero (%.1f pts)",        v))
  if (v > -1.5) return(sprintf("negative (%.1f pts)",           v))
  return(             sprintf("strongly negative (%.1f pts)",   v))
}

# Helper: format an absence comment
abs_label <- function(pp) {
  if (is.na(pp)) return("data missing")
  if (pp >  1.5) return(sprintf("substantially worse than predicted (+%.1f pp)", pp))
  if (pp >  0.4) return(sprintf("worse than predicted (+%.1f pp)",                pp))
  if (pp > -0.4) return(sprintf("close to predicted (%.1f pp)",                   pp))
  if (pp > -1.5) return(sprintf("better than predicted (%.1f pp)",                pp))
  return(             sprintf("substantially better than predicted (%.1f pp)",   pp))
}

cat_school <- function(s, lead = "") {
  if (is.null(s)) return(invisible(NULL))
  cat(sprintf("- **%s** --- value-added %s; absence %s (observed %.1f%%, intake-predicted %.1f%%).%s\n",
              s$name,
              va_label(s$vadd_M2),
              abs_label(s$resid_pp),
              s$obs_abs, s$exp_abs,
              if (nzchar(lead)) paste0(" ", lead) else ""))
}

varndean   <- get_school("Varndean")
kings      <- get_school("^King")
baca       <- get_school("Aldridge.*Brighton|Brighton.*Aldridge|^Brighton Aldridge")
paca       <- get_school("Portslade.*Aldridge|Aldridge.*Portslade")
stringer   <- get_school("Stringer")
cardinal   <- get_school("Cardinal Newman")
blatch     <- get_school("Blatchington")
longhill   <- get_school("Longhill")
hovepark   <- get_school("Hove Park")
patcham    <- get_school("Patcham")

cat("### Two similar profiles: Varndean and King's\n\n")
cat("These two are often spoken about in the same breath in city debates. The combined view shows whether they really do behave alike on both signals.\n\n")
cat_school(varndean)
cat_school(kings)
cat("\nIf both schools land in the same value-added range *and* the same residual-absence range, the popular twinning is empirically reasonable. If one has materially better attendance management than the other given its intake, that is a real difference behind the apparent similarity.\n\n")

cat("### High-FSM city schools: BACA, PACA and the rest of the picture\n\n")
cat("BACA and PACA both serve more disadvantaged catchments than the city average, which is the kind of intake stage 1 is built to expect higher absence from. The decomposition tells us how each school is performing relative to that expectation.\n\n")
cat_school(baca,    "If `vadd_M2` is positive, BACA is adding attainment value over and above what its intake predicts (the headline finding from the main paper). The absence reading then says whether that value-added comes through unusually good attendance management or despite an attendance profile that is harder to shift.")
cat_school(paca,    "PACA's reading sits alongside BACA's --- if both have positive `vadd_M2` the city has *two* high-FSM schools punching above their weight. Differences in `resid_abs_pp` between them say whether they get there through different routes (e.g. one through attendance, one despite it).")
cat_school(stringer,   "Dorothy Stringer is the canonical 'desirable' city school. The interesting question is whether its attainment record sits comfortably *above* its intake-predicted attainment (large positive `vadd_M2`) or close to it (around-zero `vadd_M2`) --- the latter would mean its raw attainment is largely intake-driven rather than school-driven.")
cat_school(cardinal,   "Cardinal Newman has a faith-school admissions profile that often draws a different intake mix from the local catchment. A positive `vadd_M2` paired with negative `resid_abs_pp` would suggest the school's apparent strength rests partly on better attendance management than its intake would predict.")
cat_school(blatch,     "Blatchington Mill is a good test case for the joint signal: a large central catchment school where the question is whether its size and position translate into above-expected outcomes once intake is properly accounted for.")

cat("\n### A high-absence pair: Longhill and Hove Park\n\n")
cat("Both schools have featured in city debates around absence. The combined view tells us whether their high observed absence is mostly intake-driven (and so largely outside the school's day-to-day control) or whether the school is contributing to it.\n\n")
cat_school(longhill)
cat_school(hovepark)
cat("\nFor either school, a positive `resid_abs_pp` says the school is running higher absence than even its (already-disadvantaged) intake would predict --- attendance is then the cleanest single lever for local intervention. A near-zero `resid_abs_pp` says the headline absence number is largely structural and the school is doing about what its intake predicts on attendance.\n\n")

cat("### A negative-negative case: Patcham\n\n")
cat_school(patcham)
cat("\nIf both signals are negative, Patcham's mid-table raw position hides a value-added signal that is *worse* than its intake predicts, and its observed absence is *better* than predicted given intake. The two signals point in different directions: attendance is being managed unusually well, but the attainment outcome is still falling short of what the intake would imply. That is the most awkward combination for a school to face --- it cannot point at attendance as the underlying problem, and the gap shows up most clearly when the value-added is computed under the M2 specification (which gives credit for the better-than-predicted attendance).\n")
```

::: {.callout-tip}
## Why these worked examples matter

The pairings above illustrate why the *joint* reading is more informative than either signal alone:

- **Two schools with the same value-added** can have very different residual-absence stories --- one earning its value-added through attendance management, the other through pedagogy *despite* attendance.
- **Two schools with similar high observed absence** can split between "the absence is what we'd predict from this intake" (structural) and "the absence is worse than the intake predicts" (school-controllable). Only the second is a clean attendance-policy target.
- **A negative-negative case** like Patcham complicates simple reform narratives: a school that handles attendance well but underperforms on attainment cannot be "fixed" through attendance interventions, and the next-most-likely lever (curriculum, pedagogy, leadership) is harder for an LA to act on directly.
:::


## Per-year breakdown: does the decomposition behave differently year-on-year?

The models above pool all four years (2021-22 to 2024-25) and absorb between-year shifts via a year-level random intercept. But absence rose sharply during and after the pandemic, and the relationship between intake and absence may itself have changed across years. This sub-section refits stages 1 and 2 separately within each year and reports how the key quantities move.

```{r per-year-fit}
#| label: per-year-fit

years_to_fit <- levels(stage2_data$year_label)

per_year_results <- map(years_to_fit, function(yr) {

  d_y <- stage2_data %>% filter(year_label == yr) %>% droplevels()
  if (nrow(d_y) < 500) return(NULL)

  # Stage 1: single-year (no year random effect needed)
  s1_y <- try(lmer(
    log(PERCTOT) ~ log(PTFSM6CLA1A) + log(PNUMEAL) + PTPRIORLO +
      gorard_segregation +
      (1 | gor_name/LANAME),
    data = d_y, REML = TRUE,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  ), silent = TRUE)
  if (inherits(s1_y, "try-error")) return(NULL)

  d_y$log_expected_absence <- predict(s1_y, newdata = d_y,
                                       re.form = NULL, allow.new.levels = TRUE)
  d_y$residual_absence_log <- log(d_y$PERCTOT) - d_y$log_expected_absence

  # Stage 2: M1 and M2 (drop year RE since within-year)
  m1_y <- try(lmer(
    log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      remained_in_the_same_school +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d_y, REML = TRUE,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  ), silent = TRUE)

  m2_y <- try(lmer(
    log(ATT8SCR) ~ log(PTFSM6CLA1A) + log_expected_absence + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d_y, REML = TRUE,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  ), silent = TRUE)

  if (inherits(m1_y, "try-error") || inherits(m2_y, "try-error")) return(NULL)

  list(year = yr, data = d_y, s1 = s1_y, M1 = m1_y, M2 = m2_y)
}) %>% setNames(years_to_fit) %>% compact()

cat("Years successfully fit:", paste(names(per_year_results), collapse = ", "), "\n")
```

### Stage 1 coefficients by year

How much do intake → absence relationships shift across years?

```{r per-year-s1-coefs}
#| label: per-year-s1-coefs

s1_year_coefs <- map_dfr(per_year_results, function(yr_res) {
  s <- summary(yr_res$s1)$coefficients
  tibble(
    year = yr_res$year,
    term = rownames(s),
    estimate = s[, "Estimate"],
    std_error = s[, "Std. Error"],
    p_value = s[, "Pr(>|t|)"]
  )
})

s1_year_wide <- s1_year_coefs %>%
  mutate(value = sprintf("%.3f%s",
                         estimate,
                         case_when(p_value < 0.001 ~ "***",
                                   p_value < 0.01  ~ "**",
                                   p_value < 0.05  ~ "*",
                                   TRUE            ~ ""))) %>%
  select(term, year, value) %>%
  pivot_wider(names_from = year, values_from = value, values_fill = "—")

knitr::kable(s1_year_wide,
             caption = "Stage 1 (expected absence) fixed-effects coefficients by year")
```

### Stage 2 attainment coefficients by year — M1 vs M2

```{r per-year-s2-coefs}
#| label: per-year-s2-coefs

s2_year_coefs <- map_dfr(per_year_results, function(yr_res) {
  bind_rows(
    extract_fixef_table(yr_res$M1, "M1") %>% mutate(year = yr_res$year),
    extract_fixef_table(yr_res$M2, "M2") %>% mutate(year = yr_res$year)
  )
})

# Focus on the absence-related coefficient + FSM (the variable most likely to absorb absence's influence)
focus_terms <- c("log(PTFSM6CLA1A)", "log(PERCTOT)", "log_expected_absence",
                 "PTPRIORLO", "log(PNUMEAL)")

s2_focus <- s2_year_coefs %>%
  filter(term %in% focus_terms) %>%
  mutate(value = sprintf("%.3f%s",
                         estimate,
                         case_when(p_value < 0.001 ~ "***",
                                   p_value < 0.01  ~ "**",
                                   p_value < 0.05  ~ "*",
                                   TRUE            ~ ""))) %>%
  select(term, model, year, value) %>%
  pivot_wider(names_from = year, values_from = value, values_fill = "—") %>%
  arrange(term, model)

knitr::kable(s2_focus,
             caption = "Stage 2 attainment-model coefficients (M1 vs M2) on key variables, by year")
```

### Fit comparison by year

```{r per-year-fit-stats}
#| label: per-year-fit-stats

per_year_fit_stats <- map_dfr(per_year_results, function(yr_res) {
  r1 <- performance::r2(yr_res$M1)
  r2 <- performance::r2(yr_res$M2)
  rs <- performance::r2(yr_res$s1)
  tibble(
    year = yr_res$year,
    n_obs = nrow(yr_res$data),
    s1_R2_marginal    = as.numeric(rs$R2_marginal),
    s1_R2_conditional = as.numeric(rs$R2_conditional),
    M1_R2_marginal    = as.numeric(r1$R2_marginal),
    M1_R2_conditional = as.numeric(r1$R2_conditional),
    M2_R2_marginal    = as.numeric(r2$R2_marginal),
    M2_R2_conditional = as.numeric(r2$R2_conditional)
  )
})

knitr::kable(per_year_fit_stats, digits = 3,
             caption = "Per-year R² for stage 1 (absence) and stage 2 (attainment) under M1 and M2")
```

### Brighton & Hove value-added by year, M1 vs M2

```{r per-year-bh}
#| label: per-year-bh

per_year_bh_va <- map_dfr(per_year_results, function(yr_res) {
  d <- yr_res$data
  d$resid_M1 <- residuals(yr_res$M1)
  d$resid_M2 <- residuals(yr_res$M2)
  d %>%
    filter(grepl("Brighton", LANAME)) %>%
    group_by(URN, SCHNAME) %>%
    summarise(
      year = yr_res$year,
      vadd_M1 = mean_att8 <- mean(ATT8SCR, na.rm = TRUE) * (exp(mean(resid_M1)) - 1),
      vadd_M2 = mean(ATT8SCR, na.rm = TRUE) * (exp(mean(resid_M2)) - 1),
      mean_obs_abs      = mean(PERCTOT, na.rm = TRUE),
      mean_expected_abs = mean(exp(d$log_expected_absence[d$URN == first(URN)]), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(resid_abs_pp = mean_obs_abs - mean_expected_abs)
})

bh_va_year_table <- per_year_bh_va %>%
  select(SCHNAME, year, vadd_M1, vadd_M2, resid_abs_pp) %>%
  arrange(SCHNAME, year)

knitr::kable(bh_va_year_table, digits = 2,
             caption = "Brighton & Hove schools: value-added (ATT8 points) and residual absence by year, M1 vs M2")
```

### Year-on-year M1 vs M2 rank correlations

If the rank correlation between M1 and M2 is stable across years, the choice of absence specification is similarly consequential each year. If it drifts, particular years may need different treatment (most likely: the 2021-22 post-pandemic year, when absence was historically anomalous).

```{r per-year-rank-cor}
#| label: per-year-rank-cor

per_year_rank_cor <- map_dfr(per_year_results, function(yr_res) {
  d <- yr_res$data
  d$resid_M1 <- residuals(yr_res$M1)
  d$resid_M2 <- residuals(yr_res$M2)
  va_y <- d %>%
    group_by(URN) %>%
    summarise(va_M1 = mean(resid_M1), va_M2 = mean(resid_M2), .groups = "drop")
  tibble(
    year = yr_res$year,
    n_schools = nrow(va_y),
    pearson  = cor(va_y$va_M1, va_y$va_M2),
    spearman = cor(va_y$va_M1, va_y$va_M2, method = "spearman")
  )
})

knitr::kable(per_year_rank_cor, digits = 3,
             caption = "Year-by-year correlation of school value-added between M1 and M2")
```

::: {.callout-note}
## Reading the per-year results

- **Stable stage-1 coefficients across years** would suggest the intake → absence relationship is structural rather than period-specific, supporting the validity of pooling.
- **Drift in the FSM coefficient between M1 and M2 within a year** quantifies how much disadvantage signal absence is absorbing in that particular year.
- **Pandemic anomalies (2021-22)**: this was the year when post-pandemic absence patterns were most disrupted; if the M1↔M2 rank correlation drops here relative to later years, that's a signal that pooled estimates are partly driven by an unusual year.
- **B&H year-on-year**: if a school's value-added or residual absence changes substantially across years, that's information --- either a real shift in the school's performance, or genuine year-to-year volatility that single-year reporting would over-interpret.
:::
