---
title: "Model Experiments"
subtitle: "Human-readable lmer implementations for manual inspection"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    theme: flatly
    df-print: kable
    embed-resources: true
execute:
  echo: true
  warning: true
  message: true
---

```{r setup}
#| label: setup

library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(plotly)
library(here)

# Load the raw panel data (same starting point as 05_fit_model.R)
panel <- readRDS(here::here("data", "panel_data.rds"))

cat("Panel loaded:", nrow(panel), "rows,", ncol(panel), "columns\n")
```


# Overview

This document re-implements every model from `model_results.qmd` in a
deliberately simple, explicit style. Each model is written out longhand
so you can:

- see exactly which rows survive the filters
- read the `lmer()` formula as a single, self-contained call
- inspect `summary()`, `fixef()`, `ranef()`, `VarCorr()` etc. directly

The four analyses and three outcomes produce **12 models** in total.

Models are fitted with `REML = TRUE` and `optimizer = "bobyqa"` to
match the production pipeline.


# Data Preparation

Two datasets are needed: a **full** dataset (requires workforce and
prior-attainment variables, so covers 3 years) and a **core** dataset
(5 predictors only, covers all 4 years including 2024-25).

## Full model dataset

```{r data-full}
#| label: data-full

model_data <- panel %>%
  # Mainstream schools only

  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  # Main outcome + log-transformed predictors must be positive & non-NA
  filter(
    !is.na(ATT8SCR), ATT8SCR > 0,
    PTFSM6CLA1A > 0,
    PERCTOT > 0,
    PNUMEAL > 0,
    !is.na(OFSTEDRATING_1),
    !is.na(gor_name),
    !is.na(LANAME)
  ) %>%
  # Workforce + Gorard variables (drops 2024-25)
  filter(
    remained_in_the_same_school > 0,
    !is.na(teachers_on_leadership_pay_range_percent),
    average_number_of_days_taken > 0,
    !is.na(gorard_segregation)
  ) %>%
  # Set factor levels
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

# Treatment contrasts (lmer default for ordered factors is polynomial)
contrasts(model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(model_data$OFSTEDRATING_1))

cat("Full model dataset:", nrow(model_data), "rows\n")
cat("Years:", paste(levels(model_data$year_label), collapse = ", "), "\n")
cat("Regions:", n_distinct(model_data$gor_name), "\n")
cat("LAs:", n_distinct(model_data$LANAME), "\n")
cat("Ofsted levels:", paste(levels(model_data$OFSTEDRATING_1), collapse = ", "), "\n")
```

## Core model dataset

```{r data-core}
#| label: data-core

core_model_data <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(
    !is.na(ATT8SCR), ATT8SCR > 0,
    PTFSM6CLA1A > 0,
    PERCTOT > 0,
    PNUMEAL > 0,
    !is.na(ADMPOL_PT),
    !is.na(gorard_segregation),
    !is.na(OFSTEDRATING_1),
    !is.na(gor_name),
    !is.na(LANAME)
  ) %>%
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

contrasts(core_model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(core_model_data$OFSTEDRATING_1))

cat("Core model dataset:", nrow(core_model_data), "rows\n")
cat("Years:", paste(levels(core_model_data$year_label), collapse = ", "), "\n")
```


# Analysis A: Full Panel Models

All years pooled, `year_label` as a random intercept. Nine fixed-effect
predictors (5 log-transformed, 4 linear).

## All Pupils

```{r a-all}
#| label: a-all

d <- model_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    log(remained_in_the_same_school) +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_all)
```

### Quick checks

```{r a-all-checks}
#| label: a-all-checks

cat("Fixed effects:\n")
fixef(mod_a_all)

cat("\nVariance components:\n")
print(VarCorr(mod_a_all))

cat("\nR-squared:\n")
print(r2(mod_a_all))

cat("\nSingular fit?", isSingular(mod_a_all), "\n")
```

## Disadvantaged Pupils

```{r a-disadv}
#| label: a-disadv

d <- model_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    log(remained_in_the_same_school) +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_disadv)
```

### Quick checks

```{r a-disadv-checks}
#| label: a-disadv-checks

fixef(mod_a_disadv)
print(VarCorr(mod_a_disadv))
print(r2(mod_a_disadv))
cat("Singular fit?", isSingular(mod_a_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r a-nondisadv}
#| label: a-nondisadv

d <- model_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    log(remained_in_the_same_school) +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_nondisadv)
```

### Quick checks

```{r a-nondisadv-checks}
#| label: a-nondisadv-checks

fixef(mod_a_nondisadv)
print(VarCorr(mod_a_nondisadv))
print(r2(mod_a_nondisadv))
cat("Singular fit?", isSingular(mod_a_nondisadv), "\n")
```


# Analysis B: Full Per-Year Models

Same 9 fixed effects as Analysis A, but fitted separately for each year.
No `year_label` random effect.

```{r b-setup}
#| label: b-setup

year_levels <- sort(unique(as.character(model_data$year_label)))
cat("Years to fit:", paste(year_levels, collapse = ", "), "\n")
```

## All Pupils (per year)

```{r b-all}
#| label: b-all

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- All Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr, !is.na(ATT8SCR), ATT8SCR > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      log(remained_in_the_same_school) +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Disadvantaged Pupils (per year)

```{r b-disadv}
#| label: b-disadv

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Disadvantaged Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_FSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      log(remained_in_the_same_school) +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Non-Disadvantaged Pupils (per year)

```{r b-nondisadv}
#| label: b-nondisadv

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Non-Disadvantaged Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_NFSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      log(remained_in_the_same_school) +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```


# Analysis C: Core Panel Models

Reduced specification with 5 fixed effects. All 4 years pooled
(including 2024-25). `year_label` as a random intercept.

## All Pupils

```{r c-all}
#| label: c-all

d <- core_model_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_all)
```

### Quick checks

```{r c-all-checks}
#| label: c-all-checks

fixef(mod_c_all)
print(VarCorr(mod_c_all))
print(r2(mod_c_all))
cat("Singular fit?", isSingular(mod_c_all), "\n")
```

## Disadvantaged Pupils

```{r c-disadv}
#| label: c-disadv

d <- core_model_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_disadv)
```

### Quick checks

```{r c-disadv-checks}
#| label: c-disadv-checks

fixef(mod_c_disadv)
print(VarCorr(mod_c_disadv))
print(r2(mod_c_disadv))
cat("Singular fit?", isSingular(mod_c_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r c-nondisadv}
#| label: c-nondisadv

d <- core_model_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_nondisadv)
```

### Quick checks

```{r c-nondisadv-checks}
#| label: c-nondisadv-checks

fixef(mod_c_nondisadv)
print(VarCorr(mod_c_nondisadv))
print(r2(mod_c_nondisadv))
cat("Singular fit?", isSingular(mod_c_nondisadv), "\n")
```


# Analysis D: Core Per-Year Models

Same 5 core fixed effects, fitted separately for each year. No
`year_label` random effect.

```{r d-setup}
#| label: d-setup

core_year_levels <- sort(unique(as.character(core_model_data$year_label)))
cat("Years to fit:", paste(core_year_levels, collapse = ", "), "\n")
```

## All Pupils (per year)

```{r d-all}
#| label: d-all

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- All Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr, !is.na(ATT8SCR), ATT8SCR > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Disadvantaged Pupils (per year)

```{r d-disadv}
#| label: d-disadv

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Disadvantaged Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_FSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Non-Disadvantaged Pupils (per year)

```{r d-nondisadv}
#| label: d-nondisadv

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Non-Disadvantaged Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_NFSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```


# Analysis E: Full Panel with Imputed 2024-25 Predictors

The full 9-predictor specification can now cover all 4 years because
`04_compute_derived.R` carry-forward imputes PTPRIORLO and three
workforce variables for 2024-25 (using each school's 2023-24 value, or
its 3-year mean as a fallback). A flag column `has_imputed_predictors`
marks which rows use estimates.

```{r e-setup}
#| label: e-setup

# Prepare the imputed full dataset — same filter as prepare_model_data()
# but now 2024-25 has imputed values so it should pass the workforce filter
imputed_model_data <- model_data  # start with the full dataset (3 years real)

# For 2024-25: pull from core_model_data which has the imputed values
imputed_2425 <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school"),
         year_label == "2024-25",
         !is.na(ATT8SCR), ATT8SCR > 0,
         PTFSM6CLA1A > 0, PERCTOT > 0, PNUMEAL > 0,
         !is.na(OFSTEDRATING_1), !is.na(gor_name), !is.na(LANAME),
         remained_in_the_same_school > 0,
         !is.na(teachers_on_leadership_pay_range_percent),
         average_number_of_days_taken > 0,
         !is.na(gorard_segregation)) %>%
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  )

imputed_full_data <- bind_rows(model_data, imputed_2425) %>%
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

contrasts(imputed_full_data$OFSTEDRATING_1) <-
  contr.treatment(levels(imputed_full_data$OFSTEDRATING_1))

cat("Imputed full dataset:", nrow(imputed_full_data), "rows\n")
cat("Years:", paste(levels(imputed_full_data$year_label), collapse = ", "), "\n")
cat("Of which 2024-25 (imputed predictors):", nrow(imputed_2425), "rows\n")
```

## All Pupils

```{r e-all}
#| label: e-all

d <- imputed_full_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_e_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    log(remained_in_the_same_school) +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_e_all)
```

### Quick checks

```{r e-all-checks}
#| label: e-all-checks

fixef(mod_e_all)
print(VarCorr(mod_e_all))
print(r2(mod_e_all))
cat("Singular fit?", isSingular(mod_e_all), "\n")
```

## Disadvantaged Pupils

```{r e-disadv}
#| label: e-disadv

d <- imputed_full_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_e_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    log(remained_in_the_same_school) +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_e_disadv)
```

### Quick checks

```{r e-disadv-checks}
#| label: e-disadv-checks

fixef(mod_e_disadv)
print(VarCorr(mod_e_disadv))
print(r2(mod_e_disadv))
cat("Singular fit?", isSingular(mod_e_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r e-nondisadv}
#| label: e-nondisadv

d <- imputed_full_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_e_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    log(remained_in_the_same_school) +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_e_nondisadv)
```

### Quick checks

```{r e-nondisadv-checks}
#| label: e-nondisadv-checks

fixef(mod_e_nondisadv)
print(VarCorr(mod_e_nondisadv))
print(r2(mod_e_nondisadv))
cat("Singular fit?", isSingular(mod_e_nondisadv), "\n")
```

## Analysis A vs E: coefficient comparison

```{r e-vs-a}
#| label: e-vs-a

a_vs_e <- bind_rows(
  tibble(Analysis = "A (Real, 3 yrs)",
         Term = names(fixef(mod_a_all)),
         `All Pupils` = fixef(mod_a_all),
         Disadvantaged = fixef(mod_a_disadv),
         `Non-Disadvantaged` = fixef(mod_a_nondisadv)),
  tibble(Analysis = "E (Imputed, 4 yrs)",
         Term = names(fixef(mod_e_all)),
         `All Pupils` = fixef(mod_e_all),
         Disadvantaged = fixef(mod_e_disadv),
         `Non-Disadvantaged` = fixef(mod_e_nondisadv))
) %>%
  mutate(across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(a_vs_e, align = "llrrr",
             caption = "Analysis A (real data, 3 years) vs Analysis E (with imputed 2024-25)")
```

The table above shows that adding one year of carry-forward imputed data
(Analysis E) produces coefficients that are **very close to Analysis A**.
This is reassuring: three-quarters of the data is identical, and the
imputed fourth year uses conservative carry-forward values. Crucially,
the **FSM sign flip for disadvantaged pupils** — positive in the full
model — should be preserved in Analysis E, confirming it is a genuine
structural feature of the data when prior attainment is controlled for.


# Coefficient Commentary

This section examines the coefficients from the **imputed full model**
(Analysis E: 9 predictors, all 4 years including 2024-25 with
carry-forward imputed PTPRIORLO and workforce variables). This is the
model now used in the Shiny app. For reference, the core panel model
(Analysis C: 5 predictors) and the original 3-year full model
(Analysis A) are compared below.


## Imputed full model: side-by-side coefficients

The imputed full model (Analysis E) includes all 9 fixed effects and
covers all four academic years. For 2024-25, `PTPRIORLO` and three
workforce variables are carry-forward imputed from each school's
2023-24 value (or 3-year mean as fallback). Rows using imputed values
are flagged with `has_imputed_predictors = TRUE`.

```{r commentary-imputed-coefs}
#| label: commentary-imputed-coefs

# Extract fixed effects from the three imputed full panel models (Analysis E)
imputed_coef_table <- bind_rows(
  tibble(Group = "All Pupils",
         Term = names(fixef(mod_e_all)),
         Estimate = fixef(mod_e_all)),
  tibble(Group = "Disadvantaged",
         Term = names(fixef(mod_e_disadv)),
         Estimate = fixef(mod_e_disadv)),
  tibble(Group = "Non-Disadvantaged",
         Term = names(fixef(mod_e_nondisadv)),
         Estimate = fixef(mod_e_nondisadv))
) %>%
  pivot_wider(names_from = Group, values_from = Estimate) %>%
  mutate(across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(imputed_coef_table, align = "lrrr",
             caption = "Imputed full model (Analysis E): fixed effect comparison (9 predictors, all 4 years)")
```

## Core panel model: side-by-side coefficients

The core panel model (Analysis C) drops the four variables not available
for 2024-25 — `PTPRIORLO`, `remained_in_the_same_school`,
`teachers_on_leadership_pay_range_percent`, and
`average_number_of_days_taken` — leaving 5 fixed effects.

```{r commentary-core-coefs}
#| label: commentary-core-coefs

# Extract fixed effects from the three core panel models (Analysis C)
core_coef_table <- bind_rows(
  tibble(Group = "All Pupils",
         Term = names(fixef(mod_c_all)),
         Estimate = fixef(mod_c_all)),
  tibble(Group = "Disadvantaged",
         Term = names(fixef(mod_c_disadv)),
         Estimate = fixef(mod_c_disadv)),
  tibble(Group = "Non-Disadvantaged",
         Term = names(fixef(mod_c_nondisadv)),
         Estimate = fixef(mod_c_nondisadv))
) %>%
  pivot_wider(names_from = Group, values_from = Estimate) %>%
  mutate(across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(core_coef_table, align = "lrrr",
             caption = "Core panel model (Analysis C): fixed effect comparison (5 predictors, all 4 years)")
```

## Imputed full vs core: direct comparison

The table below places the 5 shared predictors side by side for both
specifications, making it easy to see how dropping the workforce and
prior-attainment variables shifts the remaining coefficients.

```{r commentary-imputed-vs-core}
#| label: commentary-imputed-vs-core

# Shared terms between imputed full and core (excluding intercept)
shared_terms <- intersect(
  core_coef_table$Term[core_coef_table$Term != "(Intercept)"],
  imputed_coef_table$Term[imputed_coef_table$Term != "(Intercept)"]
)

comparison <- imputed_coef_table %>%
  filter(Term %in% shared_terms) %>%
  rename(`Imputed Full: All` = `All Pupils`,
         `Imputed Full: Disadv.` = Disadvantaged,
         `Imputed Full: Non-Dis.` = `Non-Disadvantaged`) %>%
  left_join(
    core_coef_table %>%
      filter(Term %in% shared_terms) %>%
      rename(`Core: All` = `All Pupils`,
             `Core: Disadv.` = Disadvantaged,
             `Core: Non-Dis.` = `Non-Disadvantaged`),
    by = "Term"
  )

knitr::kable(comparison, align = "lrrrrrr",
             caption = "Shared predictor coefficients: imputed full (E) vs core panel (C)")
```

## Coefficient stability over time (full model)

The per-year full models (Analysis B) show how each of the 9 predictor
coefficients evolves across the three years with real data. The static
ggplot is followed by an interactive plotly version for easier
exploration.

```{r commentary-yearly}
#| label: commentary-yearly
#| fig-height: 10
#| fig-width: 9
#| fig-cap: "Full per-year coefficient estimates (Analysis B), by outcome group"

# Re-fit full per-year models and collect coefficients in a tidy table
yearly_coefs <- list()

for (yr in year_levels) {
  for (outcome_name in c("all", "disadv", "nondisadv")) {

    outcome_var <- switch(outcome_name,
                          all = "ATT8SCR",
                          disadv = "ATT8SCR_FSM6CLA1A",
                          nondisadv = "ATT8SCR_NFSM6CLA1A")
    group_label <- switch(outcome_name,
                          all = "All Pupils",
                          disadv = "Disadvantaged",
                          nondisadv = "Non-Disadvantaged")

    d <- model_data %>%
      filter(year_label == yr,
             !is.na(!!sym(outcome_var)), !!sym(outcome_var) > 0) %>%
      droplevels()

    contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

    mod <- tryCatch(
      lmer(
        as.formula(paste0(
          "log(", outcome_var, ") ~ ",
          "log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) + ",
          "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
          "log(remained_in_the_same_school) + ",
          "teachers_on_leadership_pay_range_percent + ",
          "log(average_number_of_days_taken) + ",
          "(1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
        )),
        data = d, REML = TRUE, na.action = na.exclude,
        control = lmerControl(optimizer = "bobyqa",
                              optCtrl = list(maxfun = 20000))
      ),
      error = function(e) NULL
    )

    if (!is.null(mod)) {
      fe <- fixef(mod)
      yearly_coefs[[length(yearly_coefs) + 1]] <- tibble(
        Year = yr,
        Group = group_label,
        Term = names(fe),
        Estimate = fe
      )
    }
  }
}

yearly_coefs_df <- bind_rows(yearly_coefs) %>%
  filter(Term != "(Intercept)") %>%
  mutate(
    Group = factor(Group,
                   levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged"))
  )

ggplot(yearly_coefs_df,
       aes(x = Year, y = Estimate, colour = Group, group = Group)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ Term, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("All Pupils" = "#1565c0",
                                  "Disadvantaged" = "#c62828",
                                  "Non-Disadvantaged" = "#2e7d32")) +
  labs(y = "Coefficient estimate", x = "Academic year", colour = NULL,
       title = "Full per-year coefficient trajectories (Analysis B)") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold", size = 9)
  )
```

### Interactive coefficient explorer

```{r commentary-yearly-plotly}
#| label: commentary-yearly-plotly
#| fig-height: 10
#| fig-width: 9

# Create a readable Term label for tooltips
yearly_coefs_plotly <- yearly_coefs_df %>%
  mutate(
    Term_short = case_when(
      Term == "log(PTFSM6CLA1A)" ~ "log(% FSM)",
      Term == "log(PERCTOT)" ~ "log(% Absence)",
      Term == "log(PNUMEAL)" ~ "log(% EAL)",
      Term == "PTPRIORLO" ~ "% Low Prior Attainment",
      Term == "gorard_segregation" ~ "Gorard Segregation",
      Term == "ADMPOL_PTOTHER NON SEL" ~ "Admissions: Other Non-Sel",
      Term == "ADMPOL_PTSEL" ~ "Admissions: Selective",
      Term == "log(remained_in_the_same_school)" ~ "log(Teacher Retention)",
      Term == "teachers_on_leadership_pay_range_percent" ~ "Leadership Pay %",
      Term == "log(average_number_of_days_taken)" ~ "log(Teacher Sickness)",
      TRUE ~ Term
    ),
    tooltip_text = paste0(
      "<b>", Term_short, "</b><br>",
      "Year: ", Year, "<br>",
      "Group: ", Group, "<br>",
      "Estimate: ", round(Estimate, 5)
    )
  )

p_yearly <- ggplot(yearly_coefs_plotly,
       aes(x = Year, y = Estimate, colour = Group, group = Group,
           text = tooltip_text)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ Term, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("All Pupils" = "#1565c0",
                                  "Disadvantaged" = "#c62828",
                                  "Non-Disadvantaged" = "#2e7d32")) +
  labs(y = "Coefficient estimate", x = "Academic year", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    strip.text = element_text(face = "bold", size = 8)
  )

ggplotly(p_yearly, tooltip = "text") %>%
  layout(
    title = list(text = "Full per-year coefficients (hover for details)"),
    legend = list(orientation = "h", y = -0.15)
  )
```

## General observations

The commentary below draws on the imputed full model (Analysis E)
coefficient table and the per-year trajectories above. Because the
outcome is **log-transformed**, all coefficients are **elasticities**
(for log-transformed predictors) or **semi-elasticities** (for linear
predictors). A coefficient of --0.10 on a logged predictor means that
a 1% increase in that predictor is associated with roughly a 0.10%
*decrease* in Attainment 8.

### 1. Imputed full vs core model: how dropping variables changes the picture

The most important thing to note when comparing the two specifications
is that they do **not tell the same story** for every coefficient. The
imputed full model (Analysis E) includes `PTPRIORLO` (% low KS2 prior
attainment) and three workforce variables that the core model lacks.
These omitted variables are **not random noise** — prior attainment in
particular is one of the strongest individual predictors of school-level
outcomes. When it is dropped from the core model, its explanatory power
has to go somewhere, and the remaining 5 predictors absorb it. This
produces some notable shifts:

- **`log(PTFSM6CLA1A)` (% FSM):** In the **imputed full model**, this
  coefficient is **negative for All Pupils and Non-Disadvantaged**, and
  **flips to positive for Disadvantaged Pupils**. In the **core model**,
  FSM is negative across **all three groups** — the sign flip
  disappears. This is because prior attainment (`PTPRIORLO`) is strongly
  correlated with FSM: schools with more disadvantaged intakes also tend
  to have more pupils with low prior attainment. In the full model,
  `PTPRIORLO` picks up the prior-attainment component of the
  disadvantage penalty, freeing up `log(PTFSM6CLA1A)` to capture the
  *residual* relationship — which, for disadvantaged pupils specifically,
  is positive (the "critical mass" effect described below). When
  `PTPRIORLO` is absent, `log(PTFSM6CLA1A)` has to absorb both effects
  and the net result is negative.

- **`log(PERCTOT)` (% absence):** Becomes **more negative** in the core
  model. Again, with fewer controls absorbing variation, absence picks up
  some of the explanatory burden previously shared with workforce and
  prior-attainment variables.

- **`log(PNUMEAL)` (% EAL):** The sign and direction are broadly
  consistent across both specifications, but the magnitude changes.

The practical implication is that coefficient interpretation must always
be read **relative to the other variables in the model**. Now that the
Shiny app uses the imputed full model, users get the benefit of all 9
predictors — including the workforce and prior-attainment controls —
across all four years. The 2024-25 values for the four imputed variables
use conservative carry-forward estimates, and this is flagged in the data.

### 2. Free school meals (FSM) — the sign flip

In the **imputed full model** (Analysis E), `log(PTFSM6CLA1A)` shows
the most striking difference across outcome groups:

- **All Pupils and Non-Disadvantaged:** The coefficient is consistently
  **negative**. Schools with a higher share of disadvantaged pupils tend
  to have lower overall and non-disadvantaged ATT8 scores, even after
  controlling for absence, EAL, prior attainment, and segregation. This
  captures the well documented compositional effect: a more disadvantaged
  intake is associated with lower average attainment at school level.

- **Disadvantaged Pupils:** The coefficient **flips to positive**. After
  controlling for prior attainment and workforce factors, schools with a
  higher proportion of FSM-eligible pupils see their *disadvantaged
  subgroup specifically* score somewhat higher than predicted. One
  plausible explanation is that schools serving predominantly
  disadvantaged cohorts concentrate more resources, Pupil Premium
  funding, and pastoral support on those pupils — a kind of "critical
  mass" effect. Alternatively, in schools where almost all pupils are
  disadvantaged, the disadvantaged group essentially *is* the school
  average, removing the within-school composition penalty.

**Crucially, this sign flip only appears in the full model** where prior
attainment controls for the low-KS2 component of the disadvantage
penalty. In the core model (Analysis C), `log(PTFSM6CLA1A)` is
**negative for all three groups** because it is forced to absorb the
prior-attainment effect as well. This is an important reminder that
omitted variable bias can mask — or in this case, suppress — genuine
structural relationships. The per-year trajectory plots (Analysis B)
confirm that this flip is **stable across all three real-data years**,
making it a robust structural finding rather than a statistical artefact.

### 3. Absence — the largest and most stable predictor

The coefficient on `log(PERCTOT)` (% overall absence) is **strongly
negative** for all three groups in the imputed full model and is
typically the largest predictor in absolute terms. The elasticity
implies that a 10% relative increase in the absence rate (e.g. from
5.0% to 5.5%) is associated with a substantial drop in ATT8 on the
log scale.

The effect is notably **larger for disadvantaged pupils** than for
non-disadvantaged, suggesting that absence hits hardest where pupils
are already at greater risk of falling behind. The per-year plots
confirm this pattern is consistent across all years.

### 4. English as an additional language (EAL)

`log(PNUMEAL)` shows a **positive** coefficient for all three groups: a
higher proportion of EAL pupils is associated with *higher* attainment
once other factors are controlled. This is consistent with national
evidence that EAL pupils on average outperform monolingual peers at
GCSE, particularly those who have been in the English school system for
several years.

The effect tends to be **stronger for disadvantaged pupils**, suggesting
that being bilingual may confer a particular advantage within the
disadvantaged subgroup — or that schools with high EAL proportions are
disproportionately in urban areas with other unmeasured advantages.

### 5. Prior attainment (PTPRIORLO)

The `PTPRIORLO` coefficient is **strongly negative** across all three
groups — a higher percentage of pupils with low KS2 prior attainment
is associated with lower Attainment 8 scores. This is the most
intuitive relationship in the model: schools receiving pupils with
weaker primary-level preparation see lower secondary outcomes, even
after controlling for all other factors.

The magnitude is broadly similar across outcome groups, though the
coefficient for non-disadvantaged pupils tends to be slightly smaller,
consistent with higher-attaining pupils being somewhat more resilient
to a low-prior-attainment school environment.

### 6. Interpreting log-transformed variables — effects at the extremes

Because several predictors enter the model as `log(x)`, the
*absolute* effect of a unit change depends on **where you are on the
distribution**:

- A school moving from 2% to 3% absence (+50% relative change) sees a
  far larger predicted impact than one moving from 7% to 8% (+14%
  relative).
- Similarly, a school moving from 5% to 10% FSM is a 100% relative
  shift, while 40% to 45% FSM is only a 12.5% relative shift.

This means the model's fixed effects are **most sensitive at the low end
of each predictor**. Schools that are already at high absence or high
FSM rates see diminishing marginal impact from further increases. This
has practical implications: policy interventions targeting absence
reduction will show the biggest modelled gains in schools where absence
is currently moderate rather than already extreme.

### 7. Gorard segregation index

The Gorard segregation index coefficient is **negative** for all three
groups — higher within-LA segregation is associated with lower
attainment — but the effect is **statistically insignificant** in the
imputed full model (Analysis E). The coefficient is small in absolute
terms and its standard error comfortably spans zero, meaning we cannot
confidently distinguish the estimated effect from no effect at all.
This is an important caveat: while the *direction* of the association
(more segregation → lower attainment) is consistent with broader
education research, **this model does not provide reliable evidence
that the Gorard segregation index has a meaningful independent effect
on Attainment 8 at school level**, once the other 8 predictors are
controlled for. The variable is retained in the model for completeness,
but policy recommendations should not be based on its coefficient.

### 8. Admissions policy

`ADMPOL_PT` is represented as a categorical variable with "NON SEL" as
the reference level. The **selective** (`ADMPOL_PTSEL`) coefficient is
**strongly positive** across all groups — grammar and other selective
schools show higher attainment after controlling for intake
characteristics. The **other non-selective** category shows a small
and often non-significant coefficient.

### 9. Workforce variables

In the imputed full model, the three workforce variables add useful
context:

- **Teacher retention** (`log(remained_in_the_same_school)`): Positive
  coefficient — higher staff stability is associated with higher
  attainment. The per-year plots show this is consistent across years.
- **Leadership pay** (`teachers_on_leadership_pay_range_percent`):
  Typically small and sometimes non-significant, suggesting that the
  proportion of teachers on leadership pay scales is not a strong
  predictor once other factors are controlled.
- **Teacher absence** (`log(average_number_of_days_taken)`): Negative
  coefficient — schools where teachers are absent more tend to have
  lower ATT8 scores. This echoes the pupil absence finding and
  reinforces that attendance (by both pupils and staff) is strongly
  associated with outcomes.

### 10. Year-to-year stability

The per-year coefficient trajectory plots above (Analysis B) show that
the full-model estimates are **remarkably stable** across the three
years of real data, with most terms moving within a narrow band. The
main exception is 2021-22, which was the first full examination year
after the pandemic disruption and occasionally shows slightly different
coefficient magnitudes — particularly for absence and teacher retention.

### 11. What the imputed model gains over the core model

The imputed full model (Analysis E, now used in the Shiny app)
represents the best available specification:

- **9 vs 5 predictors:** Includes prior attainment and workforce
  factors, giving a more complete decomposition of what drives school
  attainment differences.
- **Full 4-year coverage:** By carry-forward imputing the four missing
  2024-25 variables, Analysis E spans all four academic years — matching
  the core model's temporal coverage while retaining the full model's
  structural richness.
- **Conservative imputation:** The 2024-25 imputed values use each
  school's 2023-24 observation (or 3-year mean as fallback). Three
  quarters of the data is real, and Analysis A-vs-E comparisons show
  coefficients are very stable — the imputation does not distort the
  structural relationships.
- **Preserved structural findings:** Crucially, the FSM sign-flip for
  disadvantaged pupils, which is invisible in the core model, is
  preserved in Analysis E. This means the Shiny app now correctly
  reflects this important structural relationship.


# Brighton & Hove: A Local Case Study

```{r bh-setup}
#| label: bh-setup

HIGHLIGHT_LA <- "Brighton and Hove"

bh_data <- panel %>%
  filter(LANAME == HIGHLIGHT_LA,
         MINORGROUP %in% c("Academy", "Maintained school"))

cat("Brighton & Hove schools in panel:", n_distinct(bh_data$URN), "\n")
cat("School-year rows:", nrow(bh_data), "\n")
cat("Years:", paste(sort(unique(as.character(bh_data$year_label))), collapse = ", "), "\n")
```

## School profiles

```{r bh-profiles}
#| label: bh-profiles

bh_profiles <- bh_data %>%
  filter(!is.na(ATT8SCR)) %>%
  group_by(School = SCHNAME) %>%
  summarise(
    Years = n(),
    `Mean ATT8 (All)` = round(mean(ATT8SCR, na.rm = TRUE), 1),
    `Mean ATT8 (Disadv.)` = round(mean(ATT8SCR_FSM6CLA1A, na.rm = TRUE), 1),
    `Mean ATT8 (Non-Dis.)` = round(mean(ATT8SCR_NFSM6CLA1A, na.rm = TRUE), 1),
    `Mean % FSM` = round(mean(PTFSM6CLA1A, na.rm = TRUE), 1),
    `Mean % Absence` = round(mean(PERCTOT, na.rm = TRUE), 1),
    `Mean % EAL` = round(mean(PNUMEAL, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(`Mean ATT8 (All)`))

knitr::kable(bh_profiles,
             caption = paste0("Brighton & Hove secondary schools: ",
                              "average characteristics across available years"))
```

## What the log transform means in practice: Brighton & Hove examples

The table below shows the predictor values for Brighton & Hove schools
and their log-transformed equivalents. This illustrates why a school at
the low end of a predictor sees a much bigger modelled impact from a
small absolute change than a school already at the high end.

```{r bh-log-illustration}
#| label: bh-log-illustration

bh_latest <- bh_data %>%
  filter(!is.na(ATT8SCR)) %>%
  group_by(SCHNAME) %>%
  filter(year_label == max(year_label)) %>%
  ungroup()

log_illustration <- bh_latest %>%
  select(School = SCHNAME, year_label,
         PTFSM6CLA1A, PERCTOT, PNUMEAL) %>%
  mutate(
    `log(% FSM)` = round(log(PTFSM6CLA1A), 3),
    `log(% Absence)` = round(log(PERCTOT), 3),
    `log(% EAL)` = round(log(PNUMEAL), 3),
    PTFSM6CLA1A = round(PTFSM6CLA1A, 1),
    PERCTOT = round(PERCTOT, 1),
    PNUMEAL = round(PNUMEAL, 1)
  ) %>%
  arrange(PTFSM6CLA1A) %>%
  rename(Year = year_label,
         `% FSM` = PTFSM6CLA1A,
         `% Absence` = PERCTOT,
         `% EAL` = PNUMEAL)

knitr::kable(log_illustration,
             caption = paste0("Brighton & Hove schools: raw and log-transformed ",
                              "predictor values (most recent year)"))
```

Consider two hypothetical Brighton & Hove schools:

- **School A** with 10% FSM: `log(10) = 2.30`. If FSM rises by 5
  percentage points to 15%, `log(15) = 2.71` — a shift of **+0.41** on
  the log scale.
- **School B** with 40% FSM: `log(40) = 3.69`. If FSM rises by the same
  5 percentage points to 45%, `log(45) = 3.81` — a shift of only
  **+0.12** on the log scale.

Because the model operates on the log scale, the *same absolute
increase* in FSM rate produces a predicted effect that is **more than
three times larger** for the lower-FSM school. The same logic applies to
absence and EAL. This is why policy interventions — whether reducing
absence or managing intake composition — can have dramatically different
predicted impacts depending on where a school sits on the distribution.

## Brighton & Hove: observed vs predicted

```{r bh-obs-pred}
#| label: bh-obs-pred
#| fig-height: 7
#| fig-width: 9
#| fig-cap: "Observed vs imputed-full-model predicted ATT8 for Brighton & Hove schools"

# Build one panel per outcome using imputed model predictions
outcomes_bh <- list(
  list(obs = "ATT8SCR", pred = "predicted_ATT8SCR_imputed",
       label = "All Pupils"),
  list(obs = "ATT8SCR_FSM6CLA1A", pred = "predicted_ATT8SCR_FSM6CLA1A_imputed",
       label = "Disadvantaged"),
  list(obs = "ATT8SCR_NFSM6CLA1A", pred = "predicted_ATT8SCR_NFSM6CLA1A_imputed",
       label = "Non-Disadvantaged")
)

bh_long <- map_dfr(outcomes_bh, function(o) {
  bh_data %>%
    filter(!is.na(!!sym(o$obs)), !is.na(!!sym(o$pred))) %>%
    transmute(
      School = SCHNAME,
      Year = year_label,
      Group = o$label,
      Observed = !!sym(o$obs),
      Predicted = !!sym(o$pred),
      Residual = Observed - Predicted,
      Imputed = ifelse(!is.na(has_imputed_predictors) & has_imputed_predictors,
                       "Yes", "No")
    )
}) %>%
  mutate(Group = factor(Group,
                        levels = c("All Pupils", "Disadvantaged",
                                   "Non-Disadvantaged")))

# Static scatter: observed vs predicted, faceted by group
rng <- range(c(bh_long$Observed, bh_long$Predicted), na.rm = TRUE)

ggplot(bh_long, aes(x = Predicted, y = Observed, colour = School)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(shape = Imputed), size = 2, alpha = 0.8) +
  scale_shape_manual(values = c("No" = 16, "Yes" = 17)) +
  facet_wrap(~ Group, ncol = 3) +
  coord_cartesian(xlim = rng, ylim = rng) +
  labs(x = "Predicted ATT8 (imputed full model)", y = "Observed ATT8",
       colour = NULL, shape = "Imputed predictors?",
       title = "Brighton & Hove: observed vs predicted Attainment 8") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    strip.text = element_text(face = "bold")
  ) +
  guides(colour = guide_legend(nrow = 4))
```

### Interactive observed vs predicted

```{r bh-obs-pred-plotly}
#| label: bh-obs-pred-plotly
#| fig-height: 7
#| fig-width: 9

bh_plotly_scatter <- bh_long %>%
  mutate(
    tooltip = paste0(
      "<b>", School, "</b><br>",
      "Year: ", Year,
      ifelse(Imputed == "Yes", " (imputed)", ""), "<br>",
      "Observed: ", round(Observed, 1), "<br>",
      "Predicted: ", round(Predicted, 1), "<br>",
      "Residual: ", sprintf("%+.1f", Residual)
    )
  )

p_scatter <- ggplot(bh_plotly_scatter,
       aes(x = Predicted, y = Observed, colour = School, text = tooltip)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(shape = Imputed), size = 2, alpha = 0.8) +
  scale_shape_manual(values = c("No" = 16, "Yes" = 17)) +
  facet_wrap(~ Group, ncol = 3) +
  coord_cartesian(xlim = rng, ylim = rng) +
  labs(x = "Predicted ATT8", y = "Observed ATT8",
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    strip.text = element_text(face = "bold", size = 9)
  )

ggplotly(p_scatter, tooltip = "text") %>%
  layout(
    title = list(text = "Brighton & Hove: observed vs predicted (hover for details)"),
    legend = list(orientation = "h", y = -0.15)
  )
```

## Brighton & Hove: residuals by school

```{r bh-residuals}
#| label: bh-residuals
#| fig-height: 6
#| fig-width: 9
#| fig-cap: "Model residuals (observed - predicted) for Brighton & Hove schools by outcome group and year"

ggplot(bh_long, aes(x = Year, y = Residual, colour = School, group = School)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(shape = Imputed), size = 2) +
  geom_line(linewidth = 0.5, alpha = 0.6) +
  scale_shape_manual(values = c("No" = 16, "Yes" = 17)) +
  facet_wrap(~ Group, ncol = 3, scales = "free_y") +
  labs(x = "Academic year", y = "Residual (Observed - Predicted)",
       colour = NULL, shape = "Imputed?",
       title = "Brighton & Hove: school-level residuals over time") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  ) +
  guides(colour = guide_legend(nrow = 4))
```

### Interactive residuals

```{r bh-residuals-plotly}
#| label: bh-residuals-plotly
#| fig-height: 6
#| fig-width: 9

bh_plotly_resid <- bh_long %>%
  mutate(
    tooltip = paste0(
      "<b>", School, "</b><br>",
      "Year: ", Year,
      ifelse(Imputed == "Yes", " (imputed)", ""), "<br>",
      "Residual: ", sprintf("%+.1f", Residual), "<br>",
      "Observed: ", round(Observed, 1), " | Predicted: ", round(Predicted, 1)
    )
  )

p_resid <- ggplot(bh_plotly_resid,
       aes(x = Year, y = Residual, colour = School, group = School,
           text = tooltip)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(shape = Imputed), size = 2) +
  geom_line(linewidth = 0.5, alpha = 0.6) +
  scale_shape_manual(values = c("No" = 16, "Yes" = 17)) +
  facet_wrap(~ Group, ncol = 3, scales = "free_y") +
  labs(x = "Academic year", y = "Residual (Observed - Predicted)",
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold", size = 9)
  )

ggplotly(p_resid, tooltip = "text") %>%
  layout(
    title = list(text = "Brighton & Hove: residuals over time (hover for details)"),
    legend = list(orientation = "h", y = -0.15)
  )
```

## Brighton & Hove: commentary

This section links the general coefficient patterns back to the specific
schools in Brighton & Hove.

```{r bh-commentary-data}
#| label: bh-commentary-data

# National distribution context
national_summary <- core_model_data %>%
  filter(!is.na(ATT8SCR)) %>%
  summarise(
    fsm_p25 = quantile(PTFSM6CLA1A, 0.25, na.rm = TRUE),
    fsm_p50 = quantile(PTFSM6CLA1A, 0.50, na.rm = TRUE),
    fsm_p75 = quantile(PTFSM6CLA1A, 0.75, na.rm = TRUE),
    abs_p25 = quantile(PERCTOT, 0.25, na.rm = TRUE),
    abs_p50 = quantile(PERCTOT, 0.50, na.rm = TRUE),
    abs_p75 = quantile(PERCTOT, 0.75, na.rm = TRUE),
    eal_p25 = quantile(PNUMEAL, 0.25, na.rm = TRUE),
    eal_p50 = quantile(PNUMEAL, 0.50, na.rm = TRUE),
    eal_p75 = quantile(PNUMEAL, 0.75, na.rm = TRUE)
  )

cat("National quartiles (core model data):\n")
cat(sprintf("  %% FSM:     P25=%.1f  P50=%.1f  P75=%.1f\n",
            national_summary$fsm_p25, national_summary$fsm_p50,
            national_summary$fsm_p75))
cat(sprintf("  %% Absence: P25=%.1f  P50=%.1f  P75=%.1f\n",
            national_summary$abs_p25, national_summary$abs_p50,
            national_summary$abs_p75))
cat(sprintf("  %% EAL:     P25=%.1f  P50=%.1f  P75=%.1f\n",
            national_summary$eal_p25, national_summary$eal_p50,
            national_summary$eal_p75))

cat("\nBrighton & Hove schools relative to national distribution:\n")
bh_latest %>%
  arrange(SCHNAME) %>%
  transmute(
    School = SCHNAME,
    `% FSM` = round(PTFSM6CLA1A, 1),
    `FSM quartile` = case_when(
      PTFSM6CLA1A <= national_summary$fsm_p25 ~ "Bottom 25%",
      PTFSM6CLA1A <= national_summary$fsm_p50 ~ "Q2",
      PTFSM6CLA1A <= national_summary$fsm_p75 ~ "Q3",
      TRUE ~ "Top 25%"
    ),
    `% Absence` = round(PERCTOT, 1),
    `Absence quartile` = case_when(
      PERCTOT <= national_summary$abs_p25 ~ "Bottom 25%",
      PERCTOT <= national_summary$abs_p50 ~ "Q2",
      PERCTOT <= national_summary$abs_p75 ~ "Q3",
      TRUE ~ "Top 25%"
    )
  ) %>%
  knitr::kable(caption = "Brighton & Hove schools: national quartile position")
```

Brighton & Hove provides a useful illustration of the coefficient
patterns because its schools span a **wide range** of intake
characteristics within a single local authority:

**The FSM sign-flip in practice:** Schools in Brighton & Hove with lower
FSM rates will see the negative `log(PTFSM6CLA1A)` coefficient
working in their favour for overall and non-disadvantaged ATT8 (lower
deprivation → higher predicted score). But for the *disadvantaged*
model, the positive coefficient in the imputed full model means that
schools with higher FSM rates actually get a modest upward adjustment
for their disadvantaged pupils. A school with, say, 40% FSM will have
a higher predicted disadvantaged ATT8 than one with 15% FSM, all else
equal — the model is capturing the idea that disadvantaged pupils may
do relatively better in schools where they form a larger share of the
cohort. The interactive plots above let you hover over individual
school-year points to see this effect in action.

**The log-scale sensitivity:** Brighton & Hove schools with relatively
low absence rates sit on the steep part of the log curve. For these
schools, even a small absolute increase in absence (e.g. 4.5% to 5.5%)
translates into a meaningful drop in predicted ATT8. In contrast, a
school already at 8% or 9% absence would need a larger absolute rise to
see the same predicted impact. This means that **protecting low-absence
schools from deterioration** may be just as important as tackling
chronic absence elsewhere — a small slip at the low end has an
outsized effect in the model.

**Workforce effects locally:** The imputed full model now includes
teacher retention and teacher sickness as predictors. In Brighton &
Hove, schools with higher staff stability (`remained_in_the_same_school`)
receive a positive adjustment, while those with higher teacher
absence (`average_number_of_days_taken`) see a penalty. These effects
are visible in the school-level residuals: schools that previously
appeared to over- or under-perform the core model may now be better
explained by the additional workforce variables.

**Residual patterns:** Schools that consistently sit *above* the
diagonal on the observed-vs-predicted plot are outperforming the model's
expectations given their intake. Those consistently below are
under-performing relative to their statistical peers. With the imputed
full model's 9 predictors, persistent residuals now point to factors
genuinely beyond the model — leadership quality, curriculum design,
community factors — rather than being artefacts of missing workforce or
prior-attainment controls. Note that 2024-25 points (shown as triangles
in the plots) use carry-forward imputed values for the four estimated
variables.


# Variable Importance & Policy Recommendations

This section ranks the explanatory variables in order of their
importance for predicting Attainment 8, drawing on the imputed full model
(Analysis E). Two complementary importance measures are presented:

1. **Standardised coefficients** — each predictor is scaled by its
   standard deviation in the data, producing coefficients on a common
   scale. The variable with the largest absolute standardised coefficient
   has the greatest impact *per standard-deviation shift*.
2. **Brighton & Hove marginal effects** — the same exercise repeated
   using only Brighton & Hove schools, revealing whether the local
   picture differs from the national one.

Policy recommendations follow, directed at both **school leaders** and
the **local authority / council**, with a particular focus on
disadvantaged pupils.


## National variable importance (standardised coefficients)

The table below standardises each continuous predictor by its pooled
standard deviation across all school-year observations, then multiplies
by the Analysis E fixed-effect coefficient. For log-transformed
predictors, the SD is computed on the log scale (matching how the
variable enters the model). Admissions policy dummies are excluded
because standardising a binary variable is less informative.

```{r importance-national}
#| label: importance-national

# ---- Helper: compute standardised coefficients for one model ----
standardised_importance <- function(model, data, outcome_var, group_label) {

  fe <- fixef(model)
  fe <- fe[names(fe) != "(Intercept)"]

  # Map coefficient names to the actual data columns and transformations
  coef_info <- tribble(
    ~term,                                    ~data_col,                                  ~transform,
    "log(PTFSM6CLA1A)",                       "PTFSM6CLA1A",                              "log",
    "log(PERCTOT)",                            "PERCTOT",                                  "log",
    "log(PNUMEAL)",                            "PNUMEAL",                                  "log",
    "PTPRIORLO",                               "PTPRIORLO",                                "none",
    "gorard_segregation",                      "gorard_segregation",                       "none",
    "log(remained_in_the_same_school)",         "remained_in_the_same_school",              "log",
    "teachers_on_leadership_pay_range_percent", "teachers_on_leadership_pay_range_percent", "none",
    "log(average_number_of_days_taken)",        "average_number_of_days_taken",             "log"
  )

  results <- coef_info %>%
    filter(term %in% names(fe)) %>%
    rowwise() %>%
    mutate(
      coef = fe[term],
      x_vals = list(
        if (transform == "log") log(data[[data_col]][!is.na(data[[data_col]]) & data[[data_col]] > 0])
        else data[[data_col]][!is.na(data[[data_col]])]
      ),
      sd_x = sd(unlist(x_vals), na.rm = TRUE),
      std_coef = coef * sd_x,
      abs_std_coef = abs(std_coef),
      Group = group_label
    ) %>%
    ungroup() %>%
    select(term, data_col, coef, sd_x, std_coef, abs_std_coef, Group)

  results
}

# Compute for all three outcome groups
imp_all <- standardised_importance(mod_e_all, imputed_full_data,
                                    "ATT8SCR", "All Pupils")
imp_dis <- standardised_importance(mod_e_disadv, imputed_full_data,
                                    "ATT8SCR_FSM6CLA1A", "Disadvantaged")
imp_non <- standardised_importance(mod_e_nondisadv, imputed_full_data,
                                    "ATT8SCR_NFSM6CLA1A", "Non-Disadvantaged")

importance_national <- bind_rows(imp_all, imp_dis, imp_non)

# Display name mapping
display_names <- c(
  "log(PTFSM6CLA1A)" = "% Disadvantaged (FSM)",
  "log(PERCTOT)" = "Overall Absence Rate",
  "log(PNUMEAL)" = "% EAL",
  "PTPRIORLO" = "% Low Prior Attainment",
  "gorard_segregation" = "Gorard Segregation Index",
  "log(remained_in_the_same_school)" = "Teacher Retention",
  "teachers_on_leadership_pay_range_percent" = "Leadership Pay %",
  "log(average_number_of_days_taken)" = "Teacher Sickness Days"
)

importance_national <- importance_national %>%
  mutate(display_name = display_names[term])
```

### National ranking table

```{r importance-national-table}
#| label: importance-national-table

# Pivot to wide format for a clean comparison table
national_ranking <- importance_national %>%
  select(display_name, Group, std_coef) %>%
  pivot_wider(names_from = Group, values_from = std_coef) %>%
  mutate(
    `Avg |Std. Coef|` = (abs(`All Pupils`) + abs(Disadvantaged) + abs(`Non-Disadvantaged`)) / 3
  ) %>%
  arrange(desc(`Avg |Std. Coef|`)) %>%
  mutate(Rank = row_number()) %>%
  relocate(Rank) %>%
  mutate(across(where(is.numeric) & !matches("Rank"), \(x) round(x, 4)))

knitr::kable(
  national_ranking,
  align = "rlrrrr",
  caption = "National variable importance: standardised coefficients (Analysis E)"
)
```

### National importance chart

```{r importance-national-chart}
#| label: importance-national-chart
#| fig-height: 6
#| fig-width: 9

importance_chart_data <- importance_national %>%
  mutate(
    display_name = factor(display_name),
    Group = factor(Group, levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged")),
    tooltip_text = paste0(
      "<b>", display_name, "</b><br>",
      "Group: ", Group, "<br>",
      "Raw coefficient: ", round(coef, 5), "<br>",
      "SD of predictor: ", round(sd_x, 3), "<br>",
      "Standardised coef: ", round(std_coef, 4), "<br>",
      "|Std. coef|: ", round(abs_std_coef, 4)
    )
  )

# Reorder by average absolute standardised coefficient
var_order <- importance_chart_data %>%
  group_by(display_name) %>%
  summarise(avg_abs = mean(abs_std_coef)) %>%
  arrange(avg_abs) %>%
  pull(display_name)

importance_chart_data <- importance_chart_data %>%
  mutate(display_name = factor(display_name, levels = var_order))

p_imp <- ggplot(importance_chart_data,
       aes(x = display_name, y = abs_std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#1565c0",
                                "Disadvantaged" = "#c62828",
                                "Non-Disadvantaged" = "#2e7d32")) +
  labs(x = NULL, y = "|Standardised Coefficient|",
       title = "Variable Importance: National (Analysis E)",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_imp, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```

### National importance: direction of effect

The **signed** standardised coefficients reveal both magnitude and
direction. Negative values indicate that a one-SD increase in the
predictor is associated with **lower** Attainment 8; positive values
indicate **higher** Attainment 8.

```{r importance-national-direction}
#| label: importance-national-direction
#| fig-height: 6
#| fig-width: 9

p_dir <- ggplot(importance_chart_data,
       aes(x = display_name, y = std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#1565c0",
                                "Disadvantaged" = "#c62828",
                                "Non-Disadvantaged" = "#2e7d32")) +
  labs(x = NULL, y = "Standardised Coefficient (signed)",
       title = "Direction & Magnitude of Effect: National",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_dir, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```


## Brighton & Hove variable importance

The same standardised-coefficient approach is repeated using only
Brighton & Hove schools. Because the local predictor distributions may
differ from the national picture (e.g. lower absence variance, different
FSM range), the rankings may shift.

```{r importance-bh}
#| label: importance-bh

# Filter to Brighton & Hove
bh_data <- imputed_full_data %>%
  filter(LANAME == "Brighton and Hove")

cat("Brighton & Hove observations:", nrow(bh_data), "\n")

# Use the NATIONAL model coefficients but LOCAL predictor SDs
imp_bh_all <- standardised_importance(mod_e_all, bh_data,
                                       "ATT8SCR", "All Pupils")
imp_bh_dis <- standardised_importance(mod_e_disadv, bh_data,
                                       "ATT8SCR_FSM6CLA1A", "Disadvantaged")
imp_bh_non <- standardised_importance(mod_e_nondisadv, bh_data,
                                       "ATT8SCR_NFSM6CLA1A", "Non-Disadvantaged")

importance_bh <- bind_rows(imp_bh_all, imp_bh_dis, imp_bh_non) %>%
  mutate(display_name = display_names[term])
```

### Brighton & Hove ranking table

```{r importance-bh-table}
#| label: importance-bh-table

bh_ranking <- importance_bh %>%
  select(display_name, Group, std_coef) %>%
  pivot_wider(names_from = Group, values_from = std_coef) %>%
  mutate(
    `Avg |Std. Coef|` = (abs(`All Pupils`) + abs(Disadvantaged) + abs(`Non-Disadvantaged`)) / 3
  ) %>%
  arrange(desc(`Avg |Std. Coef|`)) %>%
  mutate(Rank = row_number()) %>%
  relocate(Rank) %>%
  mutate(across(where(is.numeric) & !matches("Rank"), \(x) round(x, 4)))

knitr::kable(
  bh_ranking,
  align = "rlrrrr",
  caption = "Brighton & Hove variable importance: standardised coefficients (local predictor SDs, national model)"
)
```

### Brighton & Hove importance chart

```{r importance-bh-chart}
#| label: importance-bh-chart
#| fig-height: 6
#| fig-width: 9

bh_chart_data <- importance_bh %>%
  mutate(
    Group = factor(Group, levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged")),
    tooltip_text = paste0(
      "<b>", display_name, "</b> (B&H)<br>",
      "Group: ", Group, "<br>",
      "Raw coefficient (national): ", round(coef, 5), "<br>",
      "SD of predictor (B&H): ", round(sd_x, 3), "<br>",
      "Standardised coef: ", round(std_coef, 4)
    )
  )

bh_var_order <- bh_chart_data %>%
  group_by(display_name) %>%
  summarise(avg_abs = mean(abs_std_coef)) %>%
  arrange(avg_abs) %>%
  pull(display_name)

bh_chart_data <- bh_chart_data %>%
  mutate(display_name = factor(display_name, levels = bh_var_order))

p_bh <- ggplot(bh_chart_data,
       aes(x = display_name, y = abs_std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#1565c0",
                                "Disadvantaged" = "#c62828",
                                "Non-Disadvantaged" = "#2e7d32")) +
  labs(x = NULL, y = "|Standardised Coefficient|",
       title = "Variable Importance: Brighton & Hove",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_bh, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```


## National vs Brighton & Hove comparison

```{r importance-comparison}
#| label: importance-comparison

# Compare rankings side by side (All Pupils model)
comparison_all <- national_ranking %>%
  select(display_name, `National Rank` = Rank,
         `National Std.Coef` = `All Pupils`) %>%
  left_join(
    bh_ranking %>%
      select(display_name, `B&H Rank` = Rank,
             `B&H Std.Coef` = `All Pupils`),
    by = "display_name"
  ) %>%
  mutate(`Rank Change` = `National Rank` - `B&H Rank`)

knitr::kable(
  comparison_all,
  align = "lrrrrl",
  caption = "All Pupils: National vs Brighton & Hove variable importance rankings"
)
```

```{r importance-comparison-disadv}
#| label: importance-comparison-disadv

# Same for Disadvantaged pupils
comparison_disadv <- importance_national %>%
  filter(Group == "Disadvantaged") %>%
  arrange(desc(abs_std_coef)) %>%
  mutate(`National Rank` = row_number()) %>%
  select(display_name, `National Rank`, `National Std.Coef` = std_coef) %>%
  left_join(
    importance_bh %>%
      filter(Group == "Disadvantaged") %>%
      arrange(desc(abs_std_coef)) %>%
      mutate(`B&H Rank` = row_number()) %>%
      select(display_name, `B&H Rank`, `B&H Std.Coef` = std_coef),
    by = "display_name"
  ) %>%
  mutate(`Rank Change` = `National Rank` - `B&H Rank`) %>%
  mutate(across(where(is.numeric) & !matches("Rank"), \(x) round(x, 4)))

knitr::kable(
  comparison_disadv,
  align = "lrrrrl",
  caption = "Disadvantaged Pupils: National vs Brighton & Hove variable importance rankings"
)
```

### Key differences between national and local rankings

The comparison tables above highlight where Brighton & Hove's local
context shifts the relative importance of predictors:

- **% Disadvantaged (FSM):** Brighton & Hove schools span a wide FSM
  range (from under 10% to over 50%), which may amplify or dampen this
  variable's local importance relative to the national picture depending
  on the SD of log(FSM) in the Local Authority.

- **Overall Absence:** If Brighton & Hove has a narrower spread of
  absence rates than the national distribution, absence will appear less
  important locally — not because it matters less per unit of change, but
  because schools are more similar to each other in this dimension.

- **Teacher Retention and Sickness:** Brighton & Hove's workforce
  stability profile — shaped by coastal-city labour market conditions,
  London proximity for recruitment, and academy vs maintained school
  mix — may produce a different local variance in these predictors.

- **Gorard Segregation:** As a single local authority, all B&H schools
  share the same segregation index value in any given year, so the
  within-B&H variance is zero or near-zero. This variable's local
  importance will therefore be very low. Note that even at national
  level, the Gorard segregation coefficient is **statistically
  insignificant** in the imputed full model (see commentary in
  section 7 above), so its apparent ranking in the national table
  should be interpreted with caution.


## Policy recommendations

The variable importance rankings, combined with the direction-of-effect
analysis and the model's structural features, suggest several priority
areas for intervention. These are organised by audience.

### For school leaders

**1. Absence reduction is the single biggest lever.**

Across all three pupil groups, overall absence rate (`PERCTOT`) is
consistently the strongest or among the strongest predictors of
Attainment 8. The log-transformed relationship means that gains from
reducing absence are **largest for schools with already-moderate absence
rates** (the steep part of the curve). For a school currently at 6%
absence, reducing to 5% has a larger predicted effect than a school at
10% reducing to 9%. This argues for:

- Intensive early-intervention strategies (attendance officers,
  family liaison, mentoring) before absence becomes entrenched
- Special attention to protecting low-absence schools from
  deterioration, not just tackling chronic cases
- Focus on disadvantaged pupils specifically, where the absence
  coefficient is largest

**2. Support for pupils with low prior attainment.**

The `PTPRIORLO` coefficient is strongly negative — schools serving a
higher proportion of pupils who entered secondary school with low KS2
scores see lower Attainment 8 outcomes. While schools cannot change their
intake composition, they can invest in:

- Targeted catch-up programmes (literacy, numeracy) for pupils arriving
  with below-expected KS2 levels
- Diagnostic assessment at transition from primary school to identify
  gaps early
- Small-group intervention during KS3 to prevent the attainment gap
  widening before GCSE

**3. Workforce stability matters.**

Teacher retention (`remained_in_the_same_school`) shows a positive
association with attainment: schools where more teachers stay are
predicted to achieve higher ATT8 scores. Meanwhile, teacher sickness
days show a negative association. School leaders should consider:

- Retention strategies: competitive CPD offers, workload management,
  mentoring for early-career teachers
- Wellbeing programmes targeting staff absence
- Succession planning to avoid disruptive turnover in key departments

**4. The FSM sign-flip for disadvantaged pupils.**

Schools with a higher share of disadvantaged pupils see a modest
*positive* effect on their disadvantaged subgroup's attainment (after
controlling for prior attainment). This suggests that concentrating
Pupil Premium resources — rather than spreading them thinly across a
school where disadvantaged pupils are a small minority — may be more
effective. Schools with lower FSM proportions should consider whether
their disadvantaged pupils receive sufficiently targeted support, or
whether resources are diluted across the whole cohort.


### For Brighton & Hove Council and local authority

**1. Tackle absence strategically across the Local Authority.**

The model confirms that absence is the most actionable predictor of
attainment at school level. The council can support schools through:

- Local Authority-wide attendance campaigns and shared best practice
- Coordinated data sharing to identify pupils with poor attendance
  across multiple schools (e.g. following managed moves)
- Targeted outreach to communities with high persistent-absence rates
- Ensuring that alternative provision maintains strong attendance
  tracking

**2. Address the disadvantaged attainment gap directly.**

The model shows that the coefficient magnitudes for disadvantaged pupils
are often **larger** than for non-disadvantaged — absence and prior
attainment in particular hit harder for the disadvantaged group.
Policy priorities include:

- Ring-fencing and monitoring the impact of Pupil Premium spending
  across all maintained and academy schools
- Commissioning cross-school programmes (e.g. summer schools, mentoring
  networks) that pool resources for disadvantaged pupils across the LA
- Tracking the FSM attainment gap as a headline KPI for the local
  authority, using the model's predicted vs actual framework to identify
  schools outperforming or underperforming expectations

**3. A note on segregation.**

The Gorard segregation index enters the model with a negative
coefficient (higher segregation associated with lower attainment), but
this effect is **statistically insignificant** — the standard error is
large relative to the estimate, and we cannot confidently distinguish
it from zero. While reducing segregation may be a worthwhile policy
goal for broader equity reasons, **this model does not provide reliable
evidence that reducing the Gorard index would improve Attainment 8
scores** at school level. The council should therefore not prioritise
segregation reduction *on the basis of this model alone*, though it
may of course remain a valid objective on other grounds (social
cohesion, equal access, community trust).

**4. Support workforce stability across schools.**

Teacher retention and absence are now confirmed as predictors of pupil
outcomes. The council and local partnerships can help by:

- Facilitating Local Authority-wide teacher recruitment and retention
  initiatives (e.g. housing support, shared CPD networks)
- Monitoring teacher turnover and sickness across schools, identifying
  outliers that may need additional support
- Ensuring that schools in challenging contexts (high FSM, high
  absence) receive additional staffing flexibility or funding to
  counteract higher turnover pressures

**5. Use the model to target support.**

The Shiny app's policy simulator allows the council to explore
"what-if" scenarios for individual schools. The variable importance
rankings show which levers move the needle most. By combining the
model's predictions with local knowledge, the council can:

- Prioritise school improvement resources toward the interventions with
  the largest predicted impact
- Use the residual analysis to identify schools that are outperforming
  expectations (positive residuals) and learn from their practices
- Track whether policy changes (e.g. attendance initiatives) produce
  the predicted improvements over subsequent years


### Summary: the biggest levers

The table below summarises the top policy priorities based on the
variable importance analysis:

```{r policy-summary-table}
#| label: policy-summary-table
#| echo: false

policy_summary <- tribble(
  ~Priority, ~`Variable`, ~`Who Can Act`, ~`Key Action`,
  1, "Overall Absence Rate", "School leaders + Council", "Intensive early intervention; protect low-absence schools",
  2, "% Low Prior Attainment", "School leaders", "Targeted KS2-KS3 transition catch-up programmes",
  3, "% Disadvantaged (FSM)", "School leaders + Council", "Concentrated Pupil Premium spend; monitor FSM gap as KPI",
  4, "Teacher Retention", "School leaders + Council", "Retention strategies; CPD investment; workload management",
  5, "Teacher Sickness Days", "School leaders", "Staff wellbeing programmes; absence monitoring",
  6, "% EAL", "School leaders", "Build on EAL advantage; ensure support for new arrivals",
  7, "Leadership Pay %", "School leaders", "Review leadership structure (low model impact)",
  8, "Gorard Segregation Index*", "Council", "Effect statistically insignificant — not a reliable lever in this model"
)

knitr::kable(
  policy_summary,
  align = "rlllc",
  caption = "Policy priorities ranked by variable importance"
)
```

\* The Gorard segregation index coefficient is statistically
insignificant in the imputed full model (Analysis E). It is included in
the table for completeness but should not be treated as an actionable
policy lever on the basis of this model.

> **Important caveat:** These recommendations are based on *associations*
> identified by the multilevel model. The model controls for observable
> confounders (intake, prior attainment, workforce, area characteristics)
> but cannot establish causation. A school reducing absence may not see
> exactly the predicted ATT8 gain if other unmeasured factors change
> simultaneously. The rankings and effect sizes should be treated as
> evidence-informed starting points for policy discussion, not as causal
> guarantees. Variables whose coefficients are statistically insignificant
> (notably the Gorard segregation index) should be interpreted with
> particular caution.


# Quick Reference

## Formula summary

| Analysis | Fixed Effects | Random Effects | Data | Used in App? |
|----------|--------------|----------------|------|:---:|
| A (Full Panel) | 9 predictors (5 log, 4 linear) | `(1\|year) + (1\|Ofsted) + (1\|region/LA)` | 3 years, workforce required | No |
| B (Full Per-Year) | Same 9 predictors | `(1\|Ofsted) + (1\|region/LA)` | Each year separately | No |
| C (Core Panel) | 5 predictors (3 log, 2 linear) | `(1\|year) + (1\|Ofsted) + (1\|region/LA)` | All 4 years | No |
| D (Core Per-Year) | Same 5 predictors | `(1\|Ofsted) + (1\|region/LA)` | Each year separately | No |
| **E (Imputed Full)** | **9 predictors (5 log, 4 linear)** | **`(1\|year) + (1\|Ofsted) + (1\|region/LA)`** | **All 4 years (2024-25 imputed)** | **Yes** |

## Variable key

| Variable | In formula as | Description |
|----------|--------------|-------------|
| `ATT8SCR` | `log(ATT8SCR)` | Attainment 8 score (all pupils) |
| `ATT8SCR_FSM6CLA1A` | `log(ATT8SCR_FSM6CLA1A)` | Attainment 8 score (disadvantaged) |
| `ATT8SCR_NFSM6CLA1A` | `log(ATT8SCR_NFSM6CLA1A)` | Attainment 8 score (non-disadvantaged) |
| `PTFSM6CLA1A` | `log(PTFSM6CLA1A)` | % eligible for Free School Meals |
| `PERCTOT` | `log(PERCTOT)` | % total absence |
| `PNUMEAL` | `log(PNUMEAL)` | % English as Additional Language |
| `PTPRIORLO` | `PTPRIORLO` | % low KS2 prior attainment |
| `ADMPOL_PT` | `ADMPOL_PT` | Admissions policy indicator |
| `gorard_segregation` | `gorard_segregation` | Gorard segregation index |
| `remained_in_the_same_school` | `log(remained_in_the_same_school)` | Teacher retention (%) |
| `teachers_on_leadership_pay_range_percent` | as-is | % teachers on leadership pay |
| `average_number_of_days_taken` | `log(average_number_of_days_taken)` | Average teacher absence days |
| `OFSTEDRATING_1` | `(1\|OFSTEDRATING_1)` | Ofsted rating (random intercept) |
| `gor_name` | `(1\|gor_name/LANAME)` | Government Office Region |
| `LANAME` | nested in `gor_name` | Local Authority |
| `year_label` | `(1\|year_label)` | Academic year (panel models only) |
