---
title: "Model Experiments"
subtitle: "Human-readable lmer implementations for manual inspection"
author: "Adam Dennett with some assistance from Claude Code"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: false
    theme: ucl-theme.scss
    df-print: kable
    embed-resources: true
    include-in-header: header-logos.html
    include-after-body: footer-logos.html
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup

library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(plotly)
library(here)

# Load the raw panel data (same starting point as 05_fit_model.R)
panel <- readRDS(here::here("data", "panel_data.rds"))

# Backfill missing gor_name from DfE LA/region lookup (belt-and-braces)
la_region_lookup <- read.csv(
  here::here("data", "meta", "Performancetables_130249",
             "2022-2023", "la_and_region_codes_meta.csv"),
  stringsAsFactors = FALSE
) %>%
  mutate(gor_name_broad = case_when(
    grepl("^North East",   REGION.NAME) ~ "North East",
    grepl("^North West",   REGION.NAME) ~ "North West",
    grepl("^North Yorkshire|^South and West Yorkshire", REGION.NAME) ~ "Yorkshire and the Humber",
    grepl("^East Midlands", REGION.NAME) ~ "East Midlands",
    grepl("^West Midlands", REGION.NAME) ~ "West Midlands",
    grepl("^East of England", REGION.NAME) ~ "East of England",
    grepl("^London",       REGION.NAME) ~ "London",
    grepl("^South East",   REGION.NAME) ~ "South East",
    grepl("^South West",   REGION.NAME) ~ "South West",
    TRUE ~ NA_character_
  ))

n_missing_before <- sum(is.na(panel$gor_name))
panel <- panel %>%
  left_join(la_region_lookup %>% select(LEA, gor_name_broad), by = "LEA") %>%
  mutate(gor_name = coalesce(gor_name, gor_name_broad)) %>%
  select(-gor_name_broad)
n_missing_after <- sum(is.na(panel$gor_name))

cat("Panel loaded:", nrow(panel), "rows,", ncol(panel), "columns\n")
cat("Region backfill: fixed", n_missing_before - n_missing_after, "of",
    n_missing_before, "missing gor_name\n")
```


# Overview

This document re-implements every model from `model_results.qmd` in a
deliberately simple, explicit style. Each model is written out longhand
so you can:

- see exactly which rows survive the filters
- read the `lmer()` formula as a single, self-contained call
- inspect `summary()`, `fixef()`, `ranef()`, `VarCorr()` etc. directly

The four analyses and three outcomes produce **12 models** in total.

Models are fitted with `REML = TRUE` and `optimizer = "bobyqa"` to
match the production pipeline.


# Data Preparation

Three datasets are needed:

1. A **full** dataset for Analysis A (9 predictors, **3 years only** — excluding 2024-25, which has carry-forward imputed workforce variables)
2. A **core** dataset (5 predictors, all 4 years including 2024-25)
3. An **imputed full** dataset for Analysis E (9 predictors, all 4 years — including 2024-25 with imputed workforce variables)

Note: `panel_data.rds` already contains carry-forward imputed workforce variables for 2024-25 (from `04_compute_derived.R`). To create a genuine 3-year dataset for Analysis A, we must explicitly exclude 2024-25.

## Full model dataset (3 years, real data only)

```{r data-full}
#| label: data-full

# All 4 years with complete workforce data (used for Analysis E)
all_model_data <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(
    !is.na(ATT8SCR), ATT8SCR > 0,
    PTFSM6CLA1A > 0,
    PERCTOT > 0,
    PNUMEAL > 0,
    !is.na(OFSTEDRATING_1),
    !is.na(gor_name),
    !is.na(LANAME)
  ) %>%
  filter(
    !is.na(remained_in_the_same_school),
    !is.na(teachers_on_leadership_pay_range_percent),
    average_number_of_days_taken > 0,
    !is.na(gorard_segregation)
  ) %>%
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

contrasts(all_model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(all_model_data$OFSTEDRATING_1))

# Analysis A: 3 years only (exclude 2024-25 which has imputed workforce values)
model_data <- all_model_data %>%
  filter(year_label != "2024-25") %>%
  droplevels()

contrasts(model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(model_data$OFSTEDRATING_1))

cat("Full model dataset (Analysis A, 3 years):", nrow(model_data), "rows\n")
cat("Years:", paste(levels(model_data$year_label), collapse = ", "), "\n")
cat("Regions:", n_distinct(model_data$gor_name), "\n")
cat("LAs:", n_distinct(model_data$LANAME), "\n")
cat("Ofsted levels:", paste(levels(model_data$OFSTEDRATING_1), collapse = ", "), "\n")
```

::: {.callout-note}
## Sample attrition: teacher sickness days

The full model dataset loses approximately 570 school-year observations
due to the `average_number_of_days_taken > 0` filter. This variable is
required to be strictly positive for the log transform used in the model.
The losses are concentrated in 2021-22 (250 rows) and 2022-23 (244 rows),
with far fewer in 2023-24 (54) and 2024-25 (22). Around 420 of these are
NA values — schools where the workforce census returned retention and
leadership data but not sickness absence — and the remaining ~150 are
genuine zeros. Academies are disproportionately affected (80% of losses),
likely reflecting patchier sickness absence reporting in the early
post-COVID workforce census returns. This is a known data limitation;
the dropped schools are spread across all regions and LAs rather than
concentrated in particular areas.
:::

## Core model dataset

```{r data-core}
#| label: data-core

core_model_data <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(
    !is.na(ATT8SCR), ATT8SCR > 0,
    PTFSM6CLA1A > 0,
    PERCTOT > 0,
    PNUMEAL > 0,
    !is.na(ADMPOL_PT),
    !is.na(gorard_segregation),
    !is.na(OFSTEDRATING_1),
    !is.na(gor_name),
    !is.na(LANAME)
  ) %>%
  mutate(
    OFSTEDRATING_1 = factor(OFSTEDRATING_1,
                            levels = c("Outstanding", "Good",
                                       "Requires Improvement", "Inadequate"),
                            ordered = TRUE),
    gor_name   = factor(gor_name),
    LANAME     = factor(LANAME),
    year_label = factor(year_label)
  ) %>%
  droplevels()

contrasts(core_model_data$OFSTEDRATING_1) <-
  contr.treatment(levels(core_model_data$OFSTEDRATING_1))

cat("Core model dataset:", nrow(core_model_data), "rows\n")
cat("Years:", paste(levels(core_model_data$year_label), collapse = ", "), "\n")
```

## Dropped-rows diagnostic

Which school-year observations in the raw panel were excluded from the
**full** model dataset, and why? The filters are applied sequentially
below so we can attribute each dropped row to the *first* filter that
removes it.

```{r dropped-rows}
#| label: dropped-rows
#| code-fold: true
#| code-summary: "Show diagnostic code"

# Start from the same mainstream filter used above
mainstream <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school"))

cat("Starting mainstream rows:", nrow(mainstream), "\n\n")

# Apply filters sequentially and tag each row with the reason it was dropped
d_diag <- mainstream %>%
  mutate(
    # Step 1: Outcome missing or zero
    fail_outcome = is.na(ATT8SCR) | ATT8SCR <= 0,
    # Step 2: FSM missing or zero (can't log-transform)
    fail_fsm = !fail_outcome & (is.na(PTFSM6CLA1A) | PTFSM6CLA1A <= 0),
    # Step 3: Absence missing or zero
    fail_absence = !fail_outcome & !fail_fsm & (is.na(PERCTOT) | PERCTOT <= 0),
    # Step 4: EAL missing or zero
    fail_eal = !fail_outcome & !fail_fsm & !fail_absence & (is.na(PNUMEAL) | PNUMEAL <= 0),
    # Step 5: Ofsted rating missing
    fail_ofsted = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal & is.na(OFSTEDRATING_1),
    # Step 6: Region or LA missing
    fail_geo = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal & !fail_ofsted &
               (is.na(gor_name) | is.na(LANAME)),
    # Step 7: Teacher retention missing (workforce — drops 2024-25)
    fail_retention = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal &
                     !fail_ofsted & !fail_geo & is.na(remained_in_the_same_school),
    # Step 8: Leadership pay missing
    fail_leadership = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal &
                      !fail_ofsted & !fail_geo & !fail_retention &
                      is.na(teachers_on_leadership_pay_range_percent),
    # Step 9: Teacher sickness days missing or zero
    fail_sickness = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal &
                    !fail_ofsted & !fail_geo & !fail_retention & !fail_leadership &
                    (is.na(average_number_of_days_taken) | average_number_of_days_taken <= 0),
    # Step 10: Gorard segregation index missing
    fail_gorard = !fail_outcome & !fail_fsm & !fail_absence & !fail_eal &
                  !fail_ofsted & !fail_geo & !fail_retention & !fail_leadership &
                  !fail_sickness & is.na(gorard_segregation),
    # Assign first reason
    drop_reason = case_when(
      fail_outcome    ~ "ATT8SCR missing/zero",
      fail_fsm        ~ "PTFSM6CLA1A missing/zero",
      fail_absence    ~ "PERCTOT missing/zero",
      fail_eal        ~ "PNUMEAL missing/zero",
      fail_ofsted     ~ "OFSTEDRATING missing",
      fail_geo        ~ "Region/LA missing",
      fail_retention  ~ "Teacher retention missing",
      fail_leadership ~ "Leadership pay missing",
      fail_sickness   ~ "Sickness days missing/zero",
      fail_gorard     ~ "Gorard index missing",
      TRUE            ~ "Retained"
    )
  )

# Summary table of drop reasons
reason_summary <- d_diag %>%
  count(drop_reason) %>%
  arrange(desc(n)) %>%
  mutate(pct = round(100 * n / sum(n), 1))

knitr::kable(reason_summary,
             col.names = c("Reason", "Rows", "% of mainstream"),
             caption = "Row attrition: first filter that removes each school-year observation")
```

```{r dropped-detail}
#| label: dropped-detail
#| code-fold: true
#| code-summary: "Show dropped-rows detail code"

dropped <- d_diag %>%
  filter(drop_reason != "Retained")

cat(nrow(dropped), "rows dropped from", n_distinct(dropped$SCHNAME),
    "unique schools across", n_distinct(dropped$LANAME), "LAs\n\n")

# Which years are most affected?
cat("Dropped rows by year:\n")
dropped %>% count(year_label, drop_reason) %>%
  pivot_wider(names_from = year_label, values_from = n, values_fill = 0) %>%
  arrange(drop_reason) %>%
  as.data.frame() %>%
  print()
cat("\n")

# Top LAs losing the most rows
la_drops <- dropped %>%
  count(LANAME, drop_reason) %>%
  group_by(LANAME) %>%
  summarise(
    total_dropped = sum(n),
    reasons = paste(sprintf("%s (%d)", drop_reason, n), collapse = "; "),
    .groups = "drop"
  ) %>%
  arrange(desc(total_dropped))

cat("Top 20 LAs by rows dropped:\n")
knitr::kable(head(la_drops, 20),
             col.names = c("LA", "Rows dropped", "Reasons (count)"),
             caption = "Local authorities losing the most observations")
```

```{r dropped-schools}
#| label: dropped-schools
#| code-fold: true
#| code-summary: "Show dropped-schools detail code"

# Schools that are *entirely* absent from the final model dataset
retained_schools <- d_diag %>% filter(drop_reason == "Retained") %>%
  distinct(SCHNAME) %>% pull(SCHNAME)

fully_dropped <- dropped %>%
  filter(!SCHNAME %in% retained_schools) %>%
  group_by(SCHNAME, LANAME) %>%
  summarise(
    years_present = paste(sort(unique(year_label)), collapse = ", "),
    n_rows = n(),
    reasons = paste(unique(drop_reason), collapse = "; "),
    mean_att8 = if (any(!is.na(ATT8SCR))) round(mean(ATT8SCR, na.rm = TRUE), 1) else NA,
    has_workforce = any(!is.na(remained_in_the_same_school)),
    has_gorard = any(!is.na(gorard_segregation)),
    ofsted = first(na.omit(OFSTEDRATING_1)),
    .groups = "drop"
  ) %>%
  arrange(LANAME, SCHNAME)

cat(nrow(fully_dropped), "schools are entirely excluded from the full model dataset:\n\n")

knitr::kable(fully_dropped,
             col.names = c("School", "LA", "Years in panel", "Rows", "Drop reasons",
                           "Mean ATT8", "Has workforce", "Has Gorard", "Ofsted"),
             caption = "Schools with zero rows in the final model dataset — entirely excluded from modelling")
```

```{r dropped-characteristics}
#| label: dropped-characteristics
#| code-fold: true
#| code-summary: "Show characteristics comparison"
#| fig-height: 5
#| fig-width: 9
#| fig-cap: "Comparison of key characteristics between retained and dropped observations. Dropped rows tend to have more missing data and different school profiles."

# Compare characteristics of retained vs dropped
compare <- d_diag %>%
  mutate(status = if_else(drop_reason == "Retained", "Retained", "Dropped")) %>%
  select(status, ATT8SCR, PTFSM6CLA1A, PERCTOT, PNUMEAL, TPUP,
         MINORGROUP, year_label) %>%
  pivot_longer(c(ATT8SCR, PTFSM6CLA1A, PERCTOT, PNUMEAL, TPUP),
               names_to = "variable", values_to = "value") %>%
  filter(!is.na(value))

var_labels <- c(
  ATT8SCR = "Attainment 8",
  PTFSM6CLA1A = "% FSM Eligible",
  PERCTOT = "% Overall Absence",
  PNUMEAL = "% EAL",
  TPUP = "Total Pupils"
)

compare <- compare %>%
  mutate(variable = factor(variable, levels = names(var_labels), labels = var_labels))

ggplot(compare, aes(x = status, y = value, fill = status)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 5) +
  scale_fill_manual(values = c("Retained" = "#2e6260", "Dropped" = "#7b132b")) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Retained vs Dropped observations: key characteristics") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))
```


# Analysis A: Full Panel Models

All years pooled, `year_label` as a random intercept. Nine fixed-effect
predictors (5 log-transformed, 4 linear).

## All Pupils

```{r a-all}
#| label: a-all

d <- model_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_all)
```

### Quick checks

```{r a-all-checks}
#| label: a-all-checks

cat("Fixed effects:\n")
fixef(mod_a_all)

cat("\nVariance components:\n")
print(VarCorr(mod_a_all))

cat("\nR-squared:\n")
print(r2(mod_a_all))

cat("\nSingular fit?", isSingular(mod_a_all), "\n")
```

## Disadvantaged Pupils

```{r a-disadv}
#| label: a-disadv

d <- model_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_disadv)
```

### Quick checks

```{r a-disadv-checks}
#| label: a-disadv-checks

fixef(mod_a_disadv)
print(VarCorr(mod_a_disadv))
print(r2(mod_a_disadv))
cat("Singular fit?", isSingular(mod_a_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r a-nondisadv}
#| label: a-nondisadv

d <- model_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_a_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_a_nondisadv)
```

### Quick checks

```{r a-nondisadv-checks}
#| label: a-nondisadv-checks

fixef(mod_a_nondisadv)
print(VarCorr(mod_a_nondisadv))
print(r2(mod_a_nondisadv))
cat("Singular fit?", isSingular(mod_a_nondisadv), "\n")
```


# Analysis B: Full Per-Year Models

Same 9 fixed effects as Analysis A, but fitted separately for each year.
No `year_label` random effect.

```{r b-setup}
#| label: b-setup

year_levels <- sort(unique(as.character(model_data$year_label)))
cat("Years to fit:", paste(year_levels, collapse = ", "), "\n")
```

## All Pupils (per year)

```{r b-all}
#| label: b-all

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- All Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr, !is.na(ATT8SCR), ATT8SCR > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      remained_in_the_same_school +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Disadvantaged Pupils (per year)

```{r b-disadv}
#| label: b-disadv

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Disadvantaged Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_FSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      remained_in_the_same_school +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Non-Disadvantaged Pupils (per year)

```{r b-nondisadv}
#| label: b-nondisadv

for (yr in year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Non-Disadvantaged Pupils\n")
  cat(strrep("=", 60), "\n")

  d <- model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_NFSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      PTPRIORLO + ADMPOL_PT + gorard_segregation +
      remained_in_the_same_school +
      teachers_on_leadership_pay_range_percent +
      log(average_number_of_days_taken) +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```


# Analysis C: Core Panel Models

Reduced specification with 5 fixed effects. All 4 years pooled
(including 2024-25). `year_label` as a random intercept.

## All Pupils

```{r c-all}
#| label: c-all

d <- core_model_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_all)
```

### Quick checks

```{r c-all-checks}
#| label: c-all-checks

fixef(mod_c_all)
print(VarCorr(mod_c_all))
print(r2(mod_c_all))
cat("Singular fit?", isSingular(mod_c_all), "\n")
```

## Disadvantaged Pupils

```{r c-disadv}
#| label: c-disadv

d <- core_model_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_disadv)
```

### Quick checks

```{r c-disadv-checks}
#| label: c-disadv-checks

fixef(mod_c_disadv)
print(VarCorr(mod_c_disadv))
print(r2(mod_c_disadv))
cat("Singular fit?", isSingular(mod_c_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r c-nondisadv}
#| label: c-nondisadv

d <- core_model_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_c_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    ADMPOL_PT + gorard_segregation +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_c_nondisadv)
```

### Quick checks

```{r c-nondisadv-checks}
#| label: c-nondisadv-checks

fixef(mod_c_nondisadv)
print(VarCorr(mod_c_nondisadv))
print(r2(mod_c_nondisadv))
cat("Singular fit?", isSingular(mod_c_nondisadv), "\n")
```


# Analysis D: Core Per-Year Models

Same 5 core fixed effects, fitted separately for each year. No
`year_label` random effect.

```{r d-setup}
#| label: d-setup

core_year_levels <- sort(unique(as.character(core_model_data$year_label)))
cat("Years to fit:", paste(core_year_levels, collapse = ", "), "\n")
```

## All Pupils (per year)

```{r d-all}
#| label: d-all

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- All Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr, !is.na(ATT8SCR), ATT8SCR > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Disadvantaged Pupils (per year)

```{r d-disadv}
#| label: d-disadv

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Disadvantaged Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_FSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```

## Non-Disadvantaged Pupils (per year)

```{r d-nondisadv}
#| label: d-nondisadv

for (yr in core_year_levels) {

  cat("\n", strrep("=", 60), "\n")
  cat("YEAR:", yr, "- Non-Disadvantaged Pupils (core)\n")
  cat(strrep("=", 60), "\n")

  d <- core_model_data %>%
    filter(year_label == yr,
           !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
    droplevels()

  contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

  cat("Observations:", nrow(d), "\n")

  mod <- lmer(
    log(ATT8SCR_NFSM6CLA1A) ~
      log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
      ADMPOL_PT + gorard_segregation +
      (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
    data = d,
    REML = TRUE,
    na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
  )

  print(summary(mod))
  cat("\nR-squared:\n")
  print(r2(mod))
  cat("\nSingular fit?", isSingular(mod), "\n")
}
```


# Analysis E: Full Panel with Imputed 2024-25 Predictors

The full 9-predictor specification can now cover all 4 years because
`04_compute_derived.R` carry-forward imputes PTPRIORLO and three
workforce variables for 2024-25 (using each school's 2023-24 value, or
its 3-year mean as a fallback). A flag column `has_imputed_predictors`
marks which rows use estimates.

```{r e-setup}
#| label: e-setup

# Analysis E uses all 4 years including 2024-25 (with carry-forward imputed
# workforce variables). This is simply `all_model_data` — no bind_rows needed
# because panel_data.rds already contains the imputed values.
imputed_full_data <- all_model_data

n_2425 <- sum(imputed_full_data$year_label == "2024-25")

cat("Imputed full dataset:", nrow(imputed_full_data), "rows\n")
cat("Years:", paste(levels(imputed_full_data$year_label), collapse = ", "), "\n")
cat("Of which 2024-25 (imputed predictors):", n_2425, "rows\n")
cat("Analysis A had:", nrow(model_data), "rows (3 years)\n")
cat("Difference (new 2024-25 rows):", nrow(imputed_full_data) - nrow(model_data), "\n")
```

## All Pupils

```{r e-all}
#| label: e-all

d <- imputed_full_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_e_all <- lmer(
  log(ATT8SCR) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_e_all)
```

### Quick checks

```{r e-all-checks}
#| label: e-all-checks

fixef(mod_e_all)
print(VarCorr(mod_e_all))
print(r2(mod_e_all))
cat("Singular fit?", isSingular(mod_e_all), "\n")
```

## Disadvantaged Pupils

```{r e-disadv}
#| label: e-disadv

d <- imputed_full_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_e_disadv <- lmer(
  log(ATT8SCR_FSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_e_disadv)
```

### Quick checks

```{r e-disadv-checks}
#| label: e-disadv-checks

fixef(mod_e_disadv)
print(VarCorr(mod_e_disadv))
print(r2(mod_e_disadv))
cat("Singular fit?", isSingular(mod_e_disadv), "\n")
```

## Non-Disadvantaged Pupils

```{r e-nondisadv}
#| label: e-nondisadv

d <- imputed_full_data %>%
  filter(!is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

cat("Observations:", nrow(d), "\n")

mod_e_nondisadv <- lmer(
  log(ATT8SCR_NFSM6CLA1A) ~
    log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
    PTPRIORLO + ADMPOL_PT + gorard_segregation +
    remained_in_the_same_school +
    teachers_on_leadership_pay_range_percent +
    log(average_number_of_days_taken) +
    (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
  data = d,
  REML = TRUE,
  na.action = na.exclude,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))
)

summary(mod_e_nondisadv)
```

### Quick checks

```{r e-nondisadv-checks}
#| label: e-nondisadv-checks

fixef(mod_e_nondisadv)
print(VarCorr(mod_e_nondisadv))
print(r2(mod_e_nondisadv))
cat("Singular fit?", isSingular(mod_e_nondisadv), "\n")
```

## Analysis A vs E: coefficient comparison

```{r e-vs-a}
#| label: e-vs-a

# Helper: relabel raw lmer coefficient names to human-readable form
relabel_terms <- function(term) {
  case_when(
    term == "(Intercept)" ~ "(Intercept)",
    term == "log(PTFSM6CLA1A)" ~ "log(% FSM)",
    term == "log(PERCTOT)" ~ "log(% Absence)",
    term == "log(PNUMEAL)" ~ "log(% EAL)",
    term == "PTPRIORLO" ~ "% Low Prior Attainment",
    term == "gorard_segregation" ~ "Gorard Segregation",
    term == "ADMPOL_PTOTHER NON SEL" ~ "Admissions: Other non-selective (ref: non-sel. in highly sel. area)",
    term == "ADMPOL_PTSEL" ~ "Admissions: Selective (ref: non-sel. in highly sel. area)",
    term == "remained_in_the_same_school" ~ "Teacher Retention",
    term == "teachers_on_leadership_pay_range_percent" ~ "Leadership Pay %",
    term == "log(average_number_of_days_taken)" ~ "log(Teacher Sickness)",
    TRUE ~ term
  )
}

a_vs_e <- bind_rows(
  tibble(Analysis = "A (Real, 3 yrs)",
         Term = names(fixef(mod_a_all)),
         `All Pupils` = fixef(mod_a_all),
         Disadvantaged = fixef(mod_a_disadv),
         `Non-Disadvantaged` = fixef(mod_a_nondisadv)),
  tibble(Analysis = "E (Imputed, 4 yrs)",
         Term = names(fixef(mod_e_all)),
         `All Pupils` = fixef(mod_e_all),
         Disadvantaged = fixef(mod_e_disadv),
         `Non-Disadvantaged` = fixef(mod_e_nondisadv))
) %>%
  mutate(Term = relabel_terms(Term),
         across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(a_vs_e, align = "llrrr",
             caption = "Analysis A (real data, 3 years) vs Analysis E (with imputed 2024-25)")
```

The table above shows that adding one year of carry-forward imputed data
(Analysis E) produces coefficients that are **very close to Analysis A**.
This is reassuring: three-quarters of the data is identical, and the
imputed fourth year uses conservative carry-forward values. Crucially,
the **FSM sign flip for disadvantaged pupils** — positive in the full
model — should be preserved in Analysis E, confirming it is a genuine
structural feature of the data when prior attainment is controlled for.


# Coefficient Commentary

This section examines the coefficients from the **imputed full model**
(Analysis E: 9 predictors, all 4 years including 2024-25 with
carry-forward imputed PTPRIORLO and workforce variables). This is the
model now used in the Shiny app. For reference, the core panel model
(Analysis C: 5 predictors) and the original 3-year full model
(Analysis A) are compared below.


## Imputed full model: side-by-side coefficients

The imputed full model (Analysis E) includes all 9 fixed effects and
covers all four academic years. For 2024-25, `PTPRIORLO` and three
workforce variables are carry-forward imputed from each school's
2023-24 value (or 3-year mean as fallback). Rows using imputed values
are flagged with `has_imputed_predictors = TRUE`.

```{r commentary-imputed-coefs}
#| label: commentary-imputed-coefs

# Extract fixed effects from the three imputed full panel models (Analysis E)
imputed_coef_table <- bind_rows(
  tibble(Group = "All Pupils",
         Term = names(fixef(mod_e_all)),
         Estimate = fixef(mod_e_all)),
  tibble(Group = "Disadvantaged",
         Term = names(fixef(mod_e_disadv)),
         Estimate = fixef(mod_e_disadv)),
  tibble(Group = "Non-Disadvantaged",
         Term = names(fixef(mod_e_nondisadv)),
         Estimate = fixef(mod_e_nondisadv))
) %>%
  pivot_wider(names_from = Group, values_from = Estimate) %>%
  mutate(Term = relabel_terms(Term),
         across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(imputed_coef_table, align = "lrrr",
             caption = "Imputed full model (Analysis E): fixed effect comparison (9 predictors, all 4 years)")
```

## Core panel model: side-by-side coefficients

The core panel model (Analysis C) drops the four variables not available
for 2024-25 — `PTPRIORLO`, `remained_in_the_same_school`,
`teachers_on_leadership_pay_range_percent`, and
`average_number_of_days_taken` — leaving 5 fixed effects.

```{r commentary-core-coefs}
#| label: commentary-core-coefs

# Extract fixed effects from the three core panel models (Analysis C)
core_coef_table <- bind_rows(
  tibble(Group = "All Pupils",
         Term = names(fixef(mod_c_all)),
         Estimate = fixef(mod_c_all)),
  tibble(Group = "Disadvantaged",
         Term = names(fixef(mod_c_disadv)),
         Estimate = fixef(mod_c_disadv)),
  tibble(Group = "Non-Disadvantaged",
         Term = names(fixef(mod_c_nondisadv)),
         Estimate = fixef(mod_c_nondisadv))
) %>%
  pivot_wider(names_from = Group, values_from = Estimate) %>%
  mutate(Term = relabel_terms(Term),
         across(where(is.numeric), \(x) round(x, 5)))

knitr::kable(core_coef_table, align = "lrrr",
             caption = "Core panel model (Analysis C): fixed effect comparison (5 predictors, all 4 years)")
```

## Imputed full vs core: direct comparison

The table below places the 5 shared predictors side by side for both
specifications, making it easy to see how dropping the workforce and
prior-attainment variables shifts the remaining coefficients.

```{r commentary-imputed-vs-core}
#| label: commentary-imputed-vs-core

# Shared terms between imputed full and core (excluding intercept)
shared_terms <- intersect(
  core_coef_table$Term[core_coef_table$Term != "(Intercept)"],
  imputed_coef_table$Term[imputed_coef_table$Term != "(Intercept)"]
)

comparison <- imputed_coef_table %>%
  filter(Term %in% shared_terms) %>%
  rename(`Imputed Full: All` = `All Pupils`,
         `Imputed Full: Disadv.` = Disadvantaged,
         `Imputed Full: Non-Dis.` = `Non-Disadvantaged`) %>%
  left_join(
    core_coef_table %>%
      filter(Term %in% shared_terms) %>%
      rename(`Core: All` = `All Pupils`,
             `Core: Disadv.` = Disadvantaged,
             `Core: Non-Dis.` = `Non-Disadvantaged`),
    by = "Term"
  )

knitr::kable(comparison, align = "lrrrrrr",
             caption = "Shared predictor coefficients: imputed full (E) vs core panel (C)")
```

## Coefficient stability over time (full model)

The per-year full models (Analysis B) show how each of the 9 predictor
coefficients evolves across the three years with real data. The static
ggplot is followed by an interactive plotly version for easier
exploration.

```{r commentary-yearly}
#| label: commentary-yearly
#| fig-height: 10
#| fig-width: 9
#| fig-cap: "Full per-year coefficient estimates (Analysis B), by outcome group"

# Re-fit full per-year models and collect coefficients in a tidy table
yearly_coefs <- list()

for (yr in year_levels) {
  for (outcome_name in c("all", "disadv", "nondisadv")) {

    outcome_var <- switch(outcome_name,
                          all = "ATT8SCR",
                          disadv = "ATT8SCR_FSM6CLA1A",
                          nondisadv = "ATT8SCR_NFSM6CLA1A")
    group_label <- switch(outcome_name,
                          all = "All Pupils",
                          disadv = "Disadvantaged",
                          nondisadv = "Non-Disadvantaged")

    d <- model_data %>%
      filter(year_label == yr,
             !is.na(!!sym(outcome_var)), !!sym(outcome_var) > 0) %>%
      droplevels()

    contrasts(d$OFSTEDRATING_1) <- contr.treatment(levels(d$OFSTEDRATING_1))

    mod <- tryCatch(
      lmer(
        as.formula(paste0(
          "log(", outcome_var, ") ~ ",
          "log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) + ",
          "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
          "remained_in_the_same_school + ",
          "teachers_on_leadership_pay_range_percent + ",
          "log(average_number_of_days_taken) + ",
          "(1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
        )),
        data = d, REML = TRUE, na.action = na.exclude,
        control = lmerControl(optimizer = "bobyqa",
                              optCtrl = list(maxfun = 20000))
      ),
      error = function(e) NULL
    )

    if (!is.null(mod)) {
      fe <- fixef(mod)
      yearly_coefs[[length(yearly_coefs) + 1]] <- tibble(
        Year = yr,
        Group = group_label,
        Term = names(fe),
        Estimate = fe
      )
    }
  }
}

yearly_coefs_df <- bind_rows(yearly_coefs) %>%
  filter(Term != "(Intercept)") %>%
  mutate(
    Term = relabel_terms(Term),
    Group = factor(Group,
                   levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged"))
  )

ggplot(yearly_coefs_df,
       aes(x = Year, y = Estimate, colour = Group, group = Group)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ Term, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("All Pupils" = "#2e6260",
                                  "Disadvantaged" = "#4e3c56",
                                  "Non-Disadvantaged" = "#abc766")) +
  labs(y = "Coefficient estimate", x = "Academic year", colour = NULL,
       title = "Full per-year coefficient trajectories (Analysis B)") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "bold", size = 9)
  )
```

### Interactive coefficient explorer

```{r commentary-yearly-plotly}
#| label: commentary-yearly-plotly
#| fig-height: 10
#| fig-width: 9

# Terms are already relabelled — build tooltips directly
yearly_coefs_plotly <- yearly_coefs_df %>%
  mutate(
    tooltip_text = paste0(
      "<b>", Term, "</b><br>",
      "Year: ", Year, "<br>",
      "Group: ", Group, "<br>",
      "Estimate: ", round(Estimate, 5)
    )
  )

p_yearly <- ggplot(yearly_coefs_plotly,
       aes(x = Year, y = Estimate, colour = Group, group = Group,
           text = tooltip_text)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~ Term, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("All Pupils" = "#2e6260",
                                  "Disadvantaged" = "#4e3c56",
                                  "Non-Disadvantaged" = "#abc766")) +
  labs(y = "Coefficient estimate", x = "Academic year", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    strip.text = element_text(face = "bold", size = 8)
  )

ggplotly(p_yearly, tooltip = "text") %>%
  layout(
    title = list(text = "Full per-year coefficients (hover for details)"),
    legend = list(orientation = "h", y = -0.15)
  )
```

## General observations

The commentary below draws on the imputed full model (Analysis E)
coefficient table and the per-year trajectories above. Because the
outcome is **log-transformed**, all coefficients are **elasticities**
(for log-transformed predictors) or **semi-elasticities** (for linear
predictors). A coefficient of --0.10 on a logged predictor means that
a 1% increase in that predictor is associated with roughly a 0.10%
*decrease* in Attainment 8.

### 1. Imputed full vs core model: how dropping variables changes the picture

The most important thing to note when comparing the two specifications
is that they do **not tell the same story** for every coefficient. The
imputed full model (Analysis E) includes `PTPRIORLO` (% low KS2 prior
attainment) and three workforce variables that the core model lacks.
These omitted variables are **not random noise** — prior attainment in
particular is one of the strongest individual predictors of school-level
outcomes. When it is dropped from the core model, its explanatory power
has to go somewhere, and the remaining 5 predictors absorb it. This
produces some notable shifts:

- **`log(PTFSM6CLA1A)` (% FSM):** In the **imputed full model**, this
  coefficient is **negative for All Pupils and Non-Disadvantaged**, and
  **flips to positive for Disadvantaged Pupils**. In the **core model**,
  FSM is negative across **all three groups** — the sign flip
  disappears. This is because prior attainment (`PTPRIORLO`) is strongly
  correlated with FSM: schools with more disadvantaged intakes also tend
  to have more pupils with low prior attainment. In the full model,
  `PTPRIORLO` picks up the prior-attainment component of the
  disadvantage penalty, freeing up `log(PTFSM6CLA1A)` to capture the
  *residual* relationship — which, for disadvantaged pupils specifically,
  is positive (the "critical mass" effect described below). When
  `PTPRIORLO` is absent, `log(PTFSM6CLA1A)` has to absorb both effects
  and the net result is negative.

- **`log(PERCTOT)` (% absence):** Becomes **more negative** in the core
  model. Again, with fewer controls absorbing variation, absence picks up
  some of the explanatory burden previously shared with workforce and
  prior-attainment variables.

- **`log(PNUMEAL)` (% EAL):** The sign and direction are broadly
  consistent across both specifications, but the magnitude changes.

The practical implication is that coefficient interpretation must always
be read **relative to the other variables in the model**. Now that the
Shiny app uses the imputed full model, users get the benefit of all 9
predictors — including the workforce and prior-attainment controls —
across all four years. The 2024-25 values for the four imputed variables
use conservative carry-forward estimates, and this is flagged in the data.

### 2. Free school meals (FSM) — the sign flip

In the **imputed full model** (Analysis E), `log(PTFSM6CLA1A)` shows
the most striking difference across outcome groups:

- **All Pupils and Non-Disadvantaged:** The coefficient is consistently
  **negative**. Schools with a higher share of disadvantaged pupils tend
  to have lower overall and non-disadvantaged ATT8 scores, even after
  controlling for absence, EAL, prior attainment, and segregation. This
  captures the well documented compositional effect: a more disadvantaged
  intake is associated with lower average attainment at school level.

- **Disadvantaged Pupils:** The coefficient **flips to positive**. After
  controlling for prior attainment and workforce factors, schools with a
  higher proportion of FSM-eligible pupils see their *disadvantaged
  subgroup specifically* score somewhat higher than predicted. One
  plausible explanation is that schools serving predominantly
  disadvantaged cohorts concentrate more resources, Pupil Premium
  funding, and pastoral support on those pupils — a kind of "critical
  mass" effect. Alternatively, in schools where almost all pupils are
  disadvantaged, the disadvantaged group essentially *is* the school
  average, removing the within-school composition penalty.

**Crucially, this sign flip only appears in the full model** where prior
attainment controls for the low-KS2 component of the disadvantage
penalty. In the core model (Analysis C), `log(PTFSM6CLA1A)` is
**negative for all three groups** because it is forced to absorb the
prior-attainment effect as well. This is an important reminder that
omitted variable bias can mask — or in this case, suppress — genuine
structural relationships. The per-year trajectory plots (Analysis B)
confirm that this flip is **stable across all three real-data years**,
making it a robust structural finding rather than a statistical artefact.

### 3. Absence — the largest and most stable predictor

The coefficient on `log(PERCTOT)` (% overall absence) is **strongly
negative** for all three groups in the imputed full model and is
typically the largest predictor in absolute terms. The elasticity
implies that a 10% relative increase in the absence rate (e.g. from
5.0% to 5.5%) is associated with a substantial drop in ATT8 on the
log scale.

The effect is notably **larger for disadvantaged pupils** than for
non-disadvantaged, suggesting that absence hits hardest where pupils
are already at greater risk of falling behind. The per-year plots
confirm this pattern is consistent across all years.

### 4. English as an additional language (EAL)

`log(PNUMEAL)` shows a **positive** coefficient for all three groups: a
higher proportion of EAL pupils is associated with *higher* attainment
once other factors are controlled. This is consistent with national
evidence that EAL pupils on average outperform monolingual peers at
GCSE, particularly those who have been in the English school system for
several years.

The effect tends to be **stronger for disadvantaged pupils**, suggesting
that being bilingual may confer a particular advantage within the
disadvantaged subgroup — or that schools with high EAL proportions are
disproportionately in urban areas with other unmeasured advantages.

### 5. Prior attainment (PTPRIORLO)

The `PTPRIORLO` coefficient is **strongly negative** across all three
groups — a higher percentage of pupils with low KS2 prior attainment
is associated with lower Attainment 8 scores. This is the most
intuitive relationship in the model: schools receiving pupils with
weaker primary-level preparation see lower secondary outcomes, even
after controlling for all other factors.

The magnitude is broadly similar across outcome groups, though the
coefficient for non-disadvantaged pupils tends to be slightly smaller,
consistent with higher-attaining pupils being somewhat more resilient
to a low-prior-attainment school environment.

### 6. Interpreting log-transformed variables — effects at the extremes

Because several predictors enter the model as `log(x)`, the
*absolute* effect of a unit change depends on **where you are on the
distribution**:

- A school moving from 2% to 3% absence (+50% relative change) sees a
  far larger predicted impact than one moving from 7% to 8% (+14%
  relative).
- Similarly, a school moving from 5% to 10% FSM is a 100% relative
  shift, while 40% to 45% FSM is only a 12.5% relative shift.

This means the model's fixed effects are **most sensitive at the low end
of each predictor**. Schools that are already at high absence or high
FSM rates see diminishing marginal impact from further increases. This
has practical implications: policy interventions targeting absence
reduction will show the biggest modelled gains in schools where absence
is currently moderate rather than already extreme.

### 7. Gorard segregation index

The Gorard segregation index coefficient is **negative** for all three
groups — higher within-LA segregation is associated with lower
attainment — but the effect is **statistically insignificant** in the
imputed full model (Analysis E). The coefficient is small in absolute
terms and its standard error comfortably spans zero, meaning we cannot
confidently distinguish the estimated effect from no effect at all.
This is an important caveat: while the *direction* of the association
(more segregation → lower attainment) is consistent with broader
education research, **this model does not provide reliable evidence
that the Gorard segregation index has a meaningful independent effect
on Attainment 8 at school level**, once the other 8 predictors are
controlled for. The variable is retained in the model for completeness,
but policy recommendations should not be based on its coefficient.

### 8. Admissions policy

`ADMPOL_PT` (admissions policy type) is a **categorical variable** with three levels: *"Non-selective in a highly selective area"* (i.e. a non-selective school in an area with grammar schools), *"Other non-selective"*, and *"Selective"* (grammar schools). Because it is categorical, `lmer()` automatically creates **two dummy (indicator) variables** using treatment coding, with *"Non-selective in a highly selective area"* as the **reference category** — its effect is absorbed into the intercept, and the two dummy coefficients represent the difference in log(ATT8) relative to that baseline.

The **selective** coefficient (`ADMPOL_PTSEL`) is **strongly positive** across all pupil groups — grammar and other selective schools show higher attainment after controlling for intake characteristics. The **other non-selective** coefficient (`ADMPOL_PTOTHER NON SEL`) shows a small and often non-significant difference from the reference group.

### 9. Workforce variables

In the imputed full model, the three workforce variables add useful
context:

- **Teacher retention** (`remained_in_the_same_school`): Positive
  coefficient — higher staff stability is associated with higher
  attainment. The per-year plots show this is consistent across years.
- **Leadership pay** (`teachers_on_leadership_pay_range_percent`):
  Typically small and sometimes non-significant, suggesting that the
  proportion of teachers on leadership pay scales is not a strong
  predictor once other factors are controlled.
- **Teacher absence** (`log(average_number_of_days_taken)`): Negative
  coefficient — schools where teachers are absent more tend to have
  lower ATT8 scores. This echoes the pupil absence finding and
  reinforces that attendance (by both pupils and staff) is strongly
  associated with outcomes.

### 10. Year-to-year stability

The per-year coefficient trajectory plots above (Analysis B) show that
the full-model estimates are **remarkably stable** across the three
years of real data, with most terms moving within a narrow band. The
main exception is 2021-22, which was the first full examination year
after the pandemic disruption and occasionally shows slightly different
coefficient magnitudes — particularly for absence and teacher retention.

### 11. What the imputed model gains over the core model

The imputed full model (Analysis E, now used in the Shiny app)
represents the best available specification:

- **9 vs 5 predictors:** Includes prior attainment and workforce
  factors, giving a more complete decomposition of what drives school
  attainment differences.
- **Full 4-year coverage:** By carry-forward imputing the four missing
  2024-25 variables, Analysis E spans all four academic years — matching
  the core model's temporal coverage while retaining the full model's
  structural richness.
- **Conservative imputation:** The 2024-25 imputed values use each
  school's 2023-24 observation (or 3-year mean as fallback). Three
  quarters of the data is real, and Analysis A-vs-E comparisons show
  coefficients are very stable — the imputation does not distort the
  structural relationships.
- **Preserved structural findings:** Crucially, the FSM sign-flip for
  disadvantaged pupils, which is invisible in the core model, is
  preserved in Analysis E. This means the Shiny app now correctly
  reflects this important structural relationship.


# Brighton & Hove: A Local Case Study

The Brighton & Hove case study — including interactive maps, detailed interpretation of non-linear effects, school-level residual analysis, and policy recommendations — has been moved to its own standalone report: **[Brighton & Hove Case Study](https://adamdennett.github.io/school_attainment_tool/brighton_case_study.html)**.


# Variable Importance & Policy Recommendations

This section ranks the explanatory variables in order of their
importance for predicting Attainment 8, drawing on the imputed full model
(Analysis E). Two complementary importance measures are presented:

1. **Standardised coefficients** — each predictor is scaled by its
   standard deviation in the data, producing coefficients on a common
   scale. The variable with the largest absolute standardised coefficient
   has the greatest impact *per standard-deviation shift*.
2. **Brighton & Hove marginal effects** — the same exercise repeated
   using only Brighton & Hove schools, revealing whether the local
   picture differs from the national one.

Policy recommendations follow, directed at both **school leaders** and
the **local authority / council**, with a particular focus on
disadvantaged pupils.


## National variable importance (standardised coefficients)

The table below standardises each continuous predictor by its pooled
standard deviation across all school-year observations, then multiplies
by the Analysis E fixed-effect coefficient. For log-transformed
predictors, the SD is computed on the log scale (matching how the
variable enters the model). Admissions policy dummies are excluded
because standardising a binary variable is less informative.

```{r importance-national}
#| label: importance-national

# ---- Helper: compute standardised coefficients for one model ----
standardised_importance <- function(model, data, outcome_var, group_label) {

  fe <- fixef(model)
  fe <- fe[names(fe) != "(Intercept)"]

  # Map coefficient names to the actual data columns and transformations
  coef_info <- tribble(
    ~term,                                    ~data_col,                                  ~transform,
    "log(PTFSM6CLA1A)",                       "PTFSM6CLA1A",                              "log",
    "log(PERCTOT)",                            "PERCTOT",                                  "log",
    "log(PNUMEAL)",                            "PNUMEAL",                                  "log",
    "PTPRIORLO",                               "PTPRIORLO",                                "none",
    "gorard_segregation",                      "gorard_segregation",                       "none",
    "remained_in_the_same_school",         "remained_in_the_same_school",              "none",
    "teachers_on_leadership_pay_range_percent", "teachers_on_leadership_pay_range_percent", "none",
    "log(average_number_of_days_taken)",        "average_number_of_days_taken",             "log"
  )

  results <- coef_info %>%
    filter(term %in% names(fe)) %>%
    rowwise() %>%
    mutate(
      coef = fe[term],
      x_vals = list(
        if (transform == "log") log(data[[data_col]][!is.na(data[[data_col]]) & data[[data_col]] > 0])
        else data[[data_col]][!is.na(data[[data_col]])]
      ),
      sd_x = sd(unlist(x_vals), na.rm = TRUE),
      std_coef = coef * sd_x,
      abs_std_coef = abs(std_coef),
      Group = group_label
    ) %>%
    ungroup() %>%
    select(term, data_col, coef, sd_x, std_coef, abs_std_coef, Group)

  results
}

# Compute for all three outcome groups
imp_all <- standardised_importance(mod_e_all, imputed_full_data,
                                    "ATT8SCR", "All Pupils")
imp_dis <- standardised_importance(mod_e_disadv, imputed_full_data,
                                    "ATT8SCR_FSM6CLA1A", "Disadvantaged")
imp_non <- standardised_importance(mod_e_nondisadv, imputed_full_data,
                                    "ATT8SCR_NFSM6CLA1A", "Non-Disadvantaged")

importance_national <- bind_rows(imp_all, imp_dis, imp_non)

# Display name mapping
display_names <- c(
  "log(PTFSM6CLA1A)" = "% Disadvantaged (FSM)",
  "log(PERCTOT)" = "Overall Absence Rate",
  "log(PNUMEAL)" = "% EAL",
  "PTPRIORLO" = "% Low Prior Attainment",
  "gorard_segregation" = "Gorard Segregation Index",
  "remained_in_the_same_school" = "Teacher Retention",
  "teachers_on_leadership_pay_range_percent" = "Leadership Pay %",
  "log(average_number_of_days_taken)" = "Teacher Sickness Days"
)

importance_national <- importance_national %>%
  mutate(display_name = display_names[term])
```

### National ranking table

```{r importance-national-table}
#| label: importance-national-table

# Pivot to wide format for a clean comparison table
national_ranking <- importance_national %>%
  select(display_name, Group, std_coef) %>%
  pivot_wider(names_from = Group, values_from = std_coef) %>%
  mutate(
    `Avg |Std. Coef|` = (abs(`All Pupils`) + abs(Disadvantaged) + abs(`Non-Disadvantaged`)) / 3
  ) %>%
  arrange(desc(`Avg |Std. Coef|`)) %>%
  mutate(Rank = row_number()) %>%
  relocate(Rank) %>%
  mutate(across(where(is.numeric) & !matches("Rank"), \(x) round(x, 4)))

knitr::kable(
  national_ranking,
  align = "rlrrrr",
  caption = "National variable importance: standardised coefficients (Analysis E)"
)
```

### National importance chart

```{r importance-national-chart}
#| label: importance-national-chart
#| fig-height: 6
#| fig-width: 9

importance_chart_data <- importance_national %>%
  mutate(
    display_name = factor(display_name),
    Group = factor(Group, levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged")),
    tooltip_text = paste0(
      "<b>", display_name, "</b><br>",
      "Group: ", Group, "<br>",
      "Raw coefficient: ", round(coef, 5), "<br>",
      "SD of predictor: ", round(sd_x, 3), "<br>",
      "Standardised coef: ", round(std_coef, 4), "<br>",
      "|Std. coef|: ", round(abs_std_coef, 4)
    )
  )

# Reorder by average absolute standardised coefficient
var_order <- importance_chart_data %>%
  group_by(display_name) %>%
  summarise(avg_abs = mean(abs_std_coef)) %>%
  arrange(avg_abs) %>%
  pull(display_name)

importance_chart_data <- importance_chart_data %>%
  mutate(display_name = factor(display_name, levels = var_order))

p_imp <- ggplot(importance_chart_data,
       aes(x = display_name, y = abs_std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#2e6260",
                                "Disadvantaged" = "#4e3c56",
                                "Non-Disadvantaged" = "#abc766")) +
  labs(x = NULL, y = "|Standardised Coefficient|",
       title = "Variable Importance: National (Analysis E)",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_imp, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```

### National importance: direction of effect

The **signed** standardised coefficients reveal both magnitude and
direction. Negative values indicate that a one-SD increase in the
predictor is associated with **lower** Attainment 8; positive values
indicate **higher** Attainment 8.

```{r importance-national-direction}
#| label: importance-national-direction
#| fig-height: 6
#| fig-width: 9

p_dir <- ggplot(importance_chart_data,
       aes(x = display_name, y = std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#2e6260",
                                "Disadvantaged" = "#4e3c56",
                                "Non-Disadvantaged" = "#abc766")) +
  labs(x = NULL, y = "Standardised Coefficient (signed)",
       title = "Direction & Magnitude of Effect: National",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_dir, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```


## Brighton & Hove variable importance

The same standardised-coefficient approach is repeated using only
Brighton & Hove schools. Because the local predictor distributions may
differ from the national picture (e.g. lower absence variance, different
FSM range), the rankings may shift.

```{r importance-bh}
#| label: importance-bh

# Filter to Brighton & Hove
bh_data <- imputed_full_data %>%
  filter(LANAME == "Brighton and Hove")

cat("Brighton & Hove observations:", nrow(bh_data), "\n")

# Use the NATIONAL model coefficients but LOCAL predictor SDs
imp_bh_all <- standardised_importance(mod_e_all, bh_data,
                                       "ATT8SCR", "All Pupils")
imp_bh_dis <- standardised_importance(mod_e_disadv, bh_data,
                                       "ATT8SCR_FSM6CLA1A", "Disadvantaged")
imp_bh_non <- standardised_importance(mod_e_nondisadv, bh_data,
                                       "ATT8SCR_NFSM6CLA1A", "Non-Disadvantaged")

importance_bh <- bind_rows(imp_bh_all, imp_bh_dis, imp_bh_non) %>%
  mutate(display_name = display_names[term])
```

### Brighton & Hove ranking table

```{r importance-bh-table}
#| label: importance-bh-table

bh_ranking <- importance_bh %>%
  select(display_name, Group, std_coef) %>%
  pivot_wider(names_from = Group, values_from = std_coef) %>%
  mutate(
    `Avg |Std. Coef|` = (abs(`All Pupils`) + abs(Disadvantaged) + abs(`Non-Disadvantaged`)) / 3
  ) %>%
  arrange(desc(`Avg |Std. Coef|`)) %>%
  mutate(Rank = row_number()) %>%
  relocate(Rank) %>%
  mutate(across(where(is.numeric) & !matches("Rank"), \(x) round(x, 4)))

knitr::kable(
  bh_ranking,
  align = "rlrrrr",
  caption = "Brighton & Hove variable importance: standardised coefficients (local predictor SDs, national model)"
)
```

### Brighton & Hove importance chart

```{r importance-bh-chart}
#| label: importance-bh-chart
#| fig-height: 6
#| fig-width: 9

bh_chart_data <- importance_bh %>%
  mutate(
    Group = factor(Group, levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged")),
    tooltip_text = paste0(
      "<b>", display_name, "</b> (B&H)<br>",
      "Group: ", Group, "<br>",
      "Raw coefficient (national): ", round(coef, 5), "<br>",
      "SD of predictor (B&H): ", round(sd_x, 3), "<br>",
      "Standardised coef: ", round(std_coef, 4)
    )
  )

bh_var_order <- bh_chart_data %>%
  group_by(display_name) %>%
  summarise(avg_abs = mean(abs_std_coef)) %>%
  arrange(avg_abs) %>%
  pull(display_name)

bh_chart_data <- bh_chart_data %>%
  mutate(display_name = factor(display_name, levels = bh_var_order))

p_bh <- ggplot(bh_chart_data,
       aes(x = display_name, y = abs_std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#2e6260",
                                "Disadvantaged" = "#4e3c56",
                                "Non-Disadvantaged" = "#abc766")) +
  labs(x = NULL, y = "|Standardised Coefficient|",
       title = "Variable Importance: Brighton & Hove",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_bh, tooltip = "text") %>%
  layout(legend = list(orientation = "h", y = -0.15))
```


## National vs Brighton & Hove comparison

```{r importance-comparison}
#| label: importance-comparison

# Compare rankings side by side (All Pupils model)
comparison_all <- national_ranking %>%
  select(display_name, `National Rank` = Rank,
         `National Std.Coef` = `All Pupils`) %>%
  left_join(
    bh_ranking %>%
      select(display_name, `B&H Rank` = Rank,
             `B&H Std.Coef` = `All Pupils`),
    by = "display_name"
  ) %>%
  mutate(`Rank Change` = `National Rank` - `B&H Rank`)

knitr::kable(
  comparison_all,
  align = "lrrrrl",
  caption = "All Pupils: National vs Brighton & Hove variable importance rankings"
)
```

```{r importance-comparison-disadv}
#| label: importance-comparison-disadv

# Same for Disadvantaged pupils
comparison_disadv <- importance_national %>%
  filter(Group == "Disadvantaged") %>%
  arrange(desc(abs_std_coef)) %>%
  mutate(`National Rank` = row_number()) %>%
  select(display_name, `National Rank`, `National Std.Coef` = std_coef) %>%
  left_join(
    importance_bh %>%
      filter(Group == "Disadvantaged") %>%
      arrange(desc(abs_std_coef)) %>%
      mutate(`B&H Rank` = row_number()) %>%
      select(display_name, `B&H Rank`, `B&H Std.Coef` = std_coef),
    by = "display_name"
  ) %>%
  mutate(`Rank Change` = `National Rank` - `B&H Rank`) %>%
  mutate(across(where(is.numeric) & !matches("Rank"), \(x) round(x, 4)))

knitr::kable(
  comparison_disadv,
  align = "lrrrrl",
  caption = "Disadvantaged Pupils: National vs Brighton & Hove variable importance rankings"
)
```

### Key differences between national and local rankings

The comparison tables above highlight where Brighton & Hove's local
context shifts the relative importance of predictors:

- **% Disadvantaged (FSM):** Brighton & Hove schools span a wide FSM
  range (from under 10% to over 50%), which may amplify or dampen this
  variable's local importance relative to the national picture depending
  on the SD of log(FSM) in the Local Authority.

- **Overall Absence:** If Brighton & Hove has a narrower spread of
  absence rates than the national distribution, absence will appear less
  important locally — not because it matters less per unit of change, but
  because schools are more similar to each other in this dimension.

- **Teacher Retention and Sickness:** Brighton & Hove's workforce
  stability profile — shaped by coastal-city labour market conditions,
  London proximity for recruitment, and academy vs maintained school
  mix — may produce a different local variance in these predictors.

- **Gorard Segregation:** As a single local authority, all B&H schools
  share the same segregation index value in any given year, so the
  within-B&H variance is zero or near-zero. This variable's local
  importance will therefore be very low. Note that even at national
  level, the Gorard segregation coefficient is **statistically
  insignificant** in the imputed full model (see commentary in
  section 7 above), so its apparent ranking in the national table
  should be interpreted with caution.


## Policy recommendations

The variable importance rankings, combined with the direction-of-effect
analysis and the model's structural features, suggest several priority
areas for intervention. These are organised by audience.

### For school leaders

**1. Absence reduction is the single biggest lever.**

Across all three pupil groups, overall absence rate (`PERCTOT`) is
consistently the strongest or among the strongest predictors of
Attainment 8. The log-transformed relationship means that gains from
reducing absence are **largest for schools with already-moderate absence
rates** (the steep part of the curve). For a school currently at 6%
absence, reducing to 5% has a larger predicted effect than a school at
10% reducing to 9%. This argues for:

- Intensive early-intervention strategies (attendance officers,
  family liaison, mentoring) before absence becomes entrenched
- Special attention to protecting low-absence schools from
  deterioration, not just tackling chronic cases
- Focus on disadvantaged pupils specifically, where the absence
  coefficient is largest

**2. Support for pupils with low prior attainment.**

The `PTPRIORLO` coefficient is strongly negative — schools serving a
higher proportion of pupils who entered secondary school with low KS2
scores see lower Attainment 8 outcomes. While schools cannot change their
intake composition, they can invest in:

- Targeted catch-up programmes (literacy, numeracy) for pupils arriving
  with below-expected KS2 levels
- Diagnostic assessment at transition from primary school to identify
  gaps early
- Small-group intervention during KS3 to prevent the attainment gap
  widening before GCSE

**3. Workforce stability matters.**

Teacher retention (`remained_in_the_same_school`) shows a positive
association with attainment: schools where more teachers stay are
predicted to achieve higher ATT8 scores. Meanwhile, teacher sickness
days show a negative association. School leaders should consider:

- Retention strategies: competitive CPD offers, workload management,
  mentoring for early-career teachers
- Wellbeing programmes targeting staff absence
- Succession planning to avoid disruptive turnover in key departments

**4. The FSM sign-flip for disadvantaged pupils.**

Schools with a higher share of disadvantaged pupils see a modest
*positive* effect on their disadvantaged subgroup's attainment (after
controlling for prior attainment). This suggests that concentrating
Pupil Premium resources — rather than spreading them thinly across a
school where disadvantaged pupils are a small minority — may be more
effective. Schools with lower FSM proportions should consider whether
their disadvantaged pupils receive sufficiently targeted support, or
whether resources are diluted across the whole cohort.


### For Brighton & Hove Council and local authority

**1. Tackle absence strategically across the Local Authority.**

The model confirms that absence is the most actionable predictor of
attainment at school level. The council can support schools through:

- Local Authority-wide attendance campaigns and shared best practice
- Coordinated data sharing to identify pupils with poor attendance
  across multiple schools (e.g. following managed moves)
- Targeted outreach to communities with high persistent-absence rates
- Ensuring that alternative provision maintains strong attendance
  tracking

**2. Address the disadvantaged attainment gap directly.**

The model shows that the coefficient magnitudes for disadvantaged pupils
are often **larger** than for non-disadvantaged — absence and prior
attainment in particular hit harder for the disadvantaged group.
Policy priorities include:

- Ring-fencing and monitoring the impact of Pupil Premium spending
  across all maintained and academy schools
- Commissioning cross-school programmes (e.g. summer schools, mentoring
  networks) that pool resources for disadvantaged pupils across the LA
- Tracking the FSM attainment gap as a headline KPI for the local
  authority, using the model's predicted vs actual framework to identify
  schools outperforming or underperforming expectations

**3. A note on segregation.**

The Gorard segregation index enters the model with a negative
coefficient (higher segregation associated with lower attainment), but
this effect is **statistically insignificant** — the standard error is
large relative to the estimate, and we cannot confidently distinguish
it from zero. While reducing segregation may be a worthwhile policy
goal for broader equity reasons, **this model does not provide reliable
evidence that reducing the Gorard index would improve Attainment 8
scores** at school level. The council should therefore not prioritise
segregation reduction *on the basis of this model alone*, though it
may of course remain a valid objective on other grounds (social
cohesion, equal access, community trust).

**4. Support workforce stability across schools.**

Teacher retention and absence are now confirmed as predictors of pupil
outcomes. The council and local partnerships can help by:

- Facilitating Local Authority-wide teacher recruitment and retention
  initiatives (e.g. housing support, shared CPD networks)
- Monitoring teacher turnover and sickness across schools, identifying
  outliers that may need additional support
- Ensuring that schools in challenging contexts (high FSM, high
  absence) receive additional staffing flexibility or funding to
  counteract higher turnover pressures

**5. Use the model to target support.**

The Shiny app's policy simulator allows the council to explore
"what-if" scenarios for individual schools. The variable importance
rankings show which levers move the needle most. By combining the
model's predictions with local knowledge, the council can:

- Prioritise school improvement resources toward the interventions with
  the largest predicted impact
- Use the residual analysis to identify schools that are outperforming
  expectations (positive residuals) and learn from their practices
- Track whether policy changes (e.g. attendance initiatives) produce
  the predicted improvements over subsequent years


### Summary: the biggest levers

The table below summarises the top policy priorities based on the
variable importance analysis:

```{r policy-summary-table}
#| label: policy-summary-table
#| echo: false

policy_summary <- tribble(
  ~Priority, ~`Variable`, ~`Who Can Act`, ~`Key Action`,
  1, "Overall Absence Rate", "School leaders + Council", "Intensive early intervention; protect low-absence schools",
  2, "% Low Prior Attainment", "School leaders", "Targeted KS2-KS3 transition catch-up programmes",
  3, "% Disadvantaged (FSM)", "School leaders + Council", "Concentrated Pupil Premium spend; monitor FSM gap as KPI",
  4, "Teacher Retention", "School leaders + Council", "Retention strategies; CPD investment; workload management",
  5, "Teacher Sickness Days", "School leaders", "Staff wellbeing programmes; absence monitoring",
  6, "% EAL", "School leaders", "Build on EAL advantage; ensure support for new arrivals",
  7, "Leadership Pay %", "School leaders", "Review leadership structure (low model impact)",
  8, "Gorard Segregation Index*", "Council", "Effect statistically insignificant — not a reliable lever in this model"
)

knitr::kable(
  policy_summary,
  align = "rlllc",
  caption = "Policy priorities ranked by variable importance"
)
```

\* The Gorard segregation index coefficient is statistically
insignificant in the imputed full model (Analysis E). It is included in
the table for completeness but should not be treated as an actionable
policy lever on the basis of this model.

> **Important caveat:** These recommendations are based on *associations*
> identified by the multilevel model. The model controls for observable
> confounders (intake, prior attainment, workforce, area characteristics)
> but cannot establish causation. A school reducing absence may not see
> exactly the predicted ATT8 gain if other unmeasured factors change
> simultaneously. The rankings and effect sizes should be treated as
> evidence-informed starting points for policy discussion, not as causal
> guarantees. Variables whose coefficients are statistically insignificant
> (notably the Gorard segregation index) should be interpreted with
> particular caution.


# Quick Reference

## Formula summary

| Analysis | Fixed Effects | Random Effects | Data | Used in App? |
|----------|--------------|----------------|------|:---:|
| A (Full Panel) | 9 predictors (5 log, 4 linear) | `(1\|year) + (1\|Ofsted) + (1\|region/LA)` | 3 years, workforce required | No |
| B (Full Per-Year) | Same 9 predictors | `(1\|Ofsted) + (1\|region/LA)` | Each year separately | No |
| C (Core Panel) | 5 predictors (3 log, 2 linear) | `(1\|year) + (1\|Ofsted) + (1\|region/LA)` | All 4 years | No |
| D (Core Per-Year) | Same 5 predictors | `(1\|Ofsted) + (1\|region/LA)` | Each year separately | No |
| **E (Imputed Full)** | **9 predictors (5 log, 4 linear)** | **`(1\|year) + (1\|Ofsted) + (1\|region/LA)`** | **All 4 years (2024-25 imputed)** | **Yes** |

## Variable key

| Variable | In formula as | Description |
|----------|--------------|-------------|
| `ATT8SCR` | `log(ATT8SCR)` | Attainment 8 score (all pupils) |
| `ATT8SCR_FSM6CLA1A` | `log(ATT8SCR_FSM6CLA1A)` | Attainment 8 score (disadvantaged) |
| `ATT8SCR_NFSM6CLA1A` | `log(ATT8SCR_NFSM6CLA1A)` | Attainment 8 score (non-disadvantaged) |
| `PTFSM6CLA1A` | `log(PTFSM6CLA1A)` | % eligible for Free School Meals |
| `PERCTOT` | `log(PERCTOT)` | % total absence |
| `PNUMEAL` | `log(PNUMEAL)` | % English as Additional Language |
| `PTPRIORLO` | `PTPRIORLO` | % low KS2 prior attainment |
| `ADMPOL_PT` | `ADMPOL_PT` | Admissions policy indicator |
| `gorard_segregation` | `gorard_segregation` | Gorard segregation index |
| `remained_in_the_same_school` | `remained_in_the_same_school` | Teacher retention (%) |
| `teachers_on_leadership_pay_range_percent` | as-is | % teachers on leadership pay |
| `average_number_of_days_taken` | `log(average_number_of_days_taken)` | Average teacher absence days |
| `OFSTEDRATING_1` | `(1\|OFSTEDRATING_1)` | Ofsted rating (random intercept) |
| `gor_name` | `(1\|gor_name/LANAME)` | Government Office Region |
| `LANAME` | nested in `gor_name` | Local Authority |
| `year_label` | `(1\|year_label)` | Academic year (panel models only) |


# Disadvantaged Attainment and Concentrations of Disadvantage: Directionality Test {#sec-directionality}

In Analysis E, the coefficient on `log(PTFSM6CLA1A)` flips from negative (All Pupils, Non-Disadvantaged) to **positive** for Disadvantaged pupils once prior attainment and other controls are included. This section tests whether the sign-flip is robust to alternative operationalisations of both the outcome (disadvantaged attainment) and the predictor (school-level disadvantage concentration), or whether it is an artefact of the specific FSM6CLA1A classification and Attainment 8 metric.

**Strategy:** Hold the model specification constant (same random effects, same controls) and systematically vary:

1. The **Y variable** — alternative measures of disadvantaged pupil attainment
2. The **X variable** — alternative measures of school-level disadvantage concentration

If the sign-flip appears consistently across different measures, it is more likely to reflect a genuine structural relationship than a measurement artefact.

## Setup: baseline model and data

```{r directionality-setup}
#| label: directionality-setup

# Use the same imputed full dataset as Analysis E
d_dir <- imputed_full_data %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0,
         PTFSM6CLA1A > 0, PERCTOT > 0, PNUMEAL > 0,
         !is.na(PTPRIORLO),
         !is.na(OFSTEDRATING_1), !is.na(gor_name), !is.na(LANAME),
         !is.na(remained_in_the_same_school),
         !is.na(teachers_on_leadership_pay_range_percent),
         average_number_of_days_taken > 0,
         !is.na(gorard_segregation)) %>%
  droplevels()

contrasts(d_dir$OFSTEDRATING_1) <- contr.treatment(levels(d_dir$OFSTEDRATING_1))

cat("Directionality test dataset:", nrow(d_dir), "rows\n")
cat("Years:", paste(unique(d_dir$year_label), collapse = ", "), "\n")

# Reference: the Analysis E disadvantaged coefficient
ref_coef <- fixef(mod_e_disadv)["log(PTFSM6CLA1A)"]
cat(sprintf("\nReference (Analysis E): log(PTFSM6CLA1A) coefficient = %+.5f\n", ref_coef))
```

## Test 1: Alternative Y variables (disadvantaged attainment)

Holding `log(PTFSM6CLA1A)` as the deprivation measure, we swap the outcome variable across six alternative measures of disadvantaged attainment. For each, we report the coefficient on `log(PTFSM6CLA1A)`, its standard error, t-value, and sign.

```{r directionality-y-vars}
#| label: directionality-y-vars

# Define alternative Y variables to test
y_tests <- tribble(
  ~y_var,                      ~label,                              ~transform,  ~filter_positive,
  "ATT8SCR_FSM6CLA1A",        "ATT8 Disadvantaged (reference)",    "log",       TRUE,
  "ATT8SCRENG_FSM6CLA1A",     "ATT8 English (Disadvantaged)",      "log",       TRUE,
  "ATT8SCRMAT_FSM6CLA1A",     "ATT8 Maths (Disadvantaged)",        "log",       TRUE,
  "ATT8SCREBAC_FSM6CLA1A",    "ATT8 EBacc (Disadvantaged)",        "log",       TRUE,
  "ATT8SCROPEN_FSM6CLA1A",    "ATT8 Open (Disadvantaged)",         "log",       TRUE,
  "P8MEA_FSM6CLA1A",          "Progress 8 (Disadvantaged)",        "none",      FALSE
)

# Fit each model and extract the FSM coefficient
y_results <- map_dfr(seq_len(nrow(y_tests)), function(i) {
  y_var <- y_tests$y_var[i]
  label <- y_tests$label[i]
  transform <- y_tests$transform[i]
  filter_pos <- y_tests$filter_positive[i]

  d_test <- d_dir
  y_vals <- d_test[[y_var]]

  # Filter to valid observations
  if (filter_pos) {
    d_test <- d_test[!is.na(y_vals) & y_vals > 0, ]
  } else {
    d_test <- d_test[!is.na(y_vals), ]
  }
  d_test <- droplevels(d_test)
  contrasts(d_test$OFSTEDRATING_1) <- contr.treatment(levels(d_test$OFSTEDRATING_1))

  # Build formula
  if (transform == "log") {
    lhs <- paste0("log(", y_var, ")")
  } else {
    lhs <- y_var
  }

  formula_str <- paste0(
    lhs, " ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) + ",
    "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
    "remained_in_the_same_school + teachers_on_leadership_pay_range_percent + ",
    "log(average_number_of_days_taken) + ",
    "(1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
  )

  mod <- tryCatch(
    lmer(as.formula(formula_str), data = d_test, REML = TRUE,
         na.action = na.exclude,
         control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))),
    error = function(e) NULL
  )

  if (is.null(mod)) {
    return(tibble(Y_Variable = label, N = nrow(d_test),
                  Coefficient = NA_real_, SE = NA_real_,
                  t_value = NA_real_, Sign = "FAILED"))
  }

  coefs <- summary(mod)$coefficients
  row_name <- "log(PTFSM6CLA1A)"
  if (row_name %in% rownames(coefs)) {
    tibble(
      Y_Variable = label,
      N = nrow(d_test),
      Coefficient = coefs[row_name, "Estimate"],
      SE = coefs[row_name, "Std. Error"],
      t_value = coefs[row_name, "t value"],
      Sign = ifelse(coefs[row_name, "Estimate"] > 0, "POSITIVE", "negative")
    )
  } else {
    tibble(Y_Variable = label, N = nrow(d_test),
           Coefficient = NA_real_, SE = NA_real_,
           t_value = NA_real_, Sign = "NOT FOUND")
  }
})

knitr::kable(
  y_results %>%
    mutate(Coefficient = sprintf("%+.5f", Coefficient),
           SE = sprintf("%.5f", SE),
           t_value = sprintf("%+.2f", t_value)),
  caption = "Test 1: Coefficient on log(% FSM) across alternative disadvantaged attainment measures",
  align = "lrrrrl"
)
```

```{r directionality-y-narrative}
#| label: directionality-y-narrative
#| output: asis

n_positive <- sum(y_results$Sign == "POSITIVE", na.rm = TRUE)
n_total <- sum(!is.na(y_results$Coefficient))

cat(sprintf("Of the %d outcome measures tested, **%d show a positive coefficient** on log(%% FSM) for disadvantaged pupils.\n\n", n_total, n_positive))

# Special note on Progress 8
p8_row <- y_results %>% filter(grepl("Progress 8", Y_Variable))
if (nrow(p8_row) > 0 && !is.na(p8_row$Coefficient)) {
  cat(sprintf("**Progress 8** is particularly noteworthy because it is already a value-added measure (adjusting for prior attainment at the individual pupil level). The coefficient on log(%% FSM) in the P8 model is **%+.5f** (t = %+.2f). ",
      p8_row$Coefficient, p8_row$t_value))
  if (p8_row$Coefficient > 0) {
    cat("The fact that the sign remains positive even when the outcome itself already accounts for prior attainment reinforces the finding: this is not simply a statistical artefact of controlling for `PTPRIORLO` in the model.\n\n")
  } else {
    cat("The sign is negative here, which is expected since Progress 8 already removes the prior-attainment pathway at the individual level — the school-level `PTPRIORLO` control in our model may be double-counting.\n\n")
  }
}
```

## Test 2: Alternative X variables (school-level disadvantage concentration)

Now we hold the outcome fixed at `log(ATT8SCR_FSM6CLA1A)` and swap the deprivation predictor. Each alternative X replaces `log(PTFSM6CLA1A)` in the model while keeping all other predictors unchanged.

```{r directionality-x-vars}
#| label: directionality-x-vars

# Define alternative X variables to test
x_tests <- tribble(
  ~x_var,            ~label,                                 ~transform,
  "PTFSM6CLA1A",    "% FSM6CLA1A (reference)",              "log",
  "PNUMFSMEVER",    "% ever FSM (census, broader defn.)",   "log",
  "PSEN_ALL4",      "% any SEN",                            "log",
  "PPERSABS10",     "% persistent absentees (10%+)",        "log",
  "PTEALGRP2",      "% EAL not yet fluent",                 "log"
)

x_results <- map_dfr(seq_len(nrow(x_tests)), function(i) {
  x_var <- x_tests$x_var[i]
  label <- x_tests$label[i]
  transform <- x_tests$transform[i]

  d_test <- d_dir
  x_vals <- d_test[[x_var]]

  # Filter to valid, positive values for log transform
  d_test <- d_test[!is.na(x_vals) & x_vals > 0, ]
  d_test <- droplevels(d_test)
  contrasts(d_test$OFSTEDRATING_1) <- contr.treatment(levels(d_test$OFSTEDRATING_1))

  # Build formula: swap x_var for PTFSM6CLA1A
  x_term <- if (transform == "log") paste0("log(", x_var, ")") else x_var
  coef_name <- if (transform == "log") paste0("log(", x_var, ")") else x_var

  formula_str <- paste0(
    "log(ATT8SCR_FSM6CLA1A) ~ ", x_term, " + log(PERCTOT) + log(PNUMEAL) + ",
    "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
    "remained_in_the_same_school + teachers_on_leadership_pay_range_percent + ",
    "log(average_number_of_days_taken) + ",
    "(1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
  )

  mod <- tryCatch(
    lmer(as.formula(formula_str), data = d_test, REML = TRUE,
         na.action = na.exclude,
         control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))),
    error = function(e) NULL
  )

  if (is.null(mod)) {
    return(tibble(X_Variable = label, N = nrow(d_test),
                  Coefficient = NA_real_, SE = NA_real_,
                  t_value = NA_real_, Sign = "FAILED"))
  }

  coefs <- summary(mod)$coefficients
  if (coef_name %in% rownames(coefs)) {
    tibble(
      X_Variable = label,
      N = nrow(d_test),
      Coefficient = coefs[coef_name, "Estimate"],
      SE = coefs[coef_name, "Std. Error"],
      t_value = coefs[coef_name, "t value"],
      Sign = ifelse(coefs[coef_name, "Estimate"] > 0, "POSITIVE", "negative")
    )
  } else {
    tibble(X_Variable = label, N = nrow(d_test),
           Coefficient = NA_real_, SE = NA_real_,
           t_value = NA_real_, Sign = "NOT FOUND")
  }
})

knitr::kable(
  x_results %>%
    mutate(Coefficient = sprintf("%+.5f", Coefficient),
           SE = sprintf("%.5f", SE),
           t_value = sprintf("%+.2f", t_value)),
  caption = "Test 2: Coefficient on alternative deprivation measures predicting log(ATT8 Disadvantaged)",
  align = "lrrrrl"
)
```

```{r directionality-x-narrative}
#| label: directionality-x-narrative
#| output: asis

n_positive_x <- sum(x_results$Sign == "POSITIVE", na.rm = TRUE)
n_total_x <- sum(!is.na(x_results$Coefficient))

cat(sprintf("Of the %d deprivation measures tested, **%d show a positive coefficient** when predicting disadvantaged attainment.\n\n", n_total_x, n_positive_x))

cat("The alternative X variables test different conceptualisations of disadvantage concentration:\n\n")
cat("- **% ever FSM** uses the broader census-based definition (ever eligible in the past 6 years), which captures a wider population than the KS4-cohort FSM6CLA1A measure\n")
cat("- **% any SEN** measures a different dimension of disadvantage entirely (educational needs rather than economic deprivation)\n")
cat("- **% persistent absentees** is a behavioural proxy for disadvantage that is not directly based on FSM classification\n")
cat("- **% EAL not yet fluent** captures linguistic disadvantage, partially independent of economic deprivation\n\n")

if (n_positive_x >= 3) {
  cat("The consistency of the positive sign across multiple, conceptually distinct measures of disadvantage concentration strengthens the case that this is a **genuine structural relationship** rather than an artefact of the FSM6CLA1A classification. Schools with higher concentrations of disadvantage — however measured — tend to achieve slightly higher outcomes for their disadvantaged subgroup, once prior attainment and other controls are accounted for.\n\n")
} else if (n_positive_x >= 1) {
  cat("The sign-flip appears with some but not all alternative measures, suggesting the relationship may be **partly specific to particular operationalisations** of disadvantage concentration. The FSM-based measures and closely correlated alternatives tend to show the positive sign, while more conceptually distinct measures may not.\n\n")
} else {
  cat("The positive sign appears **only with the original FSM6CLA1A measure**, raising the possibility that the sign-flip may be partly a measurement artefact rather than a robust structural finding. Further investigation with individual pupil-level data would be needed.\n\n")
}
```

## Test 3: Cross-tabulation (Y x X combinations)

As a final check, we fit the most informative subset of Y × X combinations to see whether the pattern is consistent across the full grid.

```{r directionality-grid}
#| label: directionality-grid
#| fig-height: 5
#| fig-width: 8

# Select key Y and X combinations
grid_y <- tribble(
  ~y_var,                    ~y_label,          ~y_transform,  ~y_filter_pos,
  "ATT8SCR_FSM6CLA1A",      "ATT8 Disadv.",    "log",         TRUE,
  "ATT8SCRENG_FSM6CLA1A",   "ATT8 Eng Dis.",   "log",         TRUE,
  "ATT8SCRMAT_FSM6CLA1A",   "ATT8 Mat Dis.",   "log",         TRUE,
  "P8MEA_FSM6CLA1A",        "P8 Disadv.",      "none",        FALSE
)

grid_x <- tribble(
  ~x_var,            ~x_label,              ~x_transform,
  "PTFSM6CLA1A",    "% FSM",               "log",
  "PNUMFSMEVER",    "% ever FSM",          "log",
  "PSEN_ALL4",      "% any SEN",           "log",
  "PPERSABS10",     "% persist. absent",   "log"
)

grid_results <- expand_grid(
  y_idx = seq_len(nrow(grid_y)),
  x_idx = seq_len(nrow(grid_x))
) %>%
  mutate(result = map2(y_idx, x_idx, function(yi, xi) {
    y_var <- grid_y$y_var[yi]
    x_var <- grid_x$x_var[xi]
    y_transform <- grid_y$y_transform[yi]
    x_transform <- grid_x$x_transform[xi]
    filter_pos <- grid_y$y_filter_pos[yi]

    d_test <- d_dir
    y_vals <- d_test[[y_var]]
    x_vals <- d_test[[x_var]]

    # Filter
    valid <- !is.na(y_vals) & !is.na(x_vals) & x_vals > 0
    if (filter_pos) valid <- valid & y_vals > 0
    d_test <- d_test[valid, ] %>% droplevels()
    contrasts(d_test$OFSTEDRATING_1) <- contr.treatment(levels(d_test$OFSTEDRATING_1))

    lhs <- if (y_transform == "log") paste0("log(", y_var, ")") else y_var
    x_term <- if (x_transform == "log") paste0("log(", x_var, ")") else x_var
    coef_name <- if (x_transform == "log") paste0("log(", x_var, ")") else x_var

    formula_str <- paste0(
      lhs, " ~ ", x_term, " + log(PERCTOT) + log(PNUMEAL) + ",
      "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
      "remained_in_the_same_school + teachers_on_leadership_pay_range_percent + ",
      "log(average_number_of_days_taken) + ",
      "(1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)"
    )

    mod <- tryCatch(
      lmer(as.formula(formula_str), data = d_test, REML = TRUE,
           na.action = na.exclude,
           control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000))),
      error = function(e) NULL
    )

    if (is.null(mod)) return(tibble(coef = NA_real_, t = NA_real_, n = nrow(d_test)))

    coefs <- summary(mod)$coefficients
    if (coef_name %in% rownames(coefs)) {
      tibble(coef = coefs[coef_name, "Estimate"],
             t = coefs[coef_name, "t value"],
             n = nrow(d_test))
    } else {
      tibble(coef = NA_real_, t = NA_real_, n = nrow(d_test))
    }
  })) %>%
  unnest(result) %>%
  mutate(
    y_label = grid_y$y_label[y_idx],
    x_label = grid_x$x_label[x_idx],
    sign = case_when(
      is.na(coef) ~ "NA",
      coef > 0 ~ "+",
      TRUE ~ "–"
    )
  )

# Heatmap
ggplot(grid_results, aes(x = x_label, y = y_label, fill = coef)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = sprintf("%+.4f\n(t=%+.1f)", coef, t)),
            size = 3.2, colour = "white", fontface = "bold") +
  scale_fill_gradient2(low = "#7b132b", mid = "#f5f5f5", high = "#2e6260",
                       midpoint = 0, name = "Coefficient") +
  labs(title = "Disadvantage concentration coefficient across Y x X combinations",
       subtitle = "Green = positive (sign-flip present); Red = negative",
       x = "Deprivation measure (X variable)",
       y = "Disadvantaged attainment measure (Y variable)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        panel.grid = element_blank())
```

```{r directionality-grid-narrative}
#| label: directionality-grid-narrative
#| output: asis

n_pos_grid <- sum(grid_results$coef > 0, na.rm = TRUE)
n_total_grid <- sum(!is.na(grid_results$coef))

cat(sprintf("\nAcross the **%d-cell grid** of Y × X combinations, **%d (%d%%)** show a positive coefficient on the deprivation concentration measure.\n\n",
    n_total_grid, n_pos_grid, round(100 * n_pos_grid / n_total_grid)))

cat("## Summary\n\n")

if (n_pos_grid / n_total_grid >= 0.75) {
  cat("::: {.callout-important}\n")
  cat("## The sign-flip is robust\n\n")
  cat("The positive relationship between disadvantage concentration and disadvantaged attainment holds across the large majority of alternative operationalisations tested. This is consistent with a genuine structural relationship — likely reflecting resource concentration (Pupil Premium scaling), institutional focus, and/or composition effects — rather than a measurement artefact of the FSM6CLA1A classification or the Attainment 8 metric.\n\n")
  cat("The finding that this pattern persists when using % ever FSM (broader economic measure), % any SEN (different dimension of disadvantage), and % persistent absentees (behavioural proxy) as the concentration measure provides triangulating evidence that schools serving more disadvantaged intakes generate modestly better outcomes for those pupils, once prior attainment is accounted for.\n")
  cat(":::\n")
} else if (n_pos_grid / n_total_grid >= 0.4) {
  cat("::: {.callout-note}\n")
  cat("## The sign-flip is partially robust\n\n")
  cat("The positive relationship appears in a substantial minority of combinations but is not universal. It is strongest with FSM-based deprivation measures and ATT8-based outcomes, suggesting the effect is real but may be moderated by the specific dimensions of disadvantage and attainment being measured. The core finding holds, but its magnitude and significance vary with operationalisation.\n")
  cat(":::\n")
} else {
  cat("::: {.callout-warning}\n")
  cat("## The sign-flip may be fragile\n\n")
  cat("The positive coefficient appears in fewer than 40% of combinations tested. While not necessarily an artefact, the sign-flip may be more sensitive to operationalisation than the Analysis E results alone would suggest. Individual pupil-level data would be needed to disentangle whether this reflects a genuine structural relationship or a compositional effect specific to the FSM6CLA1A classification.\n")
  cat(":::\n")
}
```
