---
title: "School Attainment Multilevel Model Results"
subtitle: "Multilevel analysis of KS4 Attainment 8 scores across English secondary schools"
author: ""
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: flatly
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
    df-print: kable
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 5
    geometry: margin=2.5cm
    keep-tex: false
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(here)

source(here::here("R", "helpers.R"))

# ---- Load all saved outputs ----
panel_data       <- readRDS(here::here("data", "panel_data.rds"))
model_data       <- readRDS(here::here("data", "model_data.rds"))
panel_models     <- readRDS(here::here("data", "models.rds"))
panel_diag       <- readRDS(here::here("data", "model_diagnostics.rds"))
yearly_models    <- readRDS(here::here("data", "models_yearly.rds"))
yearly_diag      <- readRDS(here::here("data", "model_diagnostics_yearly.rds"))

# Core models (reduced specification, all 4 years including 2024-25)
core_model_data  <- readRDS(here::here("data", "model_data_core.rds"))
core_models      <- readRDS(here::here("data", "models_core.rds"))
core_diag        <- readRDS(here::here("data", "model_diagnostics_core.rds"))
core_yearly_models <- readRDS(here::here("data", "models_yearly_core.rds"))
core_yearly_diag <- readRDS(here::here("data", "model_diagnostics_yearly_core.rds"))

# Outcome configuration
OUTCOMES <- list(
  all               = "ATT8SCR",
  disadvantaged     = "ATT8SCR_FSM6CLA1A",
  non_disadvantaged = "ATT8SCR_NFSM6CLA1A"
)

outcome_labels <- c(
  all               = "All Pupils",
  disadvantaged     = "Disadvantaged",
  non_disadvantaged = "Non-Disadvantaged"
)

# Ensure year_label is a factor for consistent ordering
model_data <- model_data %>%
  mutate(year_label = factor(year_label))

# Theme for publication-quality plots
theme_pub <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```


# Data Overview

## Sample Construction

The analysis uses data from the Department for Education Performance Tables for Key Stage 4, linked with school census, absence, Ofsted, and School Workforce data across four academic years.

```{r tbl-sample}
#| label: tbl-sample
#| tbl-cap: "Sample construction by academic year"

sample_summary <- model_data %>%
  group_by(`Academic Year` = year_label) %>%
  summarise(
    Schools = n_distinct(URN),
    `School-Year Obs.` = n(),
    `Mean ATT8 (All)` = round(mean(ATT8SCR, na.rm = TRUE), 1),
    `SD ATT8` = round(sd(ATT8SCR, na.rm = TRUE), 1),
    `Mean % Disadvantaged` = round(mean(PTFSM6CLA1A, na.rm = TRUE), 1),
    `Mean % Absence` = round(mean(PERCTOT, na.rm = TRUE), 1),
    .groups = "drop"
  )

# Add totals row
totals <- tibble(
  `Academic Year` = "Total / Mean",
  Schools = n_distinct(model_data$URN),
  `School-Year Obs.` = nrow(model_data),
  `Mean ATT8 (All)` = round(mean(model_data$ATT8SCR, na.rm = TRUE), 1),
  `SD ATT8` = round(sd(model_data$ATT8SCR, na.rm = TRUE), 1),
  `Mean % Disadvantaged` = round(mean(model_data$PTFSM6CLA1A, na.rm = TRUE), 1),
  `Mean % Absence` = round(mean(model_data$PERCTOT, na.rm = TRUE), 1)
)

bind_rows(sample_summary, totals) %>%
  kable(align = "lrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(nrow(sample_summary) + 1, bold = TRUE)
```


## Descriptive Statistics for Model Variables

```{r tbl-descriptives}
#| label: tbl-descriptives
#| tbl-cap: "Descriptive statistics for variables included in the multilevel model"

desc_vars <- c(
  "ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A",
  "PTFSM6CLA1A", "PERCTOT", "PNUMEAL", "PTPRIORLO",
  "gorard_segregation",
  "remained_in_the_same_school",
  "teachers_on_leadership_pay_range_percent",
  "average_number_of_days_taken"
)
desc_vars <- intersect(desc_vars, names(model_data))

desc_table <- map_dfr(desc_vars, function(v) {
  x <- model_data[[v]]
  tibble(
    Variable = var_label(v),
    N = sum(!is.na(x)),
    Mean = round(mean(x, na.rm = TRUE), 2),
    SD = round(sd(x, na.rm = TRUE), 2),
    Min = round(min(x, na.rm = TRUE), 2),
    Q1 = round(quantile(x, 0.25, na.rm = TRUE), 2),
    Median = round(median(x, na.rm = TRUE), 2),
    Q3 = round(quantile(x, 0.75, na.rm = TRUE), 2),
    Max = round(max(x, na.rm = TRUE), 2)
  )
})

desc_table %>%
  kable(align = "lrrrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  pack_rows("Outcome variables", 1, 3) %>%
  pack_rows("Predictor variables", 4, nrow(desc_table))
```


## Outcome Distributions

```{r fig-outcome-dist}
#| label: fig-outcome-dist
#| fig-cap: "Distribution of Attainment 8 scores by pupil group and academic year"
#| fig-height: 7

outcome_long <- model_data %>%
  select(year_label, ATT8SCR, ATT8SCR_FSM6CLA1A, ATT8SCR_NFSM6CLA1A) %>%
  pivot_longer(-year_label, names_to = "outcome", values_to = "score") %>%
  filter(!is.na(score)) %>%
  mutate(
    outcome = recode(outcome,
      ATT8SCR = "All Pupils",
      ATT8SCR_FSM6CLA1A = "Disadvantaged",
      ATT8SCR_NFSM6CLA1A = "Non-Disadvantaged"
    ),
    outcome = factor(outcome, levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged"))
  )

ggplot(outcome_long, aes(x = score, fill = outcome)) +
  geom_histogram(bins = 50, alpha = 0.7, colour = "white", linewidth = 0.2) +
  facet_grid(outcome ~ year_label, scales = "free_y") +
  scale_fill_manual(values = c("#1565c0", "#c62828", "#2e7d32")) +
  labs(x = "Attainment 8 Score", y = "Frequency") +
  theme_pub +
  theme(legend.position = "none")
```


## Correlation Matrix

```{r fig-correlation}
#| label: fig-correlation
#| fig-cap: "Pairwise correlations among model variables (Pearson)"
#| fig-height: 7
#| fig-width: 8

cor_vars <- c("ATT8SCR", "PTFSM6CLA1A", "PERCTOT", "PNUMEAL", "PTPRIORLO",
              "gorard_segregation", "remained_in_the_same_school",
              "teachers_on_leadership_pay_range_percent",
              "average_number_of_days_taken")
cor_vars <- intersect(cor_vars, names(model_data))

cor_data <- model_data %>%
  select(all_of(cor_vars)) %>%
  drop_na()

cor_mat <- cor(cor_data)

# Rename rows/cols with labels
short_labels <- var_labels(cor_vars)
rownames(cor_mat) <- short_labels
colnames(cor_mat) <- short_labels

# Melt for ggplot
cor_long <- cor_mat %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "r") %>%
  mutate(
    var1 = factor(var1, levels = rev(short_labels)),
    var2 = factor(var2, levels = short_labels)
  )

ggplot(cor_long, aes(x = var2, y = var1, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 2.5) +
  scale_fill_gradient2(low = "#c62828", mid = "white", high = "#1565c0",
                       midpoint = 0, limits = c(-1, 1), name = "Pearson r") +
  labs(x = NULL, y = NULL) +
  theme_pub +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 8))
```


# Panel Models (Analysis A)

The panel models pool all four academic years, with `year_label` entering as a random intercept alongside Ofsted rating and local authority (nested within region). This specification treats year-to-year variation as a variance component rather than a fixed trend.

$$
\log(y_{ij}) = \mathbf{X}_{ij}\boldsymbol{\beta} + u_{\text{year}} + u_{\text{Ofsted}} + u_{\text{region}} + u_{\text{LA}} + \varepsilon_{ij}
$$


## Fixed Effects

```{r tbl-panel-fixed}
#| label: tbl-panel-fixed
#| tbl-cap: "Fixed effect estimates from panel models (all years pooled, year as random effect)"

# Build combined table
panel_fe <- map_dfr(names(panel_models), function(mn) {
  m <- panel_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    transmute(
      Model = outcome_labels[mn],
      Term = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      Estimate = round(Estimate, 4),
      `Std. Error` = round(`Std. Error`, 4),
      `t value` = round(`t value`, 2),
      `p value` = ifelse(`Pr(>|t|)` < 0.001, "< 0.001",
                         ifelse(`Pr(>|t|)` < 0.01, "< 0.01",
                                ifelse(`Pr(>|t|)` < 0.05, "< 0.05",
                                       round(`Pr(>|t|)`, 3))))
    )
})

panel_fe %>%
  kable(align = "llrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Coefficient Comparison Plot

```{r fig-panel-coefs}
#| label: fig-panel-coefs
#| fig-cap: "Fixed effect estimates with 95% confidence intervals (panel models)"
#| fig-height: 7

panel_coef_data <- map_dfr(names(panel_models), function(mn) {
  m <- panel_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      model = outcome_labels[mn],
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

ggplot(panel_coef_data,
       aes(x = estimate, y = reorder(term_label, estimate),
           colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi),
                  position = position_dodge(width = 0.6), size = 0.4) +
  scale_colour_manual(values = c("All Pupils" = "#1565c0",
                                 "Disadvantaged" = "#c62828",
                                 "Non-Disadvantaged" = "#2e7d32"),
                      name = NULL) +
  labs(x = "Estimate (log scale)", y = NULL,
       subtitle = "Points = estimate; bars = 95% CI") +
  theme_pub
```


## Random Effects & Model Fit

```{r tbl-panel-random}
#| label: tbl-panel-random
#| tbl-cap: "Random effect variance components and model fit statistics (panel models)"

panel_fit <- map_dfr(names(panel_models), function(mn) {
  m <- panel_models[[mn]]
  d <- panel_diag[[mn]]
  vc <- as.data.frame(VarCorr(m))

  # Extract variance for each grouping
  var_year   <- vc$vcov[vc$grp == "year_label"]
  var_ofsted <- vc$vcov[vc$grp == "OFSTEDRATING_1"]
  var_la     <- vc$vcov[vc$grp == "LANAME:gor_name"]
  var_region <- vc$vcov[vc$grp == "gor_name"]
  var_resid  <- d$sigma^2

  tibble(
    Model = outcome_labels[mn],
    var_year   = round(var_year, 6),
    var_ofsted = round(var_ofsted, 6),
    var_la     = round(var_la, 6),
    var_region = round(var_region, 6),
    var_resid  = round(var_resid, 6),
    r2m = round(d$r2_marginal, 4),
    r2c = round(d$r2_conditional, 4),
    N = d$n_obs
  )
})

names(panel_fit)[2:8] <- c(
  paste0("\u03c3\u00b2", "<sub>year</sub>"),
  paste0("\u03c3\u00b2", "<sub>Ofsted</sub>"),
  paste0("\u03c3\u00b2", "<sub>LA</sub>"),
  paste0("\u03c3\u00b2", "<sub>region</sub>"),
  paste0("\u03c3\u00b2", "<sub>resid</sub>"),
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>"
)

panel_fit %>%
  kable(align = "lrrrrrrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


## Residual Diagnostics

```{r fig-panel-residuals}
#| label: fig-panel-residuals
#| fig-cap: "Residual diagnostics for the All Pupils panel model"
#| fig-height: 8

m_all <- panel_models$all

resid_data <- tibble(
  fitted = fitted(m_all),
  resid  = residuals(m_all, type = "pearson"),
  resid_std = resid / sd(resid)
)

p1 <- ggplot(resid_data, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.1, size = 0.5, colour = "#1565c0") +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, colour = "red", linewidth = 0.7) +
  labs(title = "Residuals vs Fitted", x = "Fitted values (log scale)", y = "Pearson residuals") +
  theme_pub

p2 <- ggplot(resid_data, aes(sample = resid_std)) +
  stat_qq(alpha = 0.1, size = 0.5, colour = "#1565c0") +
  stat_qq_line(colour = "red", linewidth = 0.7) +
  labs(title = "Normal Q-Q", x = "Theoretical quantiles", y = "Standardised residuals") +
  theme_pub

p3 <- ggplot(resid_data, aes(x = resid)) +
  geom_histogram(bins = 80, fill = "#1565c0", alpha = 0.7, colour = "white") +
  geom_vline(xintercept = 0, colour = "red", linetype = "dashed") +
  labs(title = "Residual Distribution", x = "Pearson residuals", y = "Frequency") +
  theme_pub

p4 <- ggplot(resid_data, aes(x = fitted, y = sqrt(abs(resid_std)))) +
  geom_point(alpha = 0.1, size = 0.5, colour = "#1565c0") +
  geom_smooth(method = "loess", se = FALSE, colour = "red", linewidth = 0.7) +
  labs(title = "Scale-Location", x = "Fitted values (log scale)",
       y = expression(sqrt("|Standardised residuals|"))) +
  theme_pub

(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Panel Model Diagnostics: All Pupils",
    subtitle = "log(ATT8SCR) ~ fixed effects + (1|year) + (1|Ofsted harmonised) + (1|region/LA)"
  )
```


## Random Intercepts

```{r fig-random-intercepts}
#| label: fig-random-intercepts
#| fig-cap: "Random intercept estimates by grouping level (All Pupils panel model)"
#| fig-height: 8

re <- ranef(m_all)

# Year random effects
re_year <- tibble(
  year = rownames(re$year_label),
  intercept = re$year_label[, 1]
)

p_year <- ggplot(re_year, aes(x = reorder(year, intercept), y = intercept)) +
  geom_col(fill = "#1565c0", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title = "Year Random Intercepts", x = NULL, y = "Intercept (log scale)") +
  theme_pub

# Ofsted random effects (harmonised)
re_ofsted <- tibble(
  rating = rownames(re$OFSTEDRATING_1),
  intercept = re$OFSTEDRATING_1[, 1]
)

p_ofsted <- ggplot(re_ofsted, aes(x = reorder(rating, -intercept), y = intercept)) +
  geom_col(fill = "#2e7d32", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title = "Ofsted Rating Random Intercepts (harmonised)", x = NULL, y = "Intercept (log scale)") +
  theme_pub

# Region random effects
re_region <- tibble(
  region = rownames(re$gor_name),
  intercept = re$gor_name[, 1]
)

p_region <- ggplot(re_region, aes(x = reorder(region, intercept), y = intercept)) +
  geom_col(fill = "#c62828", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title = "Region Random Intercepts", x = NULL, y = "Intercept (log scale)") +
  theme_pub

p_year / p_ofsted / p_region
```


# Per-Year Models (Analysis B)

Separate models are fitted for each academic year with the same fixed-effect specification but without a year term. This allows all coefficients to vary freely across years, revealing whether relationships between predictors and attainment are stable or shifting over time.


## Model Fit Comparison Across Years

```{r tbl-yearly-fit}
#| label: tbl-yearly-fit
#| tbl-cap: "Model fit statistics by year and outcome"

yearly_fit <- map_dfr(names(yearly_diag), function(yr) {
  map_dfr(names(yearly_diag[[yr]]), function(mn) {
    d <- yearly_diag[[yr]][[mn]]
    tibble(
      Year = yr,
      Outcome = outcome_labels[mn],
      N = ifelse(is.na(d$r2_marginal), NA_integer_, d$n_obs),
      r2m = round(d$r2_marginal, 4),
      r2c = round(d$r2_conditional, 4),
      sigma = round(d$sigma, 4)
    )
  })
})

names(yearly_fit)[5:7] <- c(
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>",
  "\u03c3"
)

yearly_fit %>%
  kable(align = "llrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Coefficient Stability Across Years

```{r fig-yearly-coefs}
#| label: fig-yearly-coefs
#| fig-cap: "How fixed-effect estimates change across academic years (All Pupils models)"
#| fig-height: 9

yearly_coef_data <- map_dfr(names(yearly_models), function(yr) {
  m <- yearly_models[[yr]]$all
  if (is.null(m)) return(NULL)

  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      year = yr,
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

if (nrow(yearly_coef_data) > 0) {
  ggplot(yearly_coef_data,
         aes(x = year, y = estimate, group = 1)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), alpha = 0.15, fill = "#1565c0") +
    geom_line(colour = "#1565c0", linewidth = 0.7) +
    geom_point(colour = "#1565c0", size = 2) +
    facet_wrap(~ term_label, scales = "free_y", ncol = 3) +
    labs(x = "Academic Year", y = "Estimate (log scale)",
         subtitle = "Shaded band = 95% CI") +
    theme_pub +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
}
```


## Side-by-Side Coefficient Table (Per-Year, All Pupils)

```{r tbl-yearly-coefs}
#| label: tbl-yearly-coefs
#| tbl-cap: "Fixed effect estimates by year (All Pupils outcome). Significance: *** p<0.001, ** p<0.01, * p<0.05"

yearly_coef_wide <- yearly_coef_data %>%
  mutate(
    # Look up p-values from the original models
    sig = map2_chr(year, term_raw, function(yr, term) {
      m <- yearly_models[[yr]]$all
      if (is.null(m)) return("")
      fe <- as.data.frame(summary(m)$coefficients)
      p <- fe[term, "Pr(>|t|)"]
      if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else ""
    }),
    cell = paste0(round(estimate, 4), sig)
  ) %>%
  select(term_label, year, cell) %>%
  pivot_wider(names_from = year, values_from = cell)

yearly_coef_wide %>%
  rename(Variable = term_label) %>%
  kable(align = c("l", rep("r", ncol(yearly_coef_wide) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


# Core Models (All 4 Years)

School Workforce Census variables (`remained_in_the_same_school`, `teachers_on_leadership_pay_range_percent`, `average_number_of_days_taken`) and KS2 prior attainment (`PTPRIORLO`) are not yet available for 2024-25. To enable cross-year comparison including the latest year, a **core specification** is fitted using only the 5 predictors available across all four years:

$$
\log(y_{ij}) = \beta_0 + \beta_1\log(\text{FSM\%}) + \beta_2\log(\text{Absence\%}) + \beta_3\log(\text{EAL\%}) + \beta_4\text{Admissions} + \beta_5\text{Gorard} + u_{\text{year}} + u_{\text{Ofsted}} + u_{\text{region}} + u_{\text{LA}} + \varepsilon_{ij}
$$


## Core Panel: Fixed Effects

```{r tbl-core-panel-fixed}
#| label: tbl-core-panel-fixed
#| tbl-cap: "Fixed effect estimates from core panel models (all 4 years, reduced specification)"

core_panel_fe <- map_dfr(names(core_models), function(mn) {
  m <- core_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    transmute(
      Model = outcome_labels[mn],
      Term = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      Estimate = round(Estimate, 4),
      `Std. Error` = round(`Std. Error`, 4),
      `t value` = round(`t value`, 2),
      `p value` = ifelse(`Pr(>|t|)` < 0.001, "< 0.001",
                         ifelse(`Pr(>|t|)` < 0.01, "< 0.01",
                                ifelse(`Pr(>|t|)` < 0.05, "< 0.05",
                                       round(`Pr(>|t|)`, 3))))
    )
})

core_panel_fe %>%
  kable(align = "llrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Core Panel: Random Effects & Model Fit

```{r tbl-core-panel-random}
#| label: tbl-core-panel-random
#| tbl-cap: "Random effect variance components and model fit statistics (core panel models)"

core_panel_fit <- map_dfr(names(core_models), function(mn) {
  m <- core_models[[mn]]
  d <- core_diag[[mn]]
  vc <- as.data.frame(VarCorr(m))

  var_year   <- vc$vcov[vc$grp == "year_label"]
  var_ofsted <- vc$vcov[vc$grp == "OFSTEDRATING_1"]
  var_la     <- vc$vcov[vc$grp == "LANAME:gor_name"]
  var_region <- vc$vcov[vc$grp == "gor_name"]
  var_resid  <- d$sigma^2

  tibble(
    Model = outcome_labels[mn],
    var_year   = round(var_year, 6),
    var_ofsted = round(var_ofsted, 6),
    var_la     = round(var_la, 6),
    var_region = round(var_region, 6),
    var_resid  = round(var_resid, 6),
    r2m = round(d$r2_marginal, 4),
    r2c = round(d$r2_conditional, 4),
    N = d$n_obs
  )
})

names(core_panel_fit)[2:8] <- c(
  paste0("\u03c3\u00b2", "<sub>year</sub>"),
  paste0("\u03c3\u00b2", "<sub>Ofsted</sub>"),
  paste0("\u03c3\u00b2", "<sub>LA</sub>"),
  paste0("\u03c3\u00b2", "<sub>region</sub>"),
  paste0("\u03c3\u00b2", "<sub>resid</sub>"),
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>"
)

core_panel_fit %>%
  kable(align = "lrrrrrrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


## Core Per-Year: Model Fit Comparison

```{r tbl-core-yearly-fit}
#| label: tbl-core-yearly-fit
#| tbl-cap: "Core model fit statistics by year and outcome (all 4 years)"

core_yearly_fit <- map_dfr(names(core_yearly_diag), function(yr) {
  map_dfr(names(core_yearly_diag[[yr]]), function(mn) {
    d <- core_yearly_diag[[yr]][[mn]]
    tibble(
      Year = yr,
      Outcome = outcome_labels[mn],
      N = ifelse(is.na(d$r2_marginal), NA_integer_, d$n_obs),
      r2m = round(d$r2_marginal, 4),
      r2c = round(d$r2_conditional, 4),
      sigma = round(d$sigma, 4)
    )
  })
})

names(core_yearly_fit)[5:7] <- c(
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>",
  "\u03c3"
)

core_yearly_fit %>%
  kable(align = "llrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Core Per-Year: Coefficient Stability

```{r fig-core-yearly-coefs}
#| label: fig-core-yearly-coefs
#| fig-cap: "How core fixed-effect estimates change across all 4 academic years (All Pupils)"
#| fig-height: 7

core_yearly_coef_data <- map_dfr(names(core_yearly_models), function(yr) {
  m <- core_yearly_models[[yr]]$all
  if (is.null(m)) return(NULL)

  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      year = yr,
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

if (nrow(core_yearly_coef_data) > 0) {
  ggplot(core_yearly_coef_data,
         aes(x = year, y = estimate, group = 1)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), alpha = 0.15, fill = "#e65100") +
    geom_line(colour = "#e65100", linewidth = 0.7) +
    geom_point(colour = "#e65100", size = 2) +
    facet_wrap(~ term_label, scales = "free_y", ncol = 3) +
    labs(x = "Academic Year", y = "Estimate (log scale)",
         subtitle = "Shaded band = 95% CI — core specification (5 predictors)") +
    theme_pub +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
}
```


## Core Per-Year: Side-by-Side Coefficients

```{r tbl-core-yearly-coefs}
#| label: tbl-core-yearly-coefs
#| tbl-cap: "Core fixed effect estimates by year (All Pupils). Significance: *** p<0.001, ** p<0.01, * p<0.05"

core_yearly_coef_wide <- core_yearly_coef_data %>%
  mutate(
    sig = map2_chr(year, term_raw, function(yr, term) {
      m <- core_yearly_models[[yr]]$all
      if (is.null(m)) return("")
      fe <- as.data.frame(summary(m)$coefficients)
      p <- fe[term, "Pr(>|t|)"]
      if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else ""
    }),
    cell = paste0(round(estimate, 4), sig)
  ) %>%
  select(term_label, year, cell) %>%
  pivot_wider(names_from = year, values_from = cell)

core_yearly_coef_wide %>%
  rename(Variable = term_label) %>%
  kable(align = c("l", rep("r", ncol(core_yearly_coef_wide) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


# Observed vs Predicted

The scatter plots below show observed Attainment 8 scores against core-model fitted values for each of the four academic years. Schools in **Brighton and Hove** are highlighted in orange. A linear regression line (red) with R² is overlaid on each panel; the dashed grey line marks perfect prediction.

## All Pupils

```{r fig-obs-vs-pred-all}
#| label: fig-obs-vs-pred-all
#| fig-cap: "Observed vs predicted Attainment 8 — All Pupils (core model, by year). Brighton and Hove schools in orange."
#| fig-height: 8
#| fig-width: 10

HIGHLIGHT_LA <- "Brighton and Hove"

# Prepare data with highlight flag
pred_data <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_core), !is.na(ATT8SCR)) %>%
  mutate(
    residual = ATT8SCR - predicted_ATT8SCR_core,
    abs_resid = abs(residual),
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA))
  ) %>%
  arrange(highlight)   # draw highlighted points on top

# Per-year R² labels
r2_labels <- pred_data %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR, predicted_ATT8SCR_core, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R\u00b2 = ", round(r2, 3), "  (n = ", scales::comma(n), ")"))

ggplot(pred_data, aes(x = predicted_ATT8SCR_core, y = ATT8SCR)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#c62828", linewidth = 0.6) +
  geom_text(data = r2_labels,
            aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.05, vjust = 1.4,
            size = 3.3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#1565c0",
                                  "Brighton and Hove" = "#e65100")) +
  scale_alpha_manual(values = c("Other" = 0.12,
                                 "Brighton and Hove" = 0.85), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 (core model)", y = "Observed ATT8",
       colour = NULL) +
  coord_equal() +
  theme_pub +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11))
```

```{r}
library(plotly)
library(dplyr)
library(scales)

# 1. Prepare data (same as your original logic)
pred_data <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_core), !is.na(ATT8SCR)) %>%
  mutate(
    residual = ATT8SCR - predicted_ATT8SCR_core,
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA)),
    # Custom tooltip text
    tooltip_text = paste0(
      "LA: ", LANAME, 
      "<br>Sch:", SCHNAME,
      "<br>Observed: ", round(ATT8SCR, 2),
      "<br>Predicted: ", round(predicted_ATT8SCR_core, 2),
      "<br>Residual: ", round(residual, 2)
    )
  ) %>%
  arrange(highlight)

# 2. Calculate R² labels
r2_labels <- pred_data %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR, predicted_ATT8SCR_core, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 3), " (n = ", comma(n), ")"))

# 3. Build the ggplot object
p <- ggplot(pred_data, aes(x = predicted_ATT8SCR_core, y = ATT8SCR)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", 
              linewidth = 0.5, linetype = "dashed") +
  # We add 'text' aesthetic here for the Plotly tooltip
  geom_point(aes(colour = highlight, alpha = highlight, text = tooltip_text), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#c62828", linewidth = 0.6) +
  # Note: geom_text can be finicky in ggplotly; using explicit coordinates helps
  geom_text(data = r2_labels,
            aes(x = 20, y = 85, label = label), 
            inherit.aes = FALSE, size = 3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#1565c0", 
                                 "Brighton and Hove" = "#e65100")) +
  scale_alpha_manual(values = c("Other" = 0.2, 
                                "Brighton and Hove" = 0.9), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 (core model)", y = "Observed ATT8", colour = NULL) +
  theme_minimal() + # Use minimal for cleaner plotly conversion
  theme(legend.position = "bottom")

# 4. Convert to Plotly
ggplotly(p, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", x = 0.4, y = -0.1),
    margin = list(t = 50) # Space for facet titles
  )

```



## Disadvantaged Pupils

```{r fig-obs-vs-pred-disadv}
#| label: fig-obs-vs-pred-disadv
#| fig-cap: "Observed vs predicted Attainment 8 — Disadvantaged Pupils (core model, by year). Brighton and Hove schools in orange."
#| fig-height: 8
#| fig-width: 10

pred_disadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_FSM6CLA1A_core), !is.na(ATT8SCR_FSM6CLA1A)) %>%
  mutate(
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA))
  ) %>%
  arrange(highlight)

r2_disadv <- pred_disadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_FSM6CLA1A, predicted_ATT8SCR_FSM6CLA1A_core,
             use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R\u00b2 = ", round(r2, 3), "  (n = ", scales::comma(n), ")"))

ggplot(pred_disadv, aes(x = predicted_ATT8SCR_FSM6CLA1A_core,
                         y = ATT8SCR_FSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#c62828", linewidth = 0.6) +
  geom_text(data = r2_disadv,
            aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.05, vjust = 1.4,
            size = 3.3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#1565c0",
                                  "Brighton and Hove" = "#e65100")) +
  scale_alpha_manual(values = c("Other" = 0.12,
                                 "Brighton and Hove" = 0.85), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Disadvantaged (core model)",
       y = "Observed ATT8 — Disadvantaged",
       colour = NULL) +
  coord_equal() +
  theme_pub +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11))
```

```{r}
library(plotly)
library(dplyr)
library(scales)

# 1. Prepare data for Disadvantaged subset
pred_disadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_FSM6CLA1A_core), !is.na(ATT8SCR_FSM6CLA1A)) %>%
  mutate(
    residual = ATT8SCR_FSM6CLA1A - predicted_ATT8SCR_FSM6CLA1A_core,
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA)),
    # Enhanced tooltip for disadvantaged context
    tooltip_text = paste0(
      "<b>", SCHNAME, "</b><br>",  # Assumes SCHNAME exists in your data
      "LA: ", LANAME, 
      "<br>Sch:", SCHNAME,
      "<br>Observed (Disadv): ", round(ATT8SCR_FSM6CLA1A, 2),
      "<br>Predicted (Disadv): ", round(predicted_ATT8SCR_FSM6CLA1A_core, 2),
      "<br>Residual: ", round(residual, 2)
    )
  ) %>%
  arrange(highlight)

# 2. Calculate R² for the Disadvantaged model
r2_disadv_stats <- pred_disadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_FSM6CLA1A, predicted_ATT8SCR_FSM6CLA1A_core, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 3), " (n = ", comma(n), ")"))

# 3. Build the ggplot object
p_disadv <- ggplot(pred_disadv, aes(x = predicted_ATT8SCR_FSM6CLA1A_core, 
                                   y = ATT8SCR_FSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", 
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight, text = tooltip_text), size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#c62828", linewidth = 0.6) +
  # Positioning labels at safe numeric coordinates (e.g., x=10, y=85)
  geom_text(data = r2_disadv_stats,
            aes(x = 10, y = 85, label = label), 
            inherit.aes = FALSE, size = 3, colour = "grey30", hjust = 0) +
  scale_colour_manual(values = c("Other" = "#1565c0", 
                                 "Brighton and Hove" = "#e65100")) +
  scale_alpha_manual(values = c("Other" = 0.15, 
                                "Brighton and Hove" = 0.9), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Disadvantaged", 
       y = "Observed ATT8 — Disadvantaged", 
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))

# 4. Final Plotly conversion
ggplotly(p_disadv, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", x = 0.3, y = -0.15),
    margin = list(t = 60)
  )
```


## Non-Disadvantaged Pupils

```{r fig-obs-vs-pred-nondisadv}
#| label: fig-obs-vs-pred-nondisadv
#| fig-cap: "Observed vs predicted Attainment 8 — Non-Disadvantaged Pupils (core model, by year). Brighton and Hove schools in orange."
#| fig-height: 8
#| fig-width: 10

pred_nondisadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_NFSM6CLA1A_core), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
  mutate(
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA))
  ) %>%
  arrange(highlight)

r2_nondisadv <- pred_nondisadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_NFSM6CLA1A, predicted_ATT8SCR_NFSM6CLA1A_core,
             use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R\u00b2 = ", round(r2, 3), "  (n = ", scales::comma(n), ")"))

ggplot(pred_nondisadv, aes(x = predicted_ATT8SCR_NFSM6CLA1A_core,
                            y = ATT8SCR_NFSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#c62828", linewidth = 0.6) +
  geom_text(data = r2_nondisadv,
            aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.05, vjust = 1.4,
            size = 3.3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#1565c0",
                                  "Brighton and Hove" = "#e65100")) +
  scale_alpha_manual(values = c("Other" = 0.12,
                                 "Brighton and Hove" = 0.85), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Non-Disadvantaged (core model)",
       y = "Observed ATT8 — Non-Disadvantaged",
       colour = NULL) +
  coord_equal() +
  theme_pub +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11))
```

```{r}
library(plotly)
library(dplyr)
library(scales)

# 1. Prepare data for Non-Disadvantaged subset
pred_nondisadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_NFSM6CLA1A_core), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
  mutate(
    residual = ATT8SCR_NFSM6CLA1A - predicted_ATT8SCR_NFSM6CLA1A_core,
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA)),
    # Custom tooltip
    tooltip_text = paste0(
      "<b>", SCHNAME, "</b><br>",
      "LA: ", LANAME, 
      "<br>Sch:", SCHNAME,
      "<br>Observed (Non-Disadv): ", round(ATT8SCR_NFSM6CLA1A, 2),
      "<br>Predicted (Non-Disadv): ", round(predicted_ATT8SCR_NFSM6CLA1A_core, 2),
      "<br>Residual: ", round(residual, 2)
    )
  ) %>%
  arrange(highlight)

# 2. Calculate R² for the Non-Disadvantaged model
r2_nondisadv_stats <- pred_nondisadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_NFSM6CLA1A, predicted_ATT8SCR_NFSM6CLA1A_core, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 3), " (n = ", comma(n), ")"))

# 3. Build the ggplot object
p_nondisadv <- ggplot(pred_nondisadv, aes(x = predicted_ATT8SCR_NFSM6CLA1A_core, 
                                         y = ATT8SCR_NFSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", 
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight, text = tooltip_text), size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#c62828", linewidth = 0.6) +
  # Placed at higher coordinates as Non-Disadv scores are typically higher
  geom_text(data = r2_nondisadv_stats,
            aes(x = 20, y = 88, label = label), 
            inherit.aes = FALSE, size = 3, colour = "grey30", hjust = 0) +
  scale_colour_manual(values = c("Other" = "#1565c0", 
                                 "Brighton and Hove" = "#e65100")) +
  scale_alpha_manual(values = c("Other" = 0.15, 
                                "Brighton and Hove" = 0.9), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Non-Disadvantaged", 
       y = "Observed ATT8 — Non-Disadvantaged", 
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))

# 4. Final Plotly conversion
ggplotly(p_nondisadv, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", x = 0.3, y = -0.15),
    margin = list(t = 60)
  )
```


# Residual Geography

```{r fig-residual-by-region}
#| label: fig-residual-by-region
#| fig-cap: "Distribution of model residuals by Government Office Region"
#| fig-height: 6

if ("gor_name" %in% names(pred_data) && sum(!is.na(pred_data$gor_name)) > 0) {

  region_resid <- pred_data %>%
    filter(!is.na(gor_name)) %>%
    group_by(gor_name) %>%
    summarise(
      mean_resid = mean(residual, na.rm = TRUE),
      median_resid = median(residual, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    arrange(mean_resid)

  ggplot(pred_data %>% filter(!is.na(gor_name)),
         aes(x = reorder(gor_name, residual, FUN = median), y = residual)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_boxplot(fill = "#1565c0", alpha = 0.3, outlier.size = 0.3, outlier.alpha = 0.2) +
    coord_flip() +
    labs(x = NULL, y = "Residual (Observed - Predicted ATT8)",
         subtitle = "Positive = outperforming model; Negative = underperforming") +
    theme_pub
}
```


# Disadvantage Gap

```{r fig-disadvantage-gap}
#| label: fig-disadvantage-gap
#| fig-cap: "Disadvantage attainment gap by academic year"
#| fig-height: 5

if (all(c("ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A") %in% names(model_data))) {
  gap_data <- model_data %>%
    filter(!is.na(ATT8SCR_FSM6CLA1A), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
    mutate(gap = ATT8SCR_NFSM6CLA1A - ATT8SCR_FSM6CLA1A)

  gap_summary <- gap_data %>%
    group_by(year_label) %>%
    summarise(
      mean_gap = mean(gap, na.rm = TRUE),
      median_gap = median(gap, na.rm = TRUE),
      q25 = quantile(gap, 0.25, na.rm = TRUE),
      q75 = quantile(gap, 0.75, na.rm = TRUE),
      .groups = "drop"
    )

  ggplot(gap_data, aes(x = year_label, y = gap)) +
    geom_boxplot(fill = "#c62828", alpha = 0.3, outlier.size = 0.3, outlier.alpha = 0.2) +
    geom_point(data = gap_summary, aes(y = mean_gap),
               colour = "#c62828", size = 3, shape = 18) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(x = "Academic Year",
         y = "ATT8 Gap (Non-Disadvantaged - Disadvantaged)",
         subtitle = "Diamond = mean gap; Box = IQR") +
    theme_pub
}
```


# Appendix: Model Formulae {.appendix}

## Panel Model (Analysis A)

All four years pooled with `year_label` as a random intercept:

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  log(remained_in_the_same_school) +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Per-Year Model (Analysis B)

Separate model for each academic year (no year term):

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  log(remained_in_the_same_school) +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Core Panel Model (Analysis C)

Reduced specification using only the 5 predictors available for all 4 years (including 2024-25). Drops workforce and prior-attainment variables:

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  ADMPOL_PT + gorard_segregation +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Core Per-Year Model (Analysis D)

Same 5 core fixed effects, separate model per year (no year term):

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  ADMPOL_PT + gorard_segregation +
  (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Variable Definitions

```{r tbl-variable-defs}
#| label: tbl-variable-defs
#| tbl-cap: "Variable definitions from DfE metadata"

model_vars <- c(
  "ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A",
  "PTFSM6CLA1A", "PERCTOT", "PNUMEAL", "PTPRIORLO",
  "ADMPOL_PT", "gorard_segregation",
  "remained_in_the_same_school",
  "teachers_on_leadership_pay_range_percent",
  "average_number_of_days_taken",
  "OFSTEDRATING_1", "gor_name", "LANAME", "year_label"
)

tibble(
  `Variable Code` = model_vars,
  Description = var_labels(model_vars),
  `In Formula As` = case_when(
    model_vars %in% c("PTFSM6CLA1A", "PERCTOT", "PNUMEAL",
                       "remained_in_the_same_school",
                       "average_number_of_days_taken",
                       "ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A") ~
      paste0("log(", model_vars, ")"),
    model_vars %in% c("OFSTEDRATING_1", "gor_name", "LANAME", "year_label") ~
      paste0("(1 | ", model_vars, ")"),
    TRUE ~ model_vars
  ),
  Source = case_when(
    model_vars %in% c("ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A",
                       "PTFSM6CLA1A", "PTPRIORLO", "ADMPOL_PT") ~ "KS4 Performance Tables",
    model_vars %in% c("PERCTOT") ~ "Absence Statistics",
    model_vars %in% c("PNUMEAL") ~ "School Census",
    model_vars %in% c("OFSTEDRATING_1") ~ "Ofsted Inspections (harmonised)",
    model_vars %in% c("gorard_segregation") ~ "Derived (Gorard index)",
    model_vars %in% c("remained_in_the_same_school",
                       "teachers_on_leadership_pay_range_percent",
                       "average_number_of_days_taken") ~ "School Workforce Census",
    model_vars %in% c("gor_name", "LANAME") ~ "Edubase",
    model_vars == "year_label" ~ "Constructed",
    TRUE ~ ""
  )
) %>%
  kable(align = "llll") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 10)
```
