---
title: "School Attainment Multilevel Model Results"
subtitle: "Multilevel analysis of KS4 Attainment 8 scores across English secondary schools"
author: "Adam Dennett with some assistance from Claude Code"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: ucl-theme.scss
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
    df-print: kable
    embed-resources: true
    include-in-header: header-logos.html
    include-after-body: footer-logos.html
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 5
    geometry: margin=2.5cm
    keep-tex: false
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| label: setup
#| include: false

library(tidyverse)
library(lme4)
library(lmerTest)
library(performance)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(here)

source(here::here("R", "helpers.R"))

# ---- Load all saved outputs ----
panel_data       <- readRDS(here::here("data", "panel_data.rds"))
model_data       <- readRDS(here::here("data", "model_data.rds"))
panel_models     <- readRDS(here::here("data", "models.rds"))
panel_diag       <- readRDS(here::here("data", "model_diagnostics.rds"))
yearly_models    <- readRDS(here::here("data", "models_yearly.rds"))
yearly_diag      <- readRDS(here::here("data", "model_diagnostics_yearly.rds"))

# Core models (reduced specification, all 4 years including 2024-25)
core_model_data  <- readRDS(here::here("data", "model_data_core.rds"))
core_models      <- readRDS(here::here("data", "models_core.rds"))
core_diag        <- readRDS(here::here("data", "model_diagnostics_core.rds"))
core_yearly_models <- readRDS(here::here("data", "models_yearly_core.rds"))
core_yearly_diag <- readRDS(here::here("data", "model_diagnostics_yearly_core.rds"))

# Imputed full models (Analysis E: 9 predictors, all 4 years with carry-forward imputed 2024-25)
imputed_model_data <- readRDS(here::here("data", "model_data_imputed.rds"))
imputed_models     <- readRDS(here::here("data", "models_imputed.rds"))
imputed_diag       <- readRDS(here::here("data", "model_diagnostics_imputed.rds"))

# Outcome configuration
OUTCOMES <- list(
  all               = "ATT8SCR",
  disadvantaged     = "ATT8SCR_FSM6CLA1A",
  non_disadvantaged = "ATT8SCR_NFSM6CLA1A"
)

outcome_labels <- c(
  all               = "All Pupils",
  disadvantaged     = "Disadvantaged",
  non_disadvantaged = "Non-Disadvantaged"
)

# Ensure year_label is a factor for consistent ordering
model_data <- model_data %>%
  mutate(year_label = factor(year_label))

# Theme for publication-quality plots
theme_pub <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(colour = "grey40", size = 10),
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )
```


# Data Overview

For full details on data sources, panel construction, variable definitions, descriptive statistics, distributions, and bivariate relationships, see the [Data Overview report](https://adamdennett.github.io/school_attainment_tool/data_overview.html).

::: {.callout-tip}
## Model code & case study
The full [`lme4`](https://cran.r-project.org/web/packages/lme4/refman/lme4.html) code for all model specifications (Analyses A–E) is available in the [Model Experiments report](https://adamdennett.github.io/school_attainment_tool/model_experiments.html), which provides human-readable, self-contained implementations for manual inspection and reproducibility. A detailed application of the model to a single local authority — including maps, non-linear effect interpretation, and policy recommendations — is presented in the [Brighton & Hove Case Study](https://adamdennett.github.io/school_attainment_tool/brighton_case_study.html).
:::


::: {.callout-note}
## Sample size note

The full 9-predictor models (Analyses A, B, and E) are fitted on
approximately 12,200 school-year observations from ~13,000 mainstream
(Academy + Maintained) school-year records. The main source of attrition
is the `average_number_of_days_taken` (teacher sickness absence) variable,
which drops ~570 observations — predominantly in 2021-22 and 2022-23
where the workforce census sickness data was patchier (likely due to
COVID-era reporting disruption). Around 420 of these are NA values and
~150 are zeros excluded by the `> 0` requirement for log transformation.
Academies account for ~80% of the losses. See the
[dropped-rows diagnostic](model_experiments.html#dropped-rows-diagnostic)
in the Model Experiments report for a full breakdown.
:::

# Panel Models (Analysis A)

The panel models pool all four academic years, with `year_label` entering as a random intercept alongside Ofsted rating and local authority (nested within region). This specification treats year-to-year variation as a variance component rather than a fixed trend.

$$
\log(y_{ij}) = \mathbf{X}_{ij}\boldsymbol{\beta} + u_{\text{year}} + u_{\text{Ofsted}} + u_{\text{region}} + u_{\text{LA}} + \varepsilon_{ij}
$$


## Fixed Effects

```{r tbl-panel-fixed}
#| label: tbl-panel-fixed
#| tbl-cap: "Fixed effect estimates from panel models (all years pooled, year as random effect)"

# Build combined table
panel_fe <- map_dfr(names(panel_models), function(mn) {
  m <- panel_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    transmute(
      Model = outcome_labels[mn],
      Term = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      Estimate = round(Estimate, 4),
      `Std. Error` = round(`Std. Error`, 4),
      `t value` = round(`t value`, 2),
      `p value` = ifelse(`Pr(>|t|)` < 0.001, "< 0.001",
                         ifelse(`Pr(>|t|)` < 0.01, "< 0.01",
                                ifelse(`Pr(>|t|)` < 0.05, "< 0.05",
                                       round(`Pr(>|t|)`, 3))))
    )
})

panel_fe %>%
  kable(align = "llrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Coefficient Comparison Plot

```{r fig-panel-coefs}
#| label: fig-panel-coefs
#| fig-cap: "Fixed effect estimates with 95% confidence intervals (panel models)"
#| fig-height: 7

panel_coef_data <- map_dfr(names(panel_models), function(mn) {
  m <- panel_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      model = outcome_labels[mn],
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

ggplot(panel_coef_data,
       aes(x = estimate, y = reorder(term_label, estimate),
           colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi),
                  position = position_dodge(width = 0.6), size = 0.4) +
  scale_colour_manual(values = c("All Pupils" = "#2e6260",
                                 "Disadvantaged" = "#4e3c56",
                                 "Non-Disadvantaged" = "#abc766"),
                      name = NULL) +
  labs(x = "Estimate (log scale)", y = NULL,
       subtitle = "Points = estimate; bars = 95% CI") +
  theme_pub
```


## Random Effects & Model Fit

The multilevel model includes four levels of **random intercepts**, each allowing the baseline log(ATT8) to vary across groups:

1. **Academic year** (`year_label`) — captures system-wide year-on-year shifts in attainment (e.g. grade inflation, COVID recovery effects) that affect all schools equally within a given year.
2. **Ofsted rating** (`OFSTEDRATING_1`, harmonised to four levels) — allows schools with different inspection outcomes to have systematically different attainment baselines, beyond what the fixed-effect predictors explain.
3. **Government Office Region** (`gor_name`) — accounts for broad regional differences in attainment (e.g. London vs. the North East), which may reflect regional labour markets, funding patterns, or cultural factors.
4. **Local authority nested within region** (`LANAME:gor_name`) — the finest grouping level, capturing LA-specific effects such as local policy environments, demographic composition, or authority-level support services, over and above the regional baseline.

The table below reports the estimated variance for each random effect, along with the residual variance and overall model fit (marginal R², which reflects fixed effects only, and conditional R², which includes both fixed and random effects).

```{r tbl-panel-random}
#| label: tbl-panel-random
#| tbl-cap: "Random effect variance components and model fit statistics (panel models)"

panel_fit <- map_dfr(names(panel_models), function(mn) {
  m <- panel_models[[mn]]
  d <- panel_diag[[mn]]
  vc <- as.data.frame(VarCorr(m))

  # Extract variance for each grouping
  var_year   <- vc$vcov[vc$grp == "year_label"]
  var_ofsted <- vc$vcov[vc$grp == "OFSTEDRATING_1"]
  var_la     <- vc$vcov[vc$grp == "LANAME:gor_name"]
  var_region <- vc$vcov[vc$grp == "gor_name"]
  var_resid  <- d$sigma^2

  tibble(
    Model = outcome_labels[mn],
    var_year   = round(var_year, 6),
    var_ofsted = round(var_ofsted, 6),
    var_la     = round(var_la, 6),
    var_region = round(var_region, 6),
    var_resid  = round(var_resid, 6),
    r2m = round(d$r2_marginal, 4),
    r2c = round(d$r2_conditional, 4),
    N = d$n_obs
  )
})

names(panel_fit)[2:8] <- c(
  paste0("\u03c3\u00b2", "<sub>year</sub>"),
  paste0("\u03c3\u00b2", "<sub>Ofsted</sub>"),
  paste0("\u03c3\u00b2", "<sub>LA</sub>"),
  paste0("\u03c3\u00b2", "<sub>region</sub>"),
  paste0("\u03c3\u00b2", "<sub>resid</sub>"),
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>"
)

panel_fit %>%
  kable(format = "html", align = "lrrrrrrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


## Random Intercepts

```{r fig-random-intercepts}
#| label: fig-random-intercepts
#| fig-cap: "Random intercept estimates by grouping level (All Pupils panel model)"
#| fig-height: 8

m_all <- panel_models$all
re <- ranef(m_all)

# Year random effects
re_year <- tibble(
  year = rownames(re$year_label),
  intercept = re$year_label[, 1]
)

p_year <- ggplot(re_year, aes(x = reorder(year, intercept), y = intercept)) +
  geom_col(fill = "#2e6260", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title = "Year Random Intercepts", x = NULL, y = "Intercept (log scale)") +
  theme_pub

# Ofsted random effects (harmonised)
re_ofsted <- tibble(
  rating = rownames(re$OFSTEDRATING_1),
  intercept = re$OFSTEDRATING_1[, 1]
)

p_ofsted <- ggplot(re_ofsted, aes(x = reorder(rating, -intercept), y = intercept)) +
  geom_col(fill = "#abc766", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title = "Ofsted Rating Random Intercepts (harmonised)", x = NULL, y = "Intercept (log scale)") +
  theme_pub

# Region random effects
re_region <- tibble(
  region = rownames(re$gor_name),
  intercept = re$gor_name[, 1]
)

p_region <- ggplot(re_region, aes(x = reorder(region, intercept), y = intercept)) +
  geom_col(fill = "#4e3c56", alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  labs(title = "Region Random Intercepts", x = NULL, y = "Intercept (log scale)") +
  theme_pub

p_year / p_ofsted / p_region
```


## Null Model: Variance Structure

The violin and box plots below show the distribution of Attainment 8 scores across the random-effect grouping variables. These visualise the between-group variance that the null (intercept-only) multilevel model captures — the same structure shown in more detail in the variance components table above.

```{r fig-null-violin-ofsted}
#| label: fig-null-violin-ofsted
#| fig-cap: "Distribution of Attainment 8 scores by Ofsted rating — visualising the null model variance structure"
#| fig-height: 6

# Fit null model: ATT8 with only Ofsted as random intercept
null_ofsted <- lmer(log(ATT8SCR) ~ 1 + (1 | OFSTEDRATING_1), data = model_data)
grand_mean_ofsted <- exp(fixef(null_ofsted)[1])

# Extract shrunken group means (BLUPs)
re_ofsted_null <- ranef(null_ofsted)$OFSTEDRATING_1 %>%
  as.data.frame() %>%
  rownames_to_column("OFSTEDRATING_1") %>%
  rename(blup = `(Intercept)`) %>%
  mutate(shrunken_mean = exp(fixef(null_ofsted)[1] + blup))

# Raw group means
raw_means_ofsted <- model_data %>%
  filter(!is.na(OFSTEDRATING_1), !is.na(ATT8SCR)) %>%
  group_by(OFSTEDRATING_1) %>%
  summarise(raw_mean = mean(ATT8SCR, na.rm = TRUE), n = n(), .groups = "drop") %>%
  left_join(re_ofsted_null, by = "OFSTEDRATING_1")

model_data %>%
  filter(!is.na(OFSTEDRATING_1), !is.na(ATT8SCR)) %>%
  ggplot(aes(x = OFSTEDRATING_1, y = ATT8SCR, fill = OFSTEDRATING_1)) +
  geom_violin(alpha = 0.3, colour = NA) +
  geom_boxplot(width = 0.15, alpha = 0.7, outlier.size = 0.5, outlier.alpha = 0.3) +
  geom_hline(yintercept = grand_mean_ofsted, linetype = "dashed",
             colour = "black", linewidth = 0.8) +
  geom_point(data = raw_means_ofsted, aes(y = raw_mean),
             shape = 4, size = 3, colour = "#7b132b", stroke = 1.2) +
  geom_point(data = raw_means_ofsted, aes(y = shrunken_mean),
             shape = 18, size = 4, colour = "#2e6260") +
  annotate("text", x = 0.55, y = grand_mean_ofsted + 1.5,
           label = paste0("Grand mean = ", round(grand_mean_ofsted, 1)),
           hjust = 0, size = 3.2, colour = "grey30") +
  labs(title = "ATT8 Distribution by Ofsted Rating (Null Model Structure)",
       subtitle = "Dashed = grand mean | Teal diamond = shrunken mean (BLUP) | Dark red cross = raw mean",
       x = "Ofsted Rating (harmonised)", y = "Attainment 8 Score") +
  theme_pub +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r fig-null-violin-region}
#| label: fig-null-violin-region
#| fig-cap: "Distribution of Attainment 8 scores by Government Office Region — null model variance structure"
#| fig-height: 6

null_region <- lmer(log(ATT8SCR) ~ 1 + (1 | gor_name), data = model_data)
grand_mean_region <- exp(fixef(null_region)[1])

re_region_null <- ranef(null_region)$gor_name %>%
  as.data.frame() %>%
  rownames_to_column("gor_name") %>%
  rename(blup = `(Intercept)`) %>%
  mutate(shrunken_mean = exp(fixef(null_region)[1] + blup))

raw_means_region <- model_data %>%
  filter(!is.na(gor_name), !is.na(ATT8SCR)) %>%
  group_by(gor_name) %>%
  summarise(raw_mean = mean(ATT8SCR, na.rm = TRUE), n = n(), .groups = "drop") %>%
  left_join(re_region_null, by = "gor_name")

model_data %>%
  filter(!is.na(gor_name), !is.na(ATT8SCR)) %>%
  ggplot(aes(x = reorder(gor_name, ATT8SCR, FUN = median),
             y = ATT8SCR, fill = gor_name)) +
  geom_violin(alpha = 0.3, colour = NA) +
  geom_boxplot(width = 0.15, alpha = 0.7, outlier.size = 0.5, outlier.alpha = 0.3) +
  geom_hline(yintercept = grand_mean_region, linetype = "dashed",
             colour = "black", linewidth = 0.8) +
  geom_point(data = raw_means_region,
             aes(x = reorder(gor_name, raw_mean), y = raw_mean),
             shape = 4, size = 3, colour = "#7b132b", stroke = 1.2) +
  geom_point(data = raw_means_region,
             aes(x = reorder(gor_name, raw_mean), y = shrunken_mean),
             shape = 18, size = 4, colour = "#2e6260") +
  coord_flip() +
  labs(title = "ATT8 Distribution by Region (Null Model Structure)",
       subtitle = "Dashed = grand mean | Teal diamond = shrunken mean (BLUP) | Dark red cross = raw mean",
       x = NULL, y = "Attainment 8 Score") +
  theme_pub +
  theme(legend.position = "none")
```

```{r fig-null-violin-ofsted-year}
#| label: fig-null-violin-ofsted-year
#| fig-cap: "Distribution of Attainment 8 by Ofsted rating, faceted by academic year"
#| fig-height: 8

model_data %>%
  filter(!is.na(OFSTEDRATING_1), !is.na(ATT8SCR)) %>%
  ggplot(aes(x = OFSTEDRATING_1, y = ATT8SCR, fill = OFSTEDRATING_1)) +
  geom_violin(alpha = 0.3, colour = NA) +
  geom_boxplot(width = 0.15, alpha = 0.7, outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_wrap(~ year_label, ncol = 2) +
  labs(title = "ATT8 by Ofsted Rating — by Academic Year",
       subtitle = "Violin + boxplot shows the distributional shape within each Ofsted category",
       x = "Ofsted Rating (harmonised)", y = "Attainment 8 Score") +
  theme_pub +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```


## Caterpillar Plot: LA Random Intercepts

The caterpillar plot below displays the estimated random intercepts for each local authority (nested within region), ordered by magnitude. Error bars show 95% confidence intervals (±1.96 posterior SE). Local authorities whose interval does not cross zero are meaningfully different from the overall mean, after accounting for fixed effects and other random effects. Brighton and Hove is highlighted in pink.

```{r fig-caterpillar-la}
#| label: fig-caterpillar-la
#| fig-cap: "Caterpillar plot of LA random intercepts (imputed full panel model, All Pupils). Top and bottom 10 LAs labelled; Brighton and Hove in pink."
#| fig-height: 8
#| fig-width: 10

library(ggrepel)

# Extract LA random effects with conditional variances
re_list_imp <- ranef(imputed_models$all, condVar = TRUE)
re_la <- as.data.frame(re_list_imp$`LANAME:gor_name`)
re_la$group <- rownames(re_list_imp$`LANAME:gor_name`)

# Posterior standard errors
se_la <- sqrt(attr(re_list_imp$`LANAME:gor_name`, "postVar")[1, 1, ])
re_la$se <- se_la

re_la <- re_la %>%
  mutate(
    la_name = sub(":.*$", "", group),
    ci_lo = `(Intercept)` - 1.96 * se,
    ci_hi = `(Intercept)` + 1.96 * se,
    sig = ci_lo > 0 | ci_hi < 0,
    highlight = ifelse(grepl("Brighton", la_name), "Brighton and Hove", "Other")
  )

# Top and bottom 10 for labelling (+ always include Brighton)
extreme_la <- c(
  head(re_la$la_name[order(re_la$`(Intercept)`)], 10),
  tail(re_la$la_name[order(re_la$`(Intercept)`)], 10)
)
if (any(grepl("Brighton", re_la$la_name))) {
  extreme_la <- unique(c(extreme_la,
                          re_la$la_name[grepl("Brighton", re_la$la_name)]))
}

ggplot(re_la, aes(x = reorder(group, `(Intercept)`), y = `(Intercept)`)) +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                width = 0, colour = "grey60", linewidth = 0.3) +
  geom_point(aes(colour = highlight), size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_text_repel(
    data = re_la %>% filter(la_name %in% extreme_la),
    aes(label = la_name),
    size = 2.5, max.overlaps = 25, force = 4, nudge_y = 0.005,
    segment.colour = "grey60", segment.size = 0.3
  ) +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                  "Brighton and Hove" = "#e16fca"),
                      guide = "none") +
  labs(title = "LA Random Intercepts — Imputed Full Panel Model (All Pupils)",
       subtitle = "Ordered by estimate; error bars = 95% CI. LAs whose CI excludes 0 differ meaningfully from average.",
       x = "Local Authority (ordered by random intercept)",
       y = "Random Intercept (log scale)") +
  theme_pub +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

```{r caterpillar-helper}
#| label: caterpillar-helper
#| include: false

# Reusable function for LA caterpillar plots
plot_la_caterpillar <- function(model, title_label) {
  re_list <- ranef(model, condVar = TRUE)
  re_df <- as.data.frame(re_list$`LANAME:gor_name`)
  re_df$group <- rownames(re_list$`LANAME:gor_name`)
  se <- sqrt(attr(re_list$`LANAME:gor_name`, "postVar")[1, 1, ])
  re_df$se <- se

  re_df <- re_df %>%
    mutate(
      la_name = sub(":.*$", "", group),
      ci_lo = `(Intercept)` - 1.96 * se,
      ci_hi = `(Intercept)` + 1.96 * se,
      sig = ci_lo > 0 | ci_hi < 0,
      highlight = ifelse(grepl("Brighton", la_name), "Brighton and Hove", "Other")
    )

  # Top and bottom 10 + always include Brighton
  extreme <- c(
    head(re_df$la_name[order(re_df$`(Intercept)`)], 10),
    tail(re_df$la_name[order(re_df$`(Intercept)`)], 10)
  )
  if (any(grepl("Brighton", re_df$la_name))) {
    extreme <- unique(c(extreme,
                         re_df$la_name[grepl("Brighton", re_df$la_name)]))
  }

  ggplot(re_df, aes(x = reorder(group, `(Intercept)`), y = `(Intercept)`)) +
    geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi),
                  width = 0, colour = "grey60", linewidth = 0.3) +
    geom_point(aes(colour = highlight), size = 1.5) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
    geom_text_repel(
      data = re_df %>% filter(la_name %in% extreme),
      aes(label = la_name),
      size = 2.5, max.overlaps = 25, force = 4, nudge_y = 0.005,
      segment.colour = "grey60", segment.size = 0.3
    ) +
    scale_colour_manual(values = c("Other" = "#49a0c4",
                                    "Brighton and Hove" = "#e16fca"),
                        guide = "none") +
    labs(title = title_label,
         subtitle = "Ordered by estimate; error bars = 95% CI. LAs whose CI excludes 0 differ meaningfully from average.",
         x = "Local Authority (ordered by random intercept)",
         y = "Random Intercept (log scale)") +
    theme_pub +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
}
```

The plots below show the same caterpillar format for **disadvantaged** and **non-disadvantaged** Attainment 8 models. The LA ordering will differ between the two — local authorities that perform well for disadvantaged pupils are not necessarily the same ones that perform well for non-disadvantaged pupils.

```{r fig-caterpillar-la-disadv}
#| label: fig-caterpillar-la-disadv
#| fig-cap: "Caterpillar plot of LA random intercepts (imputed full panel model, Disadvantaged Pupils). Top and bottom 10 LAs labelled; Brighton and Hove in pink."
#| fig-height: 8
#| fig-width: 10

plot_la_caterpillar(imputed_models$disadvantaged,
                    "LA Random Intercepts — Imputed Full Panel Model (Disadvantaged Pupils)")
```

```{r fig-caterpillar-la-nondis}
#| label: fig-caterpillar-la-nondis
#| fig-cap: "Caterpillar plot of LA random intercepts (imputed full panel model, Non-Disadvantaged Pupils). Top and bottom 10 LAs labelled; Brighton and Hove in pink."
#| fig-height: 8
#| fig-width: 10

plot_la_caterpillar(imputed_models$non_disadvantaged,
                    "LA Random Intercepts — Imputed Full Panel Model (Non-Disadvantaged Pupils)")
```


## Residual Diagnostics

Having examined the random effects structure, we now turn to the model's residuals to check whether the key assumptions of the multilevel model are met: linearity, homoscedasticity (constant variance), and approximate normality of residuals.

```{r fig-panel-residuals}
#| label: fig-panel-residuals
#| fig-cap: "Residual diagnostics for the All Pupils panel model"
#| fig-height: 8

m_all <- panel_models$all

resid_data <- tibble(
  fitted = fitted(m_all),
  resid  = residuals(m_all, type = "pearson")
) %>%
  filter(!is.na(fitted), !is.na(resid)) %>%
  mutate(resid_std = resid / sd(resid))

p1 <- ggplot(resid_data, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.1, size = 0.5, colour = "#2e6260") +
  geom_hline(yintercept = 0, colour = "#7b132b", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, colour = "#7b132b", linewidth = 0.7) +
  labs(title = "Residuals vs Fitted", x = "Fitted values (log scale)", y = "Pearson residuals") +
  theme_pub

p2 <- ggplot(resid_data, aes(sample = resid_std)) +
  stat_qq(alpha = 0.1, size = 0.5, colour = "#2e6260") +
  stat_qq_line(colour = "#7b132b", linewidth = 0.7) +
  labs(title = "Normal Q-Q", x = "Theoretical quantiles", y = "Standardised residuals") +
  theme_pub

p3 <- ggplot(resid_data, aes(x = resid)) +
  geom_histogram(bins = 80, fill = "#2e6260", alpha = 0.7, colour = "white") +
  geom_vline(xintercept = 0, colour = "#7b132b", linetype = "dashed") +
  labs(title = "Residual Distribution", x = "Pearson residuals", y = "Frequency") +
  theme_pub

p4 <- ggplot(resid_data, aes(x = fitted, y = sqrt(abs(resid_std)))) +
  geom_point(alpha = 0.1, size = 0.5, colour = "#2e6260") +
  geom_smooth(method = "loess", se = FALSE, colour = "#7b132b", linewidth = 0.7) +
  labs(title = "Scale-Location", x = "Fitted values (log scale)",
       y = expression(sqrt("|Standardised residuals|"))) +
  theme_pub

(p1 + p2) / (p3 + p4) +
  plot_annotation(
    title = "Panel Model Diagnostics: All Pupils",
    subtitle = "log(ATT8SCR) ~ fixed effects + (1|year) + (1|Ofsted harmonised) + (1|region/LA)"
  )
```


# Per-Year Models (Analysis B)

Separate models are fitted for each academic year with the same fixed-effect specification but without a year term. This allows all coefficients to vary freely across years, revealing whether relationships between predictors and attainment are stable or shifting over time.


## Model Fit Comparison Across Years

```{r tbl-yearly-fit}
#| label: tbl-yearly-fit
#| tbl-cap: "Model fit statistics by year and outcome"

yearly_fit <- map_dfr(names(yearly_diag), function(yr) {
  map_dfr(names(yearly_diag[[yr]]), function(mn) {
    d <- yearly_diag[[yr]][[mn]]
    tibble(
      Year = yr,
      Outcome = outcome_labels[mn],
      N = ifelse(is.na(d$r2_marginal), NA_integer_, d$n_obs),
      r2m = round(d$r2_marginal, 4),
      r2c = round(d$r2_conditional, 4),
      sigma = round(d$sigma, 4)
    )
  })
})

names(yearly_fit)[4:6] <- c(
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>",
  "\u03c3"
)

yearly_fit %>%
  kable(format = "html", align = "llrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Coefficient Stability Across Years

```{r fig-yearly-coefs}
#| label: fig-yearly-coefs
#| fig-cap: "How fixed-effect estimates change across academic years (All Pupils models)"
#| fig-height: 9

yearly_coef_data <- map_dfr(names(yearly_models), function(yr) {
  m <- yearly_models[[yr]]$all
  if (is.null(m)) return(NULL)

  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      year = yr,
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

if (nrow(yearly_coef_data) > 0) {
  ggplot(yearly_coef_data,
         aes(x = year, y = estimate, group = 1)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), alpha = 0.15, fill = "#2e6260") +
    geom_line(colour = "#2e6260", linewidth = 0.7) +
    geom_point(colour = "#2e6260", size = 2) +
    facet_wrap(~ term_label, scales = "free_y", ncol = 3) +
    labs(x = "Academic Year", y = "Estimate (log scale)",
         subtitle = "Shaded band = 95% CI") +
    theme_pub +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
}
```


## Side-by-Side Coefficient Table (Per-Year, All Pupils)

```{r tbl-yearly-coefs}
#| label: tbl-yearly-coefs
#| tbl-cap: "Fixed effect estimates by year (All Pupils outcome). Significance: *** p<0.001, ** p<0.01, * p<0.05"

yearly_coef_wide <- yearly_coef_data %>%
  mutate(
    # Look up p-values from the original models
    sig = map2_chr(year, term_raw, function(yr, term) {
      m <- yearly_models[[yr]]$all
      if (is.null(m)) return("")
      fe <- as.data.frame(summary(m)$coefficients)
      p <- fe[term, "Pr(>|t|)"]
      if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else ""
    }),
    cell = paste0(round(estimate, 4), sig)
  ) %>%
  select(term_label, year, cell) %>%
  pivot_wider(names_from = year, values_from = cell)

yearly_coef_wide %>%
  rename(Variable = term_label) %>%
  kable(align = c("l", rep("r", ncol(yearly_coef_wide) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


# Core Models (All 4 Years)

School Workforce Census variables (`remained_in_the_same_school`, `teachers_on_leadership_pay_range_percent`, `average_number_of_days_taken`) and KS2 prior attainment (`PTPRIORLO`) are not yet available for 2024-25. To enable cross-year comparison including the latest year, a **core specification** is fitted using only the 5 predictors available across all four years:

$$
\log(y_{ij}) = \beta_0 + \beta_1\log(\text{FSM\%}) + \beta_2\log(\text{Absence\%}) + \beta_3\log(\text{EAL\%}) + \beta_4\text{Admissions} + \beta_5\text{Gorard} + u_{\text{year}} + u_{\text{Ofsted}} + u_{\text{region}} + u_{\text{LA}} + \varepsilon_{ij}
$$


## Core Panel: Fixed Effects

```{r tbl-core-panel-fixed}
#| label: tbl-core-panel-fixed
#| tbl-cap: "Fixed effect estimates from core panel models (all 4 years, reduced specification)"

core_panel_fe <- map_dfr(names(core_models), function(mn) {
  m <- core_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    transmute(
      Model = outcome_labels[mn],
      Term = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      Estimate = round(Estimate, 4),
      `Std. Error` = round(`Std. Error`, 4),
      `t value` = round(`t value`, 2),
      `p value` = ifelse(`Pr(>|t|)` < 0.001, "< 0.001",
                         ifelse(`Pr(>|t|)` < 0.01, "< 0.01",
                                ifelse(`Pr(>|t|)` < 0.05, "< 0.05",
                                       round(`Pr(>|t|)`, 3))))
    )
})

core_panel_fe %>%
  kable(align = "llrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Core Panel: Random Effects & Model Fit

```{r tbl-core-panel-random}
#| label: tbl-core-panel-random
#| tbl-cap: "Random effect variance components and model fit statistics (core panel models)"

core_panel_fit <- map_dfr(names(core_models), function(mn) {
  m <- core_models[[mn]]
  d <- core_diag[[mn]]
  vc <- as.data.frame(VarCorr(m))

  var_year   <- vc$vcov[vc$grp == "year_label"]
  var_ofsted <- vc$vcov[vc$grp == "OFSTEDRATING_1"]
  var_la     <- vc$vcov[vc$grp == "LANAME:gor_name"]
  var_region <- vc$vcov[vc$grp == "gor_name"]
  var_resid  <- d$sigma^2

  tibble(
    Model = outcome_labels[mn],
    var_year   = round(var_year, 6),
    var_ofsted = round(var_ofsted, 6),
    var_la     = round(var_la, 6),
    var_region = round(var_region, 6),
    var_resid  = round(var_resid, 6),
    r2m = round(d$r2_marginal, 4),
    r2c = round(d$r2_conditional, 4),
    N = d$n_obs
  )
})

names(core_panel_fit)[2:8] <- c(
  paste0("\u03c3\u00b2", "<sub>year</sub>"),
  paste0("\u03c3\u00b2", "<sub>Ofsted</sub>"),
  paste0("\u03c3\u00b2", "<sub>LA</sub>"),
  paste0("\u03c3\u00b2", "<sub>region</sub>"),
  paste0("\u03c3\u00b2", "<sub>resid</sub>"),
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>"
)

core_panel_fit %>%
  kable(format = "html", align = "lrrrrrrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


## Core Per-Year: Model Fit Comparison

```{r tbl-core-yearly-fit}
#| label: tbl-core-yearly-fit
#| tbl-cap: "Core model fit statistics by year and outcome (all 4 years)"

core_yearly_fit <- map_dfr(names(core_yearly_diag), function(yr) {
  map_dfr(names(core_yearly_diag[[yr]]), function(mn) {
    d <- core_yearly_diag[[yr]][[mn]]
    tibble(
      Year = yr,
      Outcome = outcome_labels[mn],
      N = ifelse(is.na(d$r2_marginal), NA_integer_, d$n_obs),
      r2m = round(d$r2_marginal, 4),
      r2c = round(d$r2_conditional, 4),
      sigma = round(d$sigma, 4)
    )
  })
})

names(core_yearly_fit)[4:6] <- c(
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>",
  "\u03c3"
)

core_yearly_fit %>%
  kable(format = "html", align = "llrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Core Per-Year: Coefficient Stability

```{r fig-core-yearly-coefs}
#| label: fig-core-yearly-coefs
#| fig-cap: "How core fixed-effect estimates change across all 4 academic years (All Pupils)"
#| fig-height: 7

core_yearly_coef_data <- map_dfr(names(core_yearly_models), function(yr) {
  m <- core_yearly_models[[yr]]$all
  if (is.null(m)) return(NULL)

  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      year = yr,
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

if (nrow(core_yearly_coef_data) > 0) {
  ggplot(core_yearly_coef_data,
         aes(x = year, y = estimate, group = 1)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_ribbon(aes(ymin = ci_lo, ymax = ci_hi), alpha = 0.15, fill = "#e16fca") +
    geom_line(colour = "#e16fca", linewidth = 0.7) +
    geom_point(colour = "#e16fca", size = 2) +
    facet_wrap(~ term_label, scales = "free_y", ncol = 3) +
    labs(x = "Academic Year", y = "Estimate (log scale)",
         subtitle = "Shaded band = 95% CI — core specification (5 predictors)") +
    theme_pub +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
}
```


## Core Per-Year: Side-by-Side Coefficients

```{r tbl-core-yearly-coefs}
#| label: tbl-core-yearly-coefs
#| tbl-cap: "Core fixed effect estimates by year (All Pupils). Significance: *** p<0.001, ** p<0.01, * p<0.05"

core_yearly_coef_wide <- core_yearly_coef_data %>%
  mutate(
    sig = map2_chr(year, term_raw, function(yr, term) {
      m <- core_yearly_models[[yr]]$all
      if (is.null(m)) return("")
      fe <- as.data.frame(summary(m)$coefficients)
      p <- fe[term, "Pr(>|t|)"]
      if (is.na(p)) "" else if (p < 0.001) "***" else if (p < 0.01) "**" else if (p < 0.05) "*" else ""
    }),
    cell = paste0(round(estimate, 4), sig)
  ) %>%
  select(term_label, year, cell) %>%
  pivot_wider(names_from = year, values_from = cell)

core_yearly_coef_wide %>%
  rename(Variable = term_label) %>%
  kable(align = c("l", rep("r", ncol(core_yearly_coef_wide) - 1))) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


# Imputed Full Panel Models (Analysis E)

The imputed full model uses the same 9-predictor specification as Analysis A, but extends coverage to all four academic years by carry-forward imputing the four variables unavailable for 2024-25 (`PTPRIORLO`, `remained_in_the_same_school`, `teachers_on_leadership_pay_range_percent`, `average_number_of_days_taken`). Each school's 2023-24 value is used (or the 3-year mean as a fallback). This is the model used in the Shiny app.

$$
\log(y_{ij}) = \mathbf{X}_{ij}\boldsymbol{\beta} + u_{\text{year}} + u_{\text{Ofsted}} + u_{\text{region}} + u_{\text{LA}} + \varepsilon_{ij}
$$


## Imputed Panel: Fixed Effects

```{r tbl-imputed-panel-fixed}
#| label: tbl-imputed-panel-fixed
#| tbl-cap: "Fixed effect estimates from imputed full panel models (Analysis E: 9 predictors, all 4 years)"

imputed_fe <- map_dfr(names(imputed_models), function(mn) {
  m <- imputed_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    transmute(
      Model = outcome_labels[mn],
      Term = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      Estimate = round(Estimate, 4),
      `Std. Error` = round(`Std. Error`, 4),
      `t value` = round(`t value`, 2),
      `p value` = ifelse(`Pr(>|t|)` < 0.001, "< 0.001",
                         ifelse(`Pr(>|t|)` < 0.01, "< 0.01",
                                ifelse(`Pr(>|t|)` < 0.05, "< 0.05",
                                       round(`Pr(>|t|)`, 3))))
    )
})

imputed_fe %>%
  kable(align = "llrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


## Imputed Panel: Coefficient Comparison Plot

```{r fig-imputed-coefs}
#| label: fig-imputed-coefs
#| fig-cap: "Fixed effect estimates with 95% confidence intervals (imputed full panel models, Analysis E)"
#| fig-height: 7

imputed_coef_data <- map_dfr(names(imputed_models), function(mn) {
  m <- imputed_models[[mn]]
  fe <- as.data.frame(summary(m)$coefficients)
  fe$term <- rownames(fe)

  fe %>%
    filter(term != "(Intercept)") %>%
    transmute(
      model = outcome_labels[mn],
      term_raw = term,
      term_label = sapply(term, function(t) {
        raw <- gsub("^log\\((.+)\\)$", "\\1", t)
        lbl <- var_label(raw)
        if (grepl("^log\\(", t)) lbl <- paste0("log(", lbl, ")")
        lbl
      }),
      estimate = Estimate,
      se = `Std. Error`,
      ci_lo = Estimate - 1.96 * `Std. Error`,
      ci_hi = Estimate + 1.96 * `Std. Error`
    )
})

ggplot(imputed_coef_data,
       aes(x = estimate, y = reorder(term_label, estimate),
           colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_pointrange(aes(xmin = ci_lo, xmax = ci_hi),
                  position = position_dodge(width = 0.6), size = 0.4) +
  scale_colour_manual(values = c("All Pupils" = "#2e6260",
                                 "Disadvantaged" = "#4e3c56",
                                 "Non-Disadvantaged" = "#abc766"),
                      name = NULL) +
  labs(x = "Estimate (log scale)", y = NULL,
       subtitle = "Points = estimate; bars = 95% CI. Analysis E (imputed full, all 4 years)") +
  theme_pub
```


## Imputed Panel: Random Effects & Model Fit

```{r tbl-imputed-panel-random}
#| label: tbl-imputed-panel-random
#| tbl-cap: "Random effect variance components and model fit statistics (imputed full panel models, Analysis E)"

imputed_fit <- map_dfr(names(imputed_models), function(mn) {
  m <- imputed_models[[mn]]
  d <- imputed_diag[[mn]]
  vc <- as.data.frame(VarCorr(m))

  var_year   <- vc$vcov[vc$grp == "year_label"]
  var_ofsted <- vc$vcov[vc$grp == "OFSTEDRATING_1"]
  var_la     <- vc$vcov[vc$grp == "LANAME:gor_name"]
  var_region <- vc$vcov[vc$grp == "gor_name"]
  var_resid  <- d$sigma^2

  tibble(
    Model = outcome_labels[mn],
    var_year   = round(var_year, 6),
    var_ofsted = round(var_ofsted, 6),
    var_la     = round(var_la, 6),
    var_region = round(var_region, 6),
    var_resid  = round(var_resid, 6),
    r2m = round(d$r2_marginal, 4),
    r2c = round(d$r2_conditional, 4),
    N = d$n_obs
  )
})

names(imputed_fit)[2:8] <- c(
  paste0("\u03c3\u00b2", "<sub>year</sub>"),
  paste0("\u03c3\u00b2", "<sub>Ofsted</sub>"),
  paste0("\u03c3\u00b2", "<sub>LA</sub>"),
  paste0("\u03c3\u00b2", "<sub>region</sub>"),
  paste0("\u03c3\u00b2", "<sub>resid</sub>"),
  "R\u00b2<sub>m</sub>",
  "R\u00b2<sub>c</sub>"
)

imputed_fit %>%
  kable(format = "html", align = "lrrrrrrrrr", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11)
```


## Analysis A vs Analysis E: Coefficient Comparison

Adding one year of carry-forward imputed data (Analysis E) produces coefficients that are very close to the original 3-year full model (Analysis A). This is reassuring: three-quarters of the data is identical, and the imputed fourth year uses conservative carry-forward values.

```{r tbl-a-vs-e}
#| label: tbl-a-vs-e
#| tbl-cap: "Coefficient comparison: Analysis A (3-year full) vs Analysis E (4-year imputed full)"

# Helper: relabel raw lmer coefficient names to human-readable form
relabel_terms <- function(term) {
  case_when(
    term == "(Intercept)" ~ "(Intercept)",
    term == "log(PTFSM6CLA1A)" ~ "log(% FSM)",
    term == "log(PERCTOT)" ~ "log(% Absence)",
    term == "log(PNUMEAL)" ~ "log(% EAL)",
    term == "PTPRIORLO" ~ "% Low Prior Attainment",
    term == "gorard_segregation" ~ "Gorard Segregation",
    term == "ADMPOL_PTOTHER NON SEL" ~ "Admissions: Other non-selective (ref: non-sel. in highly sel. area)",
    term == "ADMPOL_PTSEL" ~ "Admissions: Selective (ref: non-sel. in highly sel. area)",
    term == "remained_in_the_same_school" ~ "Teacher Retention",
    term == "teachers_on_leadership_pay_range_percent" ~ "Leadership Pay %",
    term == "log(average_number_of_days_taken)" ~ "log(Teacher Sickness)",
    TRUE ~ term
  )
}

a_vs_e <- bind_rows(
  tibble(Analysis = "A (Full, 3 yrs)",
         Term = names(fixef(panel_models$all)),
         `All Pupils` = fixef(panel_models$all),
         Disadvantaged = fixef(panel_models$disadvantaged),
         `Non-Disadvantaged` = fixef(panel_models$non_disadvantaged)),
  tibble(Analysis = "E (Imputed, 4 yrs)",
         Term = names(fixef(imputed_models$all)),
         `All Pupils` = fixef(imputed_models$all),
         Disadvantaged = fixef(imputed_models$disadvantaged),
         `Non-Disadvantaged` = fixef(imputed_models$non_disadvantaged))
) %>%
  mutate(Term = relabel_terms(Term),
         across(where(is.numeric), \(x) round(x, 5)))

a_vs_e %>%
  kable(align = "llrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 11) %>%
  collapse_rows(columns = 1, valign = "top")
```


# Stepwise Model Building

The table below shows how the model develops as predictors are added, using the All Pupils outcome. This stepwise approach reveals **confounding** and **mediating** relationships between variables — when a coefficient changes substantially as new predictors enter, it indicates that the earlier estimate was partly capturing the effect of the omitted variable.

Five nested models are fitted to the same data (Analysis E, all 4 years):

1. **Model 1** — FSM only: a single predictor measuring school-level disadvantage
2. **Model 2** — adds overall absence rate alongside FSM
3. **Model 3** — adds % low prior attainment, isolating the mediating role of KS2 achievement
4. **Model 4** — all nine fixed-effect predictors (no random effects)
5. **Model 5** — the full multilevel model with random intercepts for year, Ofsted rating, and local authority nested within region

```{r tbl-stepwise}
#| label: tbl-stepwise
#| tbl-cap: "Stepwise model building: All Pupils (Analysis E data)"

# ---- Prepare data ----
step_data <- imputed_model_data %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0)

# ---- Fit 5 nested models ----
m1 <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A),
         data = step_data)

m2 <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT),
         data = step_data)

m3 <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + PTPRIORLO,
         data = step_data)

m4 <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
           PTPRIORLO + ADMPOL_PT + gorard_segregation +
           remained_in_the_same_school +
           teachers_on_leadership_pay_range_percent +
           log(average_number_of_days_taken),
         data = step_data)

m5 <- lmer(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
             PTPRIORLO + ADMPOL_PT + gorard_segregation +
             remained_in_the_same_school +
             teachers_on_leadership_pay_range_percent +
             log(average_number_of_days_taken) +
             (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME),
           data = step_data, REML = TRUE, na.action = na.exclude,
           control = lmerControl(optimizer = "bobyqa",
                                 optCtrl = list(maxfun = 20000)))

# ---- Helper: extract coefficients with stars ----
extract_coefs <- function(model, is_lmer = FALSE) {
  s <- summary(model)$coefficients
  df <- as.data.frame(s)
  df$term <- rownames(df)

  p_col <- if ("Pr(>|t|)" %in% names(df)) "Pr(>|t|)" else "Pr(>|z|)"
  t_col <- if ("t value" %in% names(df)) "t value" else "z value"

  df %>%
    transmute(
      term = term,
      estimate = Estimate,
      t_value = .data[[t_col]],
      p_value = .data[[p_col]],
      stars = case_when(
        p_value < 0.001 ~ "***",
        p_value < 0.01  ~ "**",
        p_value < 0.05  ~ "*",
        TRUE ~ ""
      ),
      est_display = paste0(format(round(estimate, 3), nsmall = 3), stars),
      t_display = format(round(t_value, 2), nsmall = 2)
    )
}

c1 <- extract_coefs(m1)
c2 <- extract_coefs(m2)
c3 <- extract_coefs(m3)
c4 <- extract_coefs(m4)
c5 <- extract_coefs(m5, is_lmer = TRUE)

# ---- Build unified term list (in model order) ----
# Use Model 5 (full) order, then add any terms not in it
term_order <- c5$term
all_terms <- c(term_order, setdiff(unique(c(c1$term, c2$term, c3$term, c4$term)), term_order))

# ---- Assemble fixed effects table ----
step_tbl <- tibble(term = all_terms) %>%
  left_join(c1 %>% select(term, est1 = est_display, t1 = t_display), by = "term") %>%
  left_join(c2 %>% select(term, est2 = est_display, t2 = t_display), by = "term") %>%
  left_join(c3 %>% select(term, est3 = est_display, t3 = t_display), by = "term") %>%
  left_join(c4 %>% select(term, est4 = est_display, t4 = t_display), by = "term") %>%
  left_join(c5 %>% select(term, est5 = est_display, t5 = t_display), by = "term") %>%
  mutate(across(everything(), \(x) replace_na(x, ""))) %>%
  mutate(term = relabel_terms(term))

n_fixed <- nrow(step_tbl)

# ---- Random effects (Model 5 only) ----
vc <- as.data.frame(VarCorr(m5))
# vc has columns: grp, var1, var2, vcov, sdcor

re_rows <- tibble(
  term = c("", paste0("  ", vc$grp)),
  est1 = "", t1 = "",
  est2 = "", t2 = "",
  est3 = "", t3 = "",
  est4 = "", t4 = "",
  est5 = c("", format(round(vc$sdcor, 3), nsmall = 3)),
  t5   = c("", rep("(SD)", nrow(vc)))
)
# Label the header row
re_rows$term[1] <- "Random effects"
re_rows$est5[1] <- ""
re_rows$t5[1] <- ""

# Clean up group names for display
re_rows <- re_rows %>%
  mutate(term = case_when(
    term == "Random effects" ~ "Random effects",
    grepl("LANAME:gor_name", term) ~ "  LA (nested in region)",
    grepl("gor_name", term) & !grepl("LANAME", term) ~ "  Region",
    grepl("OFSTEDRATING_1", term) ~ "  Ofsted rating",
    grepl("year_label", term) ~ "  Year",
    grepl("Residual", term) ~ "  Residual",
    TRUE ~ term
  ))

# ---- R² and N rows ----
r2_m1 <- round(summary(m1)$r.squared, 4)
r2_m2 <- round(summary(m2)$r.squared, 4)
r2_m3 <- round(summary(m3)$r.squared, 4)
r2_m4 <- round(summary(m4)$r.squared, 4)
r2_m5 <- performance::r2(m5)
r2m_m5 <- round(r2_m5$R2_marginal, 4)
r2c_m5 <- round(r2_m5$R2_conditional, 4)

n_obs <- nrow(step_data)

footer_rows <- tibble(
  term = c("", "R²", "R² (conditional)", "N"),
  est1 = c("", as.character(r2_m1), "", format(n_obs, big.mark = ",")),
  t1   = c("", "", "", ""),
  est2 = c("", as.character(r2_m2), "", format(n_obs, big.mark = ",")),
  t2   = c("", "", "", ""),
  est3 = c("", as.character(r2_m3), "", format(n_obs, big.mark = ",")),
  t3   = c("", "", "", ""),
  est4 = c("", as.character(r2_m4), "", format(n_obs, big.mark = ",")),
  t4   = c("", "", "", ""),
  est5 = c("", as.character(r2m_m5), as.character(r2c_m5), format(n_obs, big.mark = ",")),
  t5   = c("", "(marginal)", "(fixed + random)", "")
)

full_tbl <- bind_rows(step_tbl, re_rows, footer_rows)

# ---- Render with kableExtra ----
names(full_tbl) <- c("Variable",
                      "Est.", "t",
                      "Est. ", "t ",
                      "Est.  ", "t  ",
                      "Est.   ", "t   ",
                      "Est.    ", "t    ")

kbl(full_tbl, format = "html", align = "lrlrlrlrlrl", escape = FALSE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 10) %>%
  add_header_above(c(" " = 1,
                      "Model 1:\nFSM only" = 2,
                      "Model 2:\n+ Absence" = 2,
                      "Model 3:\n+ Prior att." = 2,
                      "Model 4:\nAll fixed" = 2,
                      "Model 5:\nFull multilevel" = 2)) %>%
  row_spec(n_fixed + 1, bold = TRUE, extra_css = "border-top: 2px solid #666;") %>%
  row_spec(n_fixed + nrow(re_rows) + 1, extra_css = "border-top: 2px solid #666;") %>%
  row_spec(n_fixed + nrow(re_rows) + 2, bold = TRUE) %>%
  row_spec(n_fixed + nrow(re_rows) + 3, bold = TRUE) %>%
  row_spec(n_fixed + nrow(re_rows) + 4, bold = TRUE) %>%
  footnote(general = "Significance: *** p < 0.001, ** p < 0.01, * p < 0.05. Models 1–4 are OLS; Model 5 is a multilevel model (lmer) with random intercepts. Random effects shown as standard deviations.",
           general_title = "")
```

### What the stepwise results reveal

```{r stepwise-commentary}
#| label: stepwise-commentary
#| include: false

# Extract key values for inline commentary
fsm_m1 <- round(c1$estimate[c1$term == "log(PTFSM6CLA1A)"], 3)
fsm_m2 <- round(c2$estimate[c2$term == "log(PTFSM6CLA1A)"], 3)
fsm_m3 <- round(c3$estimate[c3$term == "log(PTFSM6CLA1A)"], 3)
fsm_m4 <- round(c4$estimate[c4$term == "log(PTFSM6CLA1A)"], 3)
fsm_m5 <- round(c5$estimate[c5$term == "log(PTFSM6CLA1A)"], 3)
abs_m2 <- round(c2$estimate[c2$term == "log(PERCTOT)"], 3)
abs_m3 <- round(c3$estimate[c3$term == "log(PERCTOT)"], 3)
abs_m4 <- round(c4$estimate[c4$term == "log(PERCTOT)"], 3)
abs_m5 <- round(c5$estimate[c5$term == "log(PERCTOT)"], 3)
prior_m3 <- round(c3$estimate[c3$term == "PTPRIORLO"], 3)
```

**Model 1 → Model 2 (adding absence):** When FSM is the sole predictor, its coefficient is `r fsm_m1` — but when absence is added in Model 2, the FSM coefficient shifts to `r fsm_m2`. This change reveals **confounding**: schools with higher FSM rates also tend to have higher absence rates, so Model 1's FSM coefficient was partly capturing the absence–attainment relationship. The absence coefficient itself (`r abs_m2`) is strongly negative, confirming that higher absence is associated with lower attainment. R² increases from `r r2_m1` to `r r2_m2`.

**Model 2 → Model 3 (adding % low prior attainment):** Adding a single variable — the proportion of pupils with low KS2 prior attainment — shifts the FSM coefficient from `r fsm_m2` to `r fsm_m3`. This is one of the clearest examples of **mediation** in the model: disadvantaged schools have more pupils arriving with low prior attainment, and much of the apparent FSM–attainment relationship operates through this pathway. The prior attainment coefficient (`r prior_m3`) is strongly negative. R² jumps to `r r2_m3`, a substantial increase from a single additional predictor.

**Model 3 → Model 4 (adding remaining fixed effects):** The FSM coefficient shifts further to `r fsm_m4` as EAL, admissions policy, segregation, and workforce variables enter. The absence coefficient moves from `r abs_m3` to `r abs_m4`. R² rises to `r r2_m4`, showing that the full set of school-level characteristics explains more variation, though the incremental gain from these six variables is more modest than the jump from adding prior attainment alone.

**Model 4 → Model 5 (adding random effects):** The fixed-effect coefficients are broadly similar (R²~m~ = `r r2m_m5`), but the conditional R² (R²~c~ = `r r2c_m5`) shows a substantial jump. The difference between R²~m~ and R²~c~ represents variance explained by the **grouping structure** — knowing which Ofsted category, region, and local authority a school belongs to adds considerable explanatory power beyond the nine measured predictors. The random effect standard deviations in the table show how much the intercept varies across each grouping level. This suggests that unobserved LA-level and regional factors (funding levels, local labour markets, housing costs, historical policy choices) contribute meaningfully to attainment differences.


## Stepwise helper function

```{r stepwise-helper}
#| label: stepwise-helper
#| include: false

# Reusable function to fit 5 stepwise models and return a formatted kable table
# plus extracted coefficient objects for commentary
fit_stepwise <- function(data, outcome_var, caption_label) {

  d <- data %>%
    filter(!is.na(.data[[outcome_var]]), .data[[outcome_var]] > 0)

  fml_base <- paste0("log(", outcome_var, ") ~ ")

  m1 <- lm(as.formula(paste0(fml_base, "log(PTFSM6CLA1A)")), data = d)
  m2 <- lm(as.formula(paste0(fml_base, "log(PTFSM6CLA1A) + log(PERCTOT)")), data = d)
  m3 <- lm(as.formula(paste0(fml_base, "log(PTFSM6CLA1A) + log(PERCTOT) + PTPRIORLO")), data = d)
  m4 <- lm(as.formula(paste0(fml_base,
    "log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) + ",
    "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
    "remained_in_the_same_school + ",
    "teachers_on_leadership_pay_range_percent + ",
    "log(average_number_of_days_taken)")), data = d)
  m5 <- lmer(as.formula(paste0(fml_base,
    "log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) + ",
    "PTPRIORLO + ADMPOL_PT + gorard_segregation + ",
    "remained_in_the_same_school + ",
    "teachers_on_leadership_pay_range_percent + ",
    "log(average_number_of_days_taken) + ",
    "(1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)")),
    data = d, REML = TRUE, na.action = na.exclude,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 20000)))

  c1 <- extract_coefs(m1); c2 <- extract_coefs(m2); c3 <- extract_coefs(m3)
  c4 <- extract_coefs(m4); c5 <- extract_coefs(m5, is_lmer = TRUE)

  # Unified term order
  term_order <- c5$term
  all_terms <- c(term_order, setdiff(unique(c(c1$term, c2$term, c3$term, c4$term)), term_order))

  step_tbl <- tibble(term = all_terms) %>%
    left_join(c1 %>% select(term, est1 = est_display, t1 = t_display), by = "term") %>%
    left_join(c2 %>% select(term, est2 = est_display, t2 = t_display), by = "term") %>%
    left_join(c3 %>% select(term, est3 = est_display, t3 = t_display), by = "term") %>%
    left_join(c4 %>% select(term, est4 = est_display, t4 = t_display), by = "term") %>%
    left_join(c5 %>% select(term, est5 = est_display, t5 = t_display), by = "term") %>%
    mutate(across(everything(), \(x) replace_na(x, ""))) %>%
    mutate(term = relabel_terms(term))

  n_fixed <- nrow(step_tbl)

  # Random effects
  vc <- as.data.frame(VarCorr(m5))
  re_rows <- tibble(
    term = c("Random effects", paste0("  ", vc$grp)),
    est1 = "", t1 = "", est2 = "", t2 = "", est3 = "", t3 = "",
    est4 = "", t4 = "",
    est5 = c("", format(round(vc$sdcor, 3), nsmall = 3)),
    t5   = c("", rep("(SD)", nrow(vc)))
  ) %>%
    mutate(term = case_when(
      term == "Random effects" ~ "Random effects",
      grepl("LANAME:gor_name", term) ~ "  LA (nested in region)",
      grepl("gor_name", term) & !grepl("LANAME", term) ~ "  Region",
      grepl("OFSTEDRATING_1", term) ~ "  Ofsted rating",
      grepl("year_label", term) ~ "  Year",
      grepl("Residual", term) ~ "  Residual",
      TRUE ~ term
    ))

  # R² and N
  r2_1 <- round(summary(m1)$r.squared, 4)
  r2_2 <- round(summary(m2)$r.squared, 4)
  r2_3 <- round(summary(m3)$r.squared, 4)
  r2_4 <- round(summary(m4)$r.squared, 4)
  r2_5 <- performance::r2(m5)
  r2m_5 <- round(r2_5$R2_marginal, 4)
  r2c_5 <- round(r2_5$R2_conditional, 4)
  n_obs <- nrow(d)

  footer_rows <- tibble(
    term = c("", "R\u00b2", "R\u00b2 (conditional)", "N"),
    est1 = c("", as.character(r2_1), "", format(n_obs, big.mark = ",")),
    t1 = "", est2 = c("", as.character(r2_2), "", format(n_obs, big.mark = ",")),
    t2 = "", est3 = c("", as.character(r2_3), "", format(n_obs, big.mark = ",")),
    t3 = "", est4 = c("", as.character(r2_4), "", format(n_obs, big.mark = ",")),
    t4 = "", est5 = c("", as.character(r2m_5), as.character(r2c_5), format(n_obs, big.mark = ",")),
    t5 = c("", "(marginal)", "(fixed + random)", "")
  )

  full_tbl <- bind_rows(step_tbl, re_rows, footer_rows)

  names(full_tbl) <- c("Variable", "Est.", "t", "Est. ", "t ",
                        "Est.  ", "t  ", "Est.   ", "t   ", "Est.    ", "t    ")

  tbl_out <- kbl(full_tbl, format = "html", align = "lrlrlrlrlrl", escape = FALSE) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, font_size = 10) %>%
    add_header_above(c(" " = 1, "Model 1:\nFSM only" = 2, "Model 2:\n+ Absence" = 2,
                        "Model 3:\n+ Prior att." = 2, "Model 4:\nAll fixed" = 2,
                        "Model 5:\nFull multilevel" = 2)) %>%
    row_spec(n_fixed + 1, bold = TRUE, extra_css = "border-top: 2px solid #666;") %>%
    row_spec(n_fixed + nrow(re_rows) + 1, extra_css = "border-top: 2px solid #666;") %>%
    row_spec(n_fixed + nrow(re_rows) + 2, bold = TRUE) %>%
    row_spec(n_fixed + nrow(re_rows) + 3, bold = TRUE) %>%
    row_spec(n_fixed + nrow(re_rows) + 4, bold = TRUE) %>%
    footnote(general = "Significance: *** p < 0.001, ** p < 0.01, * p < 0.05. Models 1\u20134 are OLS; Model 5 is a multilevel model (lmer). Random effects shown as standard deviations.",
             general_title = "")

  # Return everything needed for commentary
  list(
    table = tbl_out,
    coefs = list(c1 = c1, c2 = c2, c3 = c3, c4 = c4, c5 = c5),
    r2 = list(r2_1 = r2_1, r2_2 = r2_2, r2_3 = r2_3, r2_4 = r2_4,
              r2m_5 = r2m_5, r2c_5 = r2c_5),
    n = n_obs
  )
}
```


## Disadvantaged Pupils

```{r tbl-stepwise-disadv}
#| label: tbl-stepwise-disadv
#| tbl-cap: "Stepwise model building: Disadvantaged Pupils (Analysis E data)"

sw_dis <- fit_stepwise(imputed_model_data, "ATT8SCR_FSM6CLA1A", "Disadvantaged Pupils")
sw_dis$table
```

### What the stepwise results reveal — disadvantaged pupils

```{r stepwise-disadv-commentary}
#| label: stepwise-disadv-commentary
#| include: false

# Extract key values
d_fsm_m1 <- round(sw_dis$coefs$c1$estimate[sw_dis$coefs$c1$term == "log(PTFSM6CLA1A)"], 3)
d_fsm_m2 <- round(sw_dis$coefs$c2$estimate[sw_dis$coefs$c2$term == "log(PTFSM6CLA1A)"], 3)
d_fsm_m3 <- round(sw_dis$coefs$c3$estimate[sw_dis$coefs$c3$term == "log(PTFSM6CLA1A)"], 3)
d_fsm_m4 <- round(sw_dis$coefs$c4$estimate[sw_dis$coefs$c4$term == "log(PTFSM6CLA1A)"], 3)
d_fsm_m5 <- round(sw_dis$coefs$c5$estimate[sw_dis$coefs$c5$term == "log(PTFSM6CLA1A)"], 3)
d_abs_m2 <- round(sw_dis$coefs$c2$estimate[sw_dis$coefs$c2$term == "log(PERCTOT)"], 3)
d_abs_m3 <- round(sw_dis$coefs$c3$estimate[sw_dis$coefs$c3$term == "log(PERCTOT)"], 3)
d_abs_m4 <- round(sw_dis$coefs$c4$estimate[sw_dis$coefs$c4$term == "log(PERCTOT)"], 3)
d_prior_m3 <- round(sw_dis$coefs$c3$estimate[sw_dis$coefs$c3$term == "PTPRIORLO"], 3)
```

**Model 1 → Model 2 (adding absence):** For disadvantaged pupils, the FSM coefficient starts at `r d_fsm_m1` and shifts to `r d_fsm_m2` when absence enters. As with all pupils, the confounding between FSM and absence is clear — schools with higher deprivation also tend to have higher absence. R² increases from `r sw_dis$r2$r2_1` to `r sw_dis$r2$r2_2`.

**Model 2 → Model 3 (adding % low prior attainment):** The FSM coefficient moves from `r d_fsm_m2` to `r d_fsm_m3` — the mediating role of prior attainment is substantial here too. The prior attainment coefficient (`r d_prior_m3`) is strongly negative. R² jumps to `r sw_dis$r2$r2_3`.

**Model 3 → Model 4 (adding remaining fixed effects):** The FSM coefficient shifts to `r d_fsm_m4`. This is a critical transition for the disadvantaged model — watch whether the FSM coefficient **flips sign** compared to earlier steps. When prior attainment and the full set of controls are in place, a higher school-level FSM rate may become positively associated with disadvantaged attainment, suggesting that schools with more concentrated disadvantage may develop stronger support structures for their FSM-eligible pupils. R² rises to `r sw_dis$r2$r2_4`.

**Model 4 → Model 5 (adding random effects):** The marginal R² is `r sw_dis$r2$r2m_5` and the conditional R² rises to `r sw_dis$r2$r2c_5`. The grouping structure again adds substantial explanatory power.


## Non-Disadvantaged Pupils

```{r tbl-stepwise-nondis}
#| label: tbl-stepwise-nondis
#| tbl-cap: "Stepwise model building: Non-Disadvantaged Pupils (Analysis E data)"

sw_non <- fit_stepwise(imputed_model_data, "ATT8SCR_NFSM6CLA1A", "Non-Disadvantaged Pupils")
sw_non$table
```

### What the stepwise results reveal — non-disadvantaged pupils

```{r stepwise-nondis-commentary}
#| label: stepwise-nondis-commentary
#| include: false

# Extract key values
n_fsm_m1 <- round(sw_non$coefs$c1$estimate[sw_non$coefs$c1$term == "log(PTFSM6CLA1A)"], 3)
n_fsm_m2 <- round(sw_non$coefs$c2$estimate[sw_non$coefs$c2$term == "log(PTFSM6CLA1A)"], 3)
n_fsm_m3 <- round(sw_non$coefs$c3$estimate[sw_non$coefs$c3$term == "log(PTFSM6CLA1A)"], 3)
n_fsm_m4 <- round(sw_non$coefs$c4$estimate[sw_non$coefs$c4$term == "log(PTFSM6CLA1A)"], 3)
n_fsm_m5 <- round(sw_non$coefs$c5$estimate[sw_non$coefs$c5$term == "log(PTFSM6CLA1A)"], 3)
n_abs_m2 <- round(sw_non$coefs$c2$estimate[sw_non$coefs$c2$term == "log(PERCTOT)"], 3)
n_abs_m3 <- round(sw_non$coefs$c3$estimate[sw_non$coefs$c3$term == "log(PERCTOT)"], 3)
n_abs_m4 <- round(sw_non$coefs$c4$estimate[sw_non$coefs$c4$term == "log(PERCTOT)"], 3)
n_prior_m3 <- round(sw_non$coefs$c3$estimate[sw_non$coefs$c3$term == "PTPRIORLO"], 3)
```

**Model 1 → Model 2 (adding absence):** For non-disadvantaged pupils, the FSM coefficient begins at `r n_fsm_m1` and shifts to `r n_fsm_m2` when absence is added. The pattern mirrors the all-pupils model — the confounding between FSM and absence applies regardless of pupil subgroup. R² increases from `r sw_non$r2$r2_1` to `r sw_non$r2$r2_2`.

**Model 2 → Model 3 (adding % low prior attainment):** The FSM coefficient moves from `r n_fsm_m2` to `r n_fsm_m3`. Prior attainment (`r n_prior_m3`) again absorbs a substantial part of the FSM-attainment relationship. R² jumps to `r sw_non$r2$r2_3`.

**Model 3 → Model 4 (adding remaining fixed effects):** With the full set of predictors, the FSM coefficient reaches `r n_fsm_m4`. Unlike the disadvantaged model, this coefficient typically **remains negative** — the contextual effect of school-level deprivation continues to drag down attainment for non-FSM pupils even after extensive controls. R² rises to `r sw_non$r2$r2_4`.

**Model 4 → Model 5 (adding random effects):** The marginal R² is `r sw_non$r2$r2m_5` and the conditional R² reaches `r sw_non$r2$r2c_5`.


## Comparing Disadvantaged and Non-Disadvantaged Models

```{r stepwise-comparison}
#| label: stepwise-comparison
#| include: false

# Collect Model 5 (full multilevel) coefficients for both groups
dis_c5 <- sw_dis$coefs$c5
non_c5 <- sw_non$coefs$c5

# Key comparisons
comp_fsm_dis <- round(dis_c5$estimate[dis_c5$term == "log(PTFSM6CLA1A)"], 3)
comp_fsm_non <- round(non_c5$estimate[non_c5$term == "log(PTFSM6CLA1A)"], 3)
comp_abs_dis <- round(dis_c5$estimate[dis_c5$term == "log(PERCTOT)"], 3)
comp_abs_non <- round(non_c5$estimate[non_c5$term == "log(PERCTOT)"], 3)
comp_prior_dis <- round(dis_c5$estimate[dis_c5$term == "PTPRIORLO"], 3)
comp_prior_non <- round(non_c5$estimate[non_c5$term == "PTPRIORLO"], 3)

# Selective admissions
comp_sel_dis <- round(dis_c5$estimate[dis_c5$term == "ADMPOL_PTSEL"], 3)
comp_sel_non <- round(non_c5$estimate[non_c5$term == "ADMPOL_PTSEL"], 3)

# Teacher retention
comp_ret_dis <- round(dis_c5$estimate[dis_c5$term == "remained_in_the_same_school"], 3)
comp_ret_non <- round(non_c5$estimate[non_c5$term == "remained_in_the_same_school"], 3)

# EAL
comp_eal_dis <- round(dis_c5$estimate[dis_c5$term == "log(PNUMEAL)"], 3)
comp_eal_non <- round(non_c5$estimate[non_c5$term == "log(PNUMEAL)"], 3)
```

The most striking difference between the two subgroup models is the behaviour of the **% FSM coefficient** as controls are added:

::: {.callout-important}
## The FSM sign flip
In the disadvantaged model, the FSM coefficient **flips from negative to positive** as prior attainment and other controls enter the model — from `r d_fsm_m1` (Model 1) to `r comp_fsm_dis` (Model 5). This means that, *after accounting for intake characteristics*, schools with higher concentrations of disadvantaged pupils achieve **better** results for their FSM-eligible pupils than would be expected. This is consistent with a "specialisation" effect: schools that serve many disadvantaged pupils may develop more effective support programmes, allocate Pupil Premium more strategically, or benefit from a critical mass of targeted interventions.

In contrast, the non-disadvantaged model shows the FSM coefficient remains **negative** throughout: from `r n_fsm_m1` (Model 1) to `r comp_fsm_non` (Model 5). For non-FSM pupils, attending a school with a higher proportion of disadvantaged peers is associated with lower attainment even after extensive controls — a contextual or compositional disadvantage that persists.
:::

Beyond the FSM flip, the two models differ in several other important ways:

- **Absence** has a stronger negative effect on disadvantaged pupils (`r comp_abs_dis`) than on non-disadvantaged pupils (`r comp_abs_non`). Disadvantaged pupils are more vulnerable to the effects of missing school, possibly because they have fewer resources to catch up outside the classroom.

- **Prior attainment** (% low KS2) has a large negative coefficient in both models (`r comp_prior_dis` for disadvantaged, `r comp_prior_non` for non-disadvantaged), but the relative magnitude may differ — reflecting that schools with many low-attaining pupils at entry face a steeper challenge for all pupil groups.

- **Selective admissions** benefits non-disadvantaged pupils more (`r comp_sel_non`) than disadvantaged pupils (`r comp_sel_dis`), suggesting that the grammar school attainment premium is not equally shared across the pupil population.

- **EAL** (% English as additional language) shows a positive coefficient in both models (`r comp_eal_dis` for disadvantaged, `r comp_eal_non` for non-disadvantaged). Schools with more EAL pupils tend to have higher attainment after controlling for deprivation, possibly reflecting the academic strengths that many bilingual pupils bring, or the urban location of these schools which correlates with other positive factors.

- **Teacher retention** is positive in both models (`r comp_ret_dis` for disadvantaged, `r comp_ret_non` for non-disadvantaged), confirming that staff stability benefits all pupil groups — though the size of the effect may differ, indicating whether disadvantaged or non-disadvantaged pupils are more sensitive to teacher turnover.

These patterns underscore that **the same school-level factors do not operate equally across pupil subgroups**. Policy interventions need to account for these differential effects — what benefits disadvantaged pupils most (concentrated disadvantage with strong support, low absence) is not the same as what benefits non-disadvantaged pupils most (selective admissions, school composition).

### Variable importance: standardised coefficients

The chart below puts all predictors on a **common scale** so that their relative importance can be compared directly across pupil groups. Raw model coefficients cannot be compared across variables because the predictors are measured in different units (percentages, indices, days) — a coefficient of −0.15 on log(% FSM) is not directly comparable to −0.003 on % low prior attainment.

The standardisation works as follows: for each predictor, the coefficient is multiplied by the **standard deviation** of that predictor across all schools in the dataset (computed on the log scale for log-transformed variables, matching how they enter the model). The result — the **standardised coefficient** — answers the question: *"if this predictor moved by one standard deviation across the school population, how much would log(ATT8) change?"* This is a measure of **practical importance**, not statistical significance. A large standardised coefficient means the variable has a big impact on attainment differences across schools; a small one means it contributes little to the variation we observe, even if it is statistically significant.

This differs from the **t-value** shown in the stepwise tables above, which measures statistical confidence (coefficient ÷ standard error). With 12,000+ observations, even tiny effects can be highly significant (large t-values), so t-values are poor guides to practical importance. The standardised coefficients shown here are the better tool for understanding which levers matter most.

The chart also includes the **selective admissions** dummy variable (grammar schools). Because this is a binary 0/1 indicator rather than a continuous measure, the standardisation is applied in the same way — the coefficient is multiplied by the SD of the 0/1 variable. Since only around 5% of schools are selective, the SD is small (~0.22), so the standardised coefficient reflects the effect weighted by how common selective schools are in the population. This puts it on the same "population-level importance" scale as the continuous predictors: a variable that affects every school modestly will rank higher than one with a dramatic effect on only a few schools. The "other non-selective" dummy is omitted as its coefficient is small and generally non-significant.

```{r fig-std-coef-comparison}
#| label: fig-std-coef-comparison
#| fig-cap: "Standardised coefficients: relative variable importance across pupil groups (Model 5). Bars show the change in log(ATT8) associated with a one-SD shift in each predictor."
#| fig-height: 6
#| fig-width: 9

library(plotly)

# ---- Helper: compute standardised coefficients ----
standardised_importance <- function(model, data, outcome_var, group_label) {
  fe <- fixef(model)
  fe <- fe[names(fe) != "(Intercept)"]

  coef_info <- tribble(
    ~term,                                     ~data_col,                                   ~transform,
    "log(PTFSM6CLA1A)",                        "PTFSM6CLA1A",                               "log",
    "log(PERCTOT)",                             "PERCTOT",                                   "log",
    "log(PNUMEAL)",                             "PNUMEAL",                                   "log",
    "PTPRIORLO",                                "PTPRIORLO",                                 "none",
    "ADMPOL_PTSEL",                             "ADMPOL_PT",                                 "dummy_SEL",
    "gorard_segregation",                       "gorard_segregation",                        "none",
    "remained_in_the_same_school",              "remained_in_the_same_school",               "none",
    "teachers_on_leadership_pay_range_percent",  "teachers_on_leadership_pay_range_percent",  "none",
    "log(average_number_of_days_taken)",         "average_number_of_days_taken",              "log"
  )

  results <- coef_info %>%
    filter(term %in% names(fe)) %>%
    rowwise() %>%
    mutate(
      coef = fe[term],
      x_vals = list(
        if (transform == "log") log(data[[data_col]][!is.na(data[[data_col]]) & data[[data_col]] > 0])
        else if (transform == "dummy_SEL") as.numeric(data[[data_col]] == "SEL")[!is.na(data[[data_col]])]
        else data[[data_col]][!is.na(data[[data_col]])]
      ),
      sd_x = sd(unlist(x_vals), na.rm = TRUE),
      std_coef = coef * sd_x,
      abs_std_coef = abs(std_coef),
      Group = group_label
    ) %>%
    ungroup() %>%
    select(term, data_col, coef, sd_x, std_coef, abs_std_coef, Group)

  results
}

display_names <- c(
  "log(PTFSM6CLA1A)" = "% Disadvantaged (FSM)",
  "log(PERCTOT)" = "Overall Absence Rate",
  "log(PNUMEAL)" = "% EAL",
  "PTPRIORLO" = "% Low Prior Attainment",
  "ADMPOL_PTSEL" = "Selective Admissions (grammar)",
  "gorard_segregation" = "Gorard Segregation Index",
  "remained_in_the_same_school" = "Teacher Retention",
  "teachers_on_leadership_pay_range_percent" = "Leadership Pay %",
  "log(average_number_of_days_taken)" = "Teacher Sickness Days"
)

# Compute for all three groups using the imputed full models (Analysis E)
imp_all <- standardised_importance(imputed_models$all, imputed_model_data,
                                    "ATT8SCR", "All Pupils")
imp_dis <- standardised_importance(imputed_models$disadvantaged, imputed_model_data,
                                    "ATT8SCR_FSM6CLA1A", "Disadvantaged")
imp_non <- standardised_importance(imputed_models$non_disadvantaged, imputed_model_data,
                                    "ATT8SCR_NFSM6CLA1A", "Non-Disadvantaged")

importance_national <- bind_rows(imp_all, imp_dis, imp_non) %>%
  mutate(display_name = display_names[term])

# ---- Chart ----
importance_chart_data <- importance_national %>%
  mutate(
    display_name = factor(display_name),
    Group = factor(Group, levels = c("All Pupils", "Disadvantaged", "Non-Disadvantaged")),
    tooltip_text = paste0(
      "<b>", display_name, "</b><br>",
      "Group: ", Group, "<br>",
      "Raw coefficient: ", round(coef, 5), "<br>",
      "SD of predictor: ", round(sd_x, 3), "<br>",
      "Standardised coef: ", round(std_coef, 4)
    )
  )

var_order <- importance_chart_data %>%
  group_by(display_name) %>%
  summarise(avg_abs = mean(abs_std_coef)) %>%
  arrange(avg_abs) %>%
  pull(display_name)

importance_chart_data <- importance_chart_data %>%
  mutate(display_name = factor(display_name, levels = var_order))

p_imp <- ggplot(importance_chart_data,
       aes(x = display_name, y = std_coef, fill = Group,
           text = tooltip_text)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_manual(values = c("All Pupils" = "#2e6260",
                                "Disadvantaged" = "#4e3c56",
                                "Non-Disadvantaged" = "#abc766")) +
  labs(x = NULL, y = "Standardised Coefficient (signed)",
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggplotly(p_imp, tooltip = "text") %>%
  plotly::layout(legend = list(orientation = "h", y = -0.15))
```

The chart makes several features of the model immediately visible:

- **% FSM** flips direction for disadvantaged pupils (positive bar) while remaining strongly negative for non-disadvantaged and all pupils — the sign flip discussed above.
- **Overall absence** and **% low prior attainment** are the two largest negative drivers of attainment across all groups, with absence hitting disadvantaged pupils harder.
- **% EAL** is the largest *positive* predictor in standardised terms, reflecting that schools with more EAL pupils perform better than expected after controlling for deprivation.
- **Selective admissions** (grammar schools) shows a strong positive standardised coefficient, but the effect is notably larger for non-disadvantaged pupils than for disadvantaged pupils — the grammar school premium is not equally shared.
- **Teacher retention** has a modest positive effect, slightly larger for disadvantaged pupils.
- **Gorard segregation**, **leadership pay %**, and **teacher sickness days** have small standardised effects — they are statistically significant in the model but contribute relatively little to the variation in attainment across schools.


# Observed vs Predicted

The scatter plots below show observed Attainment 8 scores against imputed-full-model fitted values (Analysis E) for each of the four academic years. Schools in **Brighton and Hove** are highlighted in pink. A linear regression line with R² is overlaid on each panel; the dashed grey line marks perfect prediction.

## All Pupils

```{r fig-obs-vs-pred-all}
#| label: fig-obs-vs-pred-all
#| fig-cap: "Observed vs predicted Attainment 8 — All Pupils (imputed full model, by year). Brighton and Hove schools in pink."
#| fig-height: 8
#| fig-width: 10

HIGHLIGHT_LA <- "Brighton and Hove"

# Prepare data with highlight flag
pred_data <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_imputed), !is.na(ATT8SCR)) %>%
  mutate(
    residual = ATT8SCR - predicted_ATT8SCR_imputed,
    abs_resid = abs(residual),
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA))
  ) %>%
  arrange(highlight)   # draw highlighted points on top

# Per-year R² labels
r2_labels <- pred_data %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR, predicted_ATT8SCR_imputed, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R\u00b2 = ", round(r2, 3), "  (n = ", scales::comma(n), ")"))

ggplot(pred_data, aes(x = predicted_ATT8SCR_imputed, y = ATT8SCR)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  geom_text(data = r2_labels,
            aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.05, vjust = 1.4,
            size = 3.3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                  "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.12,
                                 "Brighton and Hove" = 0.85), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 (imputed full model)", y = "Observed ATT8",
       colour = NULL) +
  coord_equal() +
  theme_pub +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11))
```

```{r}
library(plotly)
library(dplyr)
library(scales)

# 1. Prepare data (same as your original logic)
pred_data <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_imputed), !is.na(ATT8SCR)) %>%
  mutate(
    residual = ATT8SCR - predicted_ATT8SCR_imputed,
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA)),
    # Custom tooltip text
    tooltip_text = paste0(
      "LA: ", LANAME,
      "<br>Sch:", SCHNAME,
      "<br>Observed: ", round(ATT8SCR, 2),
      "<br>Predicted: ", round(predicted_ATT8SCR_imputed, 2),
      "<br>Residual: ", round(residual, 2)
    )
  ) %>%
  arrange(highlight)

# 2. Calculate R² labels
r2_labels <- pred_data %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR, predicted_ATT8SCR_imputed, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 3), " (n = ", comma(n), ")"))

# 3. Build the ggplot object
p <- ggplot(pred_data, aes(x = predicted_ATT8SCR_imputed, y = ATT8SCR)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  # We add 'text' aesthetic here for the Plotly tooltip
  geom_point(aes(colour = highlight, alpha = highlight, text = tooltip_text), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  # Note: geom_text can be finicky in ggplotly; using explicit coordinates helps
  geom_text(data = r2_labels,
            aes(x = 20, y = 85, label = label),
            inherit.aes = FALSE, size = 3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                 "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.2,
                                "Brighton and Hove" = 0.9), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 (imputed full model)", y = "Observed ATT8", colour = NULL) +
  theme_minimal() + # Use minimal for cleaner plotly conversion
  theme(legend.position = "bottom")

# 4. Convert to Plotly
ggplotly(p, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", x = 0.4, y = -0.1),
    margin = list(t = 50) # Space for facet titles
  )

```



## Disadvantaged Pupils

```{r fig-obs-vs-pred-disadv}
#| label: fig-obs-vs-pred-disadv
#| fig-cap: "Observed vs predicted Attainment 8 — Disadvantaged Pupils (imputed full model, by year). Brighton and Hove schools in pink."
#| fig-height: 8
#| fig-width: 10

pred_disadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_FSM6CLA1A_imputed), !is.na(ATT8SCR_FSM6CLA1A)) %>%
  mutate(
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA))
  ) %>%
  arrange(highlight)

r2_disadv <- pred_disadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_FSM6CLA1A, predicted_ATT8SCR_FSM6CLA1A_imputed,
             use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R\u00b2 = ", round(r2, 3), "  (n = ", scales::comma(n), ")"))

ggplot(pred_disadv, aes(x = predicted_ATT8SCR_FSM6CLA1A_imputed,
                         y = ATT8SCR_FSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  geom_text(data = r2_disadv,
            aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.05, vjust = 1.4,
            size = 3.3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                  "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.12,
                                 "Brighton and Hove" = 0.85), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Disadvantaged (imputed full model)",
       y = "Observed ATT8 — Disadvantaged",
       colour = NULL) +
  coord_equal() +
  theme_pub +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11))
```

```{r}
library(plotly)
library(dplyr)
library(scales)

# 1. Prepare data for Disadvantaged subset
pred_disadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_FSM6CLA1A_imputed), !is.na(ATT8SCR_FSM6CLA1A)) %>%
  mutate(
    residual = ATT8SCR_FSM6CLA1A - predicted_ATT8SCR_FSM6CLA1A_imputed,
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA)),
    # Enhanced tooltip for disadvantaged context
    tooltip_text = paste0(
      "<b>", SCHNAME, "</b><br>",  # Assumes SCHNAME exists in your data
      "LA: ", LANAME,
      "<br>Sch:", SCHNAME,
      "<br>Observed (Disadv): ", round(ATT8SCR_FSM6CLA1A, 2),
      "<br>Predicted (Disadv): ", round(predicted_ATT8SCR_FSM6CLA1A_imputed, 2),
      "<br>Residual: ", round(residual, 2)
    )
  ) %>%
  arrange(highlight)

# 2. Calculate R² for the Disadvantaged model
r2_disadv_stats <- pred_disadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_FSM6CLA1A, predicted_ATT8SCR_FSM6CLA1A_imputed, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 3), " (n = ", comma(n), ")"))

# 3. Build the ggplot object
p_disadv <- ggplot(pred_disadv, aes(x = predicted_ATT8SCR_FSM6CLA1A_imputed,
                                   y = ATT8SCR_FSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight, text = tooltip_text), size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  # Positioning labels at safe numeric coordinates (e.g., x=10, y=85)
  geom_text(data = r2_disadv_stats,
            aes(x = 10, y = 85, label = label),
            inherit.aes = FALSE, size = 3, colour = "grey30", hjust = 0) +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                 "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.15,
                                "Brighton and Hove" = 0.9), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Disadvantaged",
       y = "Observed ATT8 — Disadvantaged",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))

# 4. Final Plotly conversion
ggplotly(p_disadv, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", x = 0.3, y = -0.15),
    margin = list(t = 60)
  )
```


## Non-Disadvantaged Pupils

```{r fig-obs-vs-pred-nondisadv}
#| label: fig-obs-vs-pred-nondisadv
#| fig-cap: "Observed vs predicted Attainment 8 — Non-Disadvantaged Pupils (imputed full model, by year). Brighton and Hove schools in pink."
#| fig-height: 8
#| fig-width: 10

pred_nondisadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_NFSM6CLA1A_imputed), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
  mutate(
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA))
  ) %>%
  arrange(highlight)

r2_nondisadv <- pred_nondisadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_NFSM6CLA1A, predicted_ATT8SCR_NFSM6CLA1A_imputed,
             use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R\u00b2 = ", round(r2, 3), "  (n = ", scales::comma(n), ")"))

ggplot(pred_nondisadv, aes(x = predicted_ATT8SCR_NFSM6CLA1A_imputed,
                            y = ATT8SCR_NFSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight), size = 1) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  geom_text(data = r2_nondisadv,
            aes(x = -Inf, y = Inf, label = label),
            inherit.aes = FALSE, hjust = -0.05, vjust = 1.4,
            size = 3.3, colour = "grey30") +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                  "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.12,
                                 "Brighton and Hove" = 0.85), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Non-Disadvantaged (imputed full model)",
       y = "Observed ATT8 — Non-Disadvantaged",
       colour = NULL) +
  coord_equal() +
  theme_pub +
  theme(legend.position = "bottom",
        strip.text = element_text(size = 11))
```

```{r}
library(plotly)
library(dplyr)
library(scales)

# 1. Prepare data for Non-Disadvantaged subset
pred_nondisadv <- panel_data %>%
  filter(!is.na(predicted_ATT8SCR_NFSM6CLA1A_imputed), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
  mutate(
    residual = ATT8SCR_NFSM6CLA1A - predicted_ATT8SCR_NFSM6CLA1A_imputed,
    highlight = ifelse(LANAME == HIGHLIGHT_LA, HIGHLIGHT_LA, "Other"),
    highlight = factor(highlight, levels = c("Other", HIGHLIGHT_LA)),
    # Custom tooltip
    tooltip_text = paste0(
      "<b>", SCHNAME, "</b><br>",
      "LA: ", LANAME,
      "<br>Sch:", SCHNAME,
      "<br>Observed (Non-Disadv): ", round(ATT8SCR_NFSM6CLA1A, 2),
      "<br>Predicted (Non-Disadv): ", round(predicted_ATT8SCR_NFSM6CLA1A_imputed, 2),
      "<br>Residual: ", round(residual, 2)
    )
  ) %>%
  arrange(highlight)

# 2. Calculate R² for the Non-Disadvantaged model
r2_nondisadv_stats <- pred_nondisadv %>%
  group_by(year_label) %>%
  summarise(
    r2 = cor(ATT8SCR_NFSM6CLA1A, predicted_ATT8SCR_NFSM6CLA1A_imputed, use = "complete.obs")^2,
    n  = n(),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("R² = ", round(r2, 3), " (n = ", comma(n), ")"))

# 3. Build the ggplot object
p_nondisadv <- ggplot(pred_nondisadv, aes(x = predicted_ATT8SCR_NFSM6CLA1A_imputed,
                                         y = ATT8SCR_NFSM6CLA1A)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60",
              linewidth = 0.5, linetype = "dashed") +
  geom_point(aes(colour = highlight, alpha = highlight, text = tooltip_text), size = 1.2) +
  geom_smooth(method = "lm", se = FALSE, colour = "#7b132b", linewidth = 0.6) +
  # Placed at higher coordinates as Non-Disadv scores are typically higher
  geom_text(data = r2_nondisadv_stats,
            aes(x = 20, y = 88, label = label),
            inherit.aes = FALSE, size = 3, colour = "grey30", hjust = 0) +
  scale_colour_manual(values = c("Other" = "#49a0c4",
                                 "Brighton and Hove" = "#e16fca")) +
  scale_alpha_manual(values = c("Other" = 0.15,
                                "Brighton and Hove" = 0.9), guide = "none") +
  facet_wrap(~ year_label, ncol = 2) +
  labs(x = "Predicted ATT8 — Non-Disadvantaged",
       y = "Observed ATT8 — Non-Disadvantaged",
       colour = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        strip.text = element_text(face = "bold"))

# 4. Final Plotly conversion
ggplotly(p_nondisadv, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", x = 0.3, y = -0.15),
    margin = list(t = 60)
  )
```


# Residual Geography

```{r fig-residual-by-region}
#| label: fig-residual-by-region
#| fig-cap: "Distribution of model residuals by Government Office Region"
#| fig-height: 6

if ("gor_name" %in% names(pred_data) && sum(!is.na(pred_data$gor_name)) > 0) {

  region_resid <- pred_data %>%
    filter(!is.na(gor_name)) %>%
    group_by(gor_name) %>%
    summarise(
      mean_resid = mean(residual, na.rm = TRUE),
      median_resid = median(residual, na.rm = TRUE),
      n = n(),
      .groups = "drop"
    ) %>%
    arrange(mean_resid)

  ggplot(pred_data %>% filter(!is.na(gor_name)),
         aes(x = reorder(gor_name, residual, FUN = median), y = residual)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_boxplot(fill = "#2e6260", alpha = 0.3, outlier.size = 0.3, outlier.alpha = 0.2) +
    coord_flip() +
    labs(x = NULL, y = "Residual (Observed - Predicted ATT8)",
         subtitle = "Positive = outperforming model; Negative = underperforming") +
    theme_pub
}
```


# Disadvantage Gap

```{r fig-disadvantage-gap}
#| label: fig-disadvantage-gap
#| fig-cap: "Disadvantage attainment gap by academic year"
#| fig-height: 5

if (all(c("ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A") %in% names(model_data))) {
  gap_data <- model_data %>%
    filter(!is.na(ATT8SCR_FSM6CLA1A), !is.na(ATT8SCR_NFSM6CLA1A)) %>%
    mutate(gap = ATT8SCR_NFSM6CLA1A - ATT8SCR_FSM6CLA1A)

  gap_summary <- gap_data %>%
    group_by(year_label) %>%
    summarise(
      mean_gap = mean(gap, na.rm = TRUE),
      median_gap = median(gap, na.rm = TRUE),
      q25 = quantile(gap, 0.25, na.rm = TRUE),
      q75 = quantile(gap, 0.75, na.rm = TRUE),
      .groups = "drop"
    )

  ggplot(gap_data, aes(x = year_label, y = gap)) +
    geom_boxplot(fill = "#4e3c56", alpha = 0.3, outlier.size = 0.3, outlier.alpha = 0.2) +
    geom_point(data = gap_summary, aes(y = mean_gap),
               colour = "#4e3c56", size = 3, shape = 18) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(x = "Academic Year",
         y = "ATT8 Gap (Non-Disadvantaged - Disadvantaged)",
         subtitle = "Diamond = mean gap; Box = IQR") +
    theme_pub
}
```


# Appendix: Model Formulae {.appendix}

## Panel Model (Analysis A)

All four years pooled with `year_label` as a random intercept:

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Per-Year Model (Analysis B)

Separate model for each academic year (no year term):

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Core Panel Model (Analysis C)

Reduced specification using only the 5 predictors available for all 4 years (including 2024-25). Drops workforce and prior-attainment variables:

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  ADMPOL_PT + gorard_segregation +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Core Per-Year Model (Analysis D)

Same 5 core fixed effects, separate model per year (no year term):

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  ADMPOL_PT + gorard_segregation +
  (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Imputed Full Panel Model (Analysis E)

Same 9-predictor specification as Analysis A, but covering all 4 years (2024-25 uses carry-forward imputed workforce and prior-attainment values). This is the model used in the Shiny app and for Observed vs Predicted plots.

```
log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + log(PNUMEAL) +
  PTPRIORLO + ADMPOL_PT + gorard_segregation +
  remained_in_the_same_school +
  teachers_on_leadership_pay_range_percent +
  log(average_number_of_days_taken) +
  (1 | year_label) + (1 | OFSTEDRATING_1) + (1 | gor_name/LANAME)
```

## Variable Definitions

```{r tbl-variable-defs}
#| label: tbl-variable-defs
#| tbl-cap: "Variable definitions from DfE metadata"

model_vars <- c(
  "ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A",
  "PTFSM6CLA1A", "PERCTOT", "PNUMEAL", "PTPRIORLO",
  "ADMPOL_PT", "gorard_segregation",
  "remained_in_the_same_school",
  "teachers_on_leadership_pay_range_percent",
  "average_number_of_days_taken",
  "OFSTEDRATING_1", "gor_name", "LANAME", "year_label"
)

tibble(
  `Variable Code` = model_vars,
  Description = var_labels(model_vars),
  `In Formula As` = case_when(
    model_vars %in% c("PTFSM6CLA1A", "PERCTOT", "PNUMEAL",
                       "remained_in_the_same_school",
                       "average_number_of_days_taken",
                       "ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A") ~
      paste0("log(", model_vars, ")"),
    model_vars %in% c("OFSTEDRATING_1", "gor_name", "LANAME", "year_label") ~
      paste0("(1 | ", model_vars, ")"),
    model_vars == "ADMPOL_PT" ~
      "Two dummy variables (treatment-coded): ADMPOL_PTOTHER NON SEL, ADMPOL_PTSEL (ref: non-sel. in highly sel. area)",
    TRUE ~ model_vars
  ),
  Source = case_when(
    model_vars %in% c("ATT8SCR", "ATT8SCR_FSM6CLA1A", "ATT8SCR_NFSM6CLA1A",
                       "PTFSM6CLA1A", "PTPRIORLO", "ADMPOL_PT") ~ "KS4 Performance Tables",
    model_vars %in% c("PERCTOT") ~ "Absence Statistics",
    model_vars %in% c("PNUMEAL") ~ "School Census",
    model_vars %in% c("OFSTEDRATING_1") ~ "Ofsted Inspections (harmonised)",
    model_vars %in% c("gorard_segregation") ~ "Derived (Gorard index)",
    model_vars %in% c("remained_in_the_same_school",
                       "teachers_on_leadership_pay_range_percent",
                       "average_number_of_days_taken") ~ "School Workforce Census",
    model_vars %in% c("gor_name", "LANAME") ~ "Edubase",
    model_vars == "year_label" ~ "Constructed",
    TRUE ~ ""
  )
) %>%
  kable(align = "llll") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 10)
```
