---
title: "Shared Variance and Relative Importance"
subtitle: "How much does each predictor really explain, and does the answer depend on how we ask?"
author: "Supplementary note"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
    df-print: kable
    embed-resources: true
bibliography: ../SchoolAttainment.bib
csl: ../chicago-author-date.csl
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(lme4)

panel <- readRDS(here::here("data", "panel_data.rds"))

# The Analysis E estimation sample, restricted to complete cases on every
# predictor so that all subsets below are fitted on identical rows. This
# matters: lm() drops incomplete cases silently, so models containing
# different variables would otherwise be fitted on different samples and
# their R-squared values would not be comparable.
d <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0,
         PTFSM6CLA1A > 0, PERCTOT > 0, PNUMEAL > 0,
         !is.na(ks2_c), !is.na(ADMPOL_PT), !is.na(gorard_segregation),
         !is.na(remained_in_the_same_school),
         !is.na(teachers_on_leadership_pay_range_percent),
         average_number_of_days_taken > 0) %>%
  droplevels()

PREDS <- c(
  "log(PTFSM6CLA1A)"                         = "% FSM (disadvantage)",
  "log(PERCTOT)"                             = "% Absence",
  "ks2_c"                                    = "Mean KS2 prior attainment",
  "log(PNUMEAL)"                             = "% EAL",
  "ADMPOL_PT"                                = "Admissions policy",
  "gorard_segregation"                       = "Gorard segregation",
  "remained_in_the_same_school"              = "Teacher retention",
  "teachers_on_leadership_pay_range_percent" = "Leadership pay %",
  "log(average_number_of_days_taken)"        = "Teacher sickness days"
)

FOCAL <- c("log(PTFSM6CLA1A)", "log(PERCTOT)", "ks2_c")

r2_of <- function(vs) {
  rhs <- if (length(vs) == 0) "1" else paste(vs, collapse = " + ")
  summary(lm(as.formula(paste("log(ATT8SCR) ~", rhs)), data = d))$r.squared
}

lbl <- function(v) unname(PREDS[v])

theme_vd <- theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"),
        panel.grid.minor = element_blank())
```

## Why this note exists

The paper's stepwise table (Model 1 to Model 5) reports coefficients and
*t*-values for each predictor. It is tempting to read those *t*-values as a
ranking of importance, and an earlier draft of section 4.2 did exactly that.
That reading is not safe, for a reason worth setting out carefully.

In a multiple regression, each coefficient describes a predictor's **unique**
contribution: the association that remains once every other predictor is held
constant. Where predictors are uncorrelated, unique contribution and total
contribution are the same thing and nothing is lost. Where predictors are
correlated, a large part of the variance they explain is **shared** between
them, and the model has no way to attribute that shared part to one variable
rather than another. It simply goes unallocated. A variable that overlaps
heavily with the others will therefore show a modest *t*-value even when it
accounts for a great deal of the outcome on its own.

Our three main predictors overlap a good deal, so this is not a hypothetical
concern. This note quantifies it: it reports what each predictor explains on
its own, what every combination of predictors explains, and how the shared
variance is apportioned when we use a method designed for that purpose. It
then asks the question that actually matters --- whether the paper's
conclusions depend on which measure we choose.

The analysis sample is `r format(nrow(d), big.mark = ",")` school-years, the
same rows used for the paper's models, restricted to complete cases so that
every model below is fitted on identical data.

## How much do the predictors overlap?

```{r}
#| label: fig-corr
#| fig-cap: "Correlations between the model's continuous predictors. Strong off-diagonal values mean shared variance that a multiple regression cannot attribute to one variable rather than another."
#| fig-height: 6.5

num_preds <- setdiff(names(PREDS), "ADMPOL_PT")
X <- as.data.frame(lapply(num_preds, function(v) {
  raw <- sub("^log\\((.*)\\)$", "\\1", v)
  x <- d[[raw]]
  if (grepl("^log\\(", v)) log(x) else x
}))
names(X) <- lbl(num_preds)

cm <- cor(X, use = "complete.obs")
cm_long <- as.data.frame(as.table(cm)) %>%
  setNames(c("a", "b", "r"))

ggplot(cm_long, aes(a, b, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.2f", r),
                colour = abs(r) > 0.55), size = 3.1, show.legend = FALSE) +
  scale_fill_gradient2(low = "#c0392b", mid = "white", high = "#2e6260",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "grey20")) +
  labs(title = "Correlations between predictors",
       subtitle = "The three main predictors overlap substantially",
       x = NULL, y = NULL) +
  coord_equal() +
  theme_vd +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

```{r}
#| label: tbl-focal-corr
#| tbl-cap: "Correlations between the three main predictors."

fx <- data.frame(
  fsm = log(d$PTFSM6CLA1A), abs = log(d$PERCTOT), ks2 = d$ks2_c
)
tibble(
  Pair = c("% FSM and mean KS2", "% Absence and mean KS2", "% FSM and % Absence"),
  r = c(cor(fx$fsm, fx$ks2), cor(fx$abs, fx$ks2), cor(fx$fsm, fx$abs))
) %>%
  mutate(r = sprintf("%+.3f", r)) %>%
  kable(align = "lr")
```

Schools with more disadvantaged intakes have markedly lower prior attainment,
and schools with higher absence also have lower prior attainment. These are
not nuisance correlations to be controlled away; they are the structure of the
problem. But they do mean that a large share of the variance in attainment is
explained by these variables *jointly*, and any claim about which one matters
most has to say how the joint part is being handled.

## What each predictor explains on its own

The simplest question: fit each predictor alone against attainment, and record
the $R^2$. This is the *total* association, including everything the variable
shares with the others.

```{r}
#| label: tbl-univariate
#| tbl-cap: "Each predictor fitted on its own. This is total explanatory power, sharing included."

uni <- map_dfr(names(PREDS), function(v) {
  tibble(Predictor = lbl(v), `R²` = r2_of(v))
}) %>% arrange(desc(`R²`))

uni %>%
  mutate(`R²` = sprintf("%.3f", `R²`),
         `Variance explained` = percent(as.numeric(`R²`), accuracy = 0.1)) %>%
  kable(align = "lrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
#| label: fig-univariate
#| fig-cap: "Variance in school-level Attainment 8 explained by each predictor on its own."
#| fig-height: 4.5

uni %>%
  mutate(Predictor = fct_reorder(Predictor, `R²`)) %>%
  ggplot(aes(`R²`, Predictor)) +
  geom_col(fill = "#4e3c56", width = 0.72) +
  geom_text(aes(label = percent(`R²`, accuracy = 0.1)),
            hjust = -0.15, size = 3.2, colour = "grey25") +
  scale_x_continuous(labels = percent, limits = c(0, 0.68),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(title = "Explanatory power of each predictor on its own",
       subtitle = "Total association, including variance shared with other predictors",
       x = "R² alone", y = NULL) +
  theme_vd
```

On this measure prior attainment leads, absence follows, and disadvantage is
third. Note how different this is from the *t*-values in the paper's Model 3,
where absence appears strongest and disadvantage weakest by a wide margin. The
gap between the two pictures is precisely the shared variance.

## Every combination of the three main predictors

The table below fits all eight possible subsets of disadvantage, absence and
prior attainment, and reports both the model $R^2$ and how each coefficient
and *t*-value moves as its companions are added.

```{r}
#| label: tbl-subsets
#| tbl-cap: "All 8 subsets of the three main predictors, fitted on identical rows. Watch the FSM coefficient shrink as its companions enter."

subsets <- list(character(0), "log(PTFSM6CLA1A)", "log(PERCTOT)", "ks2_c",
                c("log(PTFSM6CLA1A)", "log(PERCTOT)"),
                c("log(PTFSM6CLA1A)", "ks2_c"),
                c("log(PERCTOT)", "ks2_c"),
                FOCAL)

fit_sub <- function(vs) {
  rhs <- if (length(vs) == 0) "1" else paste(vs, collapse = " + ")
  m <- lm(as.formula(paste("log(ATT8SCR) ~", rhs)), data = d)
  s <- summary(m)$coefficients
  get <- function(v, col) if (v %in% vs) s[v, col] else NA_real_
  tibble(
    Model = if (length(vs) == 0) "(intercept only)"
            else paste(c("FSM", "Absence", "KS2")[match(vs, FOCAL)], collapse = " + "),
    `R²`      = summary(m)$r.squared,
    `FSM β`   = get("log(PTFSM6CLA1A)", "Estimate"),
    `FSM t`   = get("log(PTFSM6CLA1A)", "t value"),
    `Abs β`   = get("log(PERCTOT)", "Estimate"),
    `Abs t`   = get("log(PERCTOT)", "t value"),
    `KS2 β`   = get("ks2_c", "Estimate"),
    `KS2 t`   = get("ks2_c", "t value")
  )
}
subs <- map_dfr(subsets, fit_sub)

subs %>%
  mutate(across(where(is.numeric), \(x) ifelse(is.na(x), "—", sprintf("%.3f", x)))) %>%
  kable(align = "lrrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

```{r}
#| label: fig-coef-drift
#| fig-cap: "How the disadvantage coefficient changes as the other predictors are added. Each bar is a model; the coefficient more than halves once absence and prior attainment are present."
#| fig-height: 4

subs %>%
  filter(!is.na(`FSM β`)) %>%
  mutate(Model = fct_reorder(Model, `FSM β`)) %>%
  ggplot(aes(`FSM β`, Model)) +
  geom_col(fill = "#c0392b", width = 0.6) +
  geom_text(aes(label = sprintf("%.3f", `FSM β`)), hjust = 1.15,
            colour = "white", size = 3.2) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0))) +
  labs(title = "The disadvantage coefficient depends heavily on what else is in the model",
       subtitle = "log(% FSM) coefficient across the four models that contain it",
       x = "Coefficient on log(% FSM)", y = NULL) +
  theme_vd
```

This is the shared-variance problem made visible. Disadvantage on its own
carries a coefficient of about −0.20; with absence and prior attainment
alongside it, that falls to roughly −0.05. Nothing about the schools has
changed --- only what else the model is being told.

Which of those numbers is "the effect of disadvantage" depends entirely on the
question. If you want the total association between a school's intake
composition and its results, the first is closer. If you want the association
that remains between two schools with the same absence and the same prior
attainment, the last is closer. The paper needs the second, because its
question is what happens if you change intake composition while everything
else stays as it is --- but it should not then describe that number as the
importance of disadvantage in general, and section 3.2 now says so explicitly.

## Apportioning the shared variance

Neither "$R^2$ alone" nor "the coefficient in the full model" is a complete
answer. The first hands each variable all of the shared variance; the second
hands it none. The standard resolution is to average over every possible
order in which the variables could enter: a variable's contribution is
computed when it enters first, second and third, and these are averaged. This
is the Shapley value from cooperative game theory, known in this context as
the LMG decomposition after @Lindeman1980, and implemented for R by
@Gromping2007.

```{r}
#| label: shapley-focal

# 3 variables = 6 orderings; done by hand to avoid a package dependency
perms3 <- list(c(1,2,3), c(1,3,2), c(2,1,3), c(2,3,1), c(3,1,2), c(3,2,1))
shap3 <- setNames(numeric(3), FOCAL)
for (p in perms3) {
  sofar <- character(0)
  for (i in p) {
    v <- FOCAL[i]
    shap3[v] <- shap3[v] + (r2_of(c(sofar, v)) - r2_of(sofar))
    sofar <- c(sofar, v)
  }
}
shap3 <- shap3 / length(perms3)
```

```{r}
#| label: tbl-shapley
#| tbl-cap: "Shapley (LMG) decomposition of the three-predictor model. Each variable's contribution is averaged over all six orders of entry, so the shared variance is split rather than assigned arbitrarily."

tibble(
  Predictor = lbl(FOCAL),
  `R² alone` = vapply(FOCAL, r2_of, numeric(1)),
  `Shapley R²` = as.numeric(shap3[FOCAL]),
  `Share of explained variance` = as.numeric(shap3[FOCAL]) / sum(shap3)
) %>%
  arrange(desc(`Shapley R²`)) %>%
  mutate(
    `R² alone` = sprintf("%.3f", `R² alone`),
    `Shapley R²` = sprintf("%.3f", `Shapley R²`),
    `Share of explained variance` = percent(`Share of explained variance`, accuracy = 0.1)
  ) %>%
  kable(align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
#| label: fig-three-rankings
#| fig-cap: "The three predictors ranked three ways. Prior attainment and absence swap places depending on the measure; disadvantage is last on all three."
#| fig-height: 4.5

m3 <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + ks2_c, data = d)
tvals <- abs(summary(m3)$coefficients[FOCAL, "t value"])

rank_df <- bind_rows(
  tibble(Predictor = lbl(FOCAL), Measure = "R² alone",
         Value = vapply(FOCAL, r2_of, numeric(1)) / sum(vapply(FOCAL, r2_of, numeric(1)))),
  tibble(Predictor = lbl(FOCAL), Measure = "|t| in the 3-variable model",
         Value = tvals / sum(tvals)),
  tibble(Predictor = lbl(FOCAL), Measure = "Shapley (LMG) share",
         Value = as.numeric(shap3[FOCAL]) / sum(shap3))
) %>%
  mutate(Measure = factor(Measure, levels = c("|t| in the 3-variable model",
                                              "R² alone", "Shapley (LMG) share")))

ggplot(rank_df, aes(Value, fct_rev(factor(Predictor)), fill = Predictor)) +
  geom_col(width = 0.68, show.legend = FALSE) +
  geom_text(aes(label = percent(Value, accuracy = 0.1)), hjust = -0.12,
            size = 3, colour = "grey25") +
  facet_wrap(~ Measure, nrow = 1) +
  scale_x_continuous(labels = percent, limits = c(0, 0.62),
                     expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = c("% FSM (disadvantage)" = "#c0392b",
                               "% Absence" = "#2e6260",
                               "Mean KS2 prior attainment" = "#4e3c56")) +
  labs(title = "Relative importance under three different measures",
       subtitle = "Each measure rescaled to shares so the three are comparable",
       x = "Share of the total", y = NULL) +
  theme_vd
```

The ranking of the top two **does** depend on the measure. By *t*-value,
absence leads; by explanatory power alone and by the Shapley decomposition,
prior attainment leads. This is exactly why the paper no longer rests any
claim on the *t*-value ordering.

What does not change is the position of disadvantage: **last under all three
measures**, whether it is credited with all of the shared variance, none of
it, or an even split. The paper's substantive argument depends only on that,
and it is robust to the choice.

## All nine predictors

The same decomposition across the full fixed-effect specification. With nine
predictors there are $2^9 = 512$ subsets and $9! = 362{,}880$ orderings, so
the Shapley values are computed from the subset lattice rather than by
enumerating permutations --- the two are mathematically identical, but the
first is tractable.

```{r}
#| label: shapley-full
#| cache: false

vars9 <- names(PREDS)
k <- length(vars9)

# R-squared for every subset, indexed by bitmask
masks <- 0:(2^k - 1)
r2_cache <- numeric(length(masks))
for (m in masks) {
  sel <- vars9[bitwAnd(m, 2^(seq_len(k) - 1)) > 0]
  r2_cache[m + 1] <- r2_of(sel)
}

# Shapley value from the lattice:
#   phi_i = sum over subsets S not containing i of
#           |S|!(k-|S|-1)!/k! * (R2(S+i) - R2(S))
lfact <- lfactorial
phi <- numeric(k)
for (i in seq_len(k)) {
  bit <- 2^(i - 1)
  for (m in masks) {
    if (bitwAnd(m, bit) == 0) {
      s <- sum(bitwAnd(m, 2^(seq_len(k) - 1)) > 0)
      w <- exp(lfact(s) + lfact(k - s - 1) - lfact(k))
      phi[i] <- phi[i] + w * (r2_cache[bitwOr(m, bit) + 1] - r2_cache[m + 1])
    }
  }
}
names(phi) <- vars9
```

```{r}
#| label: tbl-shapley-full
#| tbl-cap: "Shapley (LMG) decomposition across all nine fixed-effect predictors, computed over all 512 subsets."

full_tbl <- tibble(
  Predictor = lbl(vars9),
  `R² alone` = vapply(vars9, r2_of, numeric(1)),
  `Shapley R²` = phi,
  `Share` = phi / sum(phi)
) %>% arrange(desc(`Shapley R²`))

full_tbl %>%
  mutate(`R² alone` = sprintf("%.3f", `R² alone`),
         `Shapley R²` = sprintf("%.3f", `Shapley R²`),
         `Share` = percent(`Share`, accuracy = 0.1)) %>%
  kable(align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
#| label: fig-shapley-full
#| fig-cap: "Explanatory power alone against apportioned share, all nine predictors. Variables far above the diagonal owe much of their apparent strength to variance shared with others."
#| fig-height: 5.5

full_tbl %>%
  mutate(Predictor = fct_reorder(Predictor, `Shapley R²`)) %>%
  pivot_longer(c(`R² alone`, `Shapley R²`), names_to = "Measure", values_to = "Value") %>%
  ggplot(aes(Value, Predictor, fill = Measure)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  scale_fill_manual(values = c("R² alone" = "#abc766", "Shapley R²" = "#4e3c56")) +
  scale_x_continuous(labels = percent, expand = expansion(mult = c(0, 0.04))) +
  labs(title = "Alone versus apportioned",
       subtitle = "The gap between the two bars is variance the predictor shares with others",
       x = "Variance explained", y = NULL, fill = NULL) +
  theme_vd +
  theme(legend.position = "bottom")
```

The sum of the Shapley values is
`r sprintf("%.3f", sum(phi))`, which by construction equals the $R^2$ of the
full nine-predictor model --- the decomposition allocates all of the explained
variance and none of it twice.

## What this means for the paper

Three conclusions, in order of how much they matter.

**The importance ranking of the top two predictors is method-dependent, and
the paper should not assert one.** Section 4.2 previously ranked the three by
*t*-value and called absence the most important. That has been rewritten: it
now reports the *t*-values as measures of precision, states plainly that they
cannot rank importance where predictors are correlated, and gives the
explanatory-power and Shapley figures alongside. Section 4.3, which uses
standardised coefficients from the full model, already placed prior attainment
first; the two sections no longer contradict one another.

**The paper's argument does not depend on the choice.** Concentrations of
disadvantage rank last on every measure tried here. The claim the paper
actually makes --- that disadvantage concentration is a weak lever relative to
absence, and that an admissions policy built on it is unlikely to deliver the
gains sought --- is unaffected by how the shared variance is apportioned.

**Prior attainment is large but is not a lever.** The measures above rank
*explanatory power*, not *actionability*. Prior attainment is the single
largest source of variation on two of the three measures, but it is fixed for
any cohort already in secondary school: it describes the intake a school
receives, not something the school or its authority can change. Once it is set
aside as intake rather than lever, absence is the largest factor remaining
that is open to intervention --- which is the distinction section 4.3 draws
and the one on which the policy argument rests.

## A caveat on all of this

Every measure here is descriptive. None of them licenses a causal reading, and
the relative-importance literature is explicit that these decompositions
apportion *variance explained*, not *effect* [@Gromping2007]. Two predictors
can share variance because one causes the other, because a third thing causes
both, or because they are alternative measurements of the same underlying
disadvantage --- and no decomposition of an observational dataset can tell
these apart. That prior attainment, absence and disadvantage overlap so
heavily is itself the substantive finding: they are facets of a single
structure, and the policy question is not which is "most important" but which
of them can actually be moved.

## Disadvantaged and non-disadvantaged pupils, taken separately

Everything above treats the school's headline Attainment 8 --- the average
across all its pupils --- as the outcome. That average is a mixture. A school
with a high proportion of disadvantaged pupils scores lower partly because
disadvantaged pupils score lower on average nationally, quite apart from
anything happening inside the school. The published data lets us take the
mixture apart: DfE reports Attainment 8 separately for the disadvantaged
cohort (`ATT8SCR_FSM6CLA1A`) and the non-disadvantaged cohort
(`ATT8SCR_NFSM6CLA1A`) in each school. Repeating the whole decomposition on
each of these asks a sharper question --- not "what explains a school's
average result" but "what explains how *these pupils* do here" --- and the
answers differ in ways that matter for the policy argument.

```{r}
#| label: setup-subgroups
#| include: false

# Restrict to rows where BOTH subgroup outcomes are reported as well as the
# headline figure, so that all three outcomes below are modelled on identical
# rows and their R-squared values are directly comparable. This drops a small
# number of school-years where a cohort was too small to publish.
d3 <- d %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0,
         !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

# The rows the subgroup requirement removes, kept so the section can describe
# what is being excluded rather than waving it away as a rounding difference.
dd <- anti_join(d, d3, by = c("URN", "academic_year"))

OUTCOMES3 <- c("All pupils"        = "ATT8SCR",
               "Disadvantaged"     = "ATT8SCR_FSM6CLA1A",
               "Non-disadvantaged" = "ATT8SCR_NFSM6CLA1A")

OUT_COL <- c("All pupils"        = "#4e3c56",
             "Disadvantaged"     = "#c0392b",
             "Non-disadvantaged" = "#2e6260")

out_lvl <- names(OUTCOMES3)

# Generalised version of r2_of(): outcome as well as predictors vary.
r2_out <- function(y, vs) {
  rhs <- if (length(vs) == 0) "1" else paste(vs, collapse = " + ")
  summary(lm(as.formula(paste0("log(", y, ") ~ ", rhs)), data = d3))$r.squared
}

vars9 <- names(PREDS)
```

The comparison sample is `r format(nrow(d3), big.mark = ",")` school-years,
`r format(nrow(d) - nrow(d3), big.mark = ",")` fewer than the sample used
above, and the all-pupil figures are recomputed on it throughout this section
so that the three columns are like-for-like.

That exclusion is not random, and it matters enough to name. A school-year
drops out when DfE suppresses one of the cohort figures because the cohort is
too small to publish, which in practice means schools with very few
disadvantaged pupils. The rows dropped here have a median disadvantage share
of `r sprintf("%.0f%%", median(dd$PTFSM6CLA1A))` against
`r sprintf("%.0f%%", median(d3$PTFSM6CLA1A))` in the retained sample, and a
median Attainment 8 of `r sprintf("%.0f", median(dd$ATT8SCR))` against
`r sprintf("%.0f", median(d3$ATT8SCR))`; `r sum(dd$ADMPOL_PT == "SEL")` of the
`r nrow(dd)` are selective schools. Removing them trims the extreme
high-attainment, low-disadvantage corner of the distribution, and the effect
on the all-pupil figures is not negligible: prior attainment alone rises from
an $R^2$ of `r sprintf("%.3f", r2_of("ks2_c"))` in section 3 to
`r sprintf("%.3f", r2_out("ATT8SCR", "ks2_c"))` here, and absence from
`r sprintf("%.3f", r2_of("log(PERCTOT)"))` to
`r sprintf("%.3f", r2_out("ATT8SCR", "log(PERCTOT)"))`, while admissions
policy falls from `r sprintf("%.3f", r2_of("ADMPOL_PT"))` to
`r sprintf("%.3f", r2_out("ATT8SCR", "ADMPOL_PT"))` as the selective category
loses its most distinctive members. The all-pupil column below is the right
comparator for the two cohort columns, but it is *not* interchangeable with
the headline figures in sections 3 to 6, and the two sets should not be read
across. None of the rankings change.

```{r}
#| label: tbl-outcome-summary
#| tbl-cap: "The three outcomes on the common sample. Disadvantaged attainment is both lower and considerably more variable between schools."

tibble(
  Outcome = out_lvl,
  Mean = vapply(OUTCOMES3, \(y) mean(d3[[y]]), numeric(1)),
  `SD between schools` = vapply(OUTCOMES3, \(y) sd(d3[[y]]), numeric(1)),
  `SD of log` = vapply(OUTCOMES3, \(y) sd(log(d3[[y]])), numeric(1))
) %>%
  mutate(across(where(is.numeric), \(x) sprintf("%.2f", x))) %>%
  kable(align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

The average gap between the two cohorts within a school is
`r sprintf("%.1f", mean(d3$ATT8SCR_NFSM6CLA1A - d3$ATT8SCR_FSM6CLA1A))`
Attainment 8 points. The two subgroup scores correlate at
`r sprintf("%.2f", cor(log(d3$ATT8SCR_FSM6CLA1A), log(d3$ATT8SCR_NFSM6CLA1A)))`
across schools: schools that do well by one cohort tend to do well by the
other, but far from perfectly.

### What each predictor explains, by cohort

```{r}
#| label: tbl-uni-subgroup
#| tbl-cap: "Each predictor fitted on its own against each of the three outcomes. Total association, sharing included."

uni3 <- map_dfr(out_lvl, function(nm) {
  y <- OUTCOMES3[[nm]]
  tibble(Outcome = nm, Predictor = lbl(vars9),
         R2 = vapply(vars9, \(v) r2_out(y, v), numeric(1)))
})

uni3 %>%
  pivot_wider(names_from = Outcome, values_from = R2) %>%
  arrange(desc(`All pupils`)) %>%
  mutate(across(where(is.numeric), \(x) sprintf("%.3f", x))) %>%
  kable(align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
#| label: fig-uni-subgroup
#| fig-cap: "Explanatory power of each predictor on its own, for each cohort. Prior attainment loses most ground when the outcome is restricted to disadvantaged pupils; absence loses least."
#| fig-height: 5.5

ord <- uni3 %>% filter(Outcome == "All pupils") %>% arrange(R2) %>% pull(Predictor)

uni3 %>%
  mutate(Predictor = factor(Predictor, levels = ord),
         Outcome = factor(Outcome, levels = out_lvl)) %>%
  ggplot(aes(R2, Predictor, fill = Outcome)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.72) +
  scale_fill_manual(values = OUT_COL) +
  scale_x_continuous(labels = percent, expand = expansion(mult = c(0, 0.04))) +
  labs(title = "Explanatory power of each predictor on its own, by cohort",
       subtitle = "Same predictors, same schools, three different outcomes",
       x = "R² alone", y = NULL, fill = NULL) +
  theme_vd +
  theme(legend.position = "bottom")
```

Two things stand out immediately.

**Prior attainment is a much weaker predictor of disadvantaged pupils'
results.** On its own it explains
`r sprintf("%.0f%%", 100 * r2_out("ATT8SCR_NFSM6CLA1A", "ks2_c"))` of the
variation in non-disadvantaged attainment but only
`r sprintf("%.0f%%", 100 * r2_out("ATT8SCR_FSM6CLA1A", "ks2_c"))` of the
variation in disadvantaged attainment. Part of that is a measurement artefact
--- `ks2_c` is the school's mean prior attainment across *all* pupils, so it
describes the non-disadvantaged cohort more faithfully than the disadvantaged
one --- but the size of the gap is larger than that alone would suggest.

**Absence is the one predictor that holds up.** It explains
`r sprintf("%.0f%%", 100 * r2_out("ATT8SCR_FSM6CLA1A", "log(PERCTOT)"))` of
disadvantaged attainment against
`r sprintf("%.0f%%", 100 * r2_out("ATT8SCR_NFSM6CLA1A", "log(PERCTOT)"))` of
non-disadvantaged attainment --- almost the same. For the disadvantaged
cohort it is the strongest single predictor in the set, ahead of prior
attainment.

Note also that the proportion of disadvantaged pupils explains far less of
disadvantaged pupils' own attainment
(`r sprintf("%.0f%%", 100 * r2_out("ATT8SCR_FSM6CLA1A", "log(PTFSM6CLA1A)"))`)
than it does of the headline average
(`r sprintf("%.0f%%", 100 * r2_out("ATT8SCR", "log(PTFSM6CLA1A)"))`). That
discrepancy is the subject of the next section.

### The disadvantage coefficient changes sign

Section 4 showed the coefficient on log(% FSM) shrinking from about −0.20 to
−0.05 as absence and prior attainment were added. Repeating that exercise for
each cohort produces something more dramatic.

```{r}
#| label: tbl-fsm-drift-subgroup
#| tbl-cap: "The coefficient on log(% FSM) across the four models that contain it, for each outcome. For disadvantaged pupils it does not merely shrink — it reverses."

drift_specs <- list(
  "FSM"                 = "log(PTFSM6CLA1A)",
  "FSM + Absence"       = c("log(PTFSM6CLA1A)", "log(PERCTOT)"),
  "FSM + KS2"           = c("log(PTFSM6CLA1A)", "ks2_c"),
  "FSM + Absence + KS2" = FOCAL
)

drift <- map_dfr(out_lvl, function(nm) {
  y <- OUTCOMES3[[nm]]
  map_dfr(names(drift_specs), function(sp) {
    m <- lm(as.formula(paste0("log(", y, ") ~ ",
                              paste(drift_specs[[sp]], collapse = " + "))),
            data = d3)
    s <- summary(m)$coefficients
    tibble(Outcome = nm, Model = sp,
           beta = s["log(PTFSM6CLA1A)", "Estimate"],
           t    = s["log(PTFSM6CLA1A)", "t value"],
           R2   = summary(m)$r.squared)
  })
})

drift %>%
  transmute(Outcome, Model,
            `FSM β` = sprintf("%+.4f", beta),
            `FSM t` = sprintf("%+.1f", t),
            `R²`    = sprintf("%.3f", R2)) %>%
  select(-Outcome) %>%
  kable(align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  pack_rows(index = table(factor(drift$Outcome, levels = out_lvl)))
```

```{r}
#| label: fig-fsm-signflip
#| fig-cap: "The disadvantage coefficient by cohort and specification. Conditional on absence and prior attainment, disadvantaged pupils do better in schools with more disadvantaged pupils, not worse."
#| fig-height: 4.6

drift %>%
  mutate(Model = factor(Model, levels = rev(names(drift_specs))),
         Outcome = factor(Outcome, levels = out_lvl)) %>%
  ggplot(aes(beta, Model, fill = Outcome)) +
  geom_vline(xintercept = 0, colour = "grey35", linewidth = 0.5) +
  geom_col(position = position_dodge(width = 0.78), width = 0.72) +
  scale_fill_manual(values = OUT_COL) +
  scale_x_continuous(expand = expansion(mult = c(0.06, 0.06))) +
  labs(title = "The disadvantage coefficient reverses for the disadvantaged cohort",
       subtitle = "Coefficient on log(% FSM), by outcome and by what else is in the model",
       x = "Coefficient on log(% FSM)", y = NULL, fill = NULL) +
  theme_vd +
  theme(legend.position = "bottom")
```

Taken on its own, a higher share of disadvantaged pupils is associated with
*lower* attainment for the disadvantaged cohort
(β = `r sprintf("%+.3f", drift$beta[drift$Outcome == "Disadvantaged" & drift$Model == "FSM"])`).
Hold absence and prior attainment constant and the sign flips to
`r sprintf("%+.3f", drift$beta[drift$Outcome == "Disadvantaged" & drift$Model == "FSM + Absence + KS2"])`
--- a large and precisely estimated positive association. Compare two schools
with the same attendance and the same intake prior attainment, and the one
with the *more* disadvantaged intake gets better results out of its
disadvantaged pupils. For the non-disadvantaged cohort the coefficient stays
negative throughout but collapses to
`r sprintf("%+.4f", drift$beta[drift$Outcome == "Non-disadvantaged" & drift$Model == "FSM + Absence + KS2"])`,
close enough to zero to be of no practical consequence despite its
significance.

This is not a statistical curiosity. It says that the headline association
between intake composition and results --- the association on which a
composition-based admissions policy would rest --- survives at the level of
the school average but largely vanishes, or reverses, at the level of the
pupils it is supposed to help.

### Why the aggregate says something the cohorts do not

The reason is arithmetic, and worth making explicit. A school's headline
Attainment 8 is the FSM-weighted average of its two cohort scores:

$$\text{ATT8} \;=\; w \cdot \text{ATT8}_{\text{disadv}} \;+\; (1 - w) \cdot \text{ATT8}_{\text{non-disadv}}$$

where $w$ is the school's share of disadvantaged pupils. Raising $w$ shifts
weight onto the lower-scoring group and pulls the average down *even if
neither cohort's score changes at all*. Some of the headline FSM association
is therefore pure mixing, with no schooling content whatsoever.

```{r}
#| label: composition-arith

w  <- d3$PTFSM6CLA1A / 100
cf <- w * mean(d3$ATT8SCR_FSM6CLA1A) + (1 - w) * mean(d3$ATT8SCR_NFSM6CLA1A)

b_obs  <- coef(lm(log(ATT8SCR) ~ log(PTFSM6CLA1A), data = d3))[2]
b_comp <- coef(lm(log(cf) ~ log(PTFSM6CLA1A), data = d3))[2]
```

Fixing both cohort scores at their sample means and letting only $w$ vary
reproduces an elasticity of `r sprintf("%.3f", b_comp)` on log(% FSM), against
the `r sprintf("%.3f", b_obs)` actually observed. Roughly
**`r percent(as.numeric(b_comp / b_obs), accuracy = 1)` of the raw
disadvantage--attainment association is mechanical composition**, present
before any question of school quality, teaching or peer effects arises. The
remainder is shared with absence and prior attainment, which is what the rest
of this note has been about.

### Apportioning the shared variance, by cohort

```{r}
#| label: shapley-focal-subgroup

shap3_for <- function(y) {
  perms <- list(c(1,2,3), c(1,3,2), c(2,1,3), c(2,3,1), c(3,1,2), c(3,2,1))
  s <- setNames(numeric(3), FOCAL)
  for (p in perms) {
    sofar <- character(0)
    for (i in p) {
      v <- FOCAL[i]
      s[v] <- s[v] + (r2_out(y, c(sofar, v)) - r2_out(y, sofar))
      sofar <- c(sofar, v)
    }
  }
  s / length(perms)
}

shap3_all <- set_names(map(OUTCOMES3, shap3_for), unname(OUTCOMES3))
```

```{r}
#| label: tbl-shapley-subgroup
#| tbl-cap: "Shapley (LMG) decomposition of the three-predictor model, separately by cohort. Cells show the apportioned R² and, in brackets, its share of the three."

shap3_tbl <- map_dfr(out_lvl, function(nm) {
  s <- shap3_all[[OUTCOMES3[[nm]]]]
  tibble(Outcome = nm, Predictor = lbl(FOCAL),
         `Shapley R²` = as.numeric(s[FOCAL]),
         Share = as.numeric(s[FOCAL]) / sum(s))
})

shap3_tbl %>%
  mutate(cell = paste0(sprintf("%.3f", `Shapley R²`), " (",
                       percent(Share, accuracy = 0.1), ")")) %>%
  select(Outcome, Predictor, cell) %>%
  pivot_wider(names_from = Outcome, values_from = cell) %>%
  kable(align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
#| label: fig-shapley-subgroup
#| fig-cap: "Shares of the explained variance in the three-predictor model, by cohort. Absence overtakes prior attainment for disadvantaged pupils and falls behind it for non-disadvantaged pupils."
#| fig-height: 4.4

shap3_tbl %>%
  mutate(Outcome = factor(Outcome, levels = out_lvl),
         Predictor = factor(Predictor, levels = rev(lbl(FOCAL)))) %>%
  ggplot(aes(Share, Predictor, fill = Predictor)) +
  geom_col(width = 0.68, show.legend = FALSE) +
  geom_text(aes(label = percent(Share, accuracy = 0.1)), hjust = -0.12,
            size = 3, colour = "grey25") +
  facet_wrap(~ Outcome, nrow = 1) +
  scale_x_continuous(labels = percent, limits = c(0, 0.68),
                     expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = c("% FSM (disadvantage)" = "#c0392b",
                               "% Absence" = "#2e6260",
                               "Mean KS2 prior attainment" = "#4e3c56")) +
  labs(title = "Apportioned share of explained variance, by cohort",
       subtitle = "Shapley (LMG) decomposition of the three main predictors",
       x = "Share of the three", y = NULL) +
  theme_vd
```

For all pupils and for the non-disadvantaged cohort, prior attainment takes
the largest share. For the disadvantaged cohort the order changes: absence
takes `r percent(shap3_tbl$Share[shap3_tbl$Outcome == "Disadvantaged" & shap3_tbl$Predictor == "% Absence"], accuracy = 0.1)`
of the explained variance, more than prior attainment and more than five times
the share taken by intake composition. Whatever measure is used, and for
whichever cohort, disadvantage concentration comes last of the three.

### All nine predictors, by cohort

```{r}
#| label: shapley-full-subgroup

k <- length(vars9)
bits <- 2^(seq_len(k) - 1)
masks <- 0:(2^k - 1)

shapley_full <- function(y) {
  r2c <- numeric(length(masks))
  for (m in masks) r2c[m + 1] <- r2_out(y, vars9[bitwAnd(m, bits) > 0])
  ph <- numeric(k)
  for (i in seq_len(k)) {
    b <- bits[i]
    for (m in masks) {
      if (bitwAnd(m, b) == 0) {
        s <- sum(bitwAnd(m, bits) > 0)
        ph[i] <- ph[i] +
          exp(lfactorial(s) + lfactorial(k - s - 1) - lfactorial(k)) *
          (r2c[bitwOr(m, b) + 1] - r2c[m + 1])
      }
    }
  }
  setNames(ph, vars9)
}

phi_all <- set_names(map(OUTCOMES3, shapley_full), unname(OUTCOMES3))
```

```{r}
#| label: tbl-shapley-full-subgroup
#| tbl-cap: "Shapley (LMG) shares across all nine fixed-effect predictors, by cohort. Each column sums to 100%; the R² being apportioned is given in the final row."

phi_tbl <- map_dfr(out_lvl, function(nm) {
  ph <- phi_all[[OUTCOMES3[[nm]]]]
  tibble(Outcome = nm, Predictor = lbl(vars9),
         `Shapley R²` = as.numeric(ph), Share = as.numeric(ph) / sum(ph))
})

body <- phi_tbl %>%
  select(Outcome, Predictor, Share) %>%
  pivot_wider(names_from = Outcome, values_from = Share) %>%
  arrange(desc(`All pupils`)) %>%
  mutate(across(where(is.numeric), \(x) percent(x, accuracy = 0.1)))

totals <- c(list(Predictor = "Total R² apportioned"),
            setNames(as.list(sprintf(
              "%.3f", vapply(out_lvl, \(nm) sum(phi_all[[OUTCOMES3[[nm]]]]),
                             numeric(1)))), out_lvl))

bind_rows(body, as_tibble(totals)) %>%
  kable(align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(nrow(body) + 1, bold = TRUE)
```

```{r}
#| label: fig-shapley-full-subgroup
#| fig-cap: "Apportioned share of explained variance across all nine predictors, by cohort. Admissions policy and EAL matter appreciably more for the disadvantaged cohort; prior attainment matters appreciably less."
#| fig-height: 6

ord9 <- phi_tbl %>% filter(Outcome == "All pupils") %>% arrange(Share) %>% pull(Predictor)

phi_tbl %>%
  mutate(Predictor = factor(Predictor, levels = ord9),
         Outcome = factor(Outcome, levels = out_lvl)) %>%
  ggplot(aes(Share, Predictor, fill = Outcome)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.72) +
  scale_fill_manual(values = OUT_COL) +
  scale_x_continuous(labels = percent, expand = expansion(mult = c(0, 0.04))) +
  labs(title = "Apportioned share of explained variance, all nine predictors",
       subtitle = "Shapley (LMG) shares within each cohort's own model",
       x = "Share of that cohort's explained variance", y = NULL, fill = NULL) +
  theme_vd +
  theme(legend.position = "bottom")
```

The full model explains
`r sprintf("%.3f", sum(phi_all[["ATT8SCR"]]))` of the variance in headline
attainment and
`r sprintf("%.3f", sum(phi_all[["ATT8SCR_NFSM6CLA1A"]]))` for the
non-disadvantaged cohort, but only
`r sprintf("%.3f", sum(phi_all[["ATT8SCR_FSM6CLA1A"]]))` for the disadvantaged
cohort. Structural school characteristics predict how advantaged pupils do
considerably better than they predict how disadvantaged pupils do. The
residual variation --- what schools do, and how well they do it, over and
above their intake and their attendance --- is substantially larger for the
pupils the policy is aimed at.

Two predictors are much more important for the disadvantaged cohort than for
the others. Admissions policy takes
`r percent(phi_tbl$Share[phi_tbl$Outcome == "Disadvantaged" & phi_tbl$Predictor == "Admissions policy"], accuracy = 0.1)`
of the explained variance against
`r percent(phi_tbl$Share[phi_tbl$Outcome == "All pupils" & phi_tbl$Predictor == "Admissions policy"], accuracy = 0.1)`
for all pupils, and EAL takes
`r percent(phi_tbl$Share[phi_tbl$Outcome == "Disadvantaged" & phi_tbl$Predictor == "% EAL"], accuracy = 0.1)`
against
`r percent(phi_tbl$Share[phi_tbl$Outcome == "All pupils" & phi_tbl$Predictor == "% EAL"], accuracy = 0.1)`.
The EAL result is the familiar one that disadvantaged pupils with English as
an additional language outperform disadvantaged pupils without it; the
admissions result reflects the selective and highly-selective-area categories,
where the small disadvantaged cohorts that do get in are unrepresentative of
disadvantaged pupils generally. Neither is a lever in the sense the paper
uses the word.

## Synthesis: what holds across cohorts and what does not

```{r}
#| label: tbl-synthesis
#| tbl-cap: "Rank of the three main predictors under each measure and for each cohort. 1 = largest."

rank_tbl <- map_dfr(out_lvl, function(nm) {
  y <- OUTCOMES3[[nm]]
  m3 <- lm(as.formula(paste0("log(", y, ") ~ ",
                             paste(FOCAL, collapse = " + "))), data = d3)
  tv <- abs(summary(m3)$coefficients[FOCAL, "t value"])
  bind_rows(
    tibble(Outcome = nm, Measure = "R² alone",
           Predictor = lbl(FOCAL), Value = vapply(FOCAL, \(v) r2_out(y, v), numeric(1))),
    tibble(Outcome = nm, Measure = "Absolute t in the 3-variable model",
           Predictor = lbl(FOCAL), Value = as.numeric(tv)),
    tibble(Outcome = nm, Measure = "Shapley (LMG)",
           Predictor = lbl(FOCAL), Value = as.numeric(shap3_all[[y]][FOCAL]))
  )
}) %>%
  group_by(Outcome, Measure) %>%
  mutate(Rank = rank(-Value)) %>%
  ungroup()

rank_tbl %>%
  mutate(Measure = factor(Measure, levels = c("Absolute t in the 3-variable model",
                                              "R² alone", "Shapley (LMG)")),
         Outcome = factor(Outcome, levels = out_lvl)) %>%
  arrange(Measure, Outcome) %>%
  select(Measure, Outcome, Predictor, Rank) %>%
  pivot_wider(names_from = Predictor, values_from = Rank) %>%
  kable(align = "llrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

### Three things that hold everywhere

**Disadvantage concentration is never the leading factor.** Across three
outcomes and three measures --- nine combinations in all --- the share of
disadvantaged pupils in a school ranks last of the three main predictors in
every one. It ranks last when it is credited with all of the variance it
shares with the others, when it is credited with none of it, and when the
shared part is split evenly. The paper's central claim does not depend on
which cohort is examined or on how the sharing is resolved.

**Absence is always the largest movable factor.** Prior attainment leads on
two of the three outcomes, but it is a description of the intake rather than
something a school or an authority can act on. Set it aside and absence is the
largest remaining factor for all pupils, for disadvantaged pupils and for
non-disadvantaged pupils alike. For the disadvantaged cohort it is the largest
factor full stop.

**Shared variance is substantial in every case, so *t*-values remain unsafe as
a ranking.** The gap between "$R^2$ alone" and "Shapley" that motivated this
note is just as wide for both subgroups as it is for the headline outcome. As
it happens, the *t*-value ordering agrees with the other two measures for both
cohorts, and disagrees only for all pupils --- the reverse of what one might
expect. That agreement is a fact about this data rather than a reason to trust
*t*-values: the predictors are just as correlated in the cohort models, the
*t*-values still allocate none of the shared variance, and nothing in the
method guarantees that the orderings will coincide. Where they happen to
coincide it is luck, and section 4.2 of the paper should not lean on it.

### Three things that differ

**Absence and prior attainment change places.** For non-disadvantaged pupils,
prior attainment is comfortably the largest factor
(`r percent(shap3_tbl$Share[shap3_tbl$Outcome == "Non-disadvantaged" & shap3_tbl$Predictor == "Mean KS2 prior attainment"], accuracy = 0.1)`
of the explained variance against
`r percent(shap3_tbl$Share[shap3_tbl$Outcome == "Non-disadvantaged" & shap3_tbl$Predictor == "% Absence"], accuracy = 0.1)`
for absence). For disadvantaged pupils the order reverses and absence becomes
the dominant term
(`r percent(shap3_tbl$Share[shap3_tbl$Outcome == "Disadvantaged" & shap3_tbl$Predictor == "% Absence"], accuracy = 0.1)`
against
`r percent(shap3_tbl$Share[shap3_tbl$Outcome == "Disadvantaged" & shap3_tbl$Predictor == "Mean KS2 prior attainment"], accuracy = 0.1)`).
Part of this is that `ks2_c` measures the whole intake rather than the
disadvantaged part of it, and so tracks the non-disadvantaged cohort more
closely. But the direction of the difference points the same way as the
attendance literature: missed schooling costs disadvantaged pupils more,
because they have less capacity to make the lost ground up outside school.
This strengthens rather than weakens the paper's case for attendance as the
priority lever, since the lever bites hardest exactly where the gap is.

**The disadvantage coefficient changes sign between cohorts.** Conditional on
absence and prior attainment, a more disadvantaged intake is associated with
*better* results for disadvantaged pupils
(`r sprintf("%+.3f", drift$beta[drift$Outcome == "Disadvantaged" & drift$Model == "FSM + Absence + KS2"])`)
and with marginally worse results for non-disadvantaged pupils
(`r sprintf("%+.4f", drift$beta[drift$Outcome == "Non-disadvantaged" & drift$Model == "FSM + Absence + KS2"])`).
The negative headline association is therefore not one effect operating on
everyone: it is composition arithmetic, plus a very small negative association
for the advantaged cohort, running against a positive association for the
disadvantaged cohort. A policy that dispersed disadvantaged pupils into less
disadvantaged schools would, on these figures, be moving them into settings
where --- attendance and prior attainment held equal --- disadvantaged pupils
do slightly worse rather than better. That is a more awkward result for
composition-based admissions policy than the headline model alone suggests,
and it is worth stating in the paper.

**Disadvantaged attainment is markedly less predictable.** The full
nine-predictor model reaches an $R^2$ of
`r sprintf("%.2f", sum(phi_all[["ATT8SCR_NFSM6CLA1A"]]))` for the
non-disadvantaged cohort but only
`r sprintf("%.2f", sum(phi_all[["ATT8SCR_FSM6CLA1A"]]))` for the disadvantaged
cohort, despite disadvantaged attainment varying *more* between schools
(SD of log
`r sprintf("%.2f", sd(log(d3$ATT8SCR_FSM6CLA1A)))` against
`r sprintf("%.2f", sd(log(d3$ATT8SCR_NFSM6CLA1A)))`). More between-school
variation, less of it accounted for: around
`r percent(1 - sum(phi_all[["ATT8SCR_FSM6CLA1A"]]), accuracy = 1)` of the
variation in how disadvantaged pupils do is not explained by any of the
structural characteristics available here. Read optimistically that is
headroom --- school-level practice that the intake does not determine. Read
cautiously it is a warning that the measurable characteristics of a school
tell you least about the pupils the policy most wants to help.

### Caveats specific to the subgroup analysis

Three limitations apply to this section over and above the general caveat in
section 8.

*The prior-attainment measure is not cohort-specific.* `ks2_c` is the school's
mean KS2 score across all pupils, so it is a better proxy for the
non-disadvantaged cohort's starting point than for the disadvantaged cohort's.
Some of the apparent weakness of prior attainment in the disadvantaged column
is measurement rather than substance, and the figure for a cohort-specific
prior attainment measure would be higher. The direction of the difference is
nonetheless likely to be right, since a cohort-specific measure would raise
the disadvantaged prior-attainment figure without lowering the absence figure.

*The absence measure is likewise school-wide.* Overall absence and
disadvantaged pupils' own absence are strongly related but not identical, and
disadvantaged absence is the higher of the two in almost every school. This
makes the absence result for the disadvantaged cohort conservative if
anything: a cohort-specific absence measure would be expected to strengthen
it.

*Small cohorts are noisy.* Schools with few disadvantaged pupils produce
disadvantaged Attainment 8 figures based on a handful of pupils, which inflates
the between-school variance in that column and depresses its $R^2$. The models
here are unweighted, so those schools carry the same weight as large ones.
This is one plausible contributor to the lower explained variance for the
disadvantaged cohort, and it should be borne in mind before reading the
residual as headroom.

## A combined figure: direction and apportioned magnitude together {#sec-combined-figure}

The paper's Figure 1 plots standardised coefficients for the three pupil
groups. That shows *direction*, which matters --- the sign on disadvantage
concentration reverses between cohorts, and the reversal is one of the paper's
substantive findings --- but a standardised coefficient is a partial quantity,
so it understates any predictor that shares a lot of variance with the others.
The Shapley shares computed above have the opposite property: they apportion
the shared variance, but a variance decomposition has no sign.

Neither quantity is dispensable, and neither can be made to carry the other's
information. The figure below therefore shows both, on a shared vertical axis,
with predictors ordered by apportioned share.

```{r}
#| label: fig1-combined-data

# Standardised coefficients come from the paper's fitted multilevel models, so
# the left panel is the quantity Figure 1 already reports. Shapley shares are
# computed by OLS over the full subset lattice, as everywhere else in this
# note: 2^9 subsets per cohort is tractable with lm() and is not with lmer().
models_imputed <- readRDS(here::here("data", "models_imputed.rds"))

FIG_MODELS <- list("All pupils"        = models_imputed$all,
                   "Disadvantaged"     = models_imputed$disadvantaged,
                   "Non-disadvantaged" = models_imputed$non_disadvantaged)
FIG_OUTCOME <- c("All pupils"        = "ATT8SCR",
                 "Disadvantaged"     = "ATT8SCR_FSM6CLA1A",
                 "Non-disadvantaged" = "ATT8SCR_NFSM6CLA1A")

FIG_LABELS <- c(
  "log(PERCTOT)"                             = "Overall absence",
  "ks2_c"                                    = "Mean KS2 prior attainment",
  "log(PTFSM6CLA1A)"                         = "% disadvantaged (FSM)",
  "ADMPOL_PT"                                = "Selective admissions",
  "log(PNUMEAL)"                             = "% EAL",
  "remained_in_the_same_school"              = "Teacher retention",
  "teachers_on_leadership_pay_range_percent" = "Leadership pay %",
  "log(average_number_of_days_taken)"        = "Teacher sickness days",
  "gorard_segregation"                       = "Gorard segregation index"
)
FIG_VARS <- names(FIG_LABELS)

# The comparison needs every cohort outcome present on the same rows.
fig_d <- d %>%
  filter(!is.na(ATT8SCR_FSM6CLA1A), ATT8SCR_FSM6CLA1A > 0,
         !is.na(ATT8SCR_NFSM6CLA1A), ATT8SCR_NFSM6CLA1A > 0) %>%
  droplevels()

fig_sd <- function(v) {
  if (v == "ADMPOL_PT") return(sd(as.numeric(fig_d$ADMPOL_PT == "SEL"), na.rm = TRUE))
  raw <- sub("^log\\((.*)\\)$", "\\1", v)
  x <- if (grepl("^log\\(", v)) log(fig_d[[raw]]) else fig_d[[raw]]
  sd(x, na.rm = TRUE)
}

# signed standardised coefficient from the fitted multilevel model
fig_std <- function(model) {
  fe <- lme4::fixef(model)
  vapply(FIG_VARS, function(v) {
    nm <- if (v == "ADMPOL_PT") "ADMPOL_PTSEL" else v
    if (!nm %in% names(fe)) return(NA_real_)
    unname(fe[[nm]]) * fig_sd(v)
  }, numeric(1))
}

# Shapley over all 2^9 subsets, by OLS, on the same rows
fig_r2 <- function(y, vs) {
  rhs <- if (length(vs) == 0) "1" else paste(vs, collapse = " + ")
  summary(lm(as.formula(paste0("log(", y, ") ~ ", rhs)), data = fig_d))$r.squared
}
fig_shapley <- function(y) {
  k <- length(FIG_VARS)
  masks <- 0:(2^k - 1)
  r2c <- vapply(masks, function(m)
    fig_r2(y, FIG_VARS[bitwAnd(m, 2^(seq_len(k) - 1)) > 0]), numeric(1))
  phi <- numeric(k)
  for (i in seq_len(k)) {
    bit <- 2^(i - 1)
    for (m in masks) {
      if (bitwAnd(m, bit) == 0) {
        s <- sum(bitwAnd(m, 2^(seq_len(k) - 1)) > 0)
        w <- exp(lfactorial(s) + lfactorial(k - s - 1) - lfactorial(k))
        phi[i] <- phi[i] + w * (r2c[bitwOr(m, bit) + 1] - r2c[m + 1])
      }
    }
  }
  setNames(phi, FIG_VARS)
}

fig_dat <- purrr::map_dfr(names(FIG_MODELS), function(nm) {
  phi <- fig_shapley(FIG_OUTCOME[[nm]])
  tibble(Cohort = nm, var = FIG_VARS, Predictor = unname(FIG_LABELS[FIG_VARS]),
         share = as.numeric(phi) / sum(phi),
         std   = as.numeric(fig_std(FIG_MODELS[[nm]])))
})

fig_lev <- fig_dat %>% filter(Cohort == "All pupils") %>%
  arrange(share) %>% pull(Predictor)
fig_dat <- fig_dat %>%
  mutate(Predictor = factor(Predictor, levels = fig_lev),
         Cohort = factor(Cohort, levels = names(FIG_MODELS)))
```

```{r}
#| label: fig-combined
#| fig-cap: "Relative importance of each predictor of school-level Attainment 8, by pupil group. The left panel gives the standardised coefficient from the multilevel model, which carries the direction of each association but only its unique contribution. The right panel gives the Shapley share of explained variance, which apportions the variance predictors share but has no sign. Predictors are ordered by Shapley share for all pupils."
#| fig-width: 11
#| fig-height: 5.8

COH_COL <- c("All pupils" = "#4e3c56", "Disadvantaged" = "#c0392b",
             "Non-disadvantaged" = "#2e6260")

fig_theme <- theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        plot.subtitle = element_text(colour = "grey35", size = 9, lineheight = 1.1),
        legend.position = "bottom")

p_left <- ggplot(fig_dat, aes(std, Predictor, fill = Cohort)) +
  geom_vline(xintercept = 0, colour = "grey40") +
  geom_col(position = position_dodge(width = 0.78), width = 0.7) +
  scale_fill_manual(values = COH_COL) +
  labs(subtitle = "Direction and unique contribution\n(standardised coefficient)",
       x = "Change in log(Attainment 8) per 1 SD", y = NULL) +
  fig_theme

p_right <- ggplot(fig_dat, aes(share, Predictor, fill = Cohort)) +
  geom_col(position = position_dodge(width = 0.78), width = 0.7) +
  scale_fill_manual(values = COH_COL) +
  scale_x_continuous(labels = percent, expand = expansion(mult = c(0, 0.04))) +
  labs(subtitle = "Apportioned share of explained variance\n(Shapley, sharing resolved)",
       x = "Share of model R²", y = NULL) +
  fig_theme +
  theme(axis.text.y = element_blank())

(p_left | p_right) +
  patchwork::plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

Reading the two panels together gives the paper's argument in one view.

**Absence is the largest factor a school can act on.** Prior attainment is
larger still on the right-hand panel for two of the three groups, but it
describes the intake a school receives rather than anything it can change for a
cohort already on roll. Set it aside and absence leads everywhere --- and for
disadvantaged pupils it is the largest term outright, on both panels.

**Disadvantage concentration is small, and its direction depends on who is
being asked about.** On the right it sits third for every group. On the left the
sign reverses: negative for all pupils and for the non-disadvantaged, positive
for the disadvantaged. Neither panel could show this alone. The Shapley shares
would say only that the predictor is small; the standardised coefficients would
understate how small, because a variable sharing this much variance with the
others gets little credit in a partial measure.

**The two panels disagree about ordering, and that disagreement is the point.**
For all pupils the standardised coefficient makes absence look larger than
prior attainment, while the apportioned share reverses them. Nothing is wrong
in either: they answer different questions, one about unique contribution and
one about total contribution with the sharing split. Reporting only the first,
as the paper's stepwise table did in an earlier draft, invites a ranking it
cannot support.

A note on construction. The standardised coefficients are taken from the
multilevel models the paper fits, so the left panel reports exactly the
quantity Figure 1 reports now. The Shapley shares are computed by OLS over all
512 subsets of the nine fixed effects: with `lmer()` the lattice would need
1,536 multilevel fits and is not practical, and the ordering is in any case
governed by the fixed part. Both panels use the same school-years, restricted
to rows where all three cohort outcomes are published.
