---
title: "Shared Variance and Relative Importance"
subtitle: "How much does each predictor really explain, and does the answer depend on how we ask?"
author: "Supplementary note"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    fig-width: 9
    fig-height: 6
    fig-dpi: 150
    df-print: kable
    embed-resources: true
bibliography: ../SchoolAttainment.bib
csl: ../chicago-author-date.csl
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(scales)

panel <- readRDS(here::here("data", "panel_data.rds"))

# The Analysis E estimation sample, restricted to complete cases on every
# predictor so that all subsets below are fitted on identical rows. This
# matters: lm() drops incomplete cases silently, so models containing
# different variables would otherwise be fitted on different samples and
# their R-squared values would not be comparable.
d <- panel %>%
  filter(MINORGROUP %in% c("Academy", "Maintained school")) %>%
  filter(!is.na(ATT8SCR), ATT8SCR > 0,
         PTFSM6CLA1A > 0, PERCTOT > 0, PNUMEAL > 0,
         !is.na(ks2_c), !is.na(ADMPOL_PT), !is.na(gorard_segregation),
         !is.na(remained_in_the_same_school),
         !is.na(teachers_on_leadership_pay_range_percent),
         average_number_of_days_taken > 0) %>%
  droplevels()

PREDS <- c(
  "log(PTFSM6CLA1A)"                         = "% FSM (disadvantage)",
  "log(PERCTOT)"                             = "% Absence",
  "ks2_c"                                    = "Mean KS2 prior attainment",
  "log(PNUMEAL)"                             = "% EAL",
  "ADMPOL_PT"                                = "Admissions policy",
  "gorard_segregation"                       = "Gorard segregation",
  "remained_in_the_same_school"              = "Teacher retention",
  "teachers_on_leadership_pay_range_percent" = "Leadership pay %",
  "log(average_number_of_days_taken)"        = "Teacher sickness days"
)

FOCAL <- c("log(PTFSM6CLA1A)", "log(PERCTOT)", "ks2_c")

r2_of <- function(vs) {
  rhs <- if (length(vs) == 0) "1" else paste(vs, collapse = " + ")
  summary(lm(as.formula(paste("log(ATT8SCR) ~", rhs)), data = d))$r.squared
}

lbl <- function(v) unname(PREDS[v])

theme_vd <- theme_minimal(base_size = 11) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(colour = "grey40"),
        panel.grid.minor = element_blank())
```

## Why this note exists

The paper's stepwise table (Model 1 to Model 5) reports coefficients and
*t*-values for each predictor. It is tempting to read those *t*-values as a
ranking of importance, and an earlier draft of section 4.2 did exactly that.
That reading is not safe, for a reason worth setting out carefully.

In a multiple regression, each coefficient describes a predictor's **unique**
contribution: the association that remains once every other predictor is held
constant. Where predictors are uncorrelated, unique contribution and total
contribution are the same thing and nothing is lost. Where predictors are
correlated, a large part of the variance they explain is **shared** between
them, and the model has no way to attribute that shared part to one variable
rather than another. It simply goes unallocated. A variable that overlaps
heavily with the others will therefore show a modest *t*-value even when it
accounts for a great deal of the outcome on its own.

Our three main predictors overlap a good deal, so this is not a hypothetical
concern. This note quantifies it: it reports what each predictor explains on
its own, what every combination of predictors explains, and how the shared
variance is apportioned when we use a method designed for that purpose. It
then asks the question that actually matters --- whether the paper's
conclusions depend on which measure we choose.

The analysis sample is `r format(nrow(d), big.mark = ",")` school-years, the
same rows used for the paper's models, restricted to complete cases so that
every model below is fitted on identical data.

## How much do the predictors overlap?

```{r}
#| label: fig-corr
#| fig-cap: "Correlations between the model's continuous predictors. Strong off-diagonal values mean shared variance that a multiple regression cannot attribute to one variable rather than another."
#| fig-height: 6.5

num_preds <- setdiff(names(PREDS), "ADMPOL_PT")
X <- as.data.frame(lapply(num_preds, function(v) {
  raw <- sub("^log\\((.*)\\)$", "\\1", v)
  x <- d[[raw]]
  if (grepl("^log\\(", v)) log(x) else x
}))
names(X) <- lbl(num_preds)

cm <- cor(X, use = "complete.obs")
cm_long <- as.data.frame(as.table(cm)) %>%
  setNames(c("a", "b", "r"))

ggplot(cm_long, aes(a, b, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.2f", r),
                colour = abs(r) > 0.55), size = 3.1, show.legend = FALSE) +
  scale_fill_gradient2(low = "#c0392b", mid = "white", high = "#2e6260",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  scale_colour_manual(values = c("TRUE" = "white", "FALSE" = "grey20")) +
  labs(title = "Correlations between predictors",
       subtitle = "The three main predictors overlap substantially",
       x = NULL, y = NULL) +
  coord_equal() +
  theme_vd +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

```{r}
#| label: tbl-focal-corr
#| tbl-cap: "Correlations between the three main predictors."

fx <- data.frame(
  fsm = log(d$PTFSM6CLA1A), abs = log(d$PERCTOT), ks2 = d$ks2_c
)
tibble(
  Pair = c("% FSM and mean KS2", "% Absence and mean KS2", "% FSM and % Absence"),
  r = c(cor(fx$fsm, fx$ks2), cor(fx$abs, fx$ks2), cor(fx$fsm, fx$abs))
) %>%
  mutate(r = sprintf("%+.3f", r)) %>%
  kable(align = "lr")
```

Schools with more disadvantaged intakes have markedly lower prior attainment,
and schools with higher absence also have lower prior attainment. These are
not nuisance correlations to be controlled away; they are the structure of the
problem. But they do mean that a large share of the variance in attainment is
explained by these variables *jointly*, and any claim about which one matters
most has to say how the joint part is being handled.

## What each predictor explains on its own

The simplest question: fit each predictor alone against attainment, and record
the $R^2$. This is the *total* association, including everything the variable
shares with the others.

```{r}
#| label: tbl-univariate
#| tbl-cap: "Each predictor fitted on its own. This is total explanatory power, sharing included."

uni <- map_dfr(names(PREDS), function(v) {
  tibble(Predictor = lbl(v), `R²` = r2_of(v))
}) %>% arrange(desc(`R²`))

uni %>%
  mutate(`R²` = sprintf("%.3f", `R²`),
         `Variance explained` = percent(as.numeric(`R²`), accuracy = 0.1)) %>%
  kable(align = "lrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
#| label: fig-univariate
#| fig-cap: "Variance in school-level Attainment 8 explained by each predictor on its own."
#| fig-height: 4.5

uni %>%
  mutate(Predictor = fct_reorder(Predictor, `R²`)) %>%
  ggplot(aes(`R²`, Predictor)) +
  geom_col(fill = "#4e3c56", width = 0.72) +
  geom_text(aes(label = percent(`R²`, accuracy = 0.1)),
            hjust = -0.15, size = 3.2, colour = "grey25") +
  scale_x_continuous(labels = percent, limits = c(0, 0.68),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(title = "Explanatory power of each predictor on its own",
       subtitle = "Total association, including variance shared with other predictors",
       x = "R² alone", y = NULL) +
  theme_vd
```

On this measure prior attainment leads, absence follows, and disadvantage is
third. Note how different this is from the *t*-values in the paper's Model 3,
where absence appears strongest and disadvantage weakest by a wide margin. The
gap between the two pictures is precisely the shared variance.

## Every combination of the three main predictors

The table below fits all eight possible subsets of disadvantage, absence and
prior attainment, and reports both the model $R^2$ and how each coefficient
and *t*-value moves as its companions are added.

```{r}
#| label: tbl-subsets
#| tbl-cap: "All 8 subsets of the three main predictors, fitted on identical rows. Watch the FSM coefficient shrink as its companions enter."

subsets <- list(character(0), "log(PTFSM6CLA1A)", "log(PERCTOT)", "ks2_c",
                c("log(PTFSM6CLA1A)", "log(PERCTOT)"),
                c("log(PTFSM6CLA1A)", "ks2_c"),
                c("log(PERCTOT)", "ks2_c"),
                FOCAL)

fit_sub <- function(vs) {
  rhs <- if (length(vs) == 0) "1" else paste(vs, collapse = " + ")
  m <- lm(as.formula(paste("log(ATT8SCR) ~", rhs)), data = d)
  s <- summary(m)$coefficients
  get <- function(v, col) if (v %in% vs) s[v, col] else NA_real_
  tibble(
    Model = if (length(vs) == 0) "(intercept only)"
            else paste(c("FSM", "Absence", "KS2")[match(vs, FOCAL)], collapse = " + "),
    `R²`      = summary(m)$r.squared,
    `FSM β`   = get("log(PTFSM6CLA1A)", "Estimate"),
    `FSM t`   = get("log(PTFSM6CLA1A)", "t value"),
    `Abs β`   = get("log(PERCTOT)", "Estimate"),
    `Abs t`   = get("log(PERCTOT)", "t value"),
    `KS2 β`   = get("ks2_c", "Estimate"),
    `KS2 t`   = get("ks2_c", "t value")
  )
}
subs <- map_dfr(subsets, fit_sub)

subs %>%
  mutate(across(where(is.numeric), \(x) ifelse(is.na(x), "—", sprintf("%.3f", x)))) %>%
  kable(align = "lrrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

```{r}
#| label: fig-coef-drift
#| fig-cap: "How the disadvantage coefficient changes as the other predictors are added. Each bar is a model; the coefficient more than halves once absence and prior attainment are present."
#| fig-height: 4

subs %>%
  filter(!is.na(`FSM β`)) %>%
  mutate(Model = fct_reorder(Model, `FSM β`)) %>%
  ggplot(aes(`FSM β`, Model)) +
  geom_col(fill = "#c0392b", width = 0.6) +
  geom_text(aes(label = sprintf("%.3f", `FSM β`)), hjust = 1.15,
            colour = "white", size = 3.2) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0))) +
  labs(title = "The disadvantage coefficient depends heavily on what else is in the model",
       subtitle = "log(% FSM) coefficient across the four models that contain it",
       x = "Coefficient on log(% FSM)", y = NULL) +
  theme_vd
```

This is the shared-variance problem made visible. Disadvantage on its own
carries a coefficient of about −0.20; with absence and prior attainment
alongside it, that falls to roughly −0.05. Nothing about the schools has
changed --- only what else the model is being told.

Which of those numbers is "the effect of disadvantage" depends entirely on the
question. If you want the total association between a school's intake
composition and its results, the first is closer. If you want the association
that remains between two schools with the same absence and the same prior
attainment, the last is closer. The paper needs the second, because its
question is what happens if you change intake composition while everything
else stays as it is --- but it should not then describe that number as the
importance of disadvantage in general, and section 3.2 now says so explicitly.

## Apportioning the shared variance

Neither "$R^2$ alone" nor "the coefficient in the full model" is a complete
answer. The first hands each variable all of the shared variance; the second
hands it none. The standard resolution is to average over every possible
order in which the variables could enter: a variable's contribution is
computed when it enters first, second and third, and these are averaged. This
is the Shapley value from cooperative game theory, known in this context as
the LMG decomposition after @Lindeman1980, and implemented for R by
@Gromping2007.

```{r}
#| label: shapley-focal

# 3 variables = 6 orderings; done by hand to avoid a package dependency
perms3 <- list(c(1,2,3), c(1,3,2), c(2,1,3), c(2,3,1), c(3,1,2), c(3,2,1))
shap3 <- setNames(numeric(3), FOCAL)
for (p in perms3) {
  sofar <- character(0)
  for (i in p) {
    v <- FOCAL[i]
    shap3[v] <- shap3[v] + (r2_of(c(sofar, v)) - r2_of(sofar))
    sofar <- c(sofar, v)
  }
}
shap3 <- shap3 / length(perms3)
```

```{r}
#| label: tbl-shapley
#| tbl-cap: "Shapley (LMG) decomposition of the three-predictor model. Each variable's contribution is averaged over all six orders of entry, so the shared variance is split rather than assigned arbitrarily."

tibble(
  Predictor = lbl(FOCAL),
  `R² alone` = vapply(FOCAL, r2_of, numeric(1)),
  `Shapley R²` = as.numeric(shap3[FOCAL]),
  `Share of explained variance` = as.numeric(shap3[FOCAL]) / sum(shap3)
) %>%
  arrange(desc(`Shapley R²`)) %>%
  mutate(
    `R² alone` = sprintf("%.3f", `R² alone`),
    `Shapley R²` = sprintf("%.3f", `Shapley R²`),
    `Share of explained variance` = percent(`Share of explained variance`, accuracy = 0.1)
  ) %>%
  kable(align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
#| label: fig-three-rankings
#| fig-cap: "The three predictors ranked three ways. Prior attainment and absence swap places depending on the measure; disadvantage is last on all three."
#| fig-height: 4.5

m3 <- lm(log(ATT8SCR) ~ log(PTFSM6CLA1A) + log(PERCTOT) + ks2_c, data = d)
tvals <- abs(summary(m3)$coefficients[FOCAL, "t value"])

rank_df <- bind_rows(
  tibble(Predictor = lbl(FOCAL), Measure = "R² alone",
         Value = vapply(FOCAL, r2_of, numeric(1)) / sum(vapply(FOCAL, r2_of, numeric(1)))),
  tibble(Predictor = lbl(FOCAL), Measure = "|t| in the 3-variable model",
         Value = tvals / sum(tvals)),
  tibble(Predictor = lbl(FOCAL), Measure = "Shapley (LMG) share",
         Value = as.numeric(shap3[FOCAL]) / sum(shap3))
) %>%
  mutate(Measure = factor(Measure, levels = c("|t| in the 3-variable model",
                                              "R² alone", "Shapley (LMG) share")))

ggplot(rank_df, aes(Value, fct_rev(factor(Predictor)), fill = Predictor)) +
  geom_col(width = 0.68, show.legend = FALSE) +
  geom_text(aes(label = percent(Value, accuracy = 0.1)), hjust = -0.12,
            size = 3, colour = "grey25") +
  facet_wrap(~ Measure, nrow = 1) +
  scale_x_continuous(labels = percent, limits = c(0, 0.62),
                     expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = c("% FSM (disadvantage)" = "#c0392b",
                               "% Absence" = "#2e6260",
                               "Mean KS2 prior attainment" = "#4e3c56")) +
  labs(title = "Relative importance under three different measures",
       subtitle = "Each measure rescaled to shares so the three are comparable",
       x = "Share of the total", y = NULL) +
  theme_vd
```

The ranking of the top two **does** depend on the measure. By *t*-value,
absence leads; by explanatory power alone and by the Shapley decomposition,
prior attainment leads. This is exactly why the paper no longer rests any
claim on the *t*-value ordering.

What does not change is the position of disadvantage: **last under all three
measures**, whether it is credited with all of the shared variance, none of
it, or an even split. The paper's substantive argument depends only on that,
and it is robust to the choice.

## All nine predictors

The same decomposition across the full fixed-effect specification. With nine
predictors there are $2^9 = 512$ subsets and $9! = 362{,}880$ orderings, so
the Shapley values are computed from the subset lattice rather than by
enumerating permutations --- the two are mathematically identical, but the
first is tractable.

```{r}
#| label: shapley-full
#| cache: false

vars9 <- names(PREDS)
k <- length(vars9)

# R-squared for every subset, indexed by bitmask
masks <- 0:(2^k - 1)
r2_cache <- numeric(length(masks))
for (m in masks) {
  sel <- vars9[bitwAnd(m, 2^(seq_len(k) - 1)) > 0]
  r2_cache[m + 1] <- r2_of(sel)
}

# Shapley value from the lattice:
#   phi_i = sum over subsets S not containing i of
#           |S|!(k-|S|-1)!/k! * (R2(S+i) - R2(S))
lfact <- lfactorial
phi <- numeric(k)
for (i in seq_len(k)) {
  bit <- 2^(i - 1)
  for (m in masks) {
    if (bitwAnd(m, bit) == 0) {
      s <- sum(bitwAnd(m, 2^(seq_len(k) - 1)) > 0)
      w <- exp(lfact(s) + lfact(k - s - 1) - lfact(k))
      phi[i] <- phi[i] + w * (r2_cache[bitwOr(m, bit) + 1] - r2_cache[m + 1])
    }
  }
}
names(phi) <- vars9
```

```{r}
#| label: tbl-shapley-full
#| tbl-cap: "Shapley (LMG) decomposition across all nine fixed-effect predictors, computed over all 512 subsets."

full_tbl <- tibble(
  Predictor = lbl(vars9),
  `R² alone` = vapply(vars9, r2_of, numeric(1)),
  `Shapley R²` = phi,
  `Share` = phi / sum(phi)
) %>% arrange(desc(`Shapley R²`))

full_tbl %>%
  mutate(`R² alone` = sprintf("%.3f", `R² alone`),
         `Shapley R²` = sprintf("%.3f", `Shapley R²`),
         `Share` = percent(`Share`, accuracy = 0.1)) %>%
  kable(align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
#| label: fig-shapley-full
#| fig-cap: "Explanatory power alone against apportioned share, all nine predictors. Variables far above the diagonal owe much of their apparent strength to variance shared with others."
#| fig-height: 5.5

full_tbl %>%
  mutate(Predictor = fct_reorder(Predictor, `Shapley R²`)) %>%
  pivot_longer(c(`R² alone`, `Shapley R²`), names_to = "Measure", values_to = "Value") %>%
  ggplot(aes(Value, Predictor, fill = Measure)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  scale_fill_manual(values = c("R² alone" = "#abc766", "Shapley R²" = "#4e3c56")) +
  scale_x_continuous(labels = percent, expand = expansion(mult = c(0, 0.04))) +
  labs(title = "Alone versus apportioned",
       subtitle = "The gap between the two bars is variance the predictor shares with others",
       x = "Variance explained", y = NULL, fill = NULL) +
  theme_vd +
  theme(legend.position = "bottom")
```

The sum of the Shapley values is
`r sprintf("%.3f", sum(phi))`, which by construction equals the $R^2$ of the
full nine-predictor model --- the decomposition allocates all of the explained
variance and none of it twice.

## What this means for the paper

Three conclusions, in order of how much they matter.

**The importance ranking of the top two predictors is method-dependent, and
the paper should not assert one.** Section 4.2 previously ranked the three by
*t*-value and called absence the most important. That has been rewritten: it
now reports the *t*-values as measures of precision, states plainly that they
cannot rank importance where predictors are correlated, and gives the
explanatory-power and Shapley figures alongside. Section 4.3, which uses
standardised coefficients from the full model, already placed prior attainment
first; the two sections no longer contradict one another.

**The paper's argument does not depend on the choice.** Concentrations of
disadvantage rank last on every measure tried here. The claim the paper
actually makes --- that disadvantage concentration is a weak lever relative to
absence, and that an admissions policy built on it is unlikely to deliver the
gains sought --- is unaffected by how the shared variance is apportioned.

**Prior attainment is large but is not a lever.** The measures above rank
*explanatory power*, not *actionability*. Prior attainment is the single
largest source of variation on two of the three measures, but it is fixed for
any cohort already in secondary school: it describes the intake a school
receives, not something the school or its authority can change. Once it is set
aside as intake rather than lever, absence is the largest factor remaining
that is open to intervention --- which is the distinction section 4.3 draws
and the one on which the policy argument rests.

## A caveat on all of this

Every measure here is descriptive. None of them licenses a causal reading, and
the relative-importance literature is explicit that these decompositions
apportion *variance explained*, not *effect* [@Gromping2007]. Two predictors
can share variance because one causes the other, because a third thing causes
both, or because they are alternative measurements of the same underlying
disadvantage --- and no decomposition of an observational dataset can tell
these apart. That prior attainment, absence and disadvantage overlap so
heavily is itself the substantive finding: they are facets of a single
structure, and the policy question is not which is "most important" but which
of them can actually be moved.
